### 2015-12-08 main.tex
started paper outline and bibliography

\documentclass{article}

\usepackage[affil-it]{authblk}
\usepackage{lineno}
\usepackage[square,numbers,sort&compress]{natbib}

\title{A Comparision of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}

\author{{Matthew Grasinger}
\thanks{Email: \textit{grasingerm@pitt.edu}}}

\author{John Brigham
\thanks{Email: \textit{brigham@pitt.edu}}}
\affil{Civil and Environmental Engineering Department, University of Pittsburgh}

\begin{document}

\pagenumbering{gobble}
\maketitle
\newpage
\tableofcontents
\newpage
\pagenumbering{arabic}

%\linenumbers

\section*{Abstract}

The Lattice Boltzmann method (LBM) is a computational method that can be used for simulating fluid flow.
It is well-suited for complex flows such as non-Newtonian, free surface, and multiphase multicomponent flows.
Non-Newtonian flows are the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten plastics, ETC, are best modeled as non-Newtonian fluids.
LBM is a statistical mechanical approach and each time step consists of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
The collision operator is of interest because it has a lot of control over the physics that are simulated, e.g. constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
In this paper, variations of popular collision operators, namely the linear Bhatnagar--Gross--Krook collision operator and the multiple relaxation time collision operator, will be examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
The structure of the paper will be as follows:

\section{Introduction}

\subsection{Scratch}

Lattice Boltzmann Method (LBM) is the greatest method ever invented.

What is my story?:

\begin{itemize}
	\item Motivation for non-Newtonian flows
	\item Lattice Boltzmann Method is good for non-Newtonian flow because the collision operator is entirely local to a node.
	\item Parallelability, simplicity, etc. Allows the non-linear constitutive relationship to be solved in parallel.
	\item Lattice Boltzmann Method can be unstable sometimes :(.
	\item Literature review
	\begin{itemize}
		\item history of collision operators: BGK, MRT, etc.
		\item stability analysis
		\item methods of enhancing stability
		\item non-Newtonian flow studies
	\end{itemize}
\end{itemize}

\subsubsection{Motivation for non-Newtonian Flow Simulation}

\section{Lattice Boltzmann Method}

\subsection{Overview}

Some overview of Lattice Boltzmann Method stuff~\cite{grasinger2015simulation}.

\subsection{The Boltzmann Equation}

\subsection{Hermite Polynomial Expansion}

\subsection{Boundary Conditions}

\subsection{Applied Forces}

\subsection{Collision Operator}

\subsubsection{Bhatnagar--Gross--Krook}

\subsubsection{Multiple Relaxation Time}

\subsubsection{Entropic Collisions and Entropy Balance}

This would be sweet if I somehow found the time to study it \cite{gorban2014enhancement}.

\section{Stability Analysis}

\subsection{von Neumann Stability Analysis}

\subsection{Matrix Method of Stability Analysis}

\subsection{Other Possible Approaches}

\section{Numerical Study}

\subsection{Poiseuille Flow}

\section{Conclusion}

\section*{Acknowledgements}

\bibliographystyle{abbrv}
\bibliography{master}

\end{document}

### 2015-12-10 main.tex
started writing motivation for non-Newtonian flows

@@
 	\begin{itemize}
+		\item non-Newtonian flow studies
 		\item history of collision operators: BGK, MRT, etc.
@@
 		\item methods of enhancing stability
-		\item non-Newtonian flow studies
 	\end{itemize}
@@
 \subsubsection{Motivation for non-Newtonian Flow Simulation}
+
+There are a number of non-Newtonian flows in engineering applications.
+Cement slurries and concretes are best modeled as non-Newtonian fluids because they typically have a non-linear relationship between strain rate and shear stress.
+In particular, cement slurries and concretes can be categorized as yield stress fluids. 
+Yield stress fluids do not flow until a threshold value of stress, refered to as its yield stress, is exceeded.
+Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, in construction, and blah blah blah.
+If the cement job for an oil and gas wellbore is not completed successfully then gasses and fluids can leak through the wellbore to contaminate ground water or the atmosphere ~\cite{grasinger2015simulation}.
+The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies ~\cite{RARR}.
+There are probably some other non-Newtonian form filling problems that are super important such as molten metal and plastics filling molds.
 
@@
 
+\subsection{Lid-driven Flow}
+
 \section{Conclusion}

### 2015-12-14 main.tex
continuing to flesh out introduction

@@
 \usepackage[square,numbers,sort&compress]{natbib}
+\usepackage{soul}
+\usepackage{color}
+\usepackage[utf8]{inputenc}
 
-\title{A Comparision of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}
+\title{An Investigation into the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Fluid Flows}
+%\title{A Comparison of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}
 
@@
 
-%\linenumbers
+\linenumbers
+
+Note: anything in {\color{red}red} is just random garbage notes.
+You can ignore it for the most part.
 
@@
 In this paper, variations of popular collision operators, namely the linear Bhatnagar--Gross--Krook collision operator and the multiple relaxation time collision operator, will be examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
-The structure of the paper will be as follows:
+The structure of the paper will be as follows: blah, blah, blah.
 
-\section{Introduction}
+\section{Introduction} %MG: this intro will potentially be long. Should it be broken up into subsections (as it is now)?
 
+\subsection{Motivation for non-Newtonian Flow Simulation}
+
+{\color{red}
+Some random notes that may fit in to this section eventually:
+\begin{itemize}
+\item ``viscosity'' is a manifestation of friction between fluid particles.
+One would say substances which consist of particles that can more readily "slip" past one another have lower viscosities and substances with fluid particles that tend to have a higher resistance to sliding past one another have higher viscosities.
+
+\item ``Molten plastics, polymer solutions, dyes, varnishes, suspensions, and natural liquids like blood behave in a non-Newtonian manner.'' \cite{bohme1987non}.
+
+\item Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate.
+\end{itemize}
+}
+
+There are a number of non-Newtonian flows in engineering applications.
+Cement slurries and concretes are best modeled as non-Newtonian fluids because they typically have a non-linear relationship between strain rate and shear stress.
+In particular, cement slurries and concretes can be categorized as yield stress fluids. 
+Yield stress fluids do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
+Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, in construction, and blah blah blah.
+If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere ~\cite{grasinger2015simulation}.
+The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies ~\cite{RARR}.
+Yield stress flow is relevant in many other disciplines and applications because of the many substances that can be modeled as Bingham plastics, e.g.
+pastes, paints, muds, molten plastics and metals, and blood \cite{wang2011lattice}.
+%MG: citation is for blood as non-Newtonian
+Non-Newtonian flow is therefor relevant in the engineering and development of pastes and paints, flow and deformation of mud and clay in geotechnical engineering, understanding blood circulation in biomedical applications, materials science and manufacturing of metal and plastic parts, and in particular, the molding and casting of molten metals and plastics.
+
+Power-law fluids are cool.
+Another non-Newtonian flow of interest is shear-thinning and shear-thickening fluids.
+These fluids can be modeled by another non-linear relationship, a power-law relationship.
+Shear-thinning fluids, also known as pseudoplastics, have a lower apparent viscosity at high shear, and a higher apparent viscosity at low shear.
+Shear-thickening fluids exhibit the opposite behavior.
+Examples of shear-thinning fluids are polymer mixes and molten plastics.
+
+Developing numerical methods for approximating these non-Newtonian flows is important because of how difficult it is, and often impossible it is, to derive analytical, closed-form solutions.
+
+\subsection{So why LBM?}
+
+Analytical solutions rarely exists for even the simplest non-Newtonian flows because of the complexity that a non-linear constitutive relationship entails.
+It is more practical to approximate non-Newtonian flows using numerical methods.
+However, the non-linear constitutive equation results in certain challenges for numerical methods as well.
+Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, the general algorithm is:
+
+\begin{enumerate}
+	\item Start with initial guess for the apparent viscosity, $\mu_k = \mu_0$.
+	\item \label{step:solve-for-flow} Solve for the flow using the current value of apparent viscosity, $\dot{\gamma}_k = \tau_k / \mu_k$.
+	\item Update the apparent viscosity, $\mu_{k+1} = \mu(\dot{\gamma}_k)$.
+	\item Return to Step \ref{step:solve-for-flow} until convergence is met.
+\end{enumerate}
+
+Numerical solutions work by discretizing the equations that govern the physics of interest.
+When the solution for fluid flow problems vary in space and time, a numerical approximation requires breaking the problem up into discrete locations and discrete time steps.
+The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which becomes computationally expensive.
+
+The lattice Boltzmann method (LBM) is a numerical method for fluid flow.
+LBM has the unique advantage that the strain rate is local to each node, or each discrete location in the domain.
+This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
+Because high performance computing is shifting from single, sequential processes to massively parallel processes, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flows over other numerical methods for fluid flow.
+%MG: consider moving literature review on LBM applied to non-Newtonian flows HERE
+
+LBM does however, have its drawbacks.
+LBM can be considered as a type of finite-difference scheme for the continuous Boltzmann equation, and as such, has numerical properties in common with finite-difference schemes.
+%MG: is this precise enough?
+accurate enough?
+One such consideration associated with this view, is the possibility of numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
+%MG: are all relevant sources cited?
+Much work has been done to: (1) investigate the stability of lattice Boltzmann methods, (2) enhance the stability of lattice Boltzmann methods, and (3) apply LBM to study non-Newtonian flows.
+
+Sterling and Chen investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice ~\cite{sterling1993stability,sterling1996stability}.
+The investigation was performed by linearizing all nonlinear terms about global equilibrium values and applying a von Neumann stability analysis on the resulting linearized equations.
+The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations ~\cite{sterling1993stability}.
+Sterling and Chen concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
+%MG: this probably requires some clarification
+%(Some misc notes from this study that probably don't fit here but are important to this work: ``...
+a Courant stability condition is superceded by a more stringent stability condition on the speeds.
+Another stability boundary common for finite-difference methods requires that the viscous diffusion speed be less than the lattice spacing divided by the time step.
+This boundary is not observed for LB methods because as the viscosity increases, errors of the scheme increase due to the presence of large nonequilibirum populations, but stability is still maintained''??).
+
+Worthing et. al. ~\cite{worthing1997stability} extended the stability analysis performed by Sterling and Chen.
+Instead of linearizing the LBM equations by expanding about global equilibrium values, Worthing et. al. \ul{does something else}.
+Their analysis discover both physical and nonphysical instabilities, and that setting the BGK relaxation time, $\tau = 1$ provides the optimal accuracy in time. %MG: is this actually relevant though?
+\ul{Expand a little more about what they did and its significance here.}
+%MG: all this stability talk is nice, but I need to figure out how to tie it back to non-Newtonian flows sooner rather than later.
+The problem is a lot of these studies focus on instability due to high Reynolds number flows, not really our problem.
+We're more concerned with instability due to higher relaxation rates and apparent viscosities.
+And how does MRT address this instability?
+
+{\color{red} Bawazeer wrote a thesis on the stability and accuracy of the lattice Boltzmann method. Perhaps I'll read it.}
+
+To improve upon the stability of LB models, d'Humi\`{e}res constructed an LBE model with a multiple-relaxation-time (MRT-LB) collision operator, in moment space \cite{d1994generalized}.
+%MG: make sure you fix the inconsistencies with LB, LBE, etc.
+I think just use LBE or LBM but think it over.
+Literature seems to vary.
+Lallemand and Luo investigated the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LB-MRT method by applying von Neumann stability analysis to it\cite{lallemand2000theory}.
+They concluded LBM with the MRT collision operator was more stable than with the BGK collision operator, but with increased computational expense.
+%MG: I think they have another paper that expands on this study, find and read it.
+Note that although this increased computational expense may not be too significant for Newtonian flows (10-20\% \cite{lallemand2000theory}), the issue is magnified for non-Newtonian flows because the iterative solution for the constitutive equation can require that the non-equilibrium distribution function be mapped back and forth from the velocity space to the momentum space multiple times (the strain rate tensor is determined from the non-equilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
+%MG: clean this sentence up, and maybe break it down.
+Too dense, introducing way too many terms the reader hasn't seen before
+
+{\color{red} If you have time to study/implement entropic LBM (very exciting stuff), a brief discussion of its development would go here.}
+
+Our contribution in all of this mess is to investigate the trade-off between computational efficiency and stability/accuracy for the various LBM schemes.
+The main advantage of LBM in simulating non-Newtonian flows is the local nature of the stress--strain-rate relationship, resulting in a numerical method that can be run massively in parallel \cite{something}.
+This advantage is much less realized if the collision operator is too computationally expensive.
+The aim of this paper is to determine approximate numerical values and simulation scenarios in which one LBM method may be more advantageous than another so that scientists and engineers that make use of LBM for simulating non-Newtonian flows can do so in both a computationally efficient and stable manner.
+%MG: do better than ``simulation scenarios''.
+what does that even mean?
+
+{\color{red}
 \subsection{Scratch}
@@
 \end{itemize}
-
-\subsubsection{Motivation for non-Newtonian Flow Simulation}
-
-There are a number of non-Newtonian flows in engineering applications.
-Cement slurries and concretes are best modeled as non-Newtonian fluids because they typically have a non-linear relationship between strain rate and shear stress.
-In particular, cement slurries and concretes can be categorized as yield stress fluids. 
-Yield stress fluids do not flow until a threshold value of stress, refered to as its yield stress, is exceeded.
-Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, in construction, and blah blah blah.
-If the cement job for an oil and gas wellbore is not completed successfully then gasses and fluids can leak through the wellbore to contaminate ground water or the atmosphere ~\cite{grasinger2015simulation}.
-The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies ~\cite{RARR}.
-There are probably some other non-Newtonian form filling problems that are super important such as molten metal and plastics filling molds.
+}
 
@@
 
-This would be sweet if I somehow found the time to study it \cite{gorban2014enhancement}.
+{\color{red} This would be sweet if I somehow found the time to study it \cite{gorban2014enhancement}.}
 

### 2015-12-21 main.tex
Changed title

@@
 
-\title{An Investigation into the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Fluid Flows}
+\title{Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Fluid Flows}
 %\title{A Comparison of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}
@@
 
-There are a number of non-Newtonian flows in engineering applications.
-Cement slurries and concretes are best modeled as non-Newtonian fluids because they typically have a non-linear relationship between strain rate and shear stress.
+There are a number of non-Newtonian flows in scientific and engineering applications.
+Cement slurries and concretes are best modeled as non-Newtonian fluids because they exhibit a non-linear relationship between strain rate and shear stress.
 In particular, cement slurries and concretes can be categorized as yield stress fluids. 

### 2015-12-22 main.tex
added paragraph on viscosity

@@
 It is well-suited for complex flows such as non-Newtonian, free surface, and multiphase multicomponent flows.
-Non-Newtonian flows are the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten plastics, ETC, are best modeled as non-Newtonian fluids.
+Non-Newtonian flows are the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten plastics, ETC, are best modeled as non-Newtonian fluids, but can lead to numerical instabilities in LBM models.
 LBM is a statistical mechanical approach and each time step consists of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
@@
 }
+
+The viscosity of a fluid is a measure of its resistance to deformation and is due to friction between fluid particles and inter-particle attraction.
+There are two types of viscosity: (1) dynamic viscosity, or shear viscosity, which is the fluid resistance to shear stress, and (2) bulk viscosity, or the fluid's resistance to compressive or tensile stresses.
+Outside of applications in sound, vibrations, and shock waves, the dynamic viscosity of a fluid is of primary interest because the flows that are typically studied are shear stress induced.
+For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
+These fluids are known as Newtonian fluids.
+Non-Newtonian fluids are fluids in which their apparent dynamic viscosity, or apparent resistance to shear stress, is variable--even at a constant temperature--and can be expressed as a function of strain-rate.
+This paper will focus on the flow of non-Newtonian fluids because substances that are best modeled as non-Newtonian fluids are present in many scientific problems and engineering applications, and approximating the flow of non-Newtonian fluids has certain challenges to overcome.
 
@@
 However, the non-linear constitutive equation results in certain challenges for numerical methods as well.
-Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, the general algorithm is:
+Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, a general Picard-type algorithm is:
 

### 2015-12-22 main.tex
Some more edits in the first subsection of the intro
Add: logical break into yield stress and power law fluids before discussing each

@@
 
-The viscosity of a fluid is a measure of its resistance to deformation and is due to friction between fluid particles and inter-particle attraction.
+The viscosity of a fluid is a measure of its resistance to deformation.
 There are two types of viscosity: (1) dynamic viscosity, or shear viscosity, which is the fluid resistance to shear stress, and (2) bulk viscosity, or the fluid's resistance to compressive or tensile stresses.
+Shear resistance, or dynamic viscosity, is predominately due to friction between fluid particles where as bulk viscosity is due to the packing of fluid particles, the contact that ensues, and more weakly, inter-particle potentials.
+% MJG: be more precise here.
 Outside of applications in sound, vibrations, and shock waves, the dynamic viscosity of a fluid is of primary interest because the flows that are typically studied are shear stress induced.
@@
 Non-Newtonian fluids are fluids in which their apparent dynamic viscosity, or apparent resistance to shear stress, is variable--even at a constant temperature--and can be expressed as a function of strain-rate.
-This paper will focus on the flow of non-Newtonian fluids because substances that are best modeled as non-Newtonian fluids are present in many scientific problems and engineering applications, and approximating the flow of non-Newtonian fluids has certain challenges to overcome.
+This paper will focus on the flow of non-Newtonian fluids because substances that are best modeled as non-Newtonian fluids are present in many scientific problems and engineering applications, and approximating the flow of non-Newtonian fluids has certain challenges associated with them.
 
-There are a number of non-Newtonian flows in scientific and engineering applications.
+There are a number of complex fluids present in scientific problems and engineering applications that can be classified as non-Newtonian.
+To name a few, pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids like blood all behave in a non-Newtonian manner \cite{bohme1987non}.
+Of all of the different non-Newtonian behaviors that exist, there are two major categories under which many of the behaviors fall:
+\begin{enumerate}
+	\item \emph{Yield stress fluids}. {\color{red} synopsis of yield stress fluids here}
+	\item \emph{Power-law fluids}. {\color{red} synopsis of power-law fluids here}
+	\item \emph{Other}. Many fluids either behave in a combination of the two ways described above or in another manner entirely.
+\end{enumerate}
+
+Yield stress fluids paragraph
+
+Power law fluids pargraph (lets keep this shit brief)
+
 Cement slurries and concretes are best modeled as non-Newtonian fluids because they exhibit a non-linear relationship between strain rate and shear stress.

### 2015-12-23 main.tex
moving examples to non-Newtonian fluids breakdown enumeration

@@
 \begin{enumerate}
-	\item \emph{Yield stress fluids}. {\color{red} synopsis of yield stress fluids here}
-	\item \emph{Power-law fluids}. {\color{red} synopsis of power-law fluids here}
-	\item \emph{Other}. Many fluids either behave in a combination of the two ways described above or in another manner entirely.
+	\item \emph{Yield stress fluids}. Yield stress fluids do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
+	\item \emph{Power-law fluids}.
+Power-law behavior is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate.
+Shear-thinning fluids are also known as \emph{pseudoplastics}, some examples include polymer mixes and molten plastics.
+	Shear-thickening fluids are also known as \emph{dilatants}, some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
+	\item \emph{Other}. Many fluids have behavior that can be idealized as a combination of the two ways described above. Other fluids behave in another manner entirely.
 \end{enumerate}
@@
 
+Examples paragraph.
+
 Cement slurries and concretes are best modeled as non-Newtonian fluids because they exhibit a non-linear relationship between strain rate and shear stress.
 In particular, cement slurries and concretes can be categorized as yield stress fluids.
-Yield stress fluids do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
+
 Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, in construction, and blah blah blah.

### 2016-01-04 main.tex
Changed the title and restructured non-Newtonian break down

@@
 
-\title{Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Fluid Flows}
+\title{Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flow}
 %\title{A Comparison of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}
@@
 
-\subsection{Motivation for non-Newtonian Flow Simulation}
-
-{\color{red}
-Some random notes that may fit in to this section eventually:
-\begin{itemize}
-\item ``viscosity'' is a manifestation of friction between fluid particles.
-One would say substances which consist of particles that can more readily "slip" past one another have lower viscosities and substances with fluid particles that tend to have a higher resistance to sliding past one another have higher viscosities.
-
-\item ``Molten plastics, polymer solutions, dyes, varnishes, suspensions, and natural liquids like blood behave in a non-Newtonian manner.'' \cite{bohme1987non}.
-
-\item Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate.
-\end{itemize}
-}
+\subsection{Applications of Non-Newtonian Flow}
 
 The viscosity of a fluid is a measure of its resistance to deformation.
-There are two types of viscosity: (1) dynamic viscosity, or shear viscosity, which is the fluid resistance to shear stress, and (2) bulk viscosity, or the fluid's resistance to compressive or tensile stresses.
+There are two types of viscosity: (1) dynamic viscosity, or shear viscosity, which is the fluid's resistance to shear stress, and (2) bulk viscosity, or the fluid's resistance to compressive and tensile stresses.
 Shear resistance, or dynamic viscosity, is predominately due to friction between fluid particles where as bulk viscosity is due to the packing of fluid particles, the contact that ensues, and more weakly, inter-particle potentials.
@@
 Non-Newtonian fluids are fluids in which their apparent dynamic viscosity, or apparent resistance to shear stress, is variable--even at a constant temperature--and can be expressed as a function of strain-rate.
-This paper will focus on the flow of non-Newtonian fluids because substances that are best modeled as non-Newtonian fluids are present in many scientific problems and engineering applications, and approximating the flow of non-Newtonian fluids has certain challenges associated with them.
+This paper will focus on the flow of non-Newtonian fluids because substances that are best modeled as non-Newtonian fluids are present in many scientific problems and engineering applications, and approximating the flow of non-Newtonian fluids has certain challenges.
 
-There are a number of complex fluids present in scientific problems and engineering applications that can be classified as non-Newtonian.
+There are a number of fluids present in scientific problems and engineering applications that can be classified as non-Newtonian.
 To name a few, pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids like blood all behave in a non-Newtonian manner \cite{bohme1987non}.
-Of all of the different non-Newtonian behaviors that exist, there are two major categories under which many of the behaviors fall:
+Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized:
 \begin{enumerate}
-	\item \emph{Yield stress fluids}. Yield stress fluids do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
-	\item \emph{Power-law fluids}.
-Power-law behavior is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate.
-Shear-thinning fluids are also known as \emph{pseudoplastics}, some examples include polymer mixes and molten plastics.
-	Shear-thickening fluids are also known as \emph{dilatants}, some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
-	\item \emph{Other}. Many fluids have behavior that can be idealized as a combination of the two ways described above. Other fluids behave in another manner entirely.
-\end{enumerate}
-
-Yield stress fluids paragraph
-
-Power law fluids pargraph (lets keep this shit brief)
-
-Examples paragraph.
-
-Cement slurries and concretes are best modeled as non-Newtonian fluids because they exhibit a non-linear relationship between strain rate and shear stress.
-In particular, cement slurries and concretes can be categorized as yield stress fluids.
-
-Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, in construction, and blah blah blah.
+	\item \emph{Yield stress fluids}. Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
+	Some examples of yield stress fluids are drilling muds, fresh cement slurries and concretes, and melted metal alloys \cite{ginzburg2002free}.
+	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
 	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere ~\cite{grasinger2015simulation}.
@@
 
-Power-law fluids are cool.
-Another non-Newtonian flow of interest is shear-thinning and shear-thickening fluids.
-These fluids can be modeled by another non-linear relationship, a power-law relationship.
-Shear-thinning fluids, also known as pseudoplastics, have a lower apparent viscosity at high shear, and a higher apparent viscosity at low shear.
-Shear-thickening fluids exhibit the opposite behavior.
-Examples of shear-thinning fluids are polymer mixes and molten plastics.
+	\item \emph{Power-law fluids}.
+Power-law behavior is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate.
+Shear-thinning fluids are also known as \emph{pseudoplastics}, some examples include polymer mixes and molten plastics.
+	Shear-thickening fluids are also known as \emph{dilatants}, some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
+\end{enumerate}
 
@@
 
-\subsection{So why LBM?}
+\subsection{Lattice Boltzmann Method for non-Newtonian Flow}
 
@@
 
+%{\color{red}
+%\subsection{Scratch}
+%
+%Lattice Boltzmann Method (LBM) is the greatest method ever invented.
+%
+%What is my story?:
+%
+%\begin{itemize}
+%	\item Motivation for non-Newtonian flows
+%	\item Lattice Boltzmann Method is good for non-Newtonian flow because the collision operator is entirely local to a node.
+%	\item Parallelability, simplicity, etc. Allows the non-linear constitutive relationship to be solved in parallel.
+%	\item Lattice Boltzmann Method can be unstable sometimes :(.
+%	\item Literature review
+%	\begin{itemize}
+%		\item non-Newtonian flow studies
+%		\item history of collision operators: BGK, MRT, etc.
+%		\item stability analysis
+%		\item methods of enhancing stability
+%	\end{itemize}
+%\end{itemize}
+%}
+
 {\color{red}
-\subsection{Scratch}
+	Some random notes that may fit in to this section eventually:
+	\begin{itemize}
+		\item ``viscosity'' is a manifestation of friction between fluid particles.
+		One would say substances which consist of particles that can more readily "slip" past one another have lower viscosities and substances with fluid particles that tend to have a higher resistance to sliding past one another have higher viscosities.
 
-Lattice Boltzmann Method (LBM) is the greatest method ever invented.
+		\item ``Molten plastics, polymer solutions, dyes, varnishes, suspensions, and natural liquids like blood behave in a non-Newtonian manner.'' \cite{bohme1987non}.
 
-What is my story?:
-
-\begin{itemize}
-	\item Motivation for non-Newtonian flows
-	\item Lattice Boltzmann Method is good for non-Newtonian flow because the collision operator is entirely local to a node.
-	\item Parallelability, simplicity, etc. Allows the non-linear constitutive relationship to be solved in parallel.
-	\item Lattice Boltzmann Method can be unstable sometimes :(.
-	\item Literature review
-	\begin{itemize}
-		\item non-Newtonian flow studies
-		\item history of collision operators: BGK, MRT, etc.
-		\item stability analysis
-		\item methods of enhancing stability
-	\end{itemize}
+		\item Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate.
 	\end{itemize}

### 2016-01-05 main.tex
still fleshing out the intro

@@
 	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
-	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere ~\cite{grasinger2015simulation}.
-	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies ~\cite{RARR}.
+	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere~\cite{grasinger2015simulation}.
+	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies~\cite{RARR}.
 	Yield stress flow is relevant in many other disciplines and applications because of the many substances that can be modeled as Bingham plastics, e.g.
@@
 It is more practical to approximate non-Newtonian flows using numerical methods.
-However, the non-linear constitutive equation results in certain challenges for numerical methods as well.
+However, the non-linear constitutive equation--typically of the form $\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
+%Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate
 Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, a general Picard-type algorithm is:
@@
 The lattice Boltzmann method (LBM) is a numerical method for fluid flow.
-LBM has the unique advantage that the strain rate is local to each node, or each discrete location in the domain.
+LBM has the unique advantage that computing the strain rate is local to each node, or each discrete location in the domain.
 This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
@@
 
-Sterling and Chen investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice ~\cite{sterling1993stability,sterling1996stability}.
+Sterling and Chen investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice~\cite{sterling1993stability,sterling1996stability}.
 The investigation was performed by linearizing all nonlinear terms about global equilibrium values and applying a von Neumann stability analysis on the resulting linearized equations.
-The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations ~\cite{sterling1993stability}.
+The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations~\cite{sterling1993stability}.
 Sterling and Chen concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
-%MG: this probably requires some clarification
+%MG: this probably requires some clarification and a more concrete tieback to the highlevel goals of this paper
 %(Some misc notes from this study that probably don't fit here but are important to this work: ``...
@@
 
-Worthing et. al. ~\cite{worthing1997stability} extended the stability analysis performed by Sterling and Chen.
+Worthing et. al.~\cite{worthing1997stability} extended the stability analysis performed by Sterling and Chen.
 Instead of linearizing the LBM equations by expanding about global equilibrium values, Worthing et. al. \ul{does something else}.
@@
 And how does MRT address this instability?
+
+Although much work has been done to characterize the stability of LBM models, these analyses have the following limitations to their applicability:
+
+\begin{itemize}
+	\item von Neumann stability analysis, and linear stability in general, which is what much of is found in literature, requires modifying the governing equations of the model so that they are linear.
+	This linearization cannot be done without introducing some error into the analysis. %MG: linearization is probably not a word
+	\item Much of the analysis that exists in literature is done about a zero mean flow~\cite{worthing1997stability}. %MG: definitely cite and/or verify this more
+	\item Analyses are most often carried out assuming periodic boundary conditions.
+	Periodic boundary conditions are advantageous because a stability analysis is generally conducted by performing a Fourier analysis on the propagation of error and the periodicity of Fourier modes allows the boundary conditions to automatically be satisfied.
+	Such an analysis, however, does not generalize well to instances when modes of the system are heavily influenced by the imposed boundary conditions~\cite{worthing1997stability}.
+%MG: so many questions: like what is an example of an instance when boundary conditions have a major effect and when are they more ``negligible'' or more like periodic BCs?
+\end{itemize}
 

### 2016-01-05 main.tex
intro: added place holders for future content

@@
 
+{\color{red} maybe we can put a brief lit review on LBM applied to non-Newtonian flows here. THEN discuss some drawbacks/instabilities}
+
 LBM does however, have its drawbacks.
@@
 %MG: are all relevant sources cited?
-Much work has been done to: (1) investigate the stability of lattice Boltzmann methods, (2) enhance the stability of lattice Boltzmann methods, and (3) apply LBM to study non-Newtonian flows.
+Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
 
@@
 Our contribution in all of this mess is to investigate the trade-off between computational efficiency and stability/accuracy for the various LBM schemes.
-The main advantage of LBM in simulating non-Newtonian flows is the local nature of the stress--strain-rate relationship, resulting in a numerical method that can be run massively in parallel \cite{something}.
-This advantage is much less realized if the collision operator is too computationally expensive.
+The main advantage of LBM in simulating non-Newtonian flows is the local nature of the stress--strain-rate relationship, resulting in a numerical method that can be run massively in parallel~\cite{something}.
+This advantage is much less realized if the collision operator is too computationally expensive, or if instability ensues.
 The aim of this paper is to determine approximate numerical values and simulation scenarios in which one LBM method may be more advantageous than another so that scientists and engineers that make use of LBM for simulating non-Newtonian flows can do so in both a computationally efficient and stable manner.
@@
 what does that even mean?
-
-%{\color{red}
-%\subsection{Scratch}
-%
-%Lattice Boltzmann Method (LBM) is the greatest method ever invented.
-%
-%What is my story?:
-%
-%\begin{itemize}
-%	\item Motivation for non-Newtonian flows
-%	\item Lattice Boltzmann Method is good for non-Newtonian flow because the collision operator is entirely local to a node.
-%	\item Parallelability, simplicity, etc. Allows the non-linear constitutive relationship to be solved in parallel.
-%	\item Lattice Boltzmann Method can be unstable sometimes :(.
-%	\item Literature review
-%	\begin{itemize}
-%		\item non-Newtonian flow studies
-%		\item history of collision operators: BGK, MRT, etc.
-%		\item stability analysis
-%		\item methods of enhancing stability
-%	\end{itemize}
-%\end{itemize}
-%}
-
-{\color{red}
-	Some random notes that may fit in to this section eventually:
-	\begin{itemize}
-		\item ``viscosity'' is a manifestation of friction between fluid particles.
-		One would say substances which consist of particles that can more readily "slip" past one another have lower viscosities and substances with fluid particles that tend to have a higher resistance to sliding past one another have higher viscosities.
-
-		\item ``Molten plastics, polymer solutions, dyes, varnishes, suspensions, and natural liquids like blood behave in a non-Newtonian manner.'' \cite{bohme1987non}.
-
-		\item Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate.
-	\end{itemize}
-}
 

### 2016-01-05 main.tex
intro: added ref and brief comments from Niu et. al. 2004

@@
 Shear resistance, or dynamic viscosity, is predominately due to friction between fluid particles where as bulk viscosity is due to the packing of fluid particles, the contact that ensues, and more weakly, inter-particle potentials.
-% MJG: be more precise here.
+% MG: be more precise here.
 Outside of applications in sound, vibrations, and shock waves, the dynamic viscosity of a fluid is of primary interest because the flows that are typically studied are shear stress induced.
@@
 
-\subsection{Lattice Boltzmann Method for non-Newtonian Flow}
+\subsection{Lattice Boltzmann Method for non-Newtonian Flow} %MG: should this be broken down into subsubsections? Maybe like 'Motivation', 'Stability', 'Recent Developments to Address Stability'
 
@@
 The lattice Boltzmann method (LBM) is a numerical method for fluid flow.
-LBM has the unique advantage that computing the strain rate is local to each node, or each discrete location in the domain.
+LBM has the unique advantage that computing the strain rate is local to each node.
 This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
@@
 \begin{itemize}
-	\item von Neumann stability analysis, and linear stability in general, which is what much of is found in literature, requires modifying the governing equations of the model so that they are linear.
+	\item von Neumann stability analysis, and linear stability analysis in general, which is what much of is found in literature, requires modifying the governing equations of the model so that they are linear.
+%MG: is there even a distinction between von Neumann stability analysis and linear stability analysis?
 	This linearization cannot be done without introducing some error into the analysis. %MG: linearization is probably not a word
-	\item Much of the analysis that exists in literature is done about a zero mean flow~\cite{worthing1997stability}. %MG: definitely cite and/or verify this more
+	\item Much of the analysis that exists in literature is done about a uniform flow~\cite{worthing1997stability,niu2004investigation}. %MG: definitely cite and/or verify this more
 	\item Analyses are most often carried out assuming periodic boundary conditions.
 	Periodic boundary conditions are advantageous because a stability analysis is generally conducted by performing a Fourier analysis on the propagation of error and the periodicity of Fourier modes allows the boundary conditions to automatically be satisfied.
-	Such an analysis, however, does not generalize well to instances when modes of the system are heavily influenced by the imposed boundary conditions~\cite{worthing1997stability}.
+	Such an analysis, however, does not generalize well to instances when modes of the system are heavily influenced by the boundary conditions~\cite{worthing1997stability}.
 %MG: so many questions: like what is an example of an instance when boundary conditions have a major effect and when are they more ``negligible'' or more like periodic BCs?
@@
 Literature seems to vary.
-Lallemand and Luo investigated the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LB-MRT method by applying von Neumann stability analysis to it\cite{lallemand2000theory}.
+Lallemand and Luo investigated the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LB-MRT method by applying von Neumann stability analysis to it~\cite{lallemand2000theory}.
 They concluded LBM with the MRT collision operator was more stable than with the BGK collision operator, but with increased computational expense.
@@
 This advantage is much less realized if the collision operator is too computationally expensive, or if instability ensues.
-The aim of this paper is to determine approximate numerical values and simulation scenarios in which one LBM method may be more advantageous than another so that scientists and engineers that make use of LBM for simulating non-Newtonian flows can do so in both a computationally efficient and stable manner.
+The aim of this paper is to determine approximate numerical values and simulation scenarios in which one LBM method may be more advantageous than another so that scientists and engineers that make use of LBM for simulating non-Newtonian flows can do so git in both a computationally efficient and stable manner.
 %MG: do better than ``simulation scenarios''.
@@
 
-\subsection{Collision Operator}
+\subsection{Collision Operator} %MG: if time allows, checkout differential lattice Bolzmann equation, the interpolation-supplemented lattice Boltzmann method, and the Taylor series expansion- and least square-based lattice Boltzmann method; Niu et.
+al.
+2004
 

### 2016-01-12 main.tex
intro: Added Scott to the list of authors :)
added notes and reference for entropic lattice boltzmann method

@@
 \thanks{Email: \textit{grasingerm@pitt.edu}}}
+
+\author{{Scott Overacker}
+	\thanks{Email: \textit{sdo10@pitt.edu}}}
 
@@
 {\color{red} If you have time to study/implement entropic LBM (very exciting stuff), a brief discussion of its development would go here.}
+The price of entropic lattice Boltzmann models for fluid flow is living with a bounded transport coefficient.
+Entropically stabilized collision operators ensure that entropy always increases, i.e. that the second law of thermodynamics is never violated.
+For BGK, this means that the relaxation time is bounded from below in certain instances~\cite{boghosian2001entropic}.
+{\color{red} ``it may be argued that we are relinquishing some control over the transport coefficients'' and ``of course, making $\tau$ a nontrivial function of the incoming state will impact the hydrodynamic equations derived from the model''~\cite{boghosian2001entropic}.
+``stability is guaranteed''?
+Define Lyapunov functional or Lyapunov stability condition.}
 

### 2016-01-26 main.tex
intro edits

@@
 Literature seems to vary.
+Also verify that d'Humieres is correct citation
 Lallemand and Luo investigated the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LB-MRT method by applying von Neumann stability analysis to it~\cite{lallemand2000theory}.
@@
 %MG: I think they have another paper that expands on this study, find and read it.
-Note that although this increased computational expense may not be too significant for Newtonian flows (10-20\% \cite{lallemand2000theory}), the issue is magnified for non-Newtonian flows because the iterative solution for the constitutive equation can require that the non-equilibrium distribution function be mapped back and forth from the velocity space to the momentum space multiple times (the strain rate tensor is determined from the non-equilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
+Note that although this increased computational expense may not be too significant for Newtonian flows (10-20\%~\cite{lallemand2000theory,d2002multiple}), the issue is magnified for non-Newtonian flows because the iterative solution for the constitutive equation can require that the non-equilibrium distribution function be mapped back and forth from the velocity space to the momentum space multiple times (the strain rate tensor is determined from the non-equilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
 %MG: clean this sentence up, and maybe break it down.

### 2016-02-03 main.tex
finished non-Newtonian flow literature review

--- 2016-01-12 main.tex
@@
 \usepackage{lineno}
-\usepackage[square,numbers,sort&compress]{natbib}
 \usepackage{soul}
@@
 \usepackage[utf8]{inputenc}
+\usepackage[square,numbers,sort&compress]{natbib}
 
@@
 Because high performance computing is shifting from single, sequential processes to massively parallel processes, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flows over other numerical methods for fluid flow.
-%MG: consider moving literature review on LBM applied to non-Newtonian flows HERE
 
-{\color{red} maybe we can put a brief lit review on LBM applied to non-Newtonian flows here. THEN discuss some drawbacks/instabilities}
+The lattice Boltzmann method has been studied and applied to modeling various non-Newtonian flows.
+~\cite{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flows.
+The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow, and values from literature for lid-driven cavity flows--showing the feasibility of using LBM models for yield-stress flows.
+LBM models for power-law fluid flows~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
+
+Various schemes and strategies for incorporating the physics of a shear-rate dependent viscosity have been developed and studied.
+The simplest approach, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free}, is to make the relaxation time--which is proportional to apparent viscosity--variable and dependent on the local strain rate.
+A potential issue with the variable relaxation time approach is that as the relaxation time, $\tau$, approaches $\frac{1}{2}$ the viscosity approaches zero and the LBM approximation becomes unstable, or alternatively, if the relaxation time is much greater than one the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
+In order to ensure that the variable relaxation times did not approach values leading to numerical inaccuracies and/or instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable relaxation times.
+Although bounding the relaxation time was shown to be effective, it is also nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the relaxation time does not reflect the proper constitutive relationship.
+Another scheme for incorporating non-Newtonian effects into LBM is to use a constant relaxation rate, typically unity, and to instead incorporate the local shear-rate effect into equilibrium distribution functions.
+This means particle distribution functions will always relax toward equilibrium at the same rate, but that the definition of equilibrium is modified to represent the correct stress--strain-rate relationship.
+The equilibrium distribution function is derived for the constitutive relationship of interest using the Chapman-Enskog multiscale expansion.
+The equilibrium distribution function for Bingham plastic flow was derived, implemented, and verified in~\cite{wang2008lattice}; and the equilibrium distribution functions for power-law fluids and Carreau fluids were derived, implemented, and verified in~\cite{yoshino2007numerical}.
+The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the relaxation time is constant (at unity), the local relaxation time will not approach values that are too small or too large, and so there is no reason to bound the relaxation time in a way that is nonphysical.
+\cite{wang2011lattice} developed another constant-relaxation time LBM scheme for non-Newtonian flows by splitting the constitutive relationship into Newtonian and non-Newtonian parts--for example, considering the plastic viscosity and yield stress for a Bingham plastic separately--then incorporating the Newtonian part in the usual way, namely scaling the relaxation time to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part as a source of momentum, i.e.
+as an external forcing term, that is dependent on local shear-rate.
+The variable relaxation time scheme is used in the present study because of its simplicity, popularity, and flexibility.
 
@@
 
+% MG: list studies using BGK vs. MRT in this paragraph
 To improve upon the stability of LB models, d'Humi\`{e}res constructed an LBE model with a multiple-relaxation-time (MRT-LB) collision operator, in moment space \cite{d1994generalized}.
@@
 
-\bibliographystyle{abbrv}
+\bibliographystyle{abbrvnat}
 \bibliography{master}

### 2016-02-04 main.tex
cleaned up intro

--- 2016-01-12 main.tex
@@
 It is well-suited for complex flows such as non-Newtonian, free surface, and multiphase multicomponent flows.
-Non-Newtonian flows are the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten plastics, ETC, are best modeled as non-Newtonian fluids, but can lead to numerical instabilities in LBM models.
-LBM is a statistical mechanical approach and each time step consists of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
+Non-Newtonian flows are the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten metals and plastics, blood flows, etc., are best modeled as non-Newtonian fluids.
+LBM is a statistical mechanical approach for simulating fluid flow.
+Each time step consists of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
 The collision operator is of interest because it has a lot of control over the physics that are simulated, e.g.
 constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
-In this paper, variations of popular collision operators, namely the linear Bhatnagar--Gross--Krook collision operator and the multiple relaxation time collision operator, will be examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
+In this paper, various collision operators and methods for stability enhancement will be examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
 The structure of the paper will be as follows: blah, blah, blah.
@@
 
-The viscosity of a fluid is a measure of its resistance to deformation.
-There are two types of viscosity: (1) dynamic viscosity, or shear viscosity, which is the fluid's resistance to shear stress, and (2) bulk viscosity, or the fluid's resistance to compressive and tensile stresses.
-Shear resistance, or dynamic viscosity, is predominately due to friction between fluid particles where as bulk viscosity is due to the packing of fluid particles, the contact that ensues, and more weakly, inter-particle potentials.
-% MG: be more precise here.
-Outside of applications in sound, vibrations, and shock waves, the dynamic viscosity of a fluid is of primary interest because the flows that are typically studied are shear stress induced.
+The dynamic viscosity of a fluid is a measure of its resistance to shear deformation.
 For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
 These fluids are known as Newtonian fluids.
-Non-Newtonian fluids are fluids in which their apparent dynamic viscosity, or apparent resistance to shear stress, is variable--even at a constant temperature--and can be expressed as a function of strain-rate.
-This paper will focus on the flow of non-Newtonian fluids because substances that are best modeled as non-Newtonian fluids are present in many scientific problems and engineering applications, and approximating the flow of non-Newtonian fluids has certain challenges.
-
-There are a number of fluids present in scientific problems and engineering applications that can be classified as non-Newtonian.
-To name a few, pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids like blood all behave in a non-Newtonian manner \cite{bohme1987non}.
+Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable--even at a constant temperature--and is often a function of strain-rate.
+There are a number of fluids in science and engineering applications that can be classified as non-Newtonian; pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner \cite{bohme1987non}.
 Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized:
@@
 	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
-	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere~\cite{grasinger2015simulation}.
-	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies~\cite{RARR}.
-	Yield stress flow is relevant in many other disciplines and applications because of the many substances that can be modeled as Bingham plastics, e.g.
+	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere.
+	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies~\cite{definitely cite someone}.
+	Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g.
 pastes, paints, muds, molten plastics and metals, and blood \cite{wang2011lattice}.
 %MG: citation is for blood as non-Newtonian
-	Non-Newtonian flow is therefor relevant in the engineering and development of pastes and paints, flow and deformation of mud and clay in geotechnical engineering, understanding blood circulation in biomedical applications, materials science and manufacturing of metal and plastic parts, and in particular, the molding and casting of molten metals and plastics.
+	As a consequence, simulating yield stress fluids can help engineers develop better pastes and paints, understand the flow and deformation of mud and clay in geotechnical engineering, understand blood circulation, and better manufacture metal and plastic parts.
 
@@
 Power-law behavior is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate.
-Shear-thinning fluids are also known as \emph{pseudoplastics}, some examples include polymer mixes and molten plastics.
-	Shear-thickening fluids are also known as \emph{dilatants}, some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
-\end{enumerate}
+Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include polymer mixes and molten plastics.
+	Shear-thickening fluids are also known as \emph{dilatants}, and some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
+\end{enumerate} %MG: should HB, K-L, Casson, Carreau, and other models be mentioned?
 
-Developing numerical methods for approximating these non-Newtonian flows is important because of how difficult it is, and often impossible it is, to derive analytical, closed-form solutions.
+Developing numerical methods for approximating these non-Newtonian flows is important because of how difficult, and often impossible, it is to derive analytical, closed-form solutions.
 
@@
 Analytical solutions rarely exists for even the simplest non-Newtonian flows because of the complexity that a non-linear constitutive relationship entails.
-It is more practical to approximate non-Newtonian flows using numerical methods.
+It is generally more practical to approximate non-Newtonian flows using numerical methods.
 However, the non-linear constitutive equation--typically of the form $\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
@@
 The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which becomes computationally expensive.
+The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the unique advantage that computing the strain rate is local to each node.
+This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
+Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flows over other numerical methods.
 
-The lattice Boltzmann method (LBM) is a numerical method for fluid flow.
-LBM has the unique advantage that computing the strain rate is local to each node.
-This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
-Because high performance computing is shifting from single, sequential processes to massively parallel processes, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flows over other numerical methods for fluid flow.
-%MG: consider moving literature review on LBM applied to non-Newtonian flows HERE
-
-{\color{red} maybe we can put a brief lit review on LBM applied to non-Newtonian flows here. THEN discuss some drawbacks/instabilities}
+The lattice Boltzmann method has been studied and successively applied to modeling various non-Newtonian flows.
+~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flows.
+The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow, and values from literature for lid-driven cavity flows--showing the feasibility of using LBM models for yield-stress flows.
+LBM models for power-law fluid flows~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
 
@@
 accurate enough?
-One such consideration associated with this view, is the possibility of numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
+One such consideration associated with this view, is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
 %MG: are all relevant sources cited?
-Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
+%Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
+Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the strong nonlinear relationship between shear stress and strain-rate can lead to high non-equilibrium fluctuations.
+%MG: be more precise than 'fluctuations'?
+Various schemes and strategies for incorporating the physics of non-Newtonian flows and yet maintaining a stable numerical method have been developed and studied.
+The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free}, is to make the relaxation time--which is proportional to apparent viscosity--variable and dependent on the local strain rate.
+A potential issue with the stability of the variable relaxation time approach is that as the relaxation time, $\tau$, approaches $1/2$ the viscosity approaches zero; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
+In order to ensure that the variable relaxation times did not approach values leading to numerical instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable relaxation times.
+Although bounding the relaxation time was shown to be effective in terms of stability, it is also nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the relaxation time does not reflect the proper constitutive relationship of interest.
+Another scheme for incorporating non-Newtonian effects into LBM is to use a constant relaxation rate, typically unity, and to instead incorporate the local shear-rate effect into equilibrium distribution functions.
+This means particle distribution functions will always relax toward equilibrium at the same rate, but that the definition of equilibrium is modified to represent the correct stress--strain-rate relationship.
+The equilibrium distribution function is derived for the specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.
+The equilibrium distribution function for Bingham plastic flow was derived, implemented, and verified in~\cite{wang2008lattice}; and the equilibrium distribution functions for power-law fluids and Carreau fluids were derived, implemented, and verified in~\cite{yoshino2007numerical}.
+The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the relaxation time is constant (at unity), the local relaxation time will not approach values that are too small or too large, and so there is no reason to bound the relaxation time in a way that is nonphysical.
+\citet{wang2011lattice} developed another constant-relaxation time LBM scheme for non-Newtonian flows by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts--for example, considering the plastic viscosity and yield stress for a Bingham plastic separately--then incorporating the Newtonian part in the usual way, namely scaling the relaxation time to achieve the macroscopic (Newtonian) viscosity, and incorporating the non-Newtonian part as a source of momentum, i.e.
+as an external forcing term, that is dependent on local shear-rate.
+Although the constant-relaxation time strategies present interesting alternatives, the variable relaxation time scheme is used in the present study because of its simplicity and flexibility.
+%MG: should we define what we mean by 'flexibility'?
 
-Sterling and Chen investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice~\cite{sterling1993stability,sterling1996stability}.
+Much work has been done (not specific to non-Newtonian flows) to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
+\citet{sterling1993stability,sterling1996stability} investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice.
 The investigation was performed by linearizing all nonlinear terms about global equilibrium values and applying a von Neumann stability analysis on the resulting linearized equations.
-The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations~\cite{sterling1993stability}.
-Sterling and Chen concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
+The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations.
+\citeauthor{sterling1993stability} concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
 %MG: this probably requires some clarification and a more concrete tieback to the highlevel goals of this paper
@@
 This boundary is not observed for LB methods because as the viscosity increases, errors of the scheme increase due to the presence of large nonequilibirum populations, but stability is still maintained''??).
-
-Worthing et. al.~\cite{worthing1997stability} extended the stability analysis performed by Sterling and Chen.
-Instead of linearizing the LBM equations by expanding about global equilibrium values, Worthing et. al. \ul{does something else}.
+~\citet{worthing1997stability} extended the stability analysis performed by \citeauthor{sterling1993stability}, and instead of linearizing the LBM equations by expanding about global equilibrium values, \citeauthor{worthing1997stability} \ul{does something else}.
 Their analysis discover both physical and nonphysical instabilities, and that setting the BGK relaxation time, $\tau = 1$ provides the optimal accuracy in time. %MG: is this actually relevant though?
@@
 Although much work has been done to characterize the stability of LBM models, these analyses have the following limitations to their applicability:
-
 \begin{itemize}
-	\item von Neumann stability analysis, and linear stability analysis in general, which is what much of is found in literature, requires modifying the governing equations of the model so that they are linear.
+	\item von Neumann stability analysis, and linear stability analysis in general requires modifying the governing equations of the model so that they are linear.
 %MG: is there even a distinction between von Neumann stability analysis and linear stability analysis?
-	This linearization cannot be done without introducing some error into the analysis. %MG: linearization is probably not a word
+	This linearization cannot be done without introducing some error into the analysis.
 	\item Much of the analysis that exists in literature is done about a uniform flow~\cite{worthing1997stability,niu2004investigation}. %MG: definitely cite and/or verify this more
 	\item Analyses are most often carried out assuming periodic boundary conditions.
-	Periodic boundary conditions are advantageous because a stability analysis is generally conducted by performing a Fourier analysis on the propagation of error and the periodicity of Fourier modes allows the boundary conditions to automatically be satisfied.
+	Periodic boundary conditions are advantageous for stability analysis because the analysis is generally done by a Fourier analysis on the propagation of error.
+	The periodicity of Fourier modes allows the periodic boundary conditions to automatically be satisfied.
 	Such an analysis, however, does not generalize well to instances when modes of the system are heavily influenced by the boundary conditions~\cite{worthing1997stability}.
 %MG: so many questions: like what is an example of an instance when boundary conditions have a major effect and when are they more ``negligible'' or more like periodic BCs?
+	\item The analyses are general and do not delve into any of the nuances of non-Newtonian flows.
 \end{itemize}
+{\color{red} Bawazeer wrote a thesis on the stability and accuracy of the lattice Boltzmann method. Perhaps I'll read it. I think Jonas Latt did too.}
 
-{\color{red} Bawazeer wrote a thesis on the stability and accuracy of the lattice Boltzmann method. Perhaps I'll read it.}
+To improve upon the stability of LBM models, \citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
+\citet{lallemand2000theory} used von Neumann stability analysis to investigate the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LBM-MRT model.
+They concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
+Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain aspects of the collision operator be performed at each iteration (the strain rate tensor is determined from the non-equilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
+Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
+\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
 
-To improve upon the stability of LB models, d'Humi\`{e}res constructed an LBE model with a multiple-relaxation-time (MRT-LB) collision operator, in moment space \cite{d1994generalized}.
-%MG: make sure you fix the inconsistencies with LB, LBE, etc.
-I think just use LBE or LBM but think it over.
-Literature seems to vary.
-Lallemand and Luo investigated the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LB-MRT method by applying von Neumann stability analysis to it~\cite{lallemand2000theory}.
-They concluded LBM with the MRT collision operator was more stable than with the BGK collision operator, but with increased computational expense.
-%MG: I think they have another paper that expands on this study, find and read it.
-Note that although this increased computational expense may not be too significant for Newtonian flows (10-20\% \cite{lallemand2000theory}), the issue is magnified for non-Newtonian flows because the iterative solution for the constitutive equation can require that the non-equilibrium distribution function be mapped back and forth from the velocity space to the momentum space multiple times (the strain rate tensor is determined from the non-equilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
-%MG: clean this sentence up, and maybe break it down.
-Too dense, introducing way too many terms the reader hasn't seen before
-
-{\color{red} If you have time to study/implement entropic LBM (very exciting stuff), a brief discussion of its development would go here.}
+{\color{red} 
+ELBM and entropic filtering stuff goes here.
 The price of entropic lattice Boltzmann models for fluid flow is living with a bounded transport coefficient.
@@
 For BGK, this means that the relaxation time is bounded from below in certain instances~\cite{boghosian2001entropic}.
-{\color{red} ``it may be argued that we are relinquishing some control over the transport coefficients'' and ``of course, making $\tau$ a nontrivial function of the incoming state will impact the hydrodynamic equations derived from the model''~\cite{boghosian2001entropic}.
+``it may be argued that we are relinquishing some control over the transport coefficients'' and ``of course, making $\tau$ a nontrivial function of the incoming state will impact the hydrodynamic equations derived from the model''~\cite{boghosian2001entropic}.
 ``stability is guaranteed''?
-Define Lyapunov functional or Lyapunov stability condition.}
+Define Lyapunov functional or Lyapunov stability condition.
+}
 
-Our contribution in all of this mess is to investigate the trade-off between computational efficiency and stability/accuracy for the various LBM schemes.
-The main advantage of LBM in simulating non-Newtonian flows is the local nature of the stress--strain-rate relationship, resulting in a numerical method that can be run massively in parallel~\cite{something}.
-This advantage is much less realized if the collision operator is too computationally expensive, or if instability ensues.
-The aim of this paper is to determine approximate numerical values and simulation scenarios in which one LBM method may be more advantageous than another so that scientists and engineers that make use of LBM for simulating non-Newtonian flows can do so git in both a computationally efficient and stable manner.
-%MG: do better than ``simulation scenarios''.
-what does that even mean?
+The goal of this paper is to numerically study the implications of accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flows using LBM.
+The intention of the study is to aid scientists and engineers in understanding which strategy to choose given their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
+Advantages, such as LBM's potential to scale well in parallel, can be much less realized if the collision operator is too computationally expensive, or if numerical instabilities ensue.
+A numerical study can help to determine approximate numerical values, domains, and boundary conditions in which one LBM scheme may be more advantageous than another so that LBM may be used in a computationally efficient and stable manner.
 
@@
 \subsection{The Boltzmann Equation}
-
-\subsection{Hermite Polynomial Expansion}
 
@@
 \subsubsection{Entropic Collisions and Entropy Balance}
-
-{\color{red} This would be sweet if I somehow found the time to study it \cite{gorban2014enhancement}.}
 

### 2016-02-29 main.tex
added a few minor comments

@@
 The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which becomes computationally expensive.
-The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the unique advantage that computing the strain rate is local to each node.
+The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the advantage that computing the strain rate is second-order accurate and local to each node~\cite{kruger2010second}.
 This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
@@
 
+%MG: probably include a discussion of the work done by Conrad et. al. 2015 here
 LBM does however, have its drawbacks.

### 2016-04-27 main.tex
started methods section

@@
 
+The Lattice Boltzmann method is a numerical approach that uses statistical mechanics to represent a variety of physical processes, such as fluid flow.
+More specifically, LBM can be thought of as a special finite difference discretization of the lattice Boltzmann equation~\cite{chen1998lattice}.
+The length scale of LBM is unique in contrast to most common numerical methods, and is referred to as the mesoscale.
+In contrast to continuum based methods, LBM simulates the kinetics of microscopic particles, and so it reaches a finer length scale than the macroscopic domain of continuum mechanics; and in contrast to molecular dynamics, discrete element method, and other particle scale approaches, LBM does not deal with a complete description of the degrees of freedom for each individual particle.
+LBM instead relies on a statistical description of particle distributions, making LBM, in general, more computationally efficient and requiring less memory than other particle methods.
+In addition, the Boltzmann equation is continuous, so LBM inherits the smoothness and stability of continuum methods.
+Thus, LBM can be seen as a compromise between continuum and particle methods, combining strengths from each.
+
+LBM has some advantages over other methods of CFD.
+For example, LBM is a computationally efficient approach for some CFD.
+This efficiency is a consequence of two distinct features of LBM: (1) the convective operator is linear, as opposed to the nonlinear convection terms that appear in continuum mechanics approaches; and (2) the fluid pressure is given by an equation of state.
+Solving for the fluid pressure in traditional method sis more computationally expensive and requires special treatment such as iteration and/or relaxation~\cite{chen1998lattice}.
+
 \subsection{The Boltzmann Equation}

### 2016-04-29 main.tex
Began numerical study section

--- 2016-02-29 main.tex
@@
 \section{Numerical Study}
+Two different flow simulations, Poiseuille flow and lid-driven flow, and two different non-Newtonian behaviors, yield stress fluids and power-law fluids,  served as the basis for the numerical study. 
+\subsection{Poiseuille Flow} 
+Computational Efficiency
 
-\subsection{Poiseuille Flow}
+Numerical Stability
+\subsection{Lid-driven Flow}
+Lid-driven cavity flow was simulated using Bingham numbers of 1, 10, and 100 for Reynolds numbers of 100, 1000, 5000, and 10000.
+In the simulations, a constant velocity of $u_0$ was fixed at 0.1 across the lid.
+No slip boundary conditions were fixed to the remaining walls of the cavity.
+To validate the model used for the comparison, the results had to be compared with results from [...,...,...].
+As shown in the table, it can be seen that the results of the present simulations match the benchmark results well (hopefully).
 
-\subsection{Lid-driven Flow}
+Computational Efficiency/Time
+
+Numerical Stability
+
+Effect of entropic filtering
+
+Effect of m (stress growth exponent)
 

### 2016-05-13 main.tex
Reverted to contents of previous version
Began numerical study section of paper
added commands for LBM variables

--- 2016-04-27 main.tex
@@
 \usepackage[square,numbers,sort&compress]{natbib}
+
+\usepackage{amsthm}
+\usepackage{amssymb}
+\usepackage{mathtools}
+\usepackage{mathabx}
+\usepackage{bm}
+
+\newcommand{\pos}{\mathbf{x}}
+\newcommand{\pvel}{\boldsymbol{\xi}}
+\newcommand{\mvel}{\mathbf{u}}
+\newcommand{\copt}{\boldsymbol{\Omega}}
+\newcommand{\grad}{\boldsymbol{\nabla}}
 
@@
 This efficiency is a consequence of two distinct features of LBM: (1) the convective operator is linear, as opposed to the nonlinear convection terms that appear in continuum mechanics approaches; and (2) the fluid pressure is given by an equation of state.
-Solving for the fluid pressure in traditional method sis more computationally expensive and requires special treatment such as iteration and/or relaxation~\cite{chen1998lattice}.
+Solving for the fluid pressure in traditional method is more computationally expensive and requires special treatment such as iteration and/or relaxation~\cite{chen1998lattice}.
 
 \subsection{The Boltzmann Equation}
+The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
+The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \copt$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\copt$ is the collision operator.
+The lattice part of LBM refers to the way in which the BE is discretized.
+The lattice discretizes the spatial domain with nodes that are connected to their neighbors through discrete lattice velocity vectors.
+The velocity vectors act as pathways for particle distributions to travel along.
+Each time step in LBM consists of two distinct actions:
+
+\begin{itemize}
+\item Streaming: particle distribution propagate to their neighbors along the lattice velocity vectors.
+The particles can only move along the vectors in their specified direction and can only move at a specific speed.
+\item Collision: particle distributions meet at a node and ``collide''.
+In LBM, collisions are not simulated in a realistic sense, meaning that each individual particle does not exist and glance off of, or interact with, one another.
+Instead the collision operator is formulated in such a way that particle distributions are relaxed toward equilibrium.
+What defines equilibrium depends on the mechanics of interest to be modeled.
+\end{itemize}
+
+The D2Q9 lattice was used in the current work (shown in Figure \ref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}.
+% MG: do we even need this citation?
+The lattice is two-dimensional with nine discrete velocities at each node.
+There is a stationary particle, there are four discrete velocities of magnitude 1, $(\pm1,0)$, $(0,\pm1)$, and there are four discrete velocities of magnitude $\sqrt{2}$, $(\pm1, \pm1)$.
+
+\begin{figure}
+  \includegraphics[]{figs/d2q9}
+  \caption{D2Q9 lattice}
+  \label{fig:d2q9}
+\end{figure}
 

### 2016-05-15 main.tex
Added to the lid-driven analysis section

--- 2016-04-29 main.tex
@@
 
-\usepackage[affil-it]{authblk}
-\usepackage{lineno}
-\usepackage{soul}
-\usepackage{color}
-\usepackage[utf8]{inputenc}
-\usepackage[square,numbers,sort&compress]{natbib}
+%\usepackage[affil-it]{authblk}
+%\usepackage{lineno}
+%\usepackage{soul}
+%\usepackage{color}
+%\usepackage[utf8]{inputenc}
+%\usepackage[square,numbers,sort&compress]{natbib}
 
@@
 \section{Numerical Study}
-Two different flow simulations, Poiseuille flow and lid-driven flow, and two different non-Newtonian behaviors, yield stress fluids and power-law fluids,  served as the basis for the numerical study. 
+Two different flow simulations, Poiseuille flow and lid-driven flow, and two different non-Newtonian behaviors, yield stress fluids and power-law fluids,  served as the basis for the numerical study comparing LBM for non-Newtonian fluid flow simulation.
+Within these physical models and constitutive relationships, multiple parameters were evaluated.
 \subsection{Poiseuille Flow} 
@@
 \subsection{Lid-driven Flow}
-Lid-driven cavity flow was simulated using Bingham numbers of 1, 10, and 100 for Reynolds numbers of 100, 1000, 5000, and 10000.
 In the simulations, a constant velocity of $u_0$ was fixed at 0.1 across the lid. 
@@
 To validate the model used for the comparison, the results had to be compared with results from [...,...,...]. 
-As shown in the table, it can be seen that the results of the present simulations match the benchmark results well (hopefully).
+As shown in the table (show table of our vortex centers compared with other papers), it can be seen that the results of the present simulations match the benchmark results well.
+The previous results of [...] showed grid independence between 256x256 grids and 512x512 grids, and it was found that stability or accuracy were not markedly enhanced between a 256x256 grid and a 100x100 grid for material properties evaluated in these tests.
+For computational efficiency, subsequent tests were performed on a 100x100 grid.
+
+For the yield stress lid-driven cavity flow, Bingham numbers of 1, 10, and 100 for Reynolds numbers of 100, 1000, 5000, and 10000 were used for the simulations that varied the stress growth parameter, $m$, and utilized either entropic filtering, placing bounds of 0.501 and 8.0 on the relaxation time, or no stability enhancement strategies.
+%SDO-What are the units here, seconds?
+Stability enhancements were used on both the BGK and MRT collision operators.
+For all tests, more stable and efficient results were found when the flow was closer to laminar Newtonian flow behavior. 
+Additionally, with these fluid properties the effect of the stress growth parameter and any stability enhancement was negligible in terms of stability or accuracy.  
+All collision operators with or without stability enhancements yielded very similar streamlines, stream functions, and locations and strengths of the primary vortex.
+However, computational time between the tests varied widely, with the tests using a stress growth exponent of $10^8$ taking about 10\% longer than the tests with a stress growth exponent of $10^5$.  
+Bounding the relaxation time actually increased the computational time by about 5\% and using entropic filtering with the BGK operator decreased computational time by nearly 18\%.
+Finally, using the MRT operator with or without entropic filtering decreased the computational efficiency by about 300\%.
+
+As the Reynolds number and Bingham numbers increased when plastic viscosity was decreased and yield stress was increased, respectively, it was seen that the stability of flow models conducted with the BGK operator lost stability and accuracy while maintaining computational efficiency.
+With $Re \geq 5000$, the tests using the standard BGK operator and the BGK operator with bounds on the relaxation time were unstable to the point that the solution was divergent, regardless of the Bingham number.
+Entropic filtering stabilized the BGK collision operator tests effectively, yielding streamlines and a stream function that was mimicked the results of the tests using the MRT operator and previous results [...] well.
+%SDO-Good place to show a figure.
+However, as the Reynolds and Bingham numbers increased, the location and strength of the entropic-filtered BGK operator primary vortex deviated from that determined with the MRT operator, and the smoothness of the stream function of the MRT operator was lost.
+%SDO-Another good place for a picture, maybe a table showing the locations and strengths of the central vortices.
+As has been found in previous studies [...,...], the multiple-relaxation time operator markedly increased stability and computational inefficiency. 
+With the relatively coarse grid used for this study, the results never became unstable and the efficiency was about a third of that of the BGK operator. 
+For finer grids the margin increased further, and the MRT operator required substantially more iterations to reach a steady-state solution (actually a steady state condition was never met for any MRT test).
 

### 2016-05-18 main.tex
working through methods section

--- 2016-05-13 main.tex
@@
 \newcommand{\mvel}{\mathbf{u}}
-\newcommand{\copt}{\boldsymbol{\Omega}}
+\newcommand{\cop}{\boldsymbol{\Omega}}
 \newcommand{\grad}{\boldsymbol{\nabla}}
+\newcommand{\mmom}{\mathbf{j}}
 
@@
 The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
-The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \copt$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\copt$ is the collision operator.
+The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \cop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\cop$ is the collision operator.
 The lattice part of LBM refers to the way in which the BE is discretized.
@@
 
+The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as: $f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \cop_i(f)$, where, for D2Q9, $i = 0, 1, ..., 8$.
+
+The macroscopic variables of interest can be calculated from the particle distribution functions, $f(\pos, \pvel, t)$, by integrating moments of $f$ over velocity space.
+Due to the discrete nature of velocity in LBM, the integrals simply become summations.
+The mass density is given by the sum of the particle distributions and the momentum density is given by the first moment of the particle distributions over the velocity space:
+
+\begin{align}
+\label{eq:rho} \rho(\pos, t) =& \sum_{i} f_i(\pos, t), \\
+\label{eq:mom} \mmom(\pos, t) = \rho(\pos, t) \mvel(\pos, t) =& \sum_{i} \pvel_i(\pos, t) f_i(\pos, t)
+\end{align}
+
+The fluid pressure is related to macroscopic density through an equation of state:
+
+\begin{equation}
+\label{eq:pres} p(\pos, t) = \rho(\pos, t) c_s^2,
+\end{equation}
+
+\noindent where $c_s$ is the lattice speed of sound ($c_s = \frac{1}{\sqrt{3}}$, for D2Q9).
+
 \subsection{Boundary Conditions}
 
+No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
+For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
+For velocity or pressure boundary conditions, the method proposed by \citet{zou1997pressure} can be used.
+The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the non-equilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
+
 \subsection{Applied Forces}
+
+Incorporating external forces, such as gravity, pressure gradients, etc., is done by adding a source of particle distributions in the direction of the force.
+The increase in particle distributions leads to the desired macroscopic result--an increase in momentum.
+The LBE with external forces is:
+
+\begin{equation}
+f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \Omega_i(f) + \frac{w_i \Delta t}{c_s^2} \mathbf{F} \cdot \pvel_i
+\end{equation}
+
+\noindent where $\mathbf{F}$ is the body force vector.
 
@@
 
-\subsubsection{Bhatnagar--Gross--Krook}
+The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision) \cite{Cer90}.
+In LBM, the collision operator causes particle distributions to relax toward a local equilibrium. This equilibrium is determined by the macroscopic physical behavior of interest.
+In the case of incompressible flow, the equilibrium particle distribution, $f_i^{eq}$, is given by:
 
-\subsubsection{Multiple Relaxation Time}
+\begin{equation}
+f_i^{eq} = w_i \rho \left[1 + \frac{\pvel_i \cdot \mvel}{c_s^2} + \frac{(\pvel_i \cdot \mvel)^2}{2c_s^2} - \frac{\mvel^2}{2c_s^2} \right]
+\end{equation}
+
+\noindent where $w_i$ is the weight in the $i^{th}$ direction.
+
+\begin{equation*}
+w_i = \begin{cases}
+    \frac{4}{9}, & i = 0 \\
+    \frac{1}{9}, & i = 1, 2, 3, 4 \\
+    \frac{1}{36}, & i = 5, 6, 7, 8
+\end{cases}
+\end{equation*}
+
+Due to its simplicity and computational efficiency, the most common collision operator is the Bhatnagar--Gross--Krook (BGK) operator.
+BGK consists of a single relaxation time and is a linear relaxation of particle distributions toward equilibrium.
+The BGK is expressed as:
+
+\begin{equation}
+\Omega_i = -\frac{1}{\tau} (f_i(\pos, t) - f_i^{eq}),
+\end{equation}
+
+\noindent where $\tau$ is the relaxation time \cite{Bha54}.
+The relaxation time can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
+For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\tau - \frac{1}{2})$.
 
 \subsubsection{Entropic Collisions and Entropy Balance}
-
-\section{Stability Analysis}
-
-\subsection{von Neumann Stability Analysis}
-
-\subsection{Matrix Method of Stability Analysis}
-
-\subsection{Other Possible Approaches}
 

### 2016-06-28 main.tex
Writing LBM section

@@
 \usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \usepackage[square,numbers,sort&compress]{natbib}
@@
 \newcommand{\mmom}{\mathbf{j}}
+\newcommand{\transM}{\mathbf{M}}
 
@@
 
+An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
+%In the BGK collision scheme, high values of viscosity--such as those found in some cement slurry mixes--can lead to numerical instabilities \cite{svec2011flow,lallemand2000theory}.
+Because the apparent viscosity can take on very large values without leading to numerical instabilities, the MRT collision operator has become popular for simulating non-Newtonian flows \cite{chen2014simulations,fallah2012multiple,tang2011bingham,vikhansky2008lattice,chai2011multiple}.
+%MG: is this necessary?
+wasn't this already covered
+The MRT collision operator is given by:
+
+\begin{equation} \label{eq:mrt-col-op}
+	\cop = - {\transM}^{-1} \relaxM \transM (\mathbf{f} - \mathbf{f}^{eq}),
+\end{equation}
+
+where $\transM$ is a transformation matrix that maps the particle distribution vector, $\mathbf{f}$, and equilibrium distribution vector, $\mathbf{f}^{eq}$, from the particle distribution space into moment space, they will be denoted by $\mathbf{m}$ and $\mathbf{m}^{eq}$, respectively.
+The relationships between $\mathbf{m}$, $\transM$ and $\mathbf{f}$ can be written as follows:
+
+\begin{equation}
+\mathbf{m} = \begin{bmatrix}
+\rho \\ e \\ \epsilon \\ j_x \\ q_x \\ j_y \\ q_y \\ p_{xx} \\ p_{xy}
+\end{bmatrix} = \begin{bmatrix}
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
+\end{bmatrix}
+\end{equation}
+
 \subsubsection{Entropic Collisions and Entropy Balance}

### 2016-06-29 main.tex
more LBM methods

@@
 
-where $\transM$ is a transformation matrix that maps the particle distribution vector, $\mathbf{f}$, and equilibrium distribution vector, $\mathbf{f}^{eq}$, from the particle distribution space into moment space, they will be denoted by $\mathbf{m}$ and $\mathbf{m}^{eq}$, respectively.
+\noindent where $\transM$ is a transformation matrix that maps the particle distribution vector, $\mathbf{f}$, and equilibrium distribution vector, $\mathbf{f}^{eq}$, from the particle distribution space into moment space, they will be denoted by $\mathbf{m}$ and $\mathbf{m}^{eq}$, respectively.
 The relationships between $\mathbf{m}$, $\transM$ and $\mathbf{f}$ can be written as follows:
@@
 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
+-4 & -1 & -1 & -1 & -1 & 2 & 2 & 2 & 2 \\
+4 & -2 & -2 & -2 & -2 & 1 & 1 & 1 & 1 \\
+0 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+0 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+0 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+0 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+0 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+0 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
 \end{bmatrix}

### 2016-07-03 main.tex
finished transformation matrix
More or less finished first draft of methods section Starting results section

--- 2016-05-18 main.tex
@@
+% MG: replace \ \cite with ~\cite to ensure citations stay on same line
+% MG: need an overall consistency in 'flucuation' vs. 'oscillation'; use same word if you mean same thing.
+% MG: same note as above for 'local equilibrium' vs. 'quasiequilibrium'
+% MG: consider using collision frequency, \omega, instead of relaxation time, \tau, so that the relaxation time is not confused with shear stress
+
 \documentclass{article}
@@
 \usepackage{bm}
+\usepackage[plain]{fancyref}
 
@@
 \newcommand{\mvel}{\mathbf{u}}
-\newcommand{\cop}{\boldsymbol{\Omega}}
+\newcommand{\colop}{\boldsymbol{\Omega}}
 \newcommand{\grad}{\boldsymbol{\nabla}}
 \newcommand{\mmom}{\mathbf{j}}
+\newcommand{\transM}{\mathbf{M}}
+\newcommand{\relaxM}{\mathbf{S}}
 
-\title{Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flow}
+\newcommand{\expnumber}[2]{{#1}\mathrm{e}{#2}}
+
+\title{Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows}
 %\title{A Comparison of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}
@@
 \pagenumbering{arabic}
-
-\linenumbers
 
@@
 	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
-	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere.
+	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere~\cite{grasinger2015simulation}.
+%MG: is it tacky to cite myself here?
 	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies~\cite{definitely cite someone}.
@@
 	Shear-thickening fluids are also known as \emph{dilatants}, and some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
-\end{enumerate} %MG: should HB, K-L, Casson, Carreau, and other models be mentioned?
+\end{enumerate}
 
@@
 
-\subsection{Lattice Boltzmann Method for non-Newtonian Flow} %MG: should this be broken down into subsubsections? Maybe like 'Motivation', 'Stability', 'Recent Developments to Address Stability'
+\subsection{Lattice Boltzmann Method for non-Newtonian Flow} \label{sec:lbm-for-nnf} %MG: should this be broken down into subsubsections?
+Maybe like 'Motivation', 'Stability', 'Recent Developments to Address Stability'
 
-Analytical solutions rarely exists for even the simplest non-Newtonian flows because of the complexity that a non-linear constitutive relationship entails.
+Analytical solutions rarely exists for even the simplest non-Newtonian flows because of the complexity that a nonlinear constitutive relationship entails.
 It is generally more practical to approximate non-Newtonian flows using numerical methods.
-However, the non-linear constitutive equation--typically of the form $\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
+However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
+\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma},
+\end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
 %Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate
@@
 %Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
-Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the strong nonlinear relationship between shear stress and strain-rate can lead to high non-equilibrium fluctuations.
+Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the strong nonlinear relationship between shear stress and strain-rate can lead to high nonequilibrium fluctuations.
 %MG: be more precise than 'fluctuations'?
@@
 The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free}, is to make the relaxation time--which is proportional to apparent viscosity--variable and dependent on the local strain rate.
-A potential issue with the stability of the variable relaxation time approach is that as the relaxation time, $\tau$, approaches $1/2$ the viscosity approaches zero; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
+A potential issue with the stability of the variable relaxation time approach is that as the collision frequency, $\omega$, approaches $2$ the viscosity approaches zero; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
 In order to ensure that the variable relaxation times did not approach values leading to numerical instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable relaxation times.
@@
 ~\citet{worthing1997stability} extended the stability analysis performed by \citeauthor{sterling1993stability}, and instead of linearizing the LBM equations by expanding about global equilibrium values, \citeauthor{worthing1997stability} \ul{does something else}.
-Their analysis discover both physical and nonphysical instabilities, and that setting the BGK relaxation time, $\tau = 1$ provides the optimal accuracy in time. %MG: is this actually relevant though?
+Their analysis discover both physical and nonphysical instabilities, and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time.
+%MG: is this actually relevant though?
 \ul{Expand a little more about what they did and its significance here.}
@@
 The problem is a lot of these studies focus on instability due to high Reynolds number flows, not really our problem.
-We're more concerned with instability due to higher relaxation rates and apparent viscosities.
+We're more concerned with instability due to lower relaxation rates and apparent viscosities.
 And how does MRT address this instability?
@@
 They concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
-Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain aspects of the collision operator be performed at each iteration (the strain rate tensor is determined from the non-equilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
+Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain aspects of the collision operator be performed at each iteration (the strain rate tensor is determined from the nonequilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
 Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
@@
 For BGK, this means that the relaxation time is bounded from below in certain instances~\cite{boghosian2001entropic}.
-``it may be argued that we are relinquishing some control over the transport coefficients'' and ``of course, making $\tau$ a nontrivial function of the incoming state will impact the hydrodynamic equations derived from the model''~\cite{boghosian2001entropic}.
+``it may be argued that we are relinquishing some control over the transport coefficients'' and ``of course, making $\omega$ a nontrivial function of the incoming state will impact the hydrodynamic equations derived from the model''~\cite{boghosian2001entropic}.
 ``stability is guaranteed''?
@@
 
-\section{Lattice Boltzmann Method}
+\section{Numerical Methods}
 
-\subsection{Overview}
-
-Some overview of Lattice Boltzmann Method stuff~\cite{grasinger2015simulation}.
+\subsection{Lattice Boltzmann Method: Overview}
 
@@
 The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
-The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \cop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\cop$ is the collision operator.
+The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\colop$ is the collision operator.
 The lattice part of LBM refers to the way in which the BE is discretized.
@@
 
-The D2Q9 lattice was used in the current work (shown in Figure \ref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}.
-% MG: do we even need this citation?
+The D2Q9 lattice was used in the current work (shown in \Fref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}. % MG: do we even need this citation?
 The lattice is two-dimensional with nine discrete velocities at each node.
@@
 
-The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as: $f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \cop_i(f)$, where, for D2Q9, $i = 0, 1, ..., 8$.
+The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as: $f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \colop_i(f)$, where, for D2Q9, $i = 0, 1, ..., 8$.
 
@@
 For velocity or pressure boundary conditions, the method proposed by \citet{zou1997pressure} can be used.
-The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the non-equilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
+The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
 
@@
 
-\subsection{Collision Operator} %MG: if time allows, checkout differential lattice Bolzmann equation, the interpolation-supplemented lattice Boltzmann method, and the Taylor series expansion- and least square-based lattice Boltzmann method; Niu et.
-al.
-2004
+\subsection{Collision Operator}
+
+\subsubsection{Bhatnagar-Gross-Krook (BGK)} \label{sec:bgk}
 
 The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision) \cite{Cer90}.
-In LBM, the collision operator causes particle distributions to relax toward a local equilibrium. This equilibrium is determined by the macroscopic physical behavior of interest.
-In the case of incompressible flow, the equilibrium particle distribution, $f_i^{eq}$, is given by:
+In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
+This equilibrium is determined by the macroscopic physical behavior of interest.
+In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq}$, is often given by:
 
-\begin{equation}
-f_i^{eq} = w_i \rho \left[1 + \frac{\pvel_i \cdot \mvel}{c_s^2} + \frac{(\pvel_i \cdot \mvel)^2}{2c_s^2} - \frac{\mvel^2}{2c_s^2} \right]
+\begin{equation} \label{eq:feq}
+f_i^{eq} = w_i \rho \left[1 + \frac{\pvel_i \cdot \mvel}{c_s^2} + \frac{(\pvel_i \cdot \mvel)^2}{2c_s^2} - \frac{\mvel^2}{2c_s^2} \right],
 \end{equation}
 
-\noindent where $w_i$ is the weight in the $i^{th}$ direction.
+\noindent where $w_i$ is the weight in the $i^{th}$ direction, and %MG: maybe say something about where this comes from, e.g.
+feq is a Hermite polynomial expansion of the Maxwell-Boltzmann distribution
 
-\begin{equation*}
+\begin{equation} \label{eq:weights}
 w_i = \begin{cases}
@@
     \frac{1}{36}, & i = 5, 6, 7, 8
-\end{cases}
-\end{equation*}
+\end{cases}.
+\end{equation}
 
@@
 
-\begin{equation}
-\Omega_i = -\frac{1}{\tau} (f_i(\pos, t) - f_i^{eq}),
+\begin{equation} \label{eq:bgk}
+\Omega_i = -\omega (f_i(\pos, t) - f_i^{eq}),
 \end{equation}
 
-\noindent where $\tau$ is the relaxation time \cite{Bha54}.
-The relaxation time can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
-For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\tau - \frac{1}{2})$.
+\noindent where $\omega$ is the collision frequency \cite{Bha54}.
+The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
+For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$.
 
-\subsubsection{Entropic Collisions and Entropy Balance}
+Despite the utility of the BGK collision operator, it does have a few drawbacks.
+For example, in low viscosity flows the BGK operator results in an over relaxation of particle distributions toward quasiequilibrium.
+It is well known that when large nonequilibrium distributions exist in the LBM approximation that over relaxation can result in nonphysical oscillations that are slow to decay \cite{brownlee2007stability,dellar2003incompressible}.
+To illustrate this, consider the example shown in \Fref{fig:sharp-grad}.
+If $a >> 0$, then when $f_5(x,y)$ propagates to $f_5(x+1,y+1)$ during the streaming step (where $\pos = [x,y]$, $\pvel_5 = [1,1]$, and $\Delta t = 1$), $f_5(x+1,y+1)$ will be ``far'' from $f^{eq}_5(x+1,y+1)$ for the subsequent collision step.
+If $\nu \rightarrow 0$ then $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_5(x+1,y+1)$ still being ``far'' from $f^{eq}_5(x+1,y+1)$ but on the ``other side''.
+If, as in \Fref{fig:sharp-grad}, the LBM macroscopic variables dictate that $f^{eq}_i$ remain mostly constant (or, at the very least, far from $f_i$) in the direction of propagation for $f_i$, then this leads to oscillations in the particle distribution, $f_i$.
+An example of the nonphysical oscillations that would occur for $f_5$ in \Fref{fig:sharp-grad} can be seen in \Fref{fig:overrelaxation}.
+Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to ``blow up'' or ``pollute'' the system enough to make the results highly nonphysical and altogether useless \cite{gorban2014enhancement}.
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figs/sharp_gradient}
+    \caption{Example of an LBM approximation with a sharp gradient in the macroscopic flow velocity}
+    \label{fig:sharp-grad}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figs/overrelaxation_tau-0513}
+    \caption{Example of oscillations that can occur as a result of overrelaxation (this example is related to the example shown in \Fref{fig:sharp-grad}); the relaxation time, $\omega = 1.95$ and $\nu \approx 0.0043$}
+    \label{fig:overrelaxation}
+\end{figure}
+
+In contrast to low viscosity flows, the challenge associated with high viscosity flows (using the BGK collision operator) is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because as $\nu \rightarrow \infty$, $\omega \rightarrow 0$, resulting in extreme underrelaxation to the point of being negligible.
+%MG: this needs cleaned up, I think
+
+Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flows because of the nonlinear constitutive relationship between shear stress and strain-rate.
+The nonlinear constitutive relationship can lead to sharp gradients, and depending on the form of $\dot{\gamma} \mapsto \mu_{app}$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
+
+\subsubsection{Multiple-relaxation Time (MRT)}
+
+An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
+In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f}$, herein referred to as the ``moment space''.
+The collision is then performed in the moment space.
+There a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
+\begin{enumerate}
+    \item Physical processes in fluids can be approximately described by coupling or interacting among modes; and hydrodynamic modes are directly related to, i.e.
+linear combinations of, mass and momentum moments \cite{lallemand2000theory}.
+    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor \cite{dellar2003incompressible}.
+    Of the nine relaxation rates available, the three that correspond to the extra variables--often referred to as ``ghost variables'', and their associated modes as ``ghost modes''--can be tuned in order to dampen out their associated ghost modes, ensuring these modes do not dominate or cause numerical instabilities at the lattice scale.
+\end{enumerate}
+
+The MRT collision operator is given by:
+
+\begin{equation} \label{eq:mrt-colop}
+	\colop = - {\transM}^{-1} \relaxM \transM (\mathbf{f} - \mathbf{f}^{eq}),
+\end{equation}
+
+\noindent where $\transM$ is a transformation matrix that maps the particle distribution vector, $\mathbf{f}$, and quasiequilibrium distribution vector, $\mathbf{f}^{eq}$, from the particle distribution space into moment space.
+The result of mapping the vectors $\mathbf{f}$ and $\mathbf{f}^{eq}$ into moment space will be denoted by $\mathbf{m}$ and $\mathbf{m}^{eq}$, respectively.
+The relationships between $\mathbf{m}$, $\transM$ and $\mathbf{f}$ can be written as follows:
+
+\begin{equation} \label{eq:transformation-matrix}
+\mathbf{m} = \begin{bmatrix}
+\rho \\ e \\ \epsilon \\ j_x \\ q_x \\ j_y \\ q_y \\ p_{xx} \\ p_{xy}
+\end{bmatrix} = \begin{bmatrix}
+1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
+-4 & -1 & -1 & -1 & -1 & 2 & 2 & 2 & 2 \\
+4 & -2 & -2 & -2 & -2 & 1 & 1 & 1 & 1 \\
+0 & 1 & 0 & -1 & 0 & 1 & -1 & -1 & 1 \\
+0 & -2 & 0 & 2 & 0 & 1 & -1 & -1 & 1 \\
+0 & 0 & 1 & 0 & -1 & 1 & 1 & -1 & -1 \\
+0 & 0 & -2 & 0 & 2 & 1 & 1 & -1 & -1 \\
+0 & 1 & -1 & 1 & -1 & 0 & 0 & 0 & 0 \\
+0 & 0 & 0 & 0 & 0 & 1 & -1 & 1 & -1
+\end{bmatrix} \begin{bmatrix}
+f_0 \\
+f_1 \\
+f_2 \\
+f_3 \\
+f_4 \\
+f_5 \\
+f_6 \\
+f_7 \\
+f_8
+\end{bmatrix} = \transM \mathbf{f}.
+\end{equation}
+
+\noindent where $\epsilon$ is related to the square of the energy $e$; $q_x$ and $q_y$ correspond to the energy fluxes in the x and y directions; and $p_{xx}$ and $p_{xy}$ correspond to the diagonal and off-diagonal component of the viscous stress tensor~\cite{lallemand2000theory}.
+% MG: verify this is correct
+The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i, i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
+In the case when $s_0 = s_1 = ... = s_8 = \omega$ ($\omega$ is the collision frequency in the BGK sense), the MRT collision operator is equivalent to the BGK collision operator.
+The relaxation parameters $s_0$, $s_3$, and $s_5$ are all set to zero as mass and momentum should be conserved.
+The relaxation parameters $s_1$ and $s_7 = s_8$ are related to the bulk and shear viscosities, respectively.
+The relationship for the shear viscosity is given by,
+\begin{equation} \label{eq:mrt-s7}
+\nu = c_s^2 \Delta t (\frac{1}{s_7} - \frac{1}{2}).
+\end{equation}
+The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
+It is common practice, and \citet{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.
+
+The MRT collision operator has a greater numerical stability than its BGK counterpart~\cite{lallemand2000theory,dellar2003incompressible,d2002multiple}, and because of the challenges associated with simulating non-Newtonian flow, the MRT collision operator has become popular for simulating non-Newtonian flows~\cite{chen2014simulations,fallah2012multiple,tang2011bingham,vikhansky2008lattice,chai2011multiple,grasinger2015simulation}.
+The main drawback of the MRT collision operator is its computational expense.
+Why MRT is more computationally expense is clear when one considers that \eqref{eq:mrt-colop} requires multiple matrix multiplications and \eqref{eq:bgk} requires none.
+It has been reported that MRT is approximately 15\% slower than BGK~\cite{d2002multiple}, but this was in the context of Newtonian flow.
+As will be shown later, for certain LBM implementations and non-Newtonian flows the increase in computational expense can be much greater.
+
+\subsection{Stability Enhancement through Artificial Dissipation}
+
+To reduce nonequilibrium fluctuations in LBM, one can introduce artificial dissipation.
+The idea of artificial dissipation is to increase numerical stability, while sacrificing some physical accuracy.
+A model that has more physical justification but produces unstable and nonsensical results is much less useful than a model with some minor artificial features yet produces more stable results.
+%Lyapanov stability is the idea that solutions at equilibrium stay there.
+A practical goal then would be to use only the necessary amount of artificial dissipation in order to ensure a stable solution.
+From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation necessary?'' and ``how much artificial dissipation does one introduce when it is necessary?''.
+
+To answer the first question, consider the discussion in \Fref{sec:bgk} on problems with overrelaxation and nonphysical oscillations.
+Nonphysical oscillations due to overrelaxation would be damped out more quickly if particle distributions ``far'' from quasiequilibrium were brought closer to quasiequilibrium.
+A particle distribution vector, $\mathbf{f}$, ``far'' from equilibrium would therefor be a good candidate for artificial dissipation.
+There are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
+A metric that has been developed and used successfully for determining when artificial dissipation should be introduced at a lattice site is the so called relative nonequilibrium entropy~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy}.
+The relative nonequilibrium entropy, $\Delta S$, is given by:
+
+\begin{equation} \label{eq:rel-noneq-entropy}
+\Delta S = \sum_i f_i \ln(\frac{f_i}{f^{eq}_i}).
+\end{equation}
+
+\noindent A more computationally efficient approximation of $\Delta S$ can be achieved by instead using the second order Taylor expansion of \eqref{eq:rel-noneq-entropy},
+
+\begin{equation} \label{eq:quadratic-entropy}
+\Delta S \approx \sum_i \frac{(f_i - f^{eq}_i)^2}{2f^{eq}_i}.
+\end{equation}
+
+\noindent Note that limiting nonequilibrium entropy in LBM is analogous to what flux limiters do in finite difference, finite volume, and finite element methods~\cite{brownlee2008nonequilibrium}.
+
+An option for defining criteria for when to introduce artificial dissipation that has been used successfully is to define a threshold, $\theta$, such that dissipation is added when $\Delta S > \theta$~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium}.
+A potential drawback of defining a threshold \textit{a priori} is that in order to ensure the model still retains some physical integrity, only a small number of sites can have artificial dissipation added.
+If the threshold is too low, too many sites may have dissipation added.
+If the threshold is too high, a stable solution may not be achieved.
+The threshold can be determined on a case-by-case basis through trial-and-error or by a preliminary analysis.
+In the current work, the criteria that is used for determining whether dissipation should be added is,
+
+\begin{equation} \label{eq:dissipation-criteria}
+    \Delta S > \overline{\Delta S} + n_\sigma \cdot \sigma_{\Delta S},
+\end{equation}
+
+\noindent where $\overline{\Delta S}$ and $\sigma_{\Delta S}$ are the mean and standard of $\Delta S$, respectively--both are calculated using values over the domain for the current timestep--and $n_\sigma$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S$ must be before dissipation is added.
+The number of standard deviations, $n_\sigma$, is chosen \textit{a priori}.
+The criteria described in \ref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead can think in terms of what would be the maximum percentage of sites one would want artificial dissipation to be added to.
+
+Just as there are many ways to measure a lattice site's distance from quasiequilibrium, or from that measure define criteria for artificial dissipation, there are also many different ways of deciding how much dissipation to add.
+One method of adding dissipation is the so-called Ehrenfests' regularization~\cite{brownlee2006stabilization}, and involves setting a lattice site that is chosen for artificial dissipation to its quasiequilibrium state.
+Although this achieves the desired result, namely damping out large nonequilibrium fluctuations, it does so in an intense manner.
+An alternative, the median filter, has been used successfully in conjunction with both the BGK and MRT collision schemes for simulating one-dimensional shock tubes and lid-driven cavity flows~\cite{brownlee2008nonequilibrium,gorban2014enhancement,packwood2009entropy}.
+Median filtering is an effective noise reduction technique, often used in image processing, for ``speckle noise'' or ``salt and pepper noise''~\cite{brownlee2008nonequilibrium}.
+In other words, median filters are good at reducing high frequency noise while having a minimal effect on lower frequency noise.
+In LBM this is a desirable way to introduce dissipation because it has the potential to reduce high frequency nonequilibrium fluctuations that might lead to numerical instability while retaining the lower frequency dynamics.
+To use the median filter one performs the collision step and then checks over the domain for lattice sites with $\Delta S$ that meet the criteria for artificial dissipation; sites that meet the criteria are updated as follows:
+\begin{equation} \label{eq:median-filter}
+\mathbf{f} = \mathbf{f}^{eq} + \delta (\mathbf{f} - \mathbf{f}^{eq}),
+\end{equation}
+\noindent where $\delta = \sqrt{\Delta S_{med} / \Delta S}$ is the scaling coefficient, and $\Delta S_{med}$ is the median value of $\Delta S$ for the nearest neighbors of the lattice site.
+
+\newcommand{\srtensor}{\mathbf{D}}
+
+\subsection{Strain-rate Tensor}
+
+In order to use the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf}, one must calculate the strain-rate, $\dot{\gamma}$.
+The strain-rate is given by the second invariant of the strain-rate tensor, $\srtensor$, i.e.
+\begin{equation} \label{eq:second-invariant}
+\dot{\gamma} = \sqrt{\srtensor : \srtensor}.
+\end{equation}
+
+\noindent When using the BGK collision scheme, the strain-rate tensor is determined by,
+\begin{equation} \label{eq:srtensor-bgk}
+D_{\alpha \beta} = -\frac{\omega}{2 \rho c_s^2} \sum_i \xi_{i \alpha} \xi_{i \beta} (f_i - f^{eq}_i),
+\end{equation} % MG: verify this
+
+\noindent and for the MRT collision scheme, the strain-rate tensor is determined by
+\begin{equation} \label{eq:srtensor-mrt}
+D_{\alpha \beta} = -\frac{1}{2 \rho c_s^2 \Delta t} \sum_i \xi_{i \alpha} \xi_{i \beta} \sum_j (\transM^{-1} \relaxM \transM)_{ij}(f_i - f^{eq}_i).
+\end{equation} % MG: verify this
+
+Upon comparison of \eqref{eq:srtensor-bgk} and \eqref{eq:srtensor-mrt}, it is clear that calculating the strain-rate tensor for the MRT collision operator is more computationally expensive than it is for the BGK collision operator.
+The increase in computational expense is exacerbated by the fact that approximating the apparent viscosity, $\mu_{app}$, by the algorithm outlined in \Fref{sec:lbm-for-nnf} requires iteration and therefor multiple calculations of the strain-rate tensor.
+
+\subsection{Non-Newtonian Constitutive Equations}
+
+\subsubsection{Bingham Plastic}
+
+The Bingham plastic constitutive model is popular for yield stress flows.
+A Bingham plastic does not flow, i.e.
+the strain-rate is zero, when the shear stress is below the yield stress, and behaves in an almost Newtonian manner when the shear stress is above the yield stress.
+The Bingham plastic relationship is described mathematically as:
+\begin{equation} \label{eq:bingham}
+\begin{cases}
+\tau = \tau_y + \mu_p \dot{\gamma}, & |\tau| \geq \tau_y \\
+\dot{\gamma} = 0, & |\tau| < \tau_y
+\end{cases}
+\end{equation}
+\noindent where $\tau$ is the shear stress, $\tau_y$ is the yield stress, and $\mu_p$ is the plastic viscosity~\cite{bingham1922fluidity}.
+
+Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
+Thus, a smooth approximation to \eqref{eq:bingham} formulated by \citet{papanastasiou1987flows} is often used as:
+\begin{equation} \label{eq:papa}
+\tau = \tau_y (1 - e^{-m |\dot{\gamma}|}) + \mu_p \dot{\gamma},
+\end{equation}
+\noindent where $m$ is the stress growth exponent.
+The larger the value of $m$, the closer the approximation is to the Bingham plastic model.
+
+Alternatively, the constitutive relationship can be interpreted through the apparent viscosity.
+The apparent viscosity is the value of the viscosity assuming a Newtonian relationship between stress and strain rate, $\mu_{app} = \frac{\tau}{\dot{\gamma}}$, and is used to determine the collision frequency for the BGK collision operator or $s_7$ and $s_8$ of the relaxation matrix for the MRT collision operator.
+Rearranging \eqref{eq:papa} results in the following expression for the apparent viscosity:
+\begin{equation} \label{eq:bing-mu-app}
+\mu_{app} = \frac{\tau_y}{\dot{\gamma}} (1 - e^{-m |\dot{\gamma}|}) + \mu_p.
+\end{equation}
+
+\subsubsection{Power-law}
+
+The power-law constitutive relationship is useful for modeling fluids that experience shear-thinning or shear-thickening.
+The power-law relationship is given by,
+\begin{equation} \label{eq:pl}
+\tau = k \dot{\gamma}^n,
+\end{equation}
+\noindent where $k$ is the flow consistency index and $n$ is the flow behavior index.
+When $n = 1$, the \eqref{eq:pl} results in a Newtonian constitutive relationship with dynamic viscosity, $\mu = k$.
+A flow consistency index of $n < 1$ results in shear-thinning behavior, where as $n > 1$ results in shear-thickening.
+As with the Bingham plastic relationship, \eqref{eq:pl} can be modified to determine an apparent viscosity,
+\begin{equation} \label{eq:pl-mu-app}
+\mu_{app} = k \dot{\gamma}^{n-1}.
+\end{equation}
 
@@
 
+Some introduction to the numerical study.
+It has a purpose.
+Its purpose is to try to assess the accuracy, stability, and efficiency of the LBM models as applied to non-Newtonian flows.
+For all simulations presented in this section, the apparent viscosity is approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the maximum number of iterations set at 15 and the convergence criteria,
+\begin{equation} \label{eq:mu-app-conv}
+\frac{\left|\mu_{app}^{k+1} - \mu_{app}^{k}\right|}{\mu_{app}^{k}} < 1.0 \times 10^{-6}.
+\end{equation}
+All simulations with artificial dissipation use the median filter with $n_s = 2.7$.
+All numerical values given in this section are in lattice units unless otherwise stated.
+
+\newcommand{\pgrad}{\frac{\partial p}{\partial x}}
+
 \subsection{Poiseuille Flow}
+
+Poiseuille flow is a useful benchmark because analytical solutions exist for both Bingham plastic and power-law fluids.
+Poiseuille flow is driven by a constant pressure gradient, $\pgrad$, through a two-dimensional channel.
+A schematic is shown in \Fref{fig:poise-schematic}.
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figs/poise-schematic}
+    \caption{Schematic of Poiseuille flow; no-slip boundary conditions are enforced at the top and bottom boundaries, and periodic boundary conditions are enforced at the left and right boundaries.}
+    \label{fig:poise-schematic}
+\end{figure}
+
+Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice.
+
+\subsubsection{Bingham Plastic}
+
+The Bingham plastic simulations were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a plastic viscosity of $\mu_p = 0.2$.
+The yield stress was varied between the four different values $\tau_y = [4.0, 8.0, 12.0, 16.0] \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
+The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
+The analytical solution for Bingham plastic flow is given by:
+\begin{equation} \label{eq:analytical-bingham}
+u_x(y) = \begin{cases}
+\frac{1}{2 \mu_p} \left(-\pgrad\right) \left[h^2 - y_{\tau}^2 \right] - \frac{\tau_y}{\mu_p} \left(h - y_{\tau}\right), & 0 \leq |y| \leq y_{\tau}, \\
+\frac{1}{2 \mu_p} \left(-\pgrad\right) \left[h^2 - y^2 \right] - \frac{\tau_y}{\mu_p} \left(h - |y|\right), & y_{\tau} < |y| \leq h,
+\end{cases}
+\end{equation}
+\noindent where $y_{\tau}$ is the point at which the fluid yields and $h$ is half of the total height of the channel.
 

### 2016-07-04 main.tex
finished Bingham plastic Poiseuille results
Add: Bingham results
Add: figures of analytical vs approx
Add: figures comparing f to feq

@@
 \usepackage[plain]{fancyref}
+\usepackage{tabulary}
+\usepackage{pbox}
 
@@
 For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$.
+From this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.
 
@@
 The nonlinear constitutive relationship can lead to sharp gradients, and depending on the form of $\dot{\gamma} \mapsto \mu_{app}$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
+
+Due to the instabilities associated with the collision frequency being too high, e.g.
+approaching 2, or too low, e.g.
+approaching 0, a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flows is to simply put bounds on the values in which the collision frequency may attain.
+This simple methodology for increasing stability will henceforth be referred to bounded-relaxation time BGK, or BGK-BRT.
 
@@
 \begin{equation} \label{eq:mrt-s7}
-\nu = c_s^2 \Delta t (\frac{1}{s_7} - \frac{1}{2}).
+\nu = c_s^2 \Delta t (\frac{1}{s_7} - \frac{1}{2}),
 \end{equation}
+\noindent which is equivalent with the relationship to the collision frequency, $\omega$, in the BGK sense.
 The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
@@
 
-\subsection{Stability Enhancement through Artificial Dissipation}
+\subsection{Stability Enhancement through Artificial Dissipation: Entropic Filtering}
 
@@
 A schematic is shown in \Fref{fig:poise-schematic}.
+No-slip boundary conditions are enforced at the top and bottom boundaries and periodic boundary conditions are enforced at the left and right boundaries.
+The total height of the channel will be denoted by $H$.
+The center of the channel is $y = 0$ and $y \in [-h, h]$ where $h = \frac{H}{2}$.
 
@@
 
-Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice.
+Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps.
+For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$.
 
-\subsubsection{Bingham Plastic}
+\subsubsection{Bingham Plastic Fluids}
 
@@
 The yield stress was varied between the four different values $\tau_y = [4.0, 8.0, 12.0, 16.0] \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
+(Recall $m$ is the stress growth exponent for the Papanastasiou approximation.)
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
@@
 \end{equation}
-\noindent where $y_{\tau}$ is the point at which the fluid yields and $h$ is half of the total height of the channel.
+\noindent where $y_{\tau} = -\tau_y / \pgrad$ is the point at which the fluid yields.
+
+The results are presented in \Fref{tab:poise-bing}.
+In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
+The Reynold's number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution and $H = 2h$ is the total height of the channel.
+The Bingham number was computed by $Bn = \frac{\tau_y H}{\mu_p U}$.
+
+% MG: it might be better to normalize the time with respect to the fastest, BGK with no filtering
+\vspace{20cm}
+\begin{tabulary}{\linewidth}{l l r r r r r r r r} \label{tab:poise-bing}
+\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
+\hline \\
+BGK & No & $10^5$ & 4.0 & 6.05 & 0.68 & 0.0062 & 0.0153 & 1857 \\
+& & & 8.0 & 4.42 & 1.85 & 0.0204 & 0.0411 & 2329 \\
+& & & 12.0 & 3.04 & 4.04 & 0.0503 & 0.0891 & 3345 \\
+& & & 16.0 & 1.92 & 8.52 & 0.1161 & 0.1879 & 2029 \\
+\\
+BGK & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0109 & 0.0282 & 2831 \\
+& & & 8.0 & 4.42 & 1.85 & 0.0330 & 0.0670 & 3509 \\
+& & & 12.0 & 3.04 & 4.04 & 0.0788 & 0.1570 & 4838 \\
+& & & 16.0 & 1.92 & 8.52 & 0.1991 & 0.3539 & 3790 \\
+\\
+BGK & Yes & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0100 & 0.0273 & 2903 \\
+& & & 8.0 & 4.42 & 1.85 & 0.0361 & 0.0823 & 3567 \\
+& & & 12.0 & 3.04 & 4.04 & 0.2439 & 0.3832 & 4800 \\
+& & & 16.0 & 1.92 & 8.52 & 0.7533 & 1.0718 & 3507 \\
+\\
+MRT & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
+& & & 8.0 & 4.42 & 1.85 & 0.0018 & 0.0018 & 7908 \\
+& & & 12.0 & 3.04 & 4.04 & 0.0026 & 0.0026 & 7160 \\
+& & & 16.0 & 1.92 & 8.52 & 0.0041 & 0.0041 & 5559 \\
+\\
+\end{tabulary}
+
+As has been reported previously, for the BGK collision operator--regardless of whether or not median entropic filtering is used--using a stress growth exponent of $m = 10^5$ is more accurate with respect to the analytical solution than using a stress growth exponent of $m = 10^8$~\cite{chen2014simulations}.
+A larger stress growth exponent results in a more accurate Papastasiou approximation of the true Bingham plastic constitutive model, however, it leads to more noise that degrades the numerical solution for the BGK collision operator.
+Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helps dissipate the noise that occurs as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
+In fact, the median filter resulted in a less accurate solution in nearly all cases and rendered the solution altogether useless for the higher yield stress flows (relative errors of 25-75\%).
+
+The BGK collision operator using a stress growth exponent of $m = 10^5$ was in general the fastest.
+This can be attributed to the fact that a smoother approximation of the Bingham plastic constitutive model would lead to a solution for $\mu_{app}$ converging with less iterations.
+The BGK collision operator using a stress growth exponent of $m = 10^5$ also performed acceptably in terms of accuracy for lower yield stress flows, however, for $\tau_y = 12 \times 10^{-5}$ and $16 \times 10^{-5}$ the relative $L_\infty$ errors were 8.9\% and 19\%, respectively--larger than what would be considered acceptable for most engineering applications.
+
+The LBM approximations are plotted with the analytical solution for the BGK collision operator using $m = 10^5$ and $m = 10^8$ in \Fref{fig:bing-analyt-bgk-5} and \Fref{fig:bing-analyt-bgk-8}, respectively.
+\Fref{fig:bing-analyt-bgk-5} shows that the error for the BGK model with $m = 10^5$ is not due to noise, but instead the inaccuracy of the Papastasiou approximation with a lower stress growth exponent.
+\Fref{fig:bing-analyt-bgk-8} shows that the error for the BGK model with $m = 10^8$ is due to numerical noise.
+In order to better understand why median filtering did not eliminate the noise in high yield stress flows, but instead exacerbate the problem, it is necessary to investigate what is happening at the particle distribution scale.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figs/poise-bingham/bgk-5/analytical-vs-approx.png}
+\caption{LBM approximation using BGK and $m = 10^5$ compared to the analytical solution for $\tau_y = 16 \times 10^{-5}$.}
+\label{fig:bing-analyt-bgk-5}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{figs/poise-bingham/bgk-8/analytical-vs-approx.png}
+    \caption{LBM approximation using BGK and $m = 10^8$ compared to the analytical solution for $\tau_y = 16 \times 10^{-5}$.}
+\label{fig:bing-analyt-bgk-8}
+\end{figure}
+
+\Fref{fig:feq-vs-f_bgk} compares particle distributions to their respective quasiequilbiriums.
+It can be seen that in the BGK, $m = 10^8$ solution, the noise due to nonlinear constitutive relation pollutes the quasiequilibrium values.
+Because the noise makes its way into the quasiequilibrium values, entropic median filtering does not help dampen the noise but instead contracts the particle distributions closer to the quasiequilibrium noise--explaining why the median filtered results were less accurate for Bingham plastic Poiseuille flow.
+
+%MG: reformat these images so that text is more readable
+\begin{figure}
+\begin{tabulary}{\linewidth}{c c}
+\includegraphics[width=2.1in]{figs/poise-bingham/bgk-5/feq-vs-f_5.png}
+&
+\includegraphics[width=2.1in]{figs/poise-bingham/bgk-5/feq-vs-f_8.png}
+\\
+\includegraphics[width=2.1in]{figs/poise-bingham/bgk-8/feq-vs-f_5.png}
+&
+\includegraphics[width=2.1in]{figs/poise-bingham/bgk-8/feq-vs-f_8.png}
+\end{tabulary}
+\caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums.
+The top two plots are for the BGK with $m = 10^5$, the bottom two are for the BGK with $m = 10^8$.}
+\label{fig:feq-vs-f_bgk}
+\end{figure}
+
+On average, the LBM model using the MRT collision operator was 2.9 times slower than the fastest model for the given material parameters.
+However, the increased computational time can be justified for the MRT collision operator because it did not suffer from the same level of noise as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
+when $m = 10^8$.
+The MRT collision operator was therefor the most accurate solution for all cases.
+\Fref{fig:feq-vs-f_mrt} compares particle distributions to quasiequilibrium for the MRT model; in contrast to the BGK with $m = 10^8$, there is no noise in the quasiequilibrium distributions.
+
+% A natural question may be to ask ``if the strength of the MRT collision operator is damping out the ghost modes, then is the noise for the BGK with $m = 10^8$ due to the ghost modes?''.
+% MG: maybe plot ghost modes for BGK and MRT
+
+\begin{figure}
+    \begin{tabulary}{\linewidth}{c c}
+        \includegraphics[width=2.1in]{figs/poise-bingham/mrt/feq-vs-f_5.png}
+        &
+        \includegraphics[width=2.1in]{figs/poise-bingham/mrt/feq-vs-f_8.png}
+    \end{tabulary}
+    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \label{fig:feq-vs-f_mrt}
+\end{figure}
+
+\subsubsection{Power-law Fluids}
 

### 2016-07-05 main.tex
finished first draft

@@
 % MG: replace \ \cite with ~\cite to ensure citations stay on same line
-% MG: need an overall consistency in 'flucuation' vs. 'oscillation'; use same word if you mean same thing.
+% MG: need an overall consistency in 'flucuation' vs. 'oscillation' vs. 'noise'; use same word if you mean same thing.
 % MG: same note as above for 'local equilibrium' vs. 'quasiequilibrium'
@@
 
-\title{Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows}
+\title{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows}
 %\title{A Comparison of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}
@@
 
-Note: anything in {\color{red}red} is just random garbage notes.
-You can ignore it for the most part.
-
 \section*{Abstract}
@@
 constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
-In this paper, various collision operators and methods for stability enhancement will be examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
-The structure of the paper will be as follows: blah, blah, blah.
+In this paper, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
+The investigation was be carried out, not as a rigorous mathematical treatment, but instead as a numerical study looking for qualitative, yet practical, results.
+The numerical investigation included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluid flows.
+Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
 
@@
 When the solution for fluid flow problems vary in space and time, a numerical approximation requires breaking the problem up into discrete locations and discrete time steps.
-The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which becomes computationally expensive.
+The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which can become computationally expensive.
 The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the advantage that computing the strain rate is second-order accurate and local to each node~\cite{kruger2010second}.
 This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
-Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flows over other numerical methods.
+Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flows over some other numerical methods.
 
@@
 LBM models for power-law fluid flows~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
+%MG: or were they validated?
 
@@
 LBM can be considered as a type of finite-difference scheme for the continuous Boltzmann equation, and as such, has numerical properties in common with finite-difference schemes.
-%MG: is this precise enough?
-accurate enough?
 One such consideration associated with this view, is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
@@
 %Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
-Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the strong nonlinear relationship between shear stress and strain-rate can lead to high nonequilibrium fluctuations.
+Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the nonlinear relationship between shear stress and strain-rate can lead to high nonequilibrium fluctuations.
 %MG: be more precise than 'fluctuations'?
 Various schemes and strategies for incorporating the physics of non-Newtonian flows and yet maintaining a stable numerical method have been developed and studied.
-The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free}, is to make the relaxation time--which is proportional to apparent viscosity--variable and dependent on the local strain rate.
-A potential issue with the stability of the variable relaxation time approach is that as the collision frequency, $\omega$, approaches $2$ the viscosity approaches zero; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
-In order to ensure that the variable relaxation times did not approach values leading to numerical instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable relaxation times.
-Although bounding the relaxation time was shown to be effective in terms of stability, it is also nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the relaxation time does not reflect the proper constitutive relationship of interest.
-Another scheme for incorporating non-Newtonian effects into LBM is to use a constant relaxation rate, typically unity, and to instead incorporate the local shear-rate effect into equilibrium distribution functions.
+The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free}, is to make the collision frequency--which is proportional to apparent viscosity--variable and dependent on the local strain rate.
+A potential issue with the stability of the variable relaxation time approach is that as the collision frequency, $\omega$, approaches $2$ the viscosity approaches zero and overrelaxation occurs; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
+%MG: verify this statement
+In order to ensure that the variable collision frequencies did not approach values leading to numerical instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable collision frequencies.
+Although bounding the collision frequency was shown to be effective in terms of stability, it is also nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the collision does not reflect the proper constitutive relationship of interest.
+Another scheme for incorporating non-Newtonian effects into LBM is to use a constant collision frequency, typically unity, and to instead incorporate the local shear-rate effect into equilibrium distribution functions.
 This means particle distribution functions will always relax toward equilibrium at the same rate, but that the definition of equilibrium is modified to represent the correct stress--strain-rate relationship.
@@
 The equilibrium distribution function for Bingham plastic flow was derived, implemented, and verified in~\cite{wang2008lattice}; and the equilibrium distribution functions for power-law fluids and Carreau fluids were derived, implemented, and verified in~\cite{yoshino2007numerical}.
-The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the relaxation time is constant (at unity), the local relaxation time will not approach values that are too small or too large, and so there is no reason to bound the relaxation time in a way that is nonphysical.
-\citet{wang2011lattice} developed another constant-relaxation time LBM scheme for non-Newtonian flows by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts--for example, considering the plastic viscosity and yield stress for a Bingham plastic separately--then incorporating the Newtonian part in the usual way, namely scaling the relaxation time to achieve the macroscopic (Newtonian) viscosity, and incorporating the non-Newtonian part as a source of momentum, i.e.
+The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the relaxation time is constant (at unity), the local relaxation time will not approach values that are too small or too large, and so there is no reason to bound the collision frequency in a way that is nonphysical.
+\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flows by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts--for example, considering the plastic viscosity and yield stress for a Bingham plastic separately--then incorporating the Newtonian part in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and incorporating the non-Newtonian part as a source of momentum, i.e.
 as an external forcing term, that is dependent on local shear-rate.
-Although the constant-relaxation time strategies present interesting alternatives, the variable relaxation time scheme is used in the present study because of its simplicity and flexibility.
-%MG: should we define what we mean by 'flexibility'?
+Although the constant-collision frequency strategies present interesting alternatives, the variable collision scheme is used in the present study because of its simplicity and generality, i.e.
+it can be easily fitted to any constitutive equation without for example, performing Chapman-Enskog multiscale expansion.
 
@@
 \citeauthor{sterling1993stability} concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
-%MG: this probably requires some clarification and a more concrete tieback to the highlevel goals of this paper
-%(Some misc notes from this study that probably don't fit here but are important to this work: ``...
-a Courant stability condition is superceded by a more stringent stability condition on the speeds.
-Another stability boundary common for finite-difference methods requires that the viscous diffusion speed be less than the lattice spacing divided by the time step.
-This boundary is not observed for LB methods because as the viscosity increases, errors of the scheme increase due to the presence of large nonequilibirum populations, but stability is still maintained''??).
-~\citet{worthing1997stability} extended the stability analysis performed by \citeauthor{sterling1993stability}, and instead of linearizing the LBM equations by expanding about global equilibrium values, \citeauthor{worthing1997stability} \ul{does something else}.
-Their analysis discover both physical and nonphysical instabilities, and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time.
-%MG: is this actually relevant though?
-\ul{Expand a little more about what they did and its significance here.}
-%MG: all this stability talk is nice, but I need to figure out how to tie it back to non-Newtonian flows sooner rather than later.
-The problem is a lot of these studies focus on instability due to high Reynolds number flows, not really our problem.
-We're more concerned with instability due to lower relaxation rates and apparent viscosities.
-And how does MRT address this instability?
+~\citet{worthing1997stability} extended the stability analysis performed by \citeauthor{sterling1993stability} and discovered both physical and nonphysical instabilities; and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time.
 
@@
 \begin{itemize}
-	\item von Neumann stability analysis, and linear stability analysis in general requires modifying the governing equations of the model so that they are linear.
-%MG: is there even a distinction between von Neumann stability analysis and linear stability analysis?
-	This linearization cannot be done without introducing some error into the analysis.
+	\item von Neumann stability analysis, and linear stability analysis in general, requires modifying the governing equations of the model so that they are linear. 
+	This linearization cannot be done without introducing some error into the analysis. %MG: I don't know if I like this
 	\item Much of the analysis that exists in literature is done about a uniform flow~\cite{worthing1997stability,niu2004investigation}. %MG: definitely cite and/or verify this more
@@
 	Such an analysis, however, does not generalize well to instances when modes of the system are heavily influenced by the boundary conditions~\cite{worthing1997stability}.
-%MG: so many questions: like what is an example of an instance when boundary conditions have a major effect and when are they more ``negligible'' or more like periodic BCs?
 	\item The analyses are general and do not delve into any of the nuances of non-Newtonian flows.
 \end{itemize}
-{\color{red} Bawazeer wrote a thesis on the stability and accuracy of the lattice Boltzmann method. Perhaps I'll read it. I think Jonas Latt did too.}
 
 To improve upon the stability of LBM models, \citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
-\citet{lallemand2000theory} used von Neumann stability analysis to investigate the dispersion, dissipation, isotropy, Galilean invariance, and stability of the newly constructed LBM-MRT model.
-They concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
-Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain aspects of the collision operator be performed at each iteration (the strain rate tensor is determined from the nonequilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
+\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
+Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
 Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
 \citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
-
-{\color{red} 
-ELBM and entropic filtering stuff goes here.
-The price of entropic lattice Boltzmann models for fluid flow is living with a bounded transport coefficient.
-Entropically stabilized collision operators ensure that entropy always increases, i.e. that the second law of thermodynamics is never violated.
-For BGK, this means that the relaxation time is bounded from below in certain instances~\cite{boghosian2001entropic}.
-``it may be argued that we are relinquishing some control over the transport coefficients'' and ``of course, making $\omega$ a nontrivial function of the incoming state will impact the hydrodynamic equations derived from the model''~\cite{boghosian2001entropic}.
-``stability is guaranteed''?
-Define Lyapunov functional or Lyapunov stability condition.
-}
+However, what remains unclear is:
+\begin{itemize}
+    \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flows?
+    \item Under what conditions, e.g. material parameters, physical problem, etc., for Bingham plastic fluid flows is the MRT collision operator necessary to maintain stability and/or accuracy?
+    \item Under what conditions, e.g. material parameters, physical problem, etc, for power-law fluid flows is the MRT collision operator necessary to maintain stability and/or accuracy?
+    \item What are additional strategies for increasing stability and accuracy, and what are their associated computational costs?
+\end{itemize}
+In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
+\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations--oscillations that often lead to numerical instabilities if left unchecked.
+Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in non-Newtonian flows.
 
 The goal of this paper is to numerically study the implications of accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flows using LBM.
-The intention of the study is to aid scientists and engineers in understanding which strategy to choose given their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
+The intention of the study is to aid scientists and engineers in understanding which strategy is best suited to their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
 Advantages, such as LBM's potential to scale well in parallel, can be much less realized if the collision operator is too computationally expensive, or if numerical instabilities ensue.
@@
 \section{Numerical Methods}
+
+%MG: is the beginning of this methods section too much fluff now? Would our audiance consider it fluff?
 
@@
 \end{equation}
-\noindent which is equivalent with the relationship to the collision frequency, $\omega$, in the BGK sense.
+\noindent which is equivalent with the relationship to the collision frequency, $\omega$, in the BGK sense (when $\Delta t = 1$).
 The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
@@
 \noindent where $\delta = \sqrt{\Delta S_{med} / \Delta S}$ is the scaling coefficient, and $\Delta S_{med}$ is the median value of $\Delta S$ for the nearest neighbors of the lattice site.
+
+%MG: show equation for feq that maximizes entropy and give a brief word why we should care
 
@@
 \end{equation}
+All BGK-BRT collision schemes use $\omega \in [0.05, 1.995]$ as the bounds.
 All simulations with artificial dissipation use the median filter with $n_s = 2.7$.
@@
 
-Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps.
+Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps (although the length, $L$, seems small, the effect is negligible as the east-west boundary conditions are periodic).
+In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
 For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$.
 
-\subsubsection{Bingham Plastic Fluids}
+\subsubsection{Bingham Plastic Fluids} \label{sec:poise-bing}
 
@@
 The results are presented in \Fref{tab:poise-bing}.
-In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
 The Reynold's number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution and $H = 2h$ is the total height of the channel.
@@
 % MG: it might be better to normalize the time with respect to the fastest, BGK with no filtering
-\vspace{20cm}
-\begin{tabulary}{\linewidth}{l l r r r r r r r r} \label{tab:poise-bing}
+
+\begin{table}
+\caption{Bingham plastic Poiseuille flow}
+\vspace{0.5cm}
+\begin{tabulary}{\linewidth}{l l r r r r r r r r}
 \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
@@
 \\
+\label{tab:poise-bing}
 \end{tabulary}
+\end{table}
 
@@
 Because the noise makes its way into the quasiequilibrium values, entropic median filtering does not help dampen the noise but instead contracts the particle distributions closer to the quasiequilibrium noise--explaining why the median filtered results were less accurate for Bingham plastic Poiseuille flow.
+In order to ensure that this phenomena was a side effect of all entropic filtering and not just entropic median filtering with $n_s = 2.7$, entropic filtering was tested with median filtering and Ehrenfests' regularization for $n_s = 2.7$ and for $\Delta S$ thresholds of $\delta = 10^{-4}$ and $\delta = 10^{-5}$.
+All filtering schemes suffered from the approximately the same level of noise.
 
@@
 
+\subsubsection{Power-law Fluids} \label{sec:poise-powerlaw}
+
+The power-law simulations were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a flow consistency index of $k = 0.2$.
+The flow behavior index was varied between the four different values $n = [0.5, 0.75, 1.25, 1.50]$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $[0.05, 1.995]$.
+The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
+The analytical solution for power-law fluid flow is given by:
+\begin{equation} \label{eq:analytical-power-law}
+  u_x(y) = \frac{n}{n+1}\left(-\frac{1}{k}\pgrad\right)^{1/n}\left[h^{\frac{n+1}{n}} - |y|^{\frac{n+1}{n}}\right].
+\end{equation}
+
+The results are presented in \Fref{tab:poise-power-law}.
+In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
+The Reynold's number was computed by $Re = \frac{\rho U^{2-n} H^n}{k}$, where $U$ is the maximum velocity given by the analytical solution and $H = 2h$ is the total height of the channel.
+
+% MG: it might be better to normalize the time with respect to the fastest, BGK with no filtering
+
+\begin{table}
+\caption{Power-law Poiseuille flow}
+\vspace{0.5cm}
+\begin{tabulary}{\linewidth}{l l r r r r r}
+\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $n$ & $Re$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
+\hline \\
+BGK & No & 0.50 & 0.0007 & 28.06 & 30.37 & 5232 \\
+& & 0.75 & 0.9125 & 0.0328 & 0.0569 & 2399 \\
+& & 1.25 & 423.2 & 0.0051 & 0.0055 & 2311 \\
+& & 1.50 & 2213 & 1.0 & 1.0 & 3622 \\
+\\
+BGK & Yes & 0.50 & 0.0007 & 26.72 & 28.78 & 5347 \\
+& & 0.75 & 0.9125 & 0.0328 & 0.0569 & 2385 \\
+& & 1.25 & 423.2 & 0.0051 & 0.0055 & 2275 \\
+& & 1.50 & 2213 & 0.0570 & 0.0600 & 2357 \\
+\\
+MRT & No & 0.50 & 0.0007 & 0.1758 & 0.1401 & 4906 \\
+& & 0.75 & 0.9125 & 0.0058 & 0.0045 & 4880 \\
+& & 1.25 & 423.2 & 0.0051 & 0.0055 & 4863 \\
+& & 1.50 & 2213 & 1.0 & 1.0 & 15992 \\
+\\
+BGK-BRT & No & 0.50 & 0.0007 & 11.73 & 12.56 & 4450 \\
+& & 0.75 & 0.9125 & 0.0320 & 0.0475 & 1805 \\
+& & 1.25 & 423.2 & 0.0051 & 0.0055 & 1905 \\
+& & 1.50 & 2213 & 1.0 & 1.0 & 8194 \\
+\\
+\label{tab:poise-power-law}
+\end{tabulary}
+\end{table}
+
+There are some interesting things to note from the power-law Poiseuille flow results.
+Although for $n = 0.5, 0.75, 1.25$ there is still the same efficiency for accuracy trade off when it comes to deciding between the BGK and MRT collision operators, for $n = 1.5$ the BGK operator with median filtering was the fastest, most accurate, and only solution that remained stable.
+In fact, besides for the case of $n = 0.5$, i.e. rather extreme shear-thinning, the BGK with median filtering was both fast and accurate.
+The BGK-BRT had a moderate improvement of accuracy and efficiency--with respect to the regular BGK collision scheme--for flows where the constitutive relationship was only somewhat nonlinear, i.e.
+for $n = 0.75, 1.25$.
+Note that none of the collision schemes were sufficiently accurate for the case of $n = 0.5$, and to improve this accuracy it would be necessary for a modeler to use a finer grid, and/or a smaller pressure gradient.
+
+\subsection{Lid-driven Cavity Flow} % MG: would it be of any use to show streamline plots from some of these flows?
+
+Lid-driven cavity flow is another good benchmark for two-dimensional fluid flow because it is sufficiently simple in terms of boundary conditions and implementation, while also complex in the flows and vortices it produces.
+The lid-driven benchmark was chosen for this numerical study because there are many results available in literature in which to compare with and because the vorticity of the flow coupled with the nonlinear constitutive equations should result in a challenge in terms of stability.
+Lid-driven cavity flow is generally characterized by a square cavity where a fluid velocity is prescribed tangential to the upper boundary and the remaining boundaries have a no-slip boundary condition.
+Lid-driven cavity flow is depicted in \Fref{fig:lid-driven}.
+
+\begin{figure}
+\includegraphics{figs/lid-driven}
+\caption{Schematic of lid-driven cavity flow; a velocity is prescribed tangential to the top boundary and no-slip is enforced at the remaining boundaries.}
+\label{fig:lid-driven}
+\end{figure}
+
+The lid-driven cavity simulations presented in the section were all simulated on a coarse, $100 \times 100$ lattice; a coarse lattice was chosen in order to highlight concerns with stability and accuracy.
+The simulations were run for either 50000 timesteps, or until convergence was met.
+Convergence was defined by,
+\begin{equation} \label{eq:convergence}
+\sum_{m=1}^{100} \sum_{i, j} \frac{|u_{i, j}^k - u_{i, j}^{k-m}|}{|u_{i, j}^{k-m}|},
+\end{equation}
+\noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current timestep.
+In reference to computational time, each of the simulations in this section were run on a single core of either an Intel I7-860 Quad-Core 2.80GHz processor, or a SOME OTHER PROCESSOR I NEED TO LOOK UP.
+The lid velocity was prescribed at $u_o = 0.1$.
+For the MRT relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$.
+All coordinate values presented in this section (used to describe the location of the center of vortices) are given normalized with respect to the length of the cavity side, i.e. as $(x / L, y / L)$.
+
+\subsubsection{Bingham Plastic Fluids}
+
+For the Bingham plastic numerical tests the Reynold's number was varied, $Re = 100, 1000, 5000, 1000$, and the Bingham number was varied, $Bn = 1, 10, 100$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
+The collision schemes that were tested were: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and MRT with $m = 10^8$ and median filtering.
+Tables \ref{tab:lid-bing1}--\ref{tab:lid-bing100} compare the center location of the main vortex to literature values taken from~\citet{syrakos2014performance}.
+The main vortex location is determined by calculating the stream function using Simpson's rule and finding where it attains a maximum.
+
+\begin{table}
+\caption{Bingham plastic, lid-driven cavity flow; $Bn = 1$.}
+\vspace{0.5cm}
+\begin{tabulary}{\linewidth}{r r l l r r r r}
+$Bn$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
+\hline \\
+1 & 100 & BGK     & No  & $10^5$ & (0.63, 0.79) & (0.63, 0.79) & 17685 \\
+  &     & BGK     & No  & $10^8$ &              & (0.63, 0.79) & 19398 \\
+  &     & BGK-BRT & No  & $10^8$ &              & (0.63, 0.79) & 22656 \\
+  &     & BGK     & Yes & $10^8$ &              & (0.63, 0.79) & 21535 \\
+  &     & MRT     & No  & $10^8$ &              & (0.63, 0.79) & 79040 \\
+  &     & MRT     & Yes & $10^8$ &              & (0.63, 0.79) & 85295 \\
+\\
+1 & 1000 & BGK     & No  & $10^5$ & (0.54, 0.57) & (0.54, 0.57) & 16109 \\
+  &      & BGK     & No  & $10^8$ &              & (0.54, 0.57) & 17035 \\
+  &      & BGK-BRT & No  & $10^8$ &              & (0.54, 0.57) & 16879 \\
+  &      & BGK     & Yes & $10^8$ &              & (0.54, 0.57) & 20170 \\
+  &      & MRT     & No  & $10^8$ &              & (0.54, 0.57) & 46048 \\
+  &      & MRT     & Yes & $10^8$ &              & (0.54, 0.57) & 55818 \\
+\\
+1 & 5000 & BGK     & No  & $10^5$ & (0.52, 0.53) & - & - \\
+  &      & BGK     & No  & $10^8$ &              & - & - \\
+  &      & BGK-BRT & No  & $10^8$ &              & - & - \\
+  &      & BGK     & Yes & $10^8$ &              & (0.54, 0.53) & 17248 \\
+  &      & MRT     & No  & $10^8$ &              & (0.51, 0.55) & 50572 \\
+  &      & MRT     & Yes & $10^8$ &              & (0.52, 0.53) & 54225 \\
+\\
+1 & 10000 & BGK     & No  & $10^5$ & N/A          & - & - \\
+  &       & BGK     & No  & $10^8$ &              & - & - \\
+  &       & BGK-BRT & No  & $10^8$ &              & - & - \\
+  &       & BGK     & Yes & $10^8$ &              & (0.56, 0.60) & 18186 \\
+  &       & MRT     & No  & $10^8$ &              & (0.48, 0.48) & 43246 \\
+  &       & MRT     & Yes & $10^8$ &              & (0.46, 0.54) & 50864 \\
+\\
+\end{tabulary}
+\label{tab:lid-bing1}
+\end{table}
+\begin{table}
+    \caption{Bingham plastic, lid-driven cavity flow; $Bn = 10$.}
+    \vspace{0.5cm}
+\begin{tabulary}{\linewidth}{r r l l r r r r}
+    $Bn$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
+    \hline \\
+10 & 100 & BGK     & No  & $10^5$ & (0.53, 0.87) & (0.54, 0.87) & 25217 \\
+   &     & BGK     & No  & $10^8$ &              & (0.55, 0.87) & 38285 \\
+   &     & BGK-BRT & No  & $10^8$ &              & (0.55, 0.87) & 29539 \\
+   &     & BGK     & Yes & $10^8$ &              & (0.54, 0.87) & 35817 \\
+   &     & MRT     & No  & $10^8$ &              & (0.53, 0.88) & 143741 \\
+   &     & MRT     & Yes & $10^8$ &              & (0.54, 0.88) & 135059 \\
+\\
+10 & 1000 & BGK     & No  & $10^5$ & (0.80, 0.85) & (0.78, 0.83) & 11525 \\
+   &      & BGK     & No  & $10^8$ &              & (0.78, 0.83) & 26671 \\
+   &      & BGK-BRT & No  & $10^8$ &              & (0.78, 0.83) & 19248 \\
+   &      & BGK     & Yes & $10^8$ &              & (0.79, 0.84) & 34143 \\
+   &      & MRT     & No  & $10^8$ &              & (0.79, 0.84) & 105136 \\
+   &      & MRT     & Yes & $10^8$ &              & (0.79, 0.84) & 111942 \\
+\\
+10 & 5000 & BGK     & No  & $10^5$ & (0.60, 0.55) & - & - \\
+   &      & BGK     & No  & $10^8$ &              & - & - \\
+   &      & BGK-BRT & No  & $10^8$ &              & - & - \\
+   &      & BGK     & Yes & $10^8$ &              & (0.52, 0.55) & 34638 \\
+   &      & MRT     & No  & $10^8$ &              & (0.55, 0.55) & 111274 \\
+   &      & MRT     & Yes & $10^8$ &              & (0.55, 0.53) & 121685 \\
+\\
+10 & 10000 & BGK     & No  & $10^5$ & N/A          & - & - \\
+   &       & BGK     & No  & $10^8$ &              & - & - \\
+   &       & BGK-BRT & No  & $10^8$ &              & - & - \\
+   &       & BGK     & Yes & $10^8$ &              & (0.49, 0.54) & 19351 \\
+   &       & MRT     & No  & $10^8$ &              & (0.53, 0.54) & 69249 \\
+   &       & MRT     & Yes & $10^8$ &              & (0.53, 0.53) & 69054 \\
+\\
+\end{tabulary}
+\label{tab:lid-bing10}
+\end{table}
+\begin{table}
+    \caption{Bingham plastic, lid-driven cavity flow; $Bn = 100$.}
+    \vspace{0.5cm}
+    \begin{tabulary}{\linewidth}{r r l l r r r r}
+        $Bn$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
+        \hline \\
+100 & 100 & BGK     & No  & $10^5$ & (0.51, 0.95) & (0.51, 0.95) & 13237 \\
+    &     & BGK     & No  & $10^8$ &              & (0.54, 0.96) & 26782 \\
+    &     & BGK-BRT & No  & $10^8$ &              & (0.53, 0.96) & 33095 \\
+    &     & BGK     & Yes & $10^8$ &              & (0.58, 0.96) & 34430 \\
+    &     & MRT     & No  & $10^8$ &              & (0.49, 0.95) & 161998 \\
+    &     & MRT     & Yes & $10^8$ &              & (0.54, 0.96) & 162085 \\
+\\
+100 & 1000 & BGK     & No  & $10^5$ & (0.53, 0.95) & (0.54, 0.95) & 42233 \\
+    &      & BGK     & No  & $10^8$ &              & (0.64, 0.95) & 47364 \\
+    &      & BGK-BRT & No  & $10^8$ &              & (0.60, 0.92) & 46289 \\
+    &      & BGK     & Yes & $10^8$ &              & (0.79, 0.95) & 48225 \\
+    &      & MRT     & No  & $10^8$ &              & (0.54, 0.95) & 190119 \\
+    &      & MRT     & Yes & $10^8$ &              & (0.55, 0.95) & 188242 \\
+\\
+100 & 5000 & BGK     & No  & $10^5$ & (0.93, 0.97) & - & - \\
+    &      & BGK     & No  & $10^8$ &              & - & - \\
+    &      & BGK-BRT & No  & $10^8$ &              & - & - \\
+    &      & BGK     & Yes & $10^8$ &              & (0.91, 0.95) & 57354 \\
+    &      & MRT     & No  & $10^8$ &              & (0.92, 0.96) & 163318 \\
+    &      & MRT     & Yes & $10^8$ &              & (0.93, 0.95) & 159523 \\
+\\
+100 & 10000 & BGK     & No  & $10^5$ & (0.92, 0.94) & - & - \\
+    &       & BGK     & No  & $10^8$ &              & - & - \\
+    &       & BGK-BRT & No  & $10^8$ &              & - & - \\
+    &       & BGK     & Yes & $10^8$ &              & (0.84, 0.88) & 32080 \\
+    &       & MRT     & No  & $10^8$ &              & (0.89, 0.91) & 152707 \\
+    &       & MRT     & Yes & $10^8$ &              & (0.89, 0.91) & 145722 \\
+\\
+\end{tabulary}
+\label{tab:lid-bing100}
+\end{table}
+
+Just as before with the Bingham plastic Poiseuille flow, if a modeler is using the BGK collision operator and entropic filtering is not available, a stress growth exponent of $m = 10^5$ yields faster, more accurate results (assuming the literature values are a good measure of accuracy) than using a larger stress growth exponent.
+A smaller stress growth exponent is also more effective at producing accurate results in the BGK collision scheme than placing bounds on the relaxation time (BGK-BRT).
+However, although the BGK with $m = 10^5$ was the fastest model in all cases, it was also--like the other two BGK models without median filtering--unstable for $Re \ge 5000$.
+
+For flows with $Re \ge 5000$, the BGK with $m = 10^8$ and median filtering produced numerically stable results, but did not always seem to be consistent with literature values.
+The apparent inaccuracy with regards to BGK with $m = 10^8$ and median filtering is probably due to how the numerical stability is enhanced--namely through artificial, nonphysical dissipation.
+In contrast to the nonphysical natural in which median filtering enhances stability, the stability enhancement used by MRT does not directly affect the macroscopic hydrodynamics of interest, so it again provides a stable, and typically, the most accurate solution.
+% MG: verify this sentence is correct
+The apparent superiority in terms of stability and accuracy of the MRT collision operator still comes at a price though.
+The MRT collision operator was, in general, 5-10 times slower than any of the BGK collision schemes.
+The extreme increase in computational expense is probably not due to the collision process itself, i.e.
+\eqref{eq:mrt-colop}, but instead calculating the strain-rate \eqref{eq:srtensor-mrt} for each iteration of the solution for $\mu_{app}$.
+
+For the sake of completeness, the MRT collision operator was also tested with median filtering applied.
+There does not seem to be any reason to recommend the MRT collision operator with median filtering in practice because the MRT collision operator is stable all on its own and the artificial dissipation introduced by median filtering has the potential to adversely affect the accuracy of the results.
+
+In summary, 
+\begin{itemize}
+    \item for low Reynold's number flows when computational efficiency is important, the BGK with $m = 10^5$ is recommended,
+    \item for high Reynold's number flows when computational efficiency is important, the BGK with $m = 10^8$ and median filtering is recommended,
+    \item and for high Reynold's number flows when accuracy is more important than computational efficiency, the MRT with $m = 10^8$ is recommended.
+\end{itemize}
+
 \subsubsection{Power-law Fluids}
 
-\subsection{Lid-driven Flow}
+For the power-law numerical tests the Reynold's number was varied, $Re = 100, 1000, 5000, 1000$, and the flow behavior index was varied, $n = 0.5, 1.5$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
+The collision schemes that were tested were: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (5) MRT with $m = 10^8$ and MRT with $m = 10^8$ and median filtering.
+\Fref{tab:lid-powerlaw05} and \Fref{tab:lid-powerlaw15} compare the center location of the main vortex to literature values taken from~\citet{li2014simulation}.
+The main vortex location is determined by calculating the stream function using Simpson's rule and finding where it attains a maximum.
+
+%MG: two values still need filled in in these tables
+\begin{table}
+\caption{Power-law, lid-driven cavity flow; $n = 0.5$.}
+\vspace{0.5cm}
+\begin{tabulary}{\linewidth}{r r l l r r r}
+$n$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
+\hline \\
+0.5 & 100 & BGK     & No  & (0.72, 0.78) & (0.71, 0.77) & 21429 \\
+    &     & BGK-BRT & No  &              & (0.71, 0.77) & 14547 \\
+    &     & BGK     & Yes &              & (0.72, 0.78) & 33221 \\
+    &     & MRT     & No  &              & (0.71, 0.77) & 77287 \\
+    &     & MRT     & Yes &              & (0.71, 0.77) & 110403 \\
+\\
+0.5 & 1000 & BGK     & No  & (0.58, 0.55) & - & - \\
+    &      & BGK-BRT & No  &              & - & - \\
+    &      & BGK     & Yes &              & (0.53, 0.59) & 24935 \\
+    &      & MRT     & No  &              & (0.53, 0.54) & 72102 \\
+    &      & MRT     & Yes &              & (0.53, 0.54) & 78300 \\
+\\
+0.5 & 5000 & BGK     & No  & (0.53, 0.52) & - & - \\
+    &      & BGK-BRT & No  &              & - & - \\
+    &      & BGK     & Yes &              & (0.63, 0.68) & 23517 \\
+    &      & MRT     & No  &              &              &       \\
+    &      & MRT     & Yes &              & (0.51, 0.58) & 54225 \\
+\\
+0.5 & 10000 & BGK     & No  & (0.53, 0.52) & - & - \\
+    &       & BGK-BRT & No  &              & - & - \\
+    &       & BGK     & Yes &              & (0.50, 0.55) & 22603 \\
+    &       & MRT     & No  &              &              &       \\
+    &       & MRT     & Yes &              & (0.50, 0.54) & 63892 \\
+\\
+\end{tabulary}
+\label{tab:lid-powerlaw05}
+\end{table}
+
+\begin{table}
+\caption{Power-law, lid-driven cavity flow; $n = 1.5$.}
+\vspace{0.5cm}
+\begin{tabulary}{\linewidth}{r r l l r r r}
+$n$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
+\hline \\
+1.5 & 100 & BGK     & No  & (0.56, 0.73) & (0.57, 0.73) & 10897 \\
+    &     & BGK-BRT & No  &              & (0.57, 0.73) & 6239 \\
+    &     & BGK     & Yes &              & (0.56, 0.73) & 11847 \\
+    &     & MRT     & No  &              & (0.56, 0.73) & 39017 \\
+    &     & MRT     & Yes &              & (0.56, 0.73) & 96033 \\
+\\
+  1.5 & 1000 & BGK   & No  & (0.55, 0.64) & (0.54, 0.61) & 19764 \\
+    &      & BGK-BRT & No  &                & (0.54, 0.61) & 11949 \\
+    &      & BGK     & Yes &                & (0.54, 0.61) & 24516 \\
+    &      & MRT     & No  &                & (0.54, 0.61) & 58714 \\
+    &      & MRT     & Yes &                & (0.54, 0.61) & 69884 \\
+\\
+  1.5 & 5000 & BGK     & No  & (0.53, 0.61) & (0.52, 0.57) & 20147 \\
+    &      & BGK-BRT & No  &              & (0.52, 0.57) & 12116 \\
+    &      & BGK     & Yes &              & (0.52, 0.57) & 23140 \\
+    &      & MRT     & No  &              & (0.53, 0.57) & 60584 \\
+    &      & MRT     & Yes &              & (0.52, 0.58) & 58519 \\
+\\
+  1.5 & 10000 & BGK     & No  & (0.51, 0.55) & (0.53, 0.55) & 21570 \\
+    &       & BGK-BRT & No  &              & (0.53, 0.55) & 12841 \\
+    &       & BGK     & Yes &              & (0.53, 0.55) & 21570 \\
+    &       & MRT     & No  &              & (0.49, 0.56) & 61453 \\
+    &       & MRT     & Yes &              & (0.49, 0.55) & 67546 \\
+\\
+\end{tabulary}
+\label{tab:lid-powerlaw15}
+\end{table}
+
+For the shear-thinning results, $n = 0.5$, the BGK and BGK-BRT collision schemes were unstable when $Re \ge 1000$.
+Although there still appears to be somewhat of a trade off between computational efficiency and accuracy in regards to the BGK with median filtering and MRT collision schemes, it does not seem to be nearly as drastic as when simulating Bingham plastic lid-driven flow; e.g.
+neither collision scheme consistently agrees with the literature values; and the MRT collision operator is only 2-3 times slower than the BGK with median filtering.
+
+For the shear-thickening results, $n = 1.5$, there are no issues of stability.
+What is interesting is that all of the collision schemes produce similar results for each of the cases (with the exception of the high Reynold's number case, $Re = 10000$), and the BGK-BRT is consistently the fastest solution.
+The reason the BGK-BRT may be the fastest is that because the collision frequency, $\omega$, is bounded, and consequently $\mu_{app}$ is bounded, meaning the solution to the constitutive equation may be converging to a bound with less iterations than the other collision schemes.
+
+In summary, 
+\begin{itemize}
+    \item for shear-thinning flows, either the BGK with median filtering or MRT collision schemes are recommended for stability reasons,
+    \item for shear-thickening flows, there is much less concern for numerical instability.
+\end{itemize}
 
@@
 
+A numerical investigation into the accuracy, stability, and efficiency of LBM collision models when applied to non-Newtonian flows was presented.
+The numerical investigation included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluid flows.
+Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
+The results showed that:
+\begin{itemize}
+  \item For high yield stress fluid, Poiseuille flows only the MRT collision operator did not suffer from numerical noise.
+  \item Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flows; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
+  \item In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
+  \item To summarize plastic, lid-driven cavity flows,
+    \begin{itemize}
+  \item for low Reynold's number flows when computational efficiency is important, the BGK with $m = 10^5$ is recommended,
+  \item for high Reynold's number flows when computational efficiency is important, the BGK with $m = 10^8$ and median filtering is recommended,
+  \item and for high Reynold's number flows when accuracy is more important than computational efficiency, the MRT with $m = 10^8$ is recommended.
+    \end{itemize}
+  \item To summarize power-law, lid-driven cavity flows,
+    \begin{itemize}
+  \item for shear-thinning either the BGK with median filtering or MRT collision schemes are recommended for stability reasons,
+  \item for shear-thickening, lid-driven cavity flows, there is much less concern for numerical instability.
+    \end{itemize}
+\end{itemize}
+
 \section*{Acknowledgements}
+
+% MG: should we mention: code is open source? input files used in this study are also freely available?
 

### 2016-07-07 main.tex
added investigation into nonequilibrium moments for bing poise

@@
 In order to ensure that this phenomena was a side effect of all entropic filtering and not just entropic median filtering with $n_s = 2.7$, entropic filtering was tested with median filtering and Ehrenfests' regularization for $n_s = 2.7$ and for $\Delta S$ thresholds of $\delta = 10^{-4}$ and $\delta = 10^{-5}$.
-All filtering schemes suffered from the approximately the same level of noise.
+All filtering schemes suffered from approximately the same level of noise.
 
@@
 
-% A natural question may be to ask ``if the strength of the MRT collision operator is damping out the ghost modes, then is the noise for the BGK with $m = 10^8$ due to the ghost modes?''.
-% MG: maybe plot ghost modes for BGK and MRT
-
 \begin{figure}
@@
     \label{fig:feq-vs-f_mrt}
+\end{figure}
+
+A natural question would be to ask ``if the strength of the MRT collision operator is damping out the ghost modes, then is the noise for the BGK with $m = 10^8$ (with and without median filtering) indeed due to the ghost modes?''.
+\Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment with respect to time for each of the collision schemes and was calculated using:
+\begin{equation}
+\frac{||\boldsymbol{\epsilon} - \boldsymbol{\epsilon}^{eq}||_2}{||\boldsymbol{\epsilon}^{eq}||_2}
+\end{equation}
+\noindent where $\epsilon_j = 4f_0(\pos_j) - 2f_1(\pos_j) - 2f_2(\pos_j) - 2f_3(\pos_j) - 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) + f_7(\pos_j) + f_8(\pos_j)$ is related to the square of the energy, and $\pos_j = (16, j)$ for $j = 1, 2, ..., 64$, i.e.
+$\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
+The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT or BGK with $m = 10^5$ collision schemes--which suggests that the numerical noise the BGK with $m = 10^8$ suffered from was in part due to the $\epsilon$ moment.
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon}
+    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \label{fig:epsilon}
+\end{figure}
+
+\Fref{fig:qx} shows a measure of the nonequilibrium $q_x$ moment, or energy flux in the x-direction, with respect to time for each of the collision schemes and was calculated using:
+\begin{equation}
+    \frac{||\mathbf{q_y} - \mathbf{q_y}^{eq}||_2}{||\mathbf{q_y}^{eq}||_2}
+\end{equation}
+\noindent where $q_{y_j} = - 2f_2(\pos_j) + 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) - f_7(\pos_j) - f_8(\pos_j)$ and $\pos_j$ were taken across the height of the channel in the same manner as with the $\epsilon$ moment.
+The values measured for the BGK with $m = 10^8$ and median filtered were consistently greater than any other collision scheme.
+The $q_x$ moment was likely the cause of the significant error and noise in the BGK with $m = 10^8$ and median filtering solution.
+As expected, the $q_x^{neq}$ relative $L_2$ for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $q_x^{neq}$ relative $L_2$ was negligible for the MRT collision scheme.
+These results are consistent with the solution noise and error.
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figs/poise-bingham/qx}
+    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \label{fig:qx}
+\end{figure}
+
+It is likely that the nonequilibrium $\epsilon$ and $q_x$ moment cause oscillations at the lattice level that have a cumulative effect in time.
+\Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time.
+\Fref{fig:qx-cumulative} suggests that the $q_x^{neq}$ dominates and is the primary source of oscillations.
+If an LBM collision scheme is to be developed for simulating high yield stress flows that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it should focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon_cumulative}
+    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \label{fig:epsilon-cumulative}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figs/poise-bingham/qx_cumulative}
+    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \label{fig:qx-cumulative}
 \end{figure}
@@
   \item For high yield stress fluid, Poiseuille flows only the MRT collision operator did not suffer from numerical noise.
+  \begin{itemize}
+      \item The noise appeared to be due to oscillations arising from high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
+      \item If a collision scheme is to be developed for high yield stress, Poiseuille-type flows that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, then it should focus on dampening the nonequilibrium energy flux moment.
+    \end{itemize}
   \item Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flows; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.

### 2016-07-07 main.tex
updated captions for noneq moment figures

@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon}
-    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \caption{Evoluation of relative $L_2$, $\epsilon^{neq}$ with time.
+The norm of the $\epsilon^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $\epsilon$ moment.}
     \label{fig:epsilon}
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/qx}
-    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \caption{Evoluation of relative $L_2$, $q_x^{neq}$ with time.
+The norm of the $q_x^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $q_x$ moment.}
     \label{fig:qx}
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon_cumulative}
-    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \caption{Cumulative relative $L_2$, $\epsilon^{neq}$ with time.
+Oscillations can have a positive effect on each other, building up.
+The cumulative relative $L_2$, $\epsilon^{neq}$ is a measure of how much oscillations due to the $\epsilon$ moment may have been building in time.}
     \label{fig:epsilon-cumulative}
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/qx_cumulative}
-    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \caption{Cumulative relative $L_2$, $q_x^{neq}$ with time.
+Oscillations can have a positive effect on each other, building up.
+The cumulative relative $L_2$, $q_x^{neq}$ is a measure of how much oscillations due to the $q_x$ moment may have been building in time.}
     \label{fig:qx-cumulative}

### 2016-07-07 main.tex
added notes about accuracy of strain-rate tensor calculation and bounds on s_i

@@
 The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which can become computationally expensive.
-The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the advantage that computing the strain rate is second-order accurate and local to each node~\cite{kruger2010second}.
+The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the advantage that computing the strain rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
 This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
@@
 % MG: verify this is correct
-The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i, i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
+The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i \in [0, 2], i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
 In the case when $s_0 = s_1 = ... = s_8 = \omega$ ($\omega$ is the collision frequency in the BGK sense), the MRT collision operator is equivalent to the BGK collision operator.
@@
 \end{equation} % MG: verify this
+
+\noindent Computing the strain-rate tensor by either \eqref{eq:srtensor-bgk} or \eqref{eq:srtensor-mrt} is second order accurate in space~\cite{kruger2009shear,kruger2010second}.
 

### 2016-07-09 thesis.tex
converted to thesis format, started ch2

--- 2016-07-07 main.tex
@@
 
-\documentclass{article}
+\documentclass[pdftex,ms]{pittetd}
+
+\usepackage[pdftex]{graphicx}
 
@@
 
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{amssymb}
 \usepackage{mathtools}
-\usepackage{mathabx}
 \usepackage{bm}
@@
 \usepackage{pbox}
+
+\patch{amsmatch}
+\patch{amsthm}
 
@@
 
-\title{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows}
-%\title{A Comparison of Lattice Boltzmann Method Collision Operators for Accuracy, Stability and Efficiency in Simulating Non-Newtonian Flows}
-
-\author{{Matthew Grasinger}
-\thanks{Email: \textit{grasingerm@pitt.edu}}}
-
-\author{{Scott Overacker}
-\thanks{Email: \textit{sdo10@pitt.edu}}}
-
-\author{John Brigham
-\thanks{Email: \textit{brigham@pitt.edu}}}
-\affil{Civil and Environmental Engineering Department, University of Pittsburgh}
+\title{How the Lattice Boltzmann Method changed my life forever!}
+% The optional argument is the version of the title that will appear in Acrobat Reader's Document Info dialog box.
+\author{Matthew Grasinger}
+\degree{BS in Civil and Environmental Engineering, University of Pittsburgh, Pittsburgh, PA 2013}
+\date{July 20, 2016}
+\keywords{lattice Boltzmann method, non-Newtonian flow, free surface flow, wellbore cementing}
+\subject{Draft}
+\school{Swanson School of Engineering}
+\degreesought{Master of Science}
+
+\chapterfloats
 
 \begin{document}
-
-\pagenumbering{gobble}
 \maketitle
-\newpage
+
+% For the committee membership page, you have to provide the names and affiliations of the members. The first one will
+% be treated by pittetd as the committee chair (thesis/dissertation advisor).
+\committeemember{John C. Brigham, PhD, Senior Lecturer, School of Engineering and Computing Sciences at Durham University}
+\committeemember{Julie M. Vandenbossche, PhD, Associate Professor, Department of Civil and Environmental Engineering}
+\committeemember{Anthony T. Iannacchione, PhD, Associate Professor, Department of Civil and Environmental Engineering}
+
+\school{Swanson School of Engineering}
+\makecommittee
+\copyrightpage
+
+\begin{abstract}
+\end{abstract}
+
 \tableofcontents
-\newpage
-\pagenumbering{arabic}
+\listoftables
+\listoffigures
+
+%\preface
+% acknowledgements and such
+
+\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows}
 
@@
 Various schemes and strategies for incorporating the physics of non-Newtonian flows and yet maintaining a stable numerical method have been developed and studied.
-The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free}, is to make the collision frequency--which is proportional to apparent viscosity--variable and dependent on the local strain rate.
+The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}, is to make the collision frequency--which is proportional to apparent viscosity--variable and dependent on the local strain rate.
 A potential issue with the stability of the variable relaxation time approach is that as the collision frequency, $\omega$, approaches $2$ the viscosity approaches zero and overrelaxation occurs; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
@@
 
-The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as: $f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \colop_i(f)$, where, for D2Q9, $i = 0, 1, ..., 8$.
+The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as: $f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \Omega_i(f)$, where, for D2Q9, $i = 0, 1, ..., 8$.
 
@@
 \begin{equation}
-    \frac{||\mathbf{q_y} - \mathbf{q_y}^{eq}||_2}{||\mathbf{q_y}^{eq}||_2}
-\end{equation}
-\noindent where $q_{y_j} = - 2f_2(\pos_j) + 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) - f_7(\pos_j) - f_8(\pos_j)$ and $\pos_j$ were taken across the height of the channel in the same manner as with the $\epsilon$ moment.
+    \frac{||\mathbf{q_x} - \mathbf{q_x}^{eq}||_2}{||\mathbf{q_x}^{eq}||_2}
+\end{equation}
+\noindent where $q_{x_j} = - 2f_1(\pos_j) + 2f_3(\pos_j) + f_5(\pos_j) - f_6(\pos_j) - f_7(\pos_j) + f_8(\pos_j)$ and $\pos_j$ were taken across the height of the channel in the same manner as with the $\epsilon$ moment.
 The values measured for the BGK with $m = 10^8$ and median filtered were consistently greater than any other collision scheme.
@@
 for $n = 0.75, 1.25$.
-Note that none of the collision schemes were sufficiently accurate for the case of $n = 0.5$, and to improve this accuracy it would be necessary for a modeler to use a finer grid, and/or a smaller pressure gradient.
+Note that none of the collision schemes were sufficiently accurate for the case of $n = 0.5$, and to improve this accuracy it would be necessary for a modeler to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
 
@@
 \begin{equation} \label{eq:convergence}
-\sum_{m=1}^{100} \sum_{i, j} \frac{|u_{i, j}^k - u_{i, j}^{k-m}|}{|u_{i, j}^{k-m}|},
+\sum_{m=1}^{100} \sum_{i, j} \frac{|u_{i, j}^k - u_{i, j}^{k-m}|}{|u_{i, j}^{k-m}|} < 1.0 \times 10^{-7},
 \end{equation}
@@
 
+% MG: should we mention: code is open source? input files used in this study are also freely available?
+
 \section*{Acknowledgements}
 
-% MG: should we mention: code is open source? input files used in this study are also freely available?
+\chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method}
+
+\section*{Abstract}
+
+Wellbore cementing is
+
+\section{Introduction}
+
+% what is wellbore cementing?
+
+Cementing of oil and gas wellbores is important because if cement does not completely fill the annulus then pathways can form for fluids and gasses to flow through.
+The pathways need not be large--a pathway of SOMETHING is sufficient for enough loss of pressure at the wellhead for the well to lose SOME percent production.
+There are also environmental concerns as oils and gasses can migrate into ground water or the atmosphere causing environmental damage.
+
+Of critical importance to drilling operations of all kinds is that barriers are placed to prevent gasses and fluids from migrating from one geological zone to another (i.e., zonal isolation is achieved).
+If zonal isolation is not achieved, gasses and fluids may migrate through the well to the surface and up into the atmosphere--potentially causing pollution, or at the very least, contributing to an increase in greenhouse gasses--or, worse yet, formation gasses and fluids can migrate into aquifers, potentially harming wild life and people.
+Furthermore, it can be argued that the most important, yet vulnerable, barrier in terms of achieving zonal isolation of a wellbore is the cement annulus that is created between the casing and the formation.
+There are various mechanisms by which this cement annulus can fail.
+According to previous studies~\cite{Bit02,Che85,Lev79,Ste88,Tal93,Zul12}, the most prevalent failure mechanisms of the cement annulus are:
+\begin{itemize}
+\item Stresses developing in the cement annulus as a result of temperature gradients, moisture gradients, overburden pressure, and chemical shrinkage of the cement matrix.
+These stresses can cause cracking in the cement, or debonding at either the casing--cement or cement--formation interfaces.
+\item Poor fill of the annulus.
+If cement slurry does not fill the annulus then if drilling mud is left behind it can dry, crack, and provide a weak path for fluids and gasses under pressure to push through, or if a dry hole is is cemented and not filled properly, fluids and gasses can travel through channels and voids left in the annular space.
+\item Gas channeling through the annulus during cement curing.
+Initially, after the annulus is poured, the cement column transmits its full hydrostatic pressure against the rock formation.
+As the cement annulus hydrates, it solidifies, becoming stiffer and more self-supporting.
+When the column begins to support itself, the hydrostatic pressure of the cement column--against the rock formation--begins to drop.
+If the hydrostatic pressure of the cement column drops below the pressure of a formation gas before the cement annulus has developed adequate strength to resist it, the formation gas can penetrate int the cement annulus and form channels, or degrade the cement annulus in other ways.
+\end{itemize}
+% what research has been done that is considered classical?
+
+% why do we need modelling?
+% Before the end of this paragraph--modeling is important because...
+
+In order to better understand, and thus improve wellbore cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
+For example, \citet{Bei77} developed a mathematical model to ``describe the miscible displacement of drilling muds by cement slurries under laminar flow conditions''.
+Among their assumptions were that the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the y-coordinate.
+The model was then used to explore the effect of various parameters on the displacement efficiency of the spacer, such as the density ratio between the mud and the spacer, the viscosity ratio between the mud and the spacer, the displacement rate, the displaced phase yield stress, and the displacing phase yield stress.
+It was concluded that displacement efficiency could be improved by increasing the density of the displacing phase, increasing the viscosity of the displacing phase, (in general) decreasing the displacement rate, decreasing the yield stress of the displaced phase, and increasing the yield stress of the displacing phase.
+\citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
+The model was more general than the model developed by \citeauthor{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
+Annulus eccentricity occurs often in wellbores and can sometimes be extreme as it is difficult to keep long strings of casing centered in an imperfectly drilled hole.
+The bulk fluid motion of the spacer and drilling mud during primary cementing were investigated.
+The intended use of the model was not to make general statements about what properties of the spacer and drilling mud were desirable, but to instead be used iteratively during the design process for the cement job.
+Results showed that the model met its intended purpose, and in particular, was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
+\citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
+The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either \citeauthor{Bei77} or \citeauthor{Bit02}.
+The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
+\citeauthor{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
+The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
+% talk about new LBM study
+
+Overall, past studies have tended to idealize the rock formation as a straight wall.
+Although this is a desirable simplification to make for capturing bulk fluid penomean, it does not accurately describe what is happening locally at the surface of the rock formation.
+Flow near the surface of the formation, and in and around the imperfections of the formation geometry, could influence the bond of the cement annulus to the rock formation, and the fill of the cement slurry very near to the rock formation.
+If an adequate bond does not develop, debonding can occur as a result of thermal and mechanical stresses.
+Even if most of the annulus is filled by the cement slurry, a small gap at the cement--formation interface that stretches for a long enough span of the wellbore would still be susceptible to fluid and gas migration, i.e.
+loss of zonal isolation.
+One could even argue that the characteristic of the cement slurry flow is most important local to the cement--formation interface, and that considering imperfections in the rock formation is imperative to determing whether zonal isolation has been achieved.
+
+The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features.
+The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation, among other benefits.
+LBM also has other advantages as an approach for simulating wellbore cementing:
+\begin{itemize}
+\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
+Many fluids present in wellbore cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
+\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it results in interface capturing instead of interface tracking, i.e.
+BLAH BLAH BLAH and cite something.
+\item LBM is easily written in parallel.
+As hardware architectures shifts from one or a few cores to, especially in the case of scientific computing, several CPUs, it becomes increasingly important that code for simulating complex and computationally expense phenomena be written in parallel.
+\end{itemize}
+What follows is blahdy blah blah an LBM model foook.
+
+The goal of this paper is to apply an LBM algorithm for free surface flow to the problem of wellbore cementing, and in doing so,
+
+\section{Numerical Methods}
+
+\subsection{Lattice Boltzmann Method}
+
+\subsection{Yield Stress Flow}
+
+\subsection{Free Surface Flow}
+
+\subsection{Inlet and Outlet Boundary Conditions}
+
+\section{Wellbore Cementing}
+
+\
 

### 2016-07-11 thesis.tex
finished first draft of methods for cementing paper

@@
 \usepackage[square,numbers,sort&compress]{natbib}
+\usepackage{enumitem}
+\setlistdepth{9}
+
+\newlist{myEnumerate}{enumerate}{9}
+\setlist[myEnumerate,1]{label=(\arabic*)}
+\setlist[myEnumerate,2]{label=(\Roman*)}
+\setlist[myEnumerate,3]{label=(\Alph*)}
+\setlist[myEnumerate,4]{label=(\roman*)}
+\setlist[myEnumerate,5]{label=(\alph*)}
+\setlist[myEnumerate,6]{label=(\arabic*)}
+\setlist[myEnumerate,7]{label=(\Roman*)}
+\setlist[myEnumerate,8]{label=(\Alph*)}
+\setlist[myEnumerate,9]{label=(\roman*)}
 
@@
 The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
+\citet{zhao2016lattice} developed a model simulating wellbore cementing using the Lattice Boltzmann Method.
+\citeauthor{zhao2016lattice} used the model to investigate the displacement interface shape in a horizontal, eccentric annulus; and in order to simplify the model from a 3D to 2D flow, used the classical Hele-Shaw approach.
+It was found, in general, that a greater density ratio between the displacing phase, i.e.
+cement slurry, and displaced phase, i.e.
+drilling mud, and lower eccentricity, results in better displacement in a horizontal, eccentric annulus.
+
 % talk about new LBM study
@@
 Overall, past studies have tended to idealize the rock formation as a straight wall.
-Although this is a desirable simplification to make for capturing bulk fluid penomean, it does not accurately describe what is happening locally at the surface of the rock formation.
+Although this is a desirable simplification to make for capturing bulk fluid phenomena, it does not accurately describe what is happening locally at the surface of the rock formation.
 Flow near the surface of the formation, and in and around the imperfections of the formation geometry, could influence the bond of the cement annulus to the rock formation, and the fill of the cement slurry very near to the rock formation.
@@
 loss of zonal isolation.
-One could even argue that the characteristic of the cement slurry flow is most important local to the cement--formation interface, and that considering imperfections in the rock formation is imperative to determing whether zonal isolation has been achieved.
-
-The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features.
+One could even argue that the characteristic of the cement slurry flow is most important local to the cement--formation interface, and that considering imperfections in the rock formation is imperative to determining whether zonal isolation has been achieved.
+
+The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features (i.e., the purpose is an extension to the work present in~\citet{grasinger2015simulation}).
 The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation, among other benefits.
@@
 Many fluids present in wellbore cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
-\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it results in interface capturing instead of interface tracking, i.e.
-BLAH BLAH BLAH and cite something.
+\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components for reasons similar to why the no-slip boundary condition works well with complex geometries.
 \item LBM is easily written in parallel.
@@
 \end{itemize}
-What follows is blahdy blah blah an LBM model foook.
-
-The goal of this paper is to apply an LBM algorithm for free surface flow to the problem of wellbore cementing, and in doing so,
-
-\section{Numerical Methods}
+The current work focuses on the problem of cementing a dry hole, i.e. an annulus that is not filled with drilling mud prior to cementing.
+In order to numerically investigate this problem, an LBM model for non-Newtonian and free-surface flow is developed.
+The details of the LBM model are outlined in \Fref{sec:numerical-methods}.
+What follows in \Fref{sec:results} is a preliminary study on how different characteristics of flow, e.g.
+cement slurry yield stress, cement slurry inlet velocity, etc., perform in regards to fill of the annulus for different wellbores with imperfect rock formation geometries.
+\Fref{sec:conclusion} are conclusions resulting from the numerical study and a discussion of future work.
+
+\section{Numerical Methods} \label{sec:numerical-methods}
 
@@
 
+% We need to explain/justify why 2D vs. 3D
+
 \subsection{Yield Stress Flow}
@@
 
-\subsection{Inlet and Outlet Boundary Conditions}
-
-\section{Wellbore Cementing}
-
-\
+A free-surface flow is a flow in which two-fluids exist but the dynamics of one fluid does not need to be explicitly modeled.
+Typically, the phase of the primary fluid, or the fluid of interest, is a liquid; and the secondary phase that does not need to be explicitly modeled is a gas, often times air at atmospheric pressure.
+The primary fluid is modeled using a standard CFD method, such as LBM, and the interaction of the secondary fluid on the primary fluid is modeled as a boundary condition at the interface between the two fluids.
+The interface is what is referred to as the free-surface.
+The advantage of modeling a multiphase multicomponent flow as a free-surface is that because the dynamics of the secondary fluid are not explicitly tracked and simulated the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
+The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in \citet{korner2005lattice,thurey2005interactive}.
+
+\subsection{Capturing the Free-Surface}
+
+Consider at each node a cell of size $\text{d}x \times \text{d}x$ where $\text{d}x$ is the distance between nodes and the cell is centered at the node.
+In order to track where the primary fluid is and where it is not, i.e.
+which cells contain the primary fluid, a variable is introduced, $\epsilon$, which describes what fraction of the cell is covered by the primary fluid.
+Cells that have a fluid fraction of $\epsilon = 0$ contain no primary fluid, i.e.
+are empty; cells that have a fluid fraction of $\epsilon = 1$ are filled; and cells that have a fluid fraction $0 < \epsilon < 1$ are neither completely filled nor are they completely empty.
+The fluid fraction concept is similar to how fluids are tracked in Volume of Fluid (VOF) methods.
+
+\subsubsection{Mass Transfer}
+
+It is important to note that the particle distributions, $f_0, f_1, ..., f_8$, cannot be used to directly determine the fluid fraction of a cell.
+Neither can the zeroth velocity moment of the particle distributions, $\rho = \sum_i f_i$.
+For the algorithm used in this work, neither can be considered a measure of the amount of primary fluid in a cell, i.e. $\epsilon(\pos, t) \neq \epsilon(\mathbf{f(\pos, t)}, \rho(\pos, t))$.
+Instead, each $f_i$ is related to the amount of primary fluid advection in each of the $\pvel_i$ discrete velocity directions.
+To track the fluid fraction and how it changes, another variable is introduced, $M(\pos, t)$, which denotes the fluid mass in each cell.
+The fluid mass is related to the fluid fraction by $\epsilon = M / \rho$.
+The change in mass, $\Delta M(\pos, t)$ is related to $\mathbf{f}(\pos, t)$ and particle distribution functions in the neighborhood of $\pos$.
+Given initial conditions on $M(\pos, 0)$ and a fixed time step size $\Delta t $, the fluid mass is determined by $M(\pos, t_n) = M(\pos, 0) + \Delta t \sum_{k=1}^n \Delta M(\pos, t_k)$.
+
+\newcommand{\opidx}{{\bar{i}}}
+\newcommand{\fset}{\mathcal{F}}
+\newcommand{\iset}{\mathcal{I}}
+\newcommand{\gset}{\mathcal{G}}
+\newcommand{\domain}{\mathcal{X}}
+
+\subsubsection{Cell States} \label{sec:mass-transfer}
+
+Each of the cells in the domain are assigned one of three states: fluid, interface, or gas.
+The set of fluid, interface, and gas cells are denoted by $\fset$, $\iset$, and $\gset$, respectively.
+A fluid cell is a cell that is entirely covered by the primary fluid and a gas cell is a cell that is completely empty, or does not contain any primary fluid.
+An interface is partially covered, and there is always a layer of interface cells between fluid and gas cells so that a fluid cell and a gas cell are never neighbors.
+The change in mass is calculated across fluid and interface cells for each discrete velocity direction as follows:
+\begin{equation} \label{eq:dM}
+\Delta M_i(\pos, t) = \begin{cases}
+f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t), & \pos \in \fset \text{ or } \pos + \pvel_i \in \fset \\
+\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \in \iset \\
+0, & \pos + \pvel_i \in \gset
+\end{cases}
+\end{equation}
+\noindent where $\pvel_\opidx$ represents the discrete velocity vector that is opposite in direction to $\pvel_i$, i.e. $\pvel_\opidx = - \pvel_i$.
+\eqref{eq:dM} states that when the border of the two cells is covered with fluid, the change in mass for a discrete velocity direction is equal to the difference of the particle distribution entering the cell from the opposite direction and particle distribution leaving the cell in that direction.
+When the border of the two cells are not covered with fluid, i.e.
+when both cells are interface cells, the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the border that is covered in fluid.
+The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M(\pos, t)$.
+Finally, note that $\Delta M_i(\pos, t) = -\Delta M_\opidx(\pos + \pvel_i \Delta t, t)$ for all $\pos \in \domain$ where $\domain$ is the domain.
+As a consequence, it is clear that $\sum_{\pos \in \domain}\Delta M(\pos, t) = 0$, i.e. that mass is conserved for the mass transfer step.
+
+\newcommand{\tconst}{\delta_{trans}}
+
+Tracking the transport of mass throughout the domain is an important step in determining what parts of the domain are covered in the primary fluid and what are not.
+However, for the free-surface, i.e. layer of interface cells between the fluid cells and gas cells, to change shape and propagate, cells need to be able to transition state.
+Cell states change based on the rule set presented in \Fref{tab:cell-transition-rules} where $\tconst$ is a small, computational constant that keeps cells from transitioning back and forth between fluid and interface or interface and gas states in subsequent time steps.
+To summarize \Fref{tab:cell-transition-rules}, gas cells cannot transition to fluid cells and vice versa, interface cells transition to fluid cells when their mass has exceeded $\rho$ by a computational constant, and gas cells when their mass is less than zero minus a computational constant.
+In addition, fluid cells and gas cells transition to the interface state in order to maintain continuity of the interface, i.e.
+to ensure there is always a layer of interface cells in between fluid cells and gas cells.
+An example of a cell transitioning state is depicted in \Fref{fig:cell-transition}.
+
+\begin{table}
+\caption{Cell state transition rules.}
+\begin{tabulary}{\linewidth}{l|L|L}
+\textbf{Current State} & \textbf{Condition} & \textbf{Transition} \\
+\hline &  & \\
+$(\pos, t) \in \iset$ & $M(\pos, t) < -\tconst$ & $(\pos, t + \Delta t) \rightarrow \gset$, $M(\pos, t + \Delta t) = 0$ \\
+                      & $M(\pos, t) > \tconst + \rho(\pos, t)$ & $(\pos, t + \Delta t) \rightarrow \fset$, $M(\pos, t + \Delta t) = \rho(\pos, t + \Delta t)$ \\
+                      &\\
+$(\pos, t) \in \fset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \gset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
+&\\
+$(\pos, t) \in \gset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \fset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
+\end{tabulary}
+\label{tab:cell-transition-rules}
+\end{table}
+
+\begin{figure}
+\centering
+\includegraphics{figs/cell-transition}
+\caption{Example of cell state transitions.
+The interface cell highlighted in red on the left is transitioned to a gas cell as it has emptied.
+On the right, the three cells highlighted in red were transitioned from fluid to interface cells in order to keep the interface continuous.}
+\label{fig:cell-transition}
+\end{figure}
+
+\subsubsection{Updating Cell States and Mass Redistribution} \label{sec:cell-updates}
+
+The rules for updating cell states in \Fref{tab:cell-transition-rules} are fairly intuitive, however, there are some bookkeeping nuances that must be considered in order to ensure mass is properly conserved and interface continuity is achieved; e.g.
+note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells.
+The algorithm for updating cell states follows that which was presented in~\citet{thurey2005interactive}.
+It is presented here for the sake of completeness and for outlining minor modifications.
+\begin{enumerate}
+\item Initialize a variable that represents the global excess mass, e.g. $M_{gex} = 0$.
+\item For all cells $\pos \in \iset$, store cells that should be transitioned to $\fset$ in one data structure, $\iset_f$, and store cells that should be transitioned to $\gset$ in another, $\iset_g$.
+\item \label{enum:prep-fnbrs} Loop over cells that will be transitioned from interface to fluid cells, i.e. loop over $\iset_f$, and prepare their neighborhoods for transition:
+	\begin{enumerate}
+	\item Calculate the average density, $\rho_{avg}$, and velocity, $\mathbf{u}_{avg}$, of each neighborhood.
+	\item For all cells in the neighborhood of the new fluid cell:
+		\begin{enumerate}
+			\item If the cell is a gas cell, transition to interface state and set $f_i = f_i^{eq}(\rho_{avg}, \mathbf{u}_{avg})$, $\rho = \rho_{avg}$, and $\mathbf{u} = \mathbf{u}_{avg}$.
+			\item If the cell is an interface cell in $\iset_g$, remove it from $\iset_g$.
+		\end{enumerate}
+	\end{enumerate}
+\item \label{enum:prep-gnbrs} Loop over cells that will be transitioned from interface to gas cells, i.e. loop over $\iset_g$, and prepare their neighborhoods for transition:
+	\begin{enumerate}
+	\item For all cells in the neighborhood of the new gas cell:
+		\begin{enumerate}
+			\item If the cell is a fluid cell, transition to interface state.
+		\end{enumerate}
+	\end{enumerate}
+\item \label{enum:rloop-fluid} Loop over cells that will be transitioned from interface to fluid cells, i.e. loop over $\pos \in \iset_f$, and:
+\begin{enumerate}
+	\item Let $M_{ex} = M - \rho$.
+	\item Calculate the unit norm outward from the interface, i.e. toward empty space, $\mathbf{n}$.
+	Initialize $v_{sum} = 0$.
+	\item For all cells in the neighborhood,
+	\begin{enumerate}
+		\item if an interface cell, store in a data structure, $\iset_{r}$, and for the discrete velocity vector direction, $\pvel_i$, of the respective neighbor, let $v_i = \begin{cases}
+					\mathbf{n} \cdot \pvel_i, & \mathbf{n} \cdot \pvel_i > 0 \\
+					0, & \text{otherwise}
+				\end{cases}$.
+		\item $v_{sum} = v_{sum} + v_i$.
+	\end{enumerate}
+	\item If $v_{sum} \ge 0$, then for each cell, $\pos_j \in \iset_r$, redistribute the excess mass according to $M(\pos_j) = M(\pos_j) + M_{ex} \frac{v_i}{v_{sum}}$. \label{enum:redist-dop-f}
+	\item \label{enum:redist-neighbors-f} If $v_{sum} = 0$ and $\iset_r \neq \emptyset$, then for each cell, $\pos_j \in \iset_r$, redistributed the excess mass according to $M(\pos_j) = M(\pos_j) +  M_{ex} / n_r$ where $n_r$ is the number of cells in $\iset_r$.
+	\item Otherwise, increment the global excess mass, $M_{gex} = M_{gex} + M_{ex}$.
+	\item Lastly, transition $\pos$ from $\iset$ to $\fset$ and $M(\pos) = \rho(\pos)$.
+\end{enumerate}
+\item \label{enum:rloop-gas} Loop over cells that will be transitioned from interface to gas cells, i.e. loop over $\pos \in \iset_g$, and:
+\begin{enumerate}
+	\item Let $M_{ex} = M$.
+	\item Calculate the unit norm outward from the interface, i.e. toward empty space, $\mathbf{n}$.
+	Initialize $v_{sum} = 0$.
+	\item For all cells in the neighborhood,
+	\begin{enumerate}
+		\item if an interface cell, store in a data structure, $\iset_{r}$, and for the discrete velocity vector direction, $\pvel_i$, of the respective neighbor, let $v_i = \begin{cases}
+					-\mathbf{n} \cdot \pvel_i, & \mathbf{n} \cdot \pvel_i < 0 \\
+					0, & \text{otherwise}
+				\end{cases}$.
+		\item $v_{sum} = v_{sum} + v_i$.
+	\end{enumerate}
+	\item \label{enum:redist-dop-g} If $v_{sum} \ge 0$ and $\iset_r \neq \emptyset$, then for each cell, $\pos_j \in \iset_r$, redistribute the excess mass according to $M(\pos_j) = M(\pos_j) +  M_{ex} \frac{v_i}{v_{sum}}$.
+	\item \label{enum:redist-neighbors-g} If $v_{sum} = 0$ and $\iset \neq \emptyset$, then for each cell, $\pos_j \in \iset_r$, redistributed the excess mass according to $M(\pos_j) = M(\pos_j) + M_{ex} / n_r$ where $n_r$ is the number of cells in $\iset_r$.
+	\item Otherwise, increment the global excess mass, $M_{gex} = M_{gex} + M_{ex}$.
+	\item Lastly, transition $\pos$ from $\iset$ to $\gset$ and $M(\pos) = 0$.
+\end{enumerate}
+\item \label{enum:final-redist} The global excess mass, i.e. the mass that could not be redistributed by other means, is redistributed according to:
+\begin{enumerate}
+	\item \label{enum:final-rloop} for all $\pos \in \iset$, $M(\pos) = M(\pos) + M_{gex} / n_i$ where $n_i$ is equal to the number of cells in the interface set, $\iset$.
+\end{enumerate}
+\end{enumerate}
+The algorithm, as presented, ensures both mass conservation and continuity of the free-surface.
+Moreover, when mass redistribution inevitably occurs, it is first attempted to be redistributed and weighted in the direction in which the interface is propagating, e.g.
+as in Step \ref{enum:rloop-fluid}.\ref{enum:redist-dop-f} or \ref{enum:rloop-gas}.\ref{enum:redist-dop-g}.
+If there are no interface cells in the direction of interface propagation, it is instead redistributed to neighboring interface cells, e.g.
+Step \ref{enum:rloop-fluid}.\ref{enum:redist-neighbors-f} or \ref{enum:rloop-gas}.\ref{enum:redist-neighbors-g}.
+On the off-chance that there are no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells, e.g.
+Step \ref{enum:final-redist}.
+The hierarchy of possible redistribution steps, though seemingly complex, ensures that mass redistribution is done in the most physically meaningful way as is possible.
+% We may need to hard code references to steps in here...
+Lastly, note that the interface normal, $\mathbf{n}$, is calculated using a finite difference approximation,
+\begin{equation} \label{eq:n}
+\mathbf{n}_{ij} = \frac{1}{2} \begin{Bmatrix}
+\epsilon_{i-1, j} - \epsilon_{i+1, j} \\
+\epsilon_{i, j-1} - \epsilon_{i, j+1}
+\end{Bmatrix}
+\end{equation}
+\noindent where $i$ is the index of the node in the x-direction and $j$ is the index of the node in the y-direction.
+
+\subsection{Boundary Conditions at the Free-Surface} \label{sec:bc-at-fs}
+
+Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and a still more subtle consideration, how particle distributions that are missing from the streaming step are reconstructed.
+The degrees of freedom, i.e. $f_i$, are not tracked for gas cells.
+After the streaming step, interface cells will be missing particle distributions, $f_i$, that would have streamed from gas cells--an issue that is depicted in \Fref{fig:ps}.
+
+\begin{figure}
+\centering
+\includegraphics{figs/ps}
+\caption{Example of a scenario in which particle distributions are missing post-streaming.}
+\label{fig:ps}
+\end{figure}
+
+The missing particle distributions need reconstructed.
+Also momentum needs to be conserved at the interface.
+When the particle distributions are reconstructed, they will be reconstructed in such a way that momentum is conserved between the primary fluid and the secondary fluid.
+In order to perform the reconstruction, there are a few assumptions that need to be made:
+\begin{enumerate}
+\item The velocity of the secondary fluid is equal to the velocity of the primary fluid at the free-surface, i.e. there is no-slip between the two-fluids.
+\item The secondary fluid is at equilibrium and at constant pressure, $p_G$.
+\item There is a force balance for opposite lattice directions.
+\end{enumerate}
+Because particle distributions are a measure of momentum transport, the pressure of the secondary fluid can be converted to the particle distribution scale and the missing particle distribution can be solved for algebraically.
+Consider \Fref{fig:fbi}.
+Balancing the opposite lattice directions results in:
+\begin{eqnarray}
+f_i + f_\opidx &=& f_\opidx(\rho_G, \mathbf{u}) + f_i(\rho_G, \mathbf{u}), \\
+\label{eq:reconst} f_\opidx &=& f_\opidx(\rho_G, \mathbf{u}) + f_i(\rho_G, \mathbf{u}) - f_i.
+\end{eqnarray}
+Missing distribution functions can be reconstructed according to \eqref{eq:reconst}.
+In order for conservation of momentum to be satisfied along the free-surface, distribution functions coming from the direction of the interface normal, $\mathbf{n}$, must also be reconstructed.
+Thus distribution functions are reconstructed if coming from a gas cell, or if $\mathbf{n} \cdot \pvel_\opidx > 0$.
+
+\begin{figure}
+\centering
+\includegraphics{figs/force-bal-at-interface}
+\caption{Particle distributions at free-surface for a pair of opposing lattice directions after gas pressure, $\rho_G$, is in the particle distribution form.}
+\label{fig:fbi}
+\end{figure}
+
+\subsection{Resulting Algorithm}
+
+The details of the free-surface algorithm have not been presented in chronological order, but instead in a logical order.
+For reasons of clarity, the steps of the algorithm are presented in order:
+\begin{enumerate}
+\item Mass transfer (\Fref{sec:mass-transfer})
+\item Stream
+\item Reconstruct particle distributions (\Fref{sec:bc-at-fs})
+\item Collide
+\item Enforce boundary conditions
+\item Calculate macroscopic variables
+\item Update cell states (\Fref{sec:cell-updates})
+\end{enumerate}
+
+\section{Wellbore Cementing} \label{sec:results}
+
+The free-surface LBM model was implemented and used to simulate primary cementing in a dry annulus.
+The focus of the current study is the performance of different cement slurry mixes and characteristic flows for different imperfect wellbore geometries.
+Because the focus in regards to the wellbore is on geometrical imperfections, the entire wellbore annulus is not simulated, but instead a small section around a geometrical imperfection is considered.
+An example is shown in \Fref{fig:wellbore-cementing-schematic}.
+
+\begin{figure}
+\centering
+\includegraphics{figs/wellbore-cementing-schematic}
+\caption{Example simulation used in wellbore cementing study.
+The black rectangles on the left boundary represent notches protruding from the rock formation surface. The contours are of the primary fluid (cement slurry) mass.}
+\label{fig:wellbore-cementing-schematic}
+\end{figure}
+
+\section{Conclusion} \label{sec:conclusion}
+
+\chapter{Conclusion}
 

### 2016-07-12 thesis.tex
first draft of the thesis is finished

@@
 
-\subsection{Lattice Boltzmann Method: Overview}
+\subsection{Lattice Boltzmann Method: Overview} \label{sec:LBM}
 
@@
 \begin{figure}
+  \centering
   \includegraphics[]{figs/d2q9}
@@
 \begin{figure}
+	\centering
     \includegraphics[width=\linewidth]{figs/sharp_gradient}
@@
 \begin{figure}
+	\centering
     \includegraphics[width=\linewidth]{figs/overrelaxation_tau-0513}
@@
 
-\subsubsection{Bingham Plastic}
+\subsubsection{Bingham Plastic} \label{sec:bp}
 
@@
 \begin{figure}
+	\centering
     \includegraphics[width=\linewidth]{figs/poise-schematic}
@@
 \begin{table}
+\centering
 \caption{Bingham plastic Poiseuille flow}
@@
 \begin{figure}
+\centering
 \includegraphics[width=\textwidth]{figs/poise-bingham/bgk-5/analytical-vs-approx.png}
@@
 \begin{figure}
+	\centering
     \includegraphics[width=\textwidth]{figs/poise-bingham/bgk-8/analytical-vs-approx.png}
@@
 \begin{figure}
+\centering
 \begin{tabulary}{\linewidth}{c c}
@@
 \begin{figure}
+	\centering
     \begin{tabulary}{\linewidth}{c c}
@@
 \begin{figure}
+	\centering
     \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon}
-    \caption{Evoluation of relative $L_2$, $\epsilon^{neq}$ with time.
+    \caption{Evolution of relative $L_2$, $\epsilon^{neq}$ with time.
 The norm of the $\epsilon^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $\epsilon$ moment.}
@@
 \begin{figure}
+	\centering
     \includegraphics[width=\linewidth]{figs/poise-bingham/qx}
-    \caption{Evoluation of relative $L_2$, $q_x^{neq}$ with time.
+    \caption{Evolution of relative $L_2$, $q_x^{neq}$ with time.
 The norm of the $q_x^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $q_x$ moment.}
@@
 \begin{table}
+\centering
 \caption{Power-law Poiseuille flow}
@@
 \begin{figure}
+\centering
 \includegraphics{figs/lid-driven}
@@
 All coordinate values presented in this section (used to describe the location of the center of vortices) are given normalized with respect to the length of the cavity side, i.e. as $(x / L, y / L)$.
+Note that all results presented as ``-'' indicate that the tests were numerically unstable to the extent that the degrees of freedom all over the domain were ``NaN''.
 
@@
 \begin{table}
+\centering
 \caption{Bingham plastic, lid-driven cavity flow; $Bn = 1$.}
+\small
 \vspace{0.5cm}
@@
 \begin{table}
+\centering
     \caption{Bingham plastic, lid-driven cavity flow; $Bn = 10$.}
+    \small
     \vspace{0.5cm}
@@
 \begin{table}
+\centering
     \caption{Bingham plastic, lid-driven cavity flow; $Bn = 100$.}
+    \small
     \vspace{0.5cm}
@@
 \begin{table}
+\centering
 \caption{Power-law, lid-driven cavity flow; $n = 0.5$.}
+\small
 \vspace{0.5cm}
@@
 \begin{table}
+\centering
 \caption{Power-law, lid-driven cavity flow; $n = 1.5$.}
+\small
 \vspace{0.5cm}
@@
 
-\section*{Acknowledgements}
-
 \chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method}
@@
 
-\subsection{Yield Stress Flow}
+For the sake of computational simplicity and computational efficiency, the numerical study was carried out using a two-dimensional lattice Boltzmann method on the D2Q9 lattice.
+The collision operator used was the linear BGK collision operator.
+For more details on the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.
+
+\subsection{Cement Slurry Constitutive Relationship}
+
+Cement slurry is best modeled using a non-Newtonian constitutive relationship.
+A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
+Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
+The Bingham plastic constitutive model is difficult to work with numerically.
+Often, a smooth approximation is used instead--the Papastasiou approximation.
+Throughout this chapter the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
+The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flows.
+For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.
 
@@
 
-It is important to note that the particle distributions, $f_0, f_1, ..., f_8$, cannot be used to directly determine the fluid fraction of a cell.
+It is important to note that the particle distributions, $f_0, ..., f_8$, cannot be used to directly determine the fluid fraction of a cell.
 Neither can the zeroth velocity moment of the particle distributions, $\rho = \sum_i f_i$.
@@
 
+\newcommand{\nodesi}{50}
+\newcommand{\nodesj}{500}
+\newcommand{\anw}{$1\frac{1}{4}$"}
+\newcommand{\anh}{1'}
+\newcommand{\nsteps}{16000}
+
 \section{Wellbore Cementing} \label{sec:results}
 
-The free-surface LBM model was implemented and used to simulate primary cementing in a dry annulus.
-The focus of the current study is the performance of different cement slurry mixes and characteristic flows for different imperfect wellbore geometries.
+The free-surface LBM model was implemented and used to simulate primary cementing in a dry annulus, i.e. an annulus without drilling mud.
+The focus of the current study was the performance of different cement slurry mixes and characteristic flows when used to cement different imperfect wellbore geometries.
 Because the focus in regards to the wellbore is on geometrical imperfections, the entire wellbore annulus is not simulated, but instead a small section around a geometrical imperfection is considered.
@@
 
+The simulations were carried out on a $\nodesi \times \nodesj$ lattice, which corresponds to an annulus width of \anw \hspace{0.05cm} and a cross-sectional height of \anh.
+For the mass inlet, a \citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
+$M(\pos_i, t) = M_0 = 1.0$ for all $\pos_j$ along the inlet.
+This corresponded to a constant mass flow, $\dot{M}_{in} = M_0 u_0$, at inlet.
+For the outlet, mass was allowed to transfer out of the domain during the mass transfer step such that $f_\opidx(\pos + \pvel_i \Delta t, t) = 0$ in \eqref{eq:dM} for all $\pos + \pvel_i$ that lie outside of the domain on the other side of the mass outlet, i.e.
+there is no backflow.
+All simulations were executed for \nsteps time steps.
+
+Both the rock formation geometries and cement slurry flow properties were parameterized.
+The different imperfect geometries that were considered at the rock formation surface were: flow over a cavity that is 30\% of the annulus width, ``cav $30 \times 1$''; flow over a cavity that is 60\% of the annulus width, ``cav $60 \times 1$''; flow over a square obstacle with sides that are 30\% of the annulus width, ``obs $30 \times 1$''; flow over two square obstacles with sides that are 30\% of the annulus width, ``obs $30 \times 2$''; flow over a square obstacle with sides that are 60\% of the annulus width, ``obs $60 \times 1$''; flow over two square obstacles with sides that are 60\% of the annulus width, ``obs $60 \times 2$''; and flow over a rectangular obstacle with width that is 60\% of the annulus width and length that is four times the width, ``obs $60 \times L$''.
+The wellbore geometries are depicted in \Fref{fig:wellbore-geometries}.
+
+\begin{figure}
+\includegraphics[width=\linewidth]{figs/wellbore-geometries}
+\caption{Wellbore geometries considered in the study.}
+\label{fig:wellbore-geometries}
+\end{figure}
+
+In addition to wellbore geometry, different cement slurry material properties and cement slurry flow rates were parameterized, which are characterized by the dimensionless Reynold's number, $Re = \frac{\rho U L}{\mu_p}$, and Bingham number, $Bn = \frac{\tau_y L}{\mu_p U}$, of the flow where $L \equiv $ the width of the annulus.
+The goal of the study was to investigate which Reynold's numbers and Bingham numbers were preferable for each geometry, and which Reynold's numbers and Bingham numbers performed well for all geometries.
+The dimensionless numbers were parameterized as follows, $(Re, Bn) = (1.00, 4.00)$,  $(2.50, 0.00)$, $(2.50, 1.00)$, $(2.50, 10.0)$, $(2.50, 25.0)$, $(5.00, 5.00)$, and $(12.5, 2.0)$.
+\Fref{fig:percent-filled} shows the percentage of the \anh \hspace{0.05cm} section wellbore filled after cementing.
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{figs/percent_filled}
+\caption{Wellbore geometries considered in the study.}
+\label{fig:percent-filled}
+\end{figure}
+
+\Fref{fig:percent-filled} shows that there are some general statements to be made about what properties of the cement slurry flow are preferable.
+For example, low yield stress, i.e. low Bingham number flows, had good fill for every wellbore geometry considered.
+In addition, the good performance of the $Re = 5.00, Bn = 5.00$ and $Re = 12.5, Bn = 2.0$ slurry flows suggests that a if the ratio of $Bn / Re$ is low, that the slurry flow will perform well for cementing a dry hole.
+As $Bn / Re$ increases, it appears that, in certain cases and wellbore geometries, the slurry flow tends to perform worse.
+What is interesting to note is that for slurry flows that performed poorly in some cases, there was a lack of consistency across wellbore geometries in terms of which slurry flow performed worse and which slurry flows performed better, which suggests that generalities cannot always be made and that computational methods are necessary for understanding and designing slurry mixes for real-life wellbores.
+
 \section{Conclusion} \label{sec:conclusion}
+
+An LBM model for simulating the non-Newtonnian and free-surface flows that occurs while cementing dry annuli in oil and gas wellbores was implemented and presented.
+It was then used to study how different cement slurry flows, parameterized by different Reynold's numbers and Bingham numbers, would perform in different wellbore geometries.
+It was found that in general, it is preferable that the yield stress and $Bn / Re$ ratio of the slurry flow be low.
+However, beyond these considerations, it is not clear what slurry flow may be preferable for a given rock formation geometry.
+The results suggest that computational methods are necessary for (a) better understanding the mechanics of primary cementing and (b) for designing a cement mix for a given wellbore.
+It is a goal of the current work to identify a computational method for efficiently simulating cement slurry flow through a realistic wellbore so that optimization methods may be used to determine the optimal cement slurry flow for primary cementing.
+The lattice Boltzmann method extended for simulating non-Newtonian and free-surface flows as presented in the current work has several properties that make it a good candidate for such a design framework:
+\begin{itemize}
+\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
+\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components.
+\item LBM is easily written in parallel and as hardware architectures shifts sequential to parallel computing, it becomes increasingly important that code for simulating complex and computationally expense phenomena be written in parallel.
+\item The LBM no-slip boundary condition works well with complex geometries, such as those that occur in realistic rock formation surfaces in drilled wellbores.
+\end{itemize}
 

### 2016-07-12 thesis.tex
first rough draft

@@
 \begin{abstract}
+
+When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
+Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
+An important yet vulnerable barrier is the cement annulus.
+Creating the cement annulus, a process known as primary cementing, is difficult to perform optimally.
+Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
+Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
+Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
+The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
+
+Despite the advantages of LBM, there are considerations that must be made, as with all computational methods, in regards to the accuracy and numerical stability of the solution.
+Issues with accuracy and numerical stability are especially prevalent in non-Newtonian flows because of the nonlinear constitutive relationship.
+\Fref{cha:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LBM models in simulating non-Newtonian flows.
+
+Once the ground-work has been laid for simulating non-Newtonian flows using LBM, the algorithm was extended in \Fref{cha:lbm-cement} to incorporate simulation of free-surface flows.
+The extended LBM model was used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
+More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
+The study is followed by conclusions and a discussion of future work.
+
 \end{abstract}
@@
 
-\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows}
+\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows} \label{cha:lbm-stability}
 
@@
 In this paper, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
-The investigation was be carried out, not as a rigorous mathematical treatment, but instead as a numerical study looking for qualitative, yet practical, results.
+The investigation was carried out, not as a rigorous mathematical treatment, but instead as a numerical study looking for qualitative, yet practical, results.
 The numerical investigation included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluid flows.
@@
 
-\chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method}
+\chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method} \label{cha:lbm-cement}
 
@@
 
-Wellbore cementing is
+When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
+Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
+An important yet vulnerable barrier is the cement annulus.
+Creating the cement annulus, a process known as primary cementing, is difficult to perform optimally.
+Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
+Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
+Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
+The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
+In \Fref{sec:numerical-methods}, an algorithm for simulating non-Newtonian free-surface flow using LBM is presented.
+The algorithm was implemented and used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
+More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
+The study is followed by conclusions and a discussion of future work.
 
@@
 
-\section{Wellbore Cementing} \label{sec:results}
+\section{Primary Cementing} \label{sec:results}
 
@@
 
-\chapter{Conclusion}
-
 \bibliographystyle{abbrvnat}

### 2016-07-13 thesis.tex
minor edits

@@
-% MG: replace \ \cite with ~\cite to ensure citations stay on same line
+% MG: replace \~\cite with ~\cite to ensure citations stay on same line
 % MG: need an overall consistency in 'flucuation' vs. 'oscillation' vs. 'noise'; use same word if you mean same thing.
@@
 \date{July 20, 2016}
-\keywords{lattice Boltzmann method, non-Newtonian flow, free surface flow, wellbore cementing}
+\keywords{lattice Boltzmann method, non-Newtonian flow, free surface flow, primary cementing}
 \subject{Draft}
@@
 Issues with accuracy and numerical stability are especially prevalent in non-Newtonian flows because of the nonlinear constitutive relationship.
-\Fref{cha:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LBM models in simulating non-Newtonian flows.
-
-Once the ground-work has been laid for simulating non-Newtonian flows using LBM, the algorithm was extended in \Fref{cha:lbm-cement} to incorporate simulation of free-surface flows.
+\Fref{chap:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LBM models in simulating non-Newtonian flows.
+
+Once the ground-work has been laid for simulating non-Newtonian flows using LBM, the algorithm was extended in \Fref{chap:lbm-cement} to incorporate simulation of free-surface flows.
 The extended LBM model was used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
@@
 
-\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows} \label{cha:lbm-stability}
+\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows} \label{chap:lbm-stability}
 
@@
 Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable--even at a constant temperature--and is often a function of strain-rate.
-There are a number of fluids in science and engineering applications that can be classified as non-Newtonian; pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner \cite{bohme1987non}.
+There are a number of fluids in science and engineering applications that can be classified as non-Newtonian; pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.
 Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized:
@@
 	\item \emph{Yield stress fluids}. Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
-	Some examples of yield stress fluids are drilling muds, fresh cement slurries and concretes, and melted metal alloys \cite{ginzburg2002free}.
+	Some examples of yield stress fluids are drilling muds, fresh cement slurries and concretes, and melted metal alloys~\cite{ginzburg2002free}.
 	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
@@
 %MG: is it tacky to cite myself here?
-	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies~\cite{definitely cite someone}.
+	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies.
 	Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g.
-pastes, paints, muds, molten plastics and metals, and blood \cite{wang2011lattice}.
+pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
 %MG: citation is for blood as non-Newtonian
@@
 
-Analytical solutions rarely exists for even the simplest non-Newtonian flows because of the complexity that a nonlinear constitutive relationship entails.
+Analytical solutions rarely exist for even the simplest non-Newtonian flows because of the complexity that a nonlinear constitutive relationship entails.
 It is generally more practical to approximate non-Newtonian flows using numerical methods.
@@
 LBM can be considered as a type of finite-difference scheme for the continuous Boltzmann equation, and as such, has numerical properties in common with finite-difference schemes.
-One such consideration associated with this view, is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
+One such consideration associated with this view is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
 %MG: are all relevant sources cited?
@@
 \citeauthor{sterling1993stability} concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
-~\citet{worthing1997stability} extended the stability analysis performed by \citeauthor{sterling1993stability} and discovered both physical and nonphysical instabilities; and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time.
+~\citet{worthing1997stability} extended the stability analysis performed by~\citeauthor{sterling1993stability} and discovered both physical and nonphysical instabilities; and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time.
 
@@
 
-To improve upon the stability of LBM models, \citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
+To improve upon the stability of LBM models,~\citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
 \citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
-Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
+Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
 Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
@@
 The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
-The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\colop$ is the collision operator.
+The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, and $\colop$ is the collision operator.
 The lattice part of LBM refers to the way in which the BE is discretized.
@@
 \begin{itemize}
-\item Streaming: particle distribution propagate to their neighbors along the lattice velocity vectors.
+\item Streaming: particle distributions propagate to their neighbors along the lattice velocity vectors.
 The particles can only move along the vectors in their specified direction and can only move at a specific speed.
@@
   \centering
-  \includegraphics[]{figs/d2q9}
+  \includegraphics[width=10cm]{figs/d2q9}
   \caption{D2Q9 lattice}
@@
 For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
-For velocity or pressure boundary conditions, the method proposed by \citet{zou1997pressure} can be used.
+For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
 The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
@@
 
-The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision) \cite{Cer90}.
+The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision)~\cite{Cer90}.
 In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
@@
 
-\noindent where $\omega$ is the collision frequency \cite{Bha54}.
+\noindent where $\omega$ is the collision frequency~\cite{Bha54}.
 The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
@@
 For example, in low viscosity flows the BGK operator results in an over relaxation of particle distributions toward quasiequilibrium.
-It is well known that when large nonequilibrium distributions exist in the LBM approximation that over relaxation can result in nonphysical oscillations that are slow to decay \cite{brownlee2007stability,dellar2003incompressible}.
+It is well known that when large nonequilibrium distributions exist in the LBM approximation that over relaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
 To illustrate this, consider the example shown in \Fref{fig:sharp-grad}.
@@
 An example of the nonphysical oscillations that would occur for $f_5$ in \Fref{fig:sharp-grad} can be seen in \Fref{fig:overrelaxation}.
-Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to ``blow up'' or ``pollute'' the system enough to make the results highly nonphysical and altogether useless \cite{gorban2014enhancement}.
+Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to ``blow up'' or ``pollute'' the system enough to make the results highly nonphysical and altogether useless~\cite{gorban2014enhancement}.
 
@@
     \item Physical processes in fluids can be approximately described by coupling or interacting among modes; and hydrodynamic modes are directly related to, i.e.
-linear combinations of, mass and momentum moments \cite{lallemand2000theory}.
-    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor \cite{dellar2003incompressible}.
+linear combinations of, mass and momentum moments~\cite{lallemand2000theory}.
+    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor~\cite{dellar2003incompressible}.
     Of the nine relaxation rates available, the three that correspond to the extra variables--often referred to as ``ghost variables'', and their associated modes as ``ghost modes''--can be tuned in order to dampen out their associated ghost modes, ensuring these modes do not dominate or cause numerical instabilities at the lattice scale.
@@
 The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
-It is common practice, and \citet{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.
+It is common practice, and~\citet{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.
 
@@
 Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
-Thus, a smooth approximation to \eqref{eq:bingham} formulated by \citet{papanastasiou1987flows} is often used as:
+Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\citet{papanastasiou1987flows} is often used as:
 \begin{equation} \label{eq:papa}
@@
 
-\chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method} \label{cha:lbm-cement}
+\chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method} \label{chap:lbm-cement}
 
@@
 
-% what is wellbore cementing?
-
-Cementing of oil and gas wellbores is important because if cement does not completely fill the annulus then pathways can form for fluids and gasses to flow through.
-The pathways need not be large--a pathway of SOMETHING is sufficient for enough loss of pressure at the wellhead for the well to lose SOME percent production.
-There are also environmental concerns as oils and gasses can migrate into ground water or the atmosphere causing environmental damage.
+% what is primary cementing?
 
@@
 
-In order to better understand, and thus improve wellbore cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
-For example, \citet{Bei77} developed a mathematical model to ``describe the miscible displacement of drilling muds by cement slurries under laminar flow conditions''.
+In order to better understand, and thus improve primary cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
+For example,~\citet{Bei77} developed a mathematical model to ``describe the miscible displacement of drilling muds by cement slurries under laminar flow conditions''.
 Among their assumptions were that the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the y-coordinate.
@@
 \citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
-The model was more general than the model developed by \citeauthor{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
+The model was more general than the model developed by~\citeauthor{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
 Annulus eccentricity occurs often in wellbores and can sometimes be extreme as it is difficult to keep long strings of casing centered in an imperfectly drilled hole.
@@
 \citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
-The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either \citeauthor{Bei77} or \citeauthor{Bit02}.
+The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citeauthor{Bei77} or~\citeauthor{Bit02}.
 The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
@@
 The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
-\citet{zhao2016lattice} developed a model simulating wellbore cementing using the Lattice Boltzmann Method.
+\citet{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
 \citeauthor{zhao2016lattice} used the model to investigate the displacement interface shape in a horizontal, eccentric annulus; and in order to simplify the model from a 3D to 2D flow, used the classical Hele-Shaw approach.
@@
 
-The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features (i.e., the purpose is an extension to the work present in~\citet{grasinger2015simulation}).
+The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features (i.e., the purpose of the current work is an extension of the work present in~\citet{grasinger2015simulation}).
 The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation, among other benefits.
-LBM also has other advantages as an approach for simulating wellbore cementing:
+LBM also has other advantages as an approach for simulating primary cementing:
 \begin{itemize}
 \item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
-Many fluids present in wellbore cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
-\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components for reasons similar to why the no-slip boundary condition works well with complex geometries.
+Many fluids present in primary cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
+\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components.
 \item LBM is easily written in parallel.
-As hardware architectures shifts from one or a few cores to, especially in the case of scientific computing, several CPUs, it becomes increasingly important that code for simulating complex and computationally expense phenomena be written in parallel.
+As hardware architectures shift from one or a few cores to, especially in the case of scientific computing, several CPUs, it becomes increasingly important that code for simulating complex and computationally expense phenomena be written in parallel.
 \end{itemize}
@@
 cement slurry yield stress, cement slurry inlet velocity, etc., perform in regards to fill of the annulus for different wellbores with imperfect rock formation geometries.
-\Fref{sec:conclusion} are conclusions resulting from the numerical study and a discussion of future work.
+\Fref{sec:conclusion} outlines conclusions resulting from the numerical study and a discussion of future work.
 
@@
 
-For the sake of computational simplicity and computational efficiency, the numerical study was carried out using a two-dimensional lattice Boltzmann method on the D2Q9 lattice.
-The collision operator used was the linear BGK collision operator.
+For the sake of computational simplicity and computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
+In addition, the collision operator used was the BGK collision operator.
 For more details on the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.
@@
 Often, a smooth approximation is used instead--the Papastasiou approximation.
-Throughout this chapter the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
+Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
 The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flows.
@@
 The interface is what is referred to as the free-surface.
-The advantage of modeling a multiphase multicomponent flow as a free-surface is that because the dynamics of the secondary fluid are not explicitly tracked and simulated the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
-The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in \citet{korner2005lattice,thurey2005interactive}.
+The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
+The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice,thurey2005interactive}.
 
@@
 Neither can the zeroth velocity moment of the particle distributions, $\rho = \sum_i f_i$.
-For the algorithm used in this work, neither can be considered a measure of the amount of primary fluid in a cell, i.e. $\epsilon(\pos, t) \neq \epsilon(\mathbf{f(\pos, t)}, \rho(\pos, t))$.
+For the algorithm used in this work, $\epsilon(\pos, t) \neq \epsilon(\mathbf{f(\pos, t)}, \rho(\pos, t))$.
 Instead, each $f_i$ is related to the amount of primary fluid advection in each of the $\pvel_i$ discrete velocity directions.
-To track the fluid fraction and how it changes, another variable is introduced, $M(\pos, t)$, which denotes the fluid mass in each cell.
+To track the fluid fraction and how it changes, another variable is introduced, $M(\pos, t)$, which denotes the (primary) fluid mass in each cell.
 The fluid mass is related to the fluid fraction by $\epsilon = M / \rho$.
-The change in mass, $\Delta M(\pos, t)$ is related to $\mathbf{f}(\pos, t)$ and particle distribution functions in the neighborhood of $\pos$.
+The change in mass, $\Delta M(\pos, t)$ is related to $\mathbf{f}(\pos, t)$ and particle distribution in the neighborhood of $\pos$.
 Given initial conditions on $M(\pos, 0)$ and a fixed time step size $\Delta t $, the fluid mass is determined by $M(\pos, t_n) = M(\pos, 0) + \Delta t \sum_{k=1}^n \Delta M(\pos, t_k)$.
+What remains to track the fluid mass is to develop an expression for $\Delta M(\pos, t)$.
 
@@
 The set of fluid, interface, and gas cells are denoted by $\fset$, $\iset$, and $\gset$, respectively.
-A fluid cell is a cell that is entirely covered by the primary fluid and a gas cell is a cell that is completely empty, or does not contain any primary fluid.
+A fluid cell is a cell that is entirely covered by the primary fluid; and a gas cell is a cell that is completely empty, or does not contain any primary fluid.
 An interface is partially covered, and there is always a layer of interface cells between fluid and gas cells so that a fluid cell and a gas cell are never neighbors.
@@
 When the border of the two cells are not covered with fluid, i.e.
-when both cells are interface cells, the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the border that is covered in fluid.
+when both cells are interface cells, the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the cell border that is covered in fluid.
 The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M(\pos, t)$.
@@
 
-Tracking the transport of mass throughout the domain is an important step in determining what parts of the domain are covered in the primary fluid and what are not.
-However, for the free-surface, i.e. layer of interface cells between the fluid cells and gas cells, to change shape and propagate, cells need to be able to transition state.
-Cell states change based on the rule set presented in \Fref{tab:cell-transition-rules} where $\tconst$ is a small, computational constant that keeps cells from transitioning back and forth between fluid and interface or interface and gas states in subsequent time steps.
+Tracking the transport of mass throughout the domain is an important step in determining what parts of the domain contain the primary fluid and what do not.
+However, for the free-surface to change shape and propagate, cells need to be able to transition state.
+Cell states change based on the rule set presented in \Fref{tab:cell-transition-rules} where $\tconst$ is a small, computational constant that keeps cells from transitioning back and forth between fluid and interface, or interface and gas states in subsequent time steps.
 To summarize \Fref{tab:cell-transition-rules}, gas cells cannot transition to fluid cells and vice versa, interface cells transition to fluid cells when their mass has exceeded $\rho$ by a computational constant, and gas cells when their mass is less than zero minus a computational constant.
@@
 Step \ref{enum:final-redist}.
-The hierarchy of possible redistribution steps, though seemingly complex, ensures that mass redistribution is done in the most physically meaningful way as is possible.
-% We may need to hard code references to steps in here...
+The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...
 Lastly, note that the interface normal, $\mathbf{n}$, is calculated using a finite difference approximation,
@@
 Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and a still more subtle consideration, how particle distributions that are missing from the streaming step are reconstructed.
-The degrees of freedom, i.e. $f_i$, are not tracked for gas cells.
-After the streaming step, interface cells will be missing particle distributions, $f_i$, that would have streamed from gas cells--an issue that is depicted in \Fref{fig:ps}.
+The degrees of freedom, $f_i$, are not tracked for gas cells.
+After the streaming step, interface cells will be missing particle distributions that would have streamed from gas cells--an issue that is depicted in \Fref{fig:ps}.
 
@@
 \begin{eqnarray}
-f_i + f_\opidx &=& f_\opidx(\rho_G, \mathbf{u}) + f_i(\rho_G, \mathbf{u}), \\
-\label{eq:reconst} f_\opidx &=& f_\opidx(\rho_G, \mathbf{u}) + f_i(\rho_G, \mathbf{u}) - f_i.
+f_i + f_\opidx &=& f^{eq}_\opidx(\rho_G, \mathbf{u}) + f_i^{eq}(\rho_G, \mathbf{u}), \\
+\label{eq:reconst} f_\opidx &=& f_\opidx^{eq}(\rho_G, \mathbf{u}) + f_i^{eq}(\rho_G, \mathbf{u}) - f_i.
 \end{eqnarray}
@@
 In order for conservation of momentum to be satisfied along the free-surface, distribution functions coming from the direction of the interface normal, $\mathbf{n}$, must also be reconstructed.
-Thus distribution functions are reconstructed if coming from a gas cell, or if $\mathbf{n} \cdot \pvel_\opidx > 0$.
+Thus, distribution functions are reconstructed if coming from a gas cell, or if $\mathbf{n} \cdot \pvel_\opidx > 0$.
 
@@
 \includegraphics{figs/wellbore-cementing-schematic}
-\caption{Example simulation used in wellbore cementing study.
+\caption{Example simulation used in primary cementing study.
 The black rectangles on the left boundary represent notches protruding from the rock formation surface. The contours are of the primary fluid (cement slurry) mass.}
@@
 
-The simulations were carried out on a $\nodesi \times \nodesj$ lattice, which corresponds to an annulus width of \anw \hspace{0.05cm} and a cross-sectional height of \anh.
-For the mass inlet, a \citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
+The simulations were carried out on a $\nodesi \times \nodesj$ lattice, which corresponds to an annulus width of \anw \hspace{0.05cm} and a sectional height of \anh.
+For the mass inlet, a~\citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
 $M(\pos_i, t) = M_0 = 1.0$ for all $\pos_j$ along the inlet.
@@
 For the outlet, mass was allowed to transfer out of the domain during the mass transfer step such that $f_\opidx(\pos + \pvel_i \Delta t, t) = 0$ in \eqref{eq:dM} for all $\pos + \pvel_i$ that lie outside of the domain on the other side of the mass outlet, i.e.
-there is no backflow.
+the boundary condition at the outlet is such that there is no backflow.
 All simulations were executed for \nsteps time steps.

### 2016-07-14 thesis.tex
addressed Scotts comments

@@
 \newcommand{\mvel}{\mathbf{u}}
-\newcommand{\colop}{\boldsymbol{\Omega}}
+\newcommand{\colop}{\Omega}
 \newcommand{\grad}{\boldsymbol{\nabla}}
@@
 Despite the utility of the BGK collision operator, it does have a few drawbacks.
-For example, in low viscosity flows the BGK operator results in an over relaxation of particle distributions toward quasiequilibrium.
-It is well known that when large nonequilibrium distributions exist in the LBM approximation that over relaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
+For example, in low viscosity flows the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
+It is well known that when large nonequilibrium distributions exist in the LBM approximation that overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
 To illustrate this, consider the example shown in \Fref{fig:sharp-grad}.
@@
 The main drawback of the MRT collision operator is its computational expense.
-Why MRT is more computationally expense is clear when one considers that \eqref{eq:mrt-colop} requires multiple matrix multiplications and \eqref{eq:bgk} requires none.
+Why MRT is more computationally expensive is clear when one considers that \eqref{eq:mrt-colop} requires multiple matrix multiplications and \eqref{eq:bgk} requires none.
 It has been reported that MRT is approximately 15\% slower than BGK~\cite{d2002multiple}, but this was in the context of Newtonian flow.
@@
 Nonphysical oscillations due to overrelaxation would be damped out more quickly if particle distributions ``far'' from quasiequilibrium were brought closer to quasiequilibrium.
-A particle distribution vector, $\mathbf{f}$, ``far'' from equilibrium would therefor be a good candidate for artificial dissipation.
+A particle distribution vector, $\mathbf{f}$, ``far'' from equilibrium would therefore be a good candidate for artificial dissipation.
 There are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
@@
 The number of standard deviations, $n_\sigma$, is chosen \textit{a priori}.
-The criteria described in \ref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead can think in terms of what would be the maximum percentage of sites one would want artificial dissipation to be added to.
+The criteria described in \eqref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead can think in terms of what would be the maximum percentage of sites one would want artificial dissipation to be added to.
 
@@
 Upon comparison of \eqref{eq:srtensor-bgk} and \eqref{eq:srtensor-mrt}, it is clear that calculating the strain-rate tensor for the MRT collision operator is more computationally expensive than it is for the BGK collision operator.
-The increase in computational expense is exacerbated by the fact that approximating the apparent viscosity, $\mu_{app}$, by the algorithm outlined in \Fref{sec:lbm-for-nnf} requires iteration and therefor multiple calculations of the strain-rate tensor.
+The increase in computational expense is exacerbated by the fact that approximating the apparent viscosity, $\mu_{app}$, by the algorithm outlined in \Fref{sec:lbm-for-nnf} requires iteration and therefore multiple calculations of the strain-rate tensor.
 
@@
 \noindent where $k$ is the flow consistency index and $n$ is the flow behavior index.
-When $n = 1$, the \eqref{eq:pl} results in a Newtonian constitutive relationship with dynamic viscosity, $\mu = k$.
+When $n = 1$, \eqref{eq:pl} results in a Newtonian constitutive relationship with dynamic viscosity, $\mu = k$.
 A flow consistency index of $n < 1$ results in shear-thinning behavior, where as $n > 1$ results in shear-thickening.
@@
 when $m = 10^8$.
-The MRT collision operator was therefor the most accurate solution for all cases.
+The MRT collision operator was therefore the most accurate solution for all cases.
 \Fref{fig:feq-vs-f_mrt} compares particle distributions to quasiequilibrium for the MRT model; in contrast to the BGK with $m = 10^8$, there is no noise in the quasiequilibrium distributions.
@@
 For the Bingham plastic numerical tests the Reynold's number was varied, $Re = 100, 1000, 5000, 1000$, and the Bingham number was varied, $Bn = 1, 10, 100$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
-The collision schemes that were tested were: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and MRT with $m = 10^8$ and median filtering.
+The collision schemes that were tested were: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
 Tables \ref{tab:lid-bing1}--\ref{tab:lid-bing100} compare the center location of the main vortex to literature values taken from~\citet{syrakos2014performance}.
@@
 The apparent inaccuracy with regards to BGK with $m = 10^8$ and median filtering is probably due to how the numerical stability is enhanced--namely through artificial, nonphysical dissipation.
-In contrast to the nonphysical natural in which median filtering enhances stability, the stability enhancement used by MRT does not directly affect the macroscopic hydrodynamics of interest, so it again provides a stable, and typically, the most accurate solution.
+In contrast to the nonphysical nature in which median filtering enhances stability, the stability enhancement used by MRT does not directly affect the macroscopic hydrodynamics of interest, so it again provides a stable, and typically, the most accurate solution.
 % MG: verify this sentence is correct
@@
 For the power-law numerical tests the Reynold's number was varied, $Re = 100, 1000, 5000, 1000$, and the flow behavior index was varied, $n = 0.5, 1.5$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
-The collision schemes that were tested were: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (5) MRT with $m = 10^8$ and MRT with $m = 10^8$ and median filtering.
+The collision schemes that were tested were: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT with $m = 10^8$ and (5) MRT with $m = 10^8$ and median filtering.
 \Fref{tab:lid-powerlaw05} and \Fref{tab:lid-powerlaw15} compare the center location of the main vortex to literature values taken from~\citet{li2014simulation}.
@@
 \begin{itemize}
-  \item For high yield stress fluid, Poiseuille flows only the MRT collision operator did not suffer from numerical noise.
+  \item For high yield stress fluids in Poiseuille-type flows, only the MRT collision operator did not suffer from numerical noise.
   \begin{itemize}

### 2016-07-14 thesis.tex
through JB comments to page 12

@@
 
-\title{How the Lattice Boltzmann Method changed my life forever!}
+\title{Evaluating the Capabilities of Lattice Boltzmann Method for non-Newtonian and Free-surface Flows towards Applications in Wellbore Cementing}
 % The optional argument is the version of the title that will appear in Acrobat Reader's Document Info dialog box.
@@
 \degree{BS in Civil and Environmental Engineering, University of Pittsburgh, Pittsburgh, PA 2013}
-\date{July 20, 2016}
+\date{July 21, 2016}
 \keywords{lattice Boltzmann method, non-Newtonian flow, free surface flow, primary cementing}
@@
 
-The Lattice Boltzmann method (LBM) is a computational method that can be used for simulating fluid flow.
-It is well-suited for complex flows such as non-Newtonian, free surface, and multiphase multicomponent flows.
+The Lattice Boltzmann method (LBM) is a method based on computational statistical mechanics that is well-suited for complex flows such as non-Newtonian, free surface, and multiphase multicomponent flows.
 Non-Newtonian flows are the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten metals and plastics, blood flows, etc., are best modeled as non-Newtonian fluids.
-LBM is a statistical mechanical approach for simulating fluid flow.
-Each time step consists of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
-The collision operator is of interest because it has a lot of control over the physics that are simulated, e.g.
+LBM is typically applied to simulate flow through a series of time seps, each consisting of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
+The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated, e.g.
 constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
 In this paper, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
-The investigation was carried out, not as a rigorous mathematical treatment, but instead as a numerical study looking for qualitative, yet practical, results.
-The numerical investigation included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluid flows.
+The investigation was carried out as a numerical study looking for qualitative, yet practical, results; it included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluid flows.
 Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
+The test results are followed by recommendations for choice collision scheme given the modeler's priorities and problem type.
 
 \section{Introduction} %MG: this intro will potentially be long. Should it be broken up into subsections (as it is now)?
-
-\subsection{Applications of Non-Newtonian Flow}
 
@@
 These fluids are known as Newtonian fluids.
-Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable--even at a constant temperature--and is often a function of strain-rate.
-There are a number of fluids in science and engineering applications that can be classified as non-Newtonian; pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.
-Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized:
-\begin{enumerate}
-	\item \emph{Yield stress fluids}. Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
-	Some examples of yield stress fluids are drilling muds, fresh cement slurries and concretes, and melted metal alloys~\cite{ginzburg2002free}.
-	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
-	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere~\cite{grasinger2015simulation}.
-%MG: is it tacky to cite myself here?
-	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies.
+Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable even at a constant temperature and is often a function of strain-rate.
+There are a number of fluids in science and engineering applications that can be classified as non-Newtonian: pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.
+
+Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized: yield stress fluids and power-law fluids.
+Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
 Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g.
@@
 
-	\item \emph{Power-law fluids}.
-Power-law behavior is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate.
+There other idealized non-Newtonian behavir, power-law behavior, is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate.
 Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include polymer mixes and molten plastics.
 Shear-thickening fluids are also known as \emph{dilatants}, and some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
-\end{enumerate}
-
-Developing numerical methods for approximating these non-Newtonian flows is important because of how difficult, and often impossible, it is to derive analytical, closed-form solutions.
-
-\subsection{Lattice Boltzmann Method for non-Newtonian Flow} \label{sec:lbm-for-nnf} %MG: should this be broken down into subsubsections?
-Maybe like 'Motivation', 'Stability', 'Recent Developments to Address Stability'
+
+\label{sec:lbm-for-nnf}
 
 Analytical solutions rarely exist for even the simplest non-Newtonian flows because of the complexity that a nonlinear constitutive relationship entails.
-It is generally more practical to approximate non-Newtonian flows using numerical methods.
+It is generally more practical to approximate non-Newtonian flows using numerical methods. %MG: ask John about the following pargraph...
 However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
@@
 \end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
-%Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate
 Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, a general Picard-type algorithm is:
@@
 The lattice Boltzmann method has been studied and successively applied to modeling various non-Newtonian flows.
-~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flows.
-The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow, and values from literature for lid-driven cavity flows--showing the feasibility of using LBM models for yield-stress flows.
-LBM models for power-law fluid flows~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
-%MG: or were they validated?
+For example, ~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flows.
+The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flows which shows the feasibility of using LBM models for yield-stress flows.
+LBM models for power-law fluid flows~\citet{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\citet{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
+%MG: does citing this way appear too dense?
 
@@
 %Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
-Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the nonlinear relationship between shear stress and strain-rate can lead to high nonequilibrium fluctuations.
+Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the nonlinear relationship between shear stress and strain-rate can lead to highly nonlinear fluctuations.
 %MG: be more precise than 'fluctuations'?
-Various schemes and strategies for incorporating the physics of non-Newtonian flows and yet maintaining a stable numerical method have been developed and studied.
-The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}, is to make the collision frequency--which is proportional to apparent viscosity--variable and dependent on the local strain rate.
-A potential issue with the stability of the variable relaxation time approach is that as the collision frequency, $\omega$, approaches $2$ the viscosity approaches zero and overrelaxation occurs; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
+Various schemes and strategies for incorporating the physics of non-Newtonian flows with LBM and yet maintaining a stable numerical method have been developed and studied.
+The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}, is to make the collision frequency, which is proportional to apparent viscosity, variable and dependent on the local strain rate.
+A potential issue with the stability of the variable relaxation time approach is that as the collision frequency approaches $2$ the viscosity approaches zero and overrelaxation occurs; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
 %MG: verify this statement
@@
 Although bounding the collision frequency was shown to be effective in terms of stability, it is also nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the collision does not reflect the proper constitutive relationship of interest.
-Another scheme for incorporating non-Newtonian effects into LBM is to use a constant collision frequency, typically unity, and to instead incorporate the local shear-rate effect into equilibrium distribution functions.
+Another scheme for incorporating non-Newtonian effects into LBM is to use a constant collision frequency, typically unity, and to instead incorporate the local shear-rate effect through equilibrium distribution functions.
 This means particle distribution functions will always relax toward equilibrium at the same rate, but that the definition of equilibrium is modified to represent the correct stress--strain-rate relationship.
 The equilibrium distribution function is derived for the specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.
-The equilibrium distribution function for Bingham plastic flow was derived, implemented, and verified in~\cite{wang2008lattice}; and the equilibrium distribution functions for power-law fluids and Carreau fluids were derived, implemented, and verified in~\cite{yoshino2007numerical}.
-The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the relaxation time is constant (at unity), the local relaxation time will not approach values that are too small or too large, and so there is no reason to bound the collision frequency in a way that is nonphysical.
-\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flows by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts--for example, considering the plastic viscosity and yield stress for a Bingham plastic separately--then incorporating the Newtonian part in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and incorporating the non-Newtonian part as a source of momentum, i.e.
+The equilibrium distribution functions for Bingham plastic fluids, and for power-law and Carreau fluids were derived, implemented, and verified in~\cite{wang2008lattice} and~\cite{yoshino2007numerical}, respectively.
+The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach zero, and will not result in overrelaxation, and the collision frequency will not reach values much greater than one, and so there is no reason to bound the collision frequency in a way that is nonphysical.
+\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flows by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
+The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum, i.e.
 as an external forcing term, that is dependent on local shear-rate.
@@
 
-Much work has been done (not specific to non-Newtonian flows) to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
-\citet{sterling1993stability,sterling1996stability} investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice.
-The investigation was performed by linearizing all nonlinear terms about global equilibrium values and applying a von Neumann stability analysis on the resulting linearized equations.
-The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations.
-\citeauthor{sterling1993stability} concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
-~\citet{worthing1997stability} extended the stability analysis performed by~\citeauthor{sterling1993stability} and discovered both physical and nonphysical instabilities; and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time.
-
-Although much work has been done to characterize the stability of LBM models, these analyses have the following limitations to their applicability:
-\begin{itemize}
-	\item von Neumann stability analysis, and linear stability analysis in general, requires modifying the governing equations of the model so that they are linear. 
-	This linearization cannot be done without introducing some error into the analysis. %MG: I don't know if I like this
-	\item Much of the analysis that exists in literature is done about a uniform flow~\cite{worthing1997stability,niu2004investigation}. %MG: definitely cite and/or verify this more
-	\item Analyses are most often carried out assuming periodic boundary conditions.
-	Periodic boundary conditions are advantageous for stability analysis because the analysis is generally done by a Fourier analysis on the propagation of error.
-	The periodicity of Fourier modes allows the periodic boundary conditions to automatically be satisfied.
-	Such an analysis, however, does not generalize well to instances when modes of the system are heavily influenced by the boundary conditions~\cite{worthing1997stability}.
-	\item The analyses are general and do not delve into any of the nuances of non-Newtonian flows.
-\end{itemize}
-
 To improve upon the stability of LBM models,~\citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
-\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
+\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
 Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
-Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
 \citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
@@
 In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
-\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations--oscillations that often lead to numerical instabilities if left unchecked.
+\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
 Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in non-Newtonian flows.
@@
 
-\section{Numerical Methods}
-
 %MG: is the beginning of this methods section too much fluff now? Would our audiance consider it fluff?
 
-\subsection{Lattice Boltzmann Method: Overview} \label{sec:LBM}
+\section{Lattice Boltzmann Method for Simulation of Non-Newtonian Fluids} \label{sec:LBM}
 
 The Lattice Boltzmann method is a numerical approach that uses statistical mechanics to represent a variety of physical processes, such as fluid flow.
-More specifically, LBM can be thought of as a special finite difference discretization of the lattice Boltzmann equation~\cite{chen1998lattice}.
+More specifically, LBM can be thought of as a special finite difference discretization of the Boltzmann equation~\cite{chen1998lattice}.
 The length scale of LBM is unique in contrast to most common numerical methods, and is referred to as the mesoscale.
@@
 LBM instead relies on a statistical description of particle distributions, making LBM, in general, more computationally efficient and requiring less memory than other particle methods.
-In addition, the Boltzmann equation is continuous, so LBM inherits the smoothness and stability of continuum methods.
 Thus, LBM can be seen as a compromise between continuum and particle methods, combining strengths from each.
@@
 The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
-The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, and $\colop$ is the collision operator.
+The BE is given as 
+\begin{equation}
+\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop,
+\end{equation}
+\noindent where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, and $\colop = \colop(f)$ is the collision operator.
 The lattice part of LBM refers to the way in which the BE is discretized.
@@
 The lattice is two-dimensional with nine discrete velocities at each node.
-There is a stationary particle, there are four discrete velocities of magnitude 1, $(\pm1,0)$, $(0,\pm1)$, and there are four discrete velocities of magnitude $\sqrt{2}$, $(\pm1, \pm1)$.
-
+There is a stationary particle, there are four discrete velocities of magnitude 1, $\begin{Bmatrix}1 & 0\end{Bmatrix}^T$, $\begin{Bmatrix}0 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & 0\end{Bmatrix}^T$, $\begin{Bmatrix}0 & -1\end{Bmatrix}^T$, and there are four discrete velocities of magnitude $\sqrt{2}$, $\begin{Bmatrix}1 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & -1\end{Bmatrix}^T$, $\begin{Bmatrix}1 & -1\end{Bmatrix}^T$.
+The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as:
+\begin{equation}
+f_i(\pos + \pvel_i \Delta t, t + \Delta t) = f_i(\pos, t) + \colop_i(\pos, t),
+\end{equation}
+\noindent where, for D2Q9, $i = 0, 1, ..., 8$ is the index of the discrete velocity vector.
 \begin{figure}
@@
   \includegraphics[width=10cm]{figs/d2q9}
-  \caption{D2Q9 lattice}
+  \caption{Schematic of D2Q9 lattice, each node is connected to its neighbor by one of eight discrete velocity vectors}
   \label{fig:d2q9}
 \end{figure}
-
-The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as: $f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \Omega_i(f)$, where, for D2Q9, $i = 0, 1, ..., 8$.
 
@@
 
-The fluid pressure is related to macroscopic density through an equation of state:
+\noindent The fluid pressure is related to macroscopic density through an equation of state:
 
@@
 \noindent where $c_s$ is the lattice speed of sound ($c_s = \frac{1}{\sqrt{3}}$, for D2Q9).
-
-\subsection{Boundary Conditions}
-
-No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
-For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
-For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
-The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
-
-\subsection{Applied Forces}
-
-Incorporating external forces, such as gravity, pressure gradients, etc., is done by adding a source of particle distributions in the direction of the force.
-The increase in particle distributions leads to the desired macroscopic result--an increase in momentum.
-The LBE with external forces is:
-
-\begin{equation}
-f_i(\pos + \pvel_i, t + \Delta t) = f_i(\pos, t) + \Omega_i(f) + \frac{w_i \Delta t}{c_s^2} \mathbf{F} \cdot \pvel_i
-\end{equation}
-
-\noindent where $\mathbf{F}$ is the body force vector.
 
@@
 This equilibrium is determined by the macroscopic physical behavior of interest.
-In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq}$, is often given by:
+In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq} = f_i^{eq}(\pos, t)$, is often given by:
 
@@
 
-\noindent where $w_i$ is the weight in the $i^{th}$ direction, and %MG: maybe say something about where this comes from, e.g.
+\noindent where $\rho$ and $\mvel$ are dependent on $\pos$ and $t$, and $w_i$ is the weight in the $i^{th}$:%MG: maybe say something about where this comes from, e.g.
 feq is a Hermite polynomial expansion of the Maxwell-Boltzmann distribution
@@
 BGK consists of a single relaxation time and is a linear relaxation of particle distributions toward equilibrium.
-The BGK is expressed as:
+The BGK collison operator for the $i^{th}$ discrete velocity is expressed as:
 
 \begin{equation} \label{eq:bgk}
-\Omega_i = -\omega (f_i(\pos, t) - f_i^{eq}),
+\Omega_i(\pos, t) = -\omega (f_i(\pos, t) - f_i^{eq}(\pos, t)),
 \end{equation}
@@
 \noindent where $\omega$ is the collision frequency~\cite{Bha54}.
-The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
-For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$.
-From this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.
+The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis~\cite{wolf2000lattice}.
+For incompressible Newtonian flow, the collision frequency is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$; and from this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.
+The method for simulating non-Newtonian flow in the current work involves approximating a solution to the local value of apparent viscosity, $\mu_{app}(\pos, t)$, using \eqref{eq:non-newtonian} where $\tau$ and $\dot{\gamma}$ are dependent on $\pos$ and $t$, and then setting the value of the local collision frequency as follows:
+\begin{equation} \label{eq:colfeq-non-newtonian}
+\omega(\pos, t) = \frac{1}{\frac{\mu_{app}(\pos, t)}{c_s^2 \rho(\pos, t)} + \frac{1}{2}}
+\end{equation}
 
@@
 
+\subsection{Boundary Conditions}
+
+No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
+For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
+For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
+The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
+
+\subsection{Applied Forces}
+
+Incorporating external forces, such as gravity, pressure gradients, etc., is done by adding a source of particle distributions in the direction of the force.
+The increase in particle distributions leads to the desired macroscopic result--an increase in momentum.
+The LBE with external forces is:
+
+\begin{equation}
+f_i(\pos + \pvel_i \Delta t, t + \Delta t) = f_i(\pos, t) + \Omega_i(f) + \frac{w_i \Delta t}{c_s^2} \mathbf{F} \cdot \pvel_i
+\end{equation}
+
+\noindent where $\mathbf{F}$ is the body force vector.
+
 \newcommand{\srtensor}{\mathbf{D}}
@@
 
-Some introduction to the numerical study.
-It has a purpose.
-Its purpose is to try to assess the accuracy, stability, and efficiency of the LBM models as applied to non-Newtonian flows.
+A numerical study was carried out in order to investigate the suitability of different LBM collision schemes for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
 For all simulations presented in this section, the apparent viscosity is approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the maximum number of iterations set at 15 and the convergence criteria,
@@
 \noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current timestep.
-In reference to computational time, each of the simulations in this section were run on a single core of either an Intel I7-860 Quad-Core 2.80GHz processor, or a SOME OTHER PROCESSOR I NEED TO LOOK UP.
+In reference to computational time, each of the simulations in this section were run on a single core of either an Intel I7-860 Quad-Core 2.80GHz processor, or an Intel Xeon v2 2.60 GHz processor.
 The lid velocity was prescribed at $u_o = 0.1$.

### 2016-07-15 presentation.tex
starting presentation, comments through pg 12

\documentclass[pdf]{beamer}
\mode<presentation>{}

\title{Evaluating the Capabilities of Lattice Boltzmann Method for non-Newtonian and Free-surface Flows towards Applications in Wellbore Cementing}
\author{Matthew Grasinger}
\date{July 21st, 2016}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\newcommand{\pw1}{\linewidth/3}

\begin{frame}{non-Newtonian Flow}
  \begin{tabular}{cc}
    \includegraphics[width=\pw1]{figs/slurry-backfill} & \includegraphics[width=\pw1]{figs/molten-metal} \\
    \includegraphics[width=\pw1]{figs/varnish} & \includegraphics[width=\pw1]{figs/blood-circulation}
\end{tabular}
\end{frame}

\end{document}

### 2016-07-15 thesis.tex
starting presentation, comments through pg 12

@@
 It is well known that when large nonequilibrium distributions exist in the LBM approximation that overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
-To illustrate this, consider the example shown in \Fref{fig:sharp-grad}.
-If $a >> 0$, then when $f_5(x,y)$ propagates to $f_5(x+1,y+1)$ during the streaming step (where $\pos = [x,y]$, $\pvel_5 = [1,1]$, and $\Delta t = 1$), $f_5(x+1,y+1)$ will be ``far'' from $f^{eq}_5(x+1,y+1)$ for the subsequent collision step.
-If $\nu \rightarrow 0$ then $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_5(x+1,y+1)$ still being ``far'' from $f^{eq}_5(x+1,y+1)$ but on the ``other side''.
-If, as in \Fref{fig:sharp-grad}, the LBM macroscopic variables dictate that $f^{eq}_i$ remain mostly constant (or, at the very least, far from $f_i$) in the direction of propagation for $f_i$, then this leads to oscillations in the particle distribution, $f_i$.
-An example of the nonphysical oscillations that would occur for $f_5$ in \Fref{fig:sharp-grad} can be seen in \Fref{fig:overrelaxation}.
-Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to ``blow up'' or ``pollute'' the system enough to make the results highly nonphysical and altogether useless~\cite{gorban2014enhancement}.
-
-\begin{figure}
-	\centering
-    \includegraphics[width=\linewidth]{figs/sharp_gradient}
-    \caption{Example of an LBM approximation with a sharp gradient in the macroscopic flow velocity}
-    \label{fig:sharp-grad}
-\end{figure}
+To illustrate this, consider a flow in which there is a sharp spatial gradient in either $\rho$ or $\mvel$ at $\pos$.
+As $f_i^{eq}$ depends on both $\rho$ and $\mvel$ \eqref(eq:feq), it may be the case that $|f_i^{eq}(\pos, t) - f_i^{eq}(\pos + \pvel_i \Delta t, t + \Delta t)| >> 0$, i.e.
+there may be a large difference in the quasiequilibrium for the $i^{th}$ discrete velocity at $(\pos, t)$ and $(\pos + \pvel_i \Delta t, t + \Delta t)$.
+In this case, if $f_i$ is ``near'' to quasiequilibrium at $\pos$ it will be ``far'' from quasiequilibrium after the streaming step when it moves to the node at $\pos + \pvel_i \Delta$.
+Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \mvel \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \mvel \Delta t,t + \Delta t)$ but on the ``other side''.
+Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations such as those depicted in \Fref{fig:overrelaxation}.
+Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to diverge or ``pollute'' the system enough to make the results highly nonphysical and altogether useless~\cite{gorban2014enhancement}.
 
@@
     \includegraphics[width=\linewidth]{figs/overrelaxation_tau-0513}
-    \caption{Example of oscillations that can occur as a result of overrelaxation (this example is related to the example shown in \Fref{fig:sharp-grad}); the relaxation time, $\omega = 1.95$ and $\nu \approx 0.0043$}
+    \caption{Example of oscillations that can occur as a result of overrelaxation; the collision frequency, $\omega = 1.95$ and $\mu_{app} \approx 0.0043$}
     \label{fig:overrelaxation}
@@
 
-In contrast to low viscosity flows, the challenge associated with high viscosity flows (using the BGK collision operator) is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because as $\nu \rightarrow \infty$, $\omega \rightarrow 0$, resulting in extreme underrelaxation to the point of being negligible.
+The challenge associated with high viscosity flows is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
 %MG: this needs cleaned up, I think
@@
 Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flows because of the nonlinear constitutive relationship between shear stress and strain-rate.
-The nonlinear constitutive relationship can lead to sharp gradients, and depending on the form of $\dot{\gamma} \mapsto \mu_{app}$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
+The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
 
@@
 approaching 2, or too low, e.g.
-approaching 0, a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flows is to simply put bounds on the values in which the collision frequency may attain.
+approaching 0, a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flows is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
 This simple methodology for increasing stability will henceforth be referred to bounded-relaxation time BGK, or BGK-BRT.

### 2016-07-15 presentation.tex
starting presentation

@@
 \mode<presentation>{}
+
+\usetheme{metropolis}
+\usepackage{graphicx}
 
@@
 \date{July 21st, 2016}
+\institute{University of Pittsburgh}
 
@@
 
-\newcommand{\pw1}{\linewidth/3}
+\newcommand{\pw}{0.4\textwidth}
+\newcommand{\ph}{0.4\textheight}
 
-\begin{frame}{non-Newtonian Flow}
+\section{Introduction}
+
+\begin{frame}{Non-Newtonian Fluids}
+  \begin{table}
+    \centering
   \begin{tabular}{cc}
-    \includegraphics[width=\pw1]{figs/slurry-backfill} & \includegraphics[width=\pw1]{figs/molten-metal} \\
-    \includegraphics[width=\pw1]{figs/varnish} & \includegraphics[width=\pw1]{figs/blood-circulation}
+    \includegraphics[width=\pw, height=\ph, keepaspectratio]{figs/slurry-backfill} & \includegraphics[width=\pw, height=\ph, keepaspectratio]{figs/molten-metal} \\
+    \includegraphics[width=\pw, height=\ph, keepaspectratio]{figs/varnish} & \includegraphics[width=\pw, height=\ph, keepaspectratio]{figs/blood-circulation}
 \end{tabular}
+\end{table}
 \end{frame}

### 2016-07-18 thesis.tex
first round of edits through methods

@@
 
-%\preface
-% acknowledgements and such
+\preface
+I would like to thank my wife and best friend, Shelby, my parents, and my family.
+Without their love, support, and example, I would have nothing.
+I would like to thank my advisor, Dr. John Brigham, for his patience and for all that he has taught me. 
+I would also like to thank the other members of my committee, Dr. Julie Vandenbossche, and Dr. Anthony Iannacchione, whom have both been great guides and teachers to me.
+Lastly, I would like to thank my fellow students for helping me along the way to learn and for the comradery that every graduate student needs.
 
@@
 The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
-The BE is given as 
+The BE is given as:
 \begin{equation}
@@
 An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
-In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f}$, herein referred to as the ``moment space''.
+In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f} = \begin{Bmatrix}f_0 & f_1 & ...
+& f_8\end{Bmatrix}$, herein referred to as the ``moment space''.
 The collision is then performed in the moment space.
-There a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
+There are a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
 \begin{enumerate}
-    \item Physical processes in fluids can be approximately described by coupling or interacting among modes; and hydrodynamic modes are directly related to, i.e.
-linear combinations of, mass and momentum moments~\cite{lallemand2000theory}.
+    \item Physical processes in fluids can be approximately described by coupling or interacting among modes, and the modes are directly related to the moments~\cite{lallemand2000theory}.
     \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor~\cite{dellar2003incompressible}.
@@
 The relaxation parameters $s_1$ and $s_7 = s_8$ are related to the bulk and shear viscosities, respectively.
-The relationship for the shear viscosity is given by,
+The relationship for the shear viscosity is given by:
 \begin{equation} \label{eq:mrt-s7}
@@
 A practical goal then would be to use only the necessary amount of artificial dissipation in order to ensure a stable solution.
-From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation necessary?'' and ``how much artificial dissipation does one introduce when it is necessary?''.
+From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation is necessary?'' and ``how much artificial dissipation does one introduce when it is necessary?''.
 
@@
 Nonphysical oscillations due to overrelaxation would be damped out more quickly if particle distributions ``far'' from quasiequilibrium were brought closer to quasiequilibrium.
-A particle distribution vector, $\mathbf{f}$, ``far'' from equilibrium would therefore be a good candidate for artificial dissipation.
+A particle distribution vector, $\mathbf{f}$, ``far'' from equilibrium would be a good candidate for artificial dissipation.
 There are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
@@
 
-\noindent A more computationally efficient approximation of $\Delta S$ can be achieved by instead using the second order Taylor expansion of \eqref{eq:rel-noneq-entropy},
+\noindent A more computationally efficient approximation of $\Delta S$ can be achieved by instead using the second order Taylor expansion of \eqref{eq:rel-noneq-entropy}:
 
@@
 
-An option for defining criteria for when to introduce artificial dissipation that has been used successfully is to define a threshold, $\theta$, such that dissipation is added when $\Delta S > \theta$~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium}.
+A criteria for introducing artificial dissipation that has been used successfully is to define a threshold, $\theta$, such that dissipation is added when $\Delta S > \theta$~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium}.
 A potential drawback of defining a threshold \textit{a priori} is that in order to ensure the model still retains some physical integrity, only a small number of sites can have artificial dissipation added.
@@
 The threshold can be determined on a case-by-case basis through trial-and-error or by a preliminary analysis.
-In the current work, the criteria that is used for determining whether dissipation should be added is,
+In the current work, the criteria that is used for determining whether dissipation should be added is:
 
@@
 
-\noindent where $\overline{\Delta S}$ and $\sigma_{\Delta S}$ are the mean and standard of $\Delta S$, respectively--both are calculated using values over the domain for the current timestep--and $n_\sigma$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S$ must be before dissipation is added.
+\noindent where $\overline{\Delta S}$ and $\sigma_{\Delta S}$ are the mean and standard deviation of $\Delta S$, respectively--both are calculated using values over the domain for the current timestep--and $n_\sigma$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S$ must be before dissipation is added.
 The number of standard deviations, $n_\sigma$, is chosen \textit{a priori}.
@@
 One method of adding dissipation is the so-called Ehrenfests' regularization~\cite{brownlee2006stabilization}, and involves setting a lattice site that is chosen for artificial dissipation to its quasiequilibrium state.
-Although this achieves the desired result, namely damping out large nonequilibrium fluctuations, it does so in an intense manner.
+Although this achieves the desired result, namely damping out large nonequilibrium fluctuations, it does so in a way that is not smooth or gradual, but sharp.
 An alternative, the median filter, has been used successfully in conjunction with both the BGK and MRT collision schemes for simulating one-dimensional shock tubes and lid-driven cavity flows~\cite{brownlee2008nonequilibrium,gorban2014enhancement,packwood2009entropy}.
@@
 
-In order to use the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf}, one must calculate the strain-rate, $\dot{\gamma}$.
-The strain-rate is given by the second invariant of the strain-rate tensor, $\srtensor$, i.e.
+In order to solve for the apparent viscosity, $\mu_{app}$, using the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf}, one must calculate the strain-rate, $\dot{\gamma}$.
+The strain-rate is given by the second invariant of the strain-rate tensor, $\srtensor$, i.e.:
 \begin{equation} \label{eq:second-invariant}
@@
 
-\noindent When using the BGK collision scheme, the strain-rate tensor is determined by,
+\noindent When using the BGK collision scheme, the strain-rate tensor is determined by:
 \begin{equation} \label{eq:srtensor-bgk}
@@
 
-\noindent and for the MRT collision scheme, the strain-rate tensor is determined by
+\noindent and for the MRT collision scheme, the strain-rate tensor is determined by:
 \begin{equation} \label{eq:srtensor-mrt}
@@
 Alternatively, the constitutive relationship can be interpreted through the apparent viscosity.
-The apparent viscosity is the value of the viscosity assuming a Newtonian relationship between stress and strain rate, $\mu_{app} = \frac{\tau}{\dot{\gamma}}$, and is used to determine the collision frequency for the BGK collision operator or $s_7$ and $s_8$ of the relaxation matrix for the MRT collision operator.
-Rearranging \eqref{eq:papa} results in the following expression for the apparent viscosity:
+Noting that $\mu_{app} = \frac{\tau}{\dot{\gamma}}$ and rearranging \eqref{eq:papa} results in the following expression for the apparent viscosity:
 \begin{equation} \label{eq:bing-mu-app}
@@
 The power-law constitutive relationship is useful for modeling fluids that experience shear-thinning or shear-thickening.
-The power-law relationship is given by,
+The power-law relationship is given by:
 \begin{equation} \label{eq:pl}
@@
 A flow consistency index of $n < 1$ results in shear-thinning behavior, where as $n > 1$ results in shear-thickening.
-As with the Bingham plastic relationship, \eqref{eq:pl} can be modified to determine an apparent viscosity,
+As with the Bingham plastic relationship, \eqref{eq:pl} can be modified to determine an apparent viscosity:
 \begin{equation} \label{eq:pl-mu-app}
@@
 A numerical study was carried out in order to investigate the suitability of different LBM collision schemes for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
-For all simulations presented in this section, the apparent viscosity is approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the maximum number of iterations set at 15 and the convergence criteria,
+For all simulations presented in this section, the apparent viscosity is approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the maximum number of iterations set at 15 and the convergence criteria:
 \begin{equation} \label{eq:mu-app-conv}

### 2016-07-19 thesis.tex
added algorithm package and fancyref prefix
editted acknowledgements

@@
 \usepackage{pbox}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+
+\makeatletter
+\def\BState{\State\hskip-\ALG@thistlm}
+\def\mkfancyprefix#1#2{%
+  \@namedef{fancyref#1labelprefix}{#1}%
+  \begingroup\def\x{\endgroup\frefformat{plain}}%
+    \expandafter\x\csname fancyref#1labelprefix\endcsname
+      {\MakeLowercase{#2}\fancyrefdefaultspacing##1}%
+  \begingroup\def\x{\endgroup\Frefformat{plain}}%
+    \expandafter\x\csname fancyref#1labelprefix\endcsname
+      {#2\fancyrefdefaultspacing##1}%
+}
+\makeatother
+\mkfancyprefix{algo}{Algorithm}
 
@@
 Without their love, support, and example, I would have nothing.
-I would like to thank my advisor, Dr. John Brigham, for his patience and for all that he has taught me. 
+I would like to thank the best advisor ever, Dr. John Brigham, for his patience and for all that he has taught me. 
 I would also like to thank the other members of my committee, Dr. Julie Vandenbossche, and Dr. Anthony Iannacchione, whom have both been great guides and teachers to me.
-Lastly, I would like to thank my fellow students for helping me along the way to learn and for the comradery that every graduate student needs.
+Lastly, I would like to thank my fellow students, especially Rob, Jing, Mo, and Scott, for helping me to learn along the way and for the comradery that every graduate student needs.
 

### 2016-07-19 src/algorithm/streaming.tex
writing pseudo for algorithms

\begin{algorithm}
\caption{Streaming} \label{algo:streaming}
\begin{algorithmic}[1]
\Procedure{streaming}{$f$}
Initialize temporary 3D array the size of lattice, $f_temp$
\EndProcedure
\end{algorithmic}
\end{algorithm}

### 2016-07-19 thesis.tex
editting acknowledgements

--- 2016-07-18 thesis.tex
@@
 I would like to thank my wife and best friend, Shelby, my parents, and my family.
-Without their love, support, and example, I would have nothing.
+Their love, support, and example means everything to me.
 I would like to thank my advisor, Dr. John Brigham, for his patience and for all that he has taught me. 
 I would also like to thank the other members of my committee, Dr. Julie Vandenbossche, and Dr. Anthony Iannacchione, whom have both been great guides and teachers to me.
-Lastly, I would like to thank my fellow students for helping me along the way to learn and for the comradery that every graduate student needs.
+Lastly, I would like to thank my fellow students for helping me to learn along the way and for the comradery that every graduate student needs.
 

### 2016-07-20 thesis.tex
corrected mistakes in a few equations
added last two results for lid-driven power-law flow

--- 2016-07-19 thesis.tex
@@
 I would like to thank my wife and best friend, Shelby, my parents, and my family.
-Without their love, support, and example, I would have nothing.
+Their love, support, and example means everything to me.
 I would like to thank the best advisor ever, Dr. John Brigham, for his patience and for all that he has taught me. 
@@
 \begin{equation} \label{eq:feq}
-  f_i^{eq} = w_i \rho \left[1 + \frac{\pvel_i \cdot \mvel}{c_s^2} + \frac{(\pvel_i \cdot \mvel)^2}{2c_s^2} - \frac{\mvel^2}{2c_s^2} \right],
+  f_i^{eq} = w_i \rho \left[1 + \frac{\pvel_i \cdot \mvel}{c_s^2} + \frac{(\pvel_i \cdot \mvel)^2}{4c_s^2} - \frac{\mvel^2}{2c_s^2} \right],
 \end{equation}
@@
 
-\newcommand{\srtensor}{\mathbf{D}}
+\newcommand{\srtensor}{D_{\alpha\beta}}
 
@@
 \begin{equation} \label{eq:second-invariant}
-\dot{\gamma} = \sqrt{\srtensor : \srtensor}.
+\dot{\gamma} = \sqrt{2 \sum_{\alpha,\beta=1}^{2} \srtensor}.
 \end{equation}
@@
 
-For the Bingham plastic numerical tests the Reynold's number was varied, $Re = 100, 1000, 5000, 1000$, and the Bingham number was varied, $Bn = 1, 10, 100$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
+For the Bingham plastic numerical tests the Reynold's number was varied, $Re = 100, 1000, 5000, 10000$, and the Bingham number was varied, $Bn = 1, 10, 100$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
 The collision schemes that were tested were: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
@@
     &      & BGK     & Yes &              & (0.63, 0.68) & 23517 \\
-    &      & MRT     & No  &              &              &       \\
+    &      & MRT     & No  &              & - & - \\
     &      & MRT     & Yes &              & (0.51, 0.58) & 54225 \\
@@
     &       & BGK     & Yes &              & (0.50, 0.55) & 22603 \\
-    &       & MRT     & No  &              &              &       \\
+    &       & MRT     & No  &              & - & - \\
     &       & MRT     & Yes &              & (0.50, 0.54) & 63892 \\

### 2016-07-26 thesis.tex
fixed a typo
fixed mathematical error
Chapter 1 first round of revisions ready for review

@@
 \end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
-Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, a general Picard-type algorithm is:
+Determining the apparent viscosity and strain-rate of a flow will often require an iterative solution, a general Picard-type algorithm is:
 
@@
 
-A criteria for introducing artificial dissipation that has been used successfully is to define a threshold, $\theta$, such that dissipation is added when $\Delta S > \theta$~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium}.
+A criteria for introducing artificial dissipation that has been used successfully ~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium} is to define a threshold, $\theta$, such that dissipation is added when:
+\begin{equation} \label{eq:ds-threshold}
+  \Delta S(\pos, t) > \theta.
+\end{equation}
 A potential drawback of defining a threshold \textit{a priori} is that in order to ensure the model still retains some physical integrity, only a small number of sites can have artificial dissipation added.
@@
 The threshold can be determined on a case-by-case basis through trial-and-error or by a preliminary analysis.
-In the current work, the criteria that is used for determining whether dissipation should be added is:
+In the current work, the criteria that is used for determining whether dissipation should be added is a combination of \eqref{eq:ds-threshold} and the following:
 
 \begin{equation} \label{eq:dissipation-criteria}
-    \Delta S > \overline{\Delta S} + n_\sigma \cdot \sigma_{\Delta S},
+  \Delta S(\pos, t) > \overline{\Delta S} + n_\sigma \cdot \sigma_{\Delta S},
 \end{equation}
@@
 The criteria described in \eqref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead can think in terms of what would be the maximum percentage of sites one would want artificial dissipation to be added to.
+A disadvantage of \eqref{eq:dissipation-critera} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artifical dissipation is unnecessary.
+By required that both \eqref{eq:ds-theshold} and \eqref{eq:dissipation-criteria} be met before artifical dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
 
@@
 \begin{equation} \label{eq:second-invariant}
-\dot{\gamma} = \sqrt{2 \sum_{\alpha,\beta=1}^{2} \srtensor}.
+\dot{\gamma} = \sqrt{2 \sum_{\alpha,\beta=1}^{2} \srtensor \srtensor}.
 \end{equation}
@@
 \end{equation}
-All BGK-BRT collision schemes use $\omega \in [0.05, 1.995]$ as the bounds.
-All simulations with artificial dissipation use the median filter with $n_s = 2.7$.
+All BGK-BRT collision schemes use $\omega \in [0.05, 1.995]$ as the bounds on the collision frequency.
+All simulations with artificial dissipation use the median filter with $\theta = 1.0 \times 10^{-6}$ and $n_{\sigma} = 2.7$ where $\theta$ is the threshold that $\Delta S(\pos, t$ must exceed before dissipation is added \eqref{eq:ds-threshold} and $n_{\sigma}$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S(\pos, t)$ must be before dissipation is added \eqref{eq:dissipation-criteria}.
 All numerical values given in this section are in lattice units unless otherwise stated.
@@
 \newcommand{\pgrad}{\frac{\partial p}{\partial x}}
+\newcommand{\unorm}{\mathbf{\hat{n}}}
 
@@
 A schematic is shown in \Fref{fig:poise-schematic}.
-No-slip boundary conditions are enforced at the top and bottom boundaries and periodic boundary conditions are enforced at the left and right boundaries.
-The total height of the channel will be denoted by $H$.
+No-slip boundary conditions are enforced at the top and bottom boundaries with a wall velocity of zero so that $\mvel \times \unorm = 0$ where $\unorm$ is the unit normal vector to the boundary. 
+Periodic boundary conditions are enforced at the left and right boundaries.
+The total height of the channel is denoted by $H$.
 The center of the channel is $y = 0$ and $y \in [-h, h]$ where $h = \frac{H}{2}$.
@@
     \includegraphics[width=\linewidth]{figs/poise-schematic}
-    \caption{Schematic of Poiseuille flow; no-slip boundary conditions are enforced at the top and bottom boundaries, and periodic boundary conditions are enforced at the left and right boundaries.}
+    \caption{Schematic of Poiseuille flow; no-slip boundary conditions are enforced at the top and bottom boundaries, and periodic boundary conditions are enforced at the left and right boundaries.
+The size of the domain is $L = 32$  and $H = 64$.}
     \label{fig:poise-schematic}
@@
 
-Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps (although the length, $L$, seems small, the effect is negligible as the east-west boundary conditions are periodic).
+Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps.
+The lattice size was chosen to be sufficiently fine for accuracy (although the length, $L = 32$, seems small, the effect is negligible as the east-west boundary conditions are periodic) and the number of time steps were set high enough to ensure a steady-state would be reached.
 In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
-For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$.
+For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$, which follows the recommendation of~\cite{lallemand2000theory} for reasons of stability, and has been successfully used to simulate non-Newtonian flows in the past~\cite{fallah2012multiple,grasinger2015simulation}.
 
@@
 The Bingham plastic simulations were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a plastic viscosity of $\mu_p = 0.2$.
-The yield stress was varied between the four different values $\tau_y = [4.0, 8.0, 12.0, 16.0] \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
-(Recall $m$ is the stress growth exponent for the Papanastasiou approximation.)
+The yield stress was varied between four different values $\tau_y = [4.0, 8.0, 12.0, 16.0] \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
+Recall that $m$ is the stress growth exponent for the Papanastasiou approximation.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
-The analytical solution for Bingham plastic flow is given by:
+The analytical solution for Poiseuille flow of a Bingham plastic fluid is given by:
 \begin{equation} \label{eq:analytical-bingham}
@@
 \end{equation}
-\noindent where $y_{\tau} = -\tau_y / \pgrad$ is the point at which the fluid yields.
-
-The results are presented in \Fref{tab:poise-bing}.
-The Reynold's number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution and $H = 2h$ is the total height of the channel.
+\noindent where $y_{\tau} = -\tau_y / \pgrad$ is the vertical location at which the fluid yields.
+
+The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing}.
+The Reynold's number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution.
 The Bingham number was computed by $Bn = \frac{\tau_y H}{\mu_p U}$.
@@
 
-As has been reported previously, for the BGK collision operator--regardless of whether or not median entropic filtering is used--using a stress growth exponent of $m = 10^5$ is more accurate with respect to the analytical solution than using a stress growth exponent of $m = 10^8$~\cite{chen2014simulations}.
-A larger stress growth exponent results in a more accurate Papastasiou approximation of the true Bingham plastic constitutive model, however, it leads to more noise that degrades the numerical solution for the BGK collision operator.
-Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helps dissipate the noise that occurs as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
-In fact, the median filter resulted in a less accurate solution in nearly all cases and rendered the solution altogether useless for the higher yield stress flows (relative errors of 25-75\%).
-
-The BGK collision operator using a stress growth exponent of $m = 10^5$ was in general the fastest.
+As has been reported previously, for the BGK collision operator, using a stress growth exponent of $m = 10^5$ is more accurate with respect to the analytical solution than using a stress growth exponent of $m = 10^8$~\cite{chen2014simulations}.
+A larger stress growth exponent results in a more accurate Papastasiou approximation of the true Bingham plastic constitutive model, however, it leads to more nonphysical oscillations that degrade the numerical solution for the BGK collision operator.
+Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helps mitigate errors that occur as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
+In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress flows (relative errors of approximately 25-75\%).
+
+In general, the BGK collision operator using a stress growth exponent of $m = 10^5$ had the lowest computational time.
 This can be attributed to the fact that a smoother approximation of the Bingham plastic constitutive model would lead to a solution for $\mu_{app}$ converging with less iterations.
-The BGK collision operator using a stress growth exponent of $m = 10^5$ also performed acceptably in terms of accuracy for lower yield stress flows, however, for $\tau_y = 12 \times 10^{-5}$ and $16 \times 10^{-5}$ the relative $L_\infty$ errors were 8.9\% and 19\%, respectively--larger than what would be considered acceptable for most engineering applications.
-
-The LBM approximations are plotted with the analytical solution for the BGK collision operator using $m = 10^5$ and $m = 10^8$ in \Fref{fig:bing-analyt-bgk-5} and \Fref{fig:bing-analyt-bgk-8}, respectively.
-\Fref{fig:bing-analyt-bgk-5} shows that the error for the BGK model with $m = 10^5$ is not due to noise, but instead the inaccuracy of the Papastasiou approximation with a lower stress growth exponent.
-\Fref{fig:bing-analyt-bgk-8} shows that the error for the BGK model with $m = 10^8$ is due to numerical noise.
-In order to better understand why median filtering did not eliminate the noise in high yield stress flows, but instead exacerbate the problem, it is necessary to investigate what is happening at the particle distribution scale.
+The BGK collision operator using a stress growth exponent of $m = 10^5$ experienced relatively low error for lower yield stress flows, however, for $\tau_y = 12 \times 10^{-5}$ and $16 \times 10^{-5}$ the relative $L_\infty$ errors were 8.9\% and 19\%, respectively, which is larger than what would be considered acceptable for most engineering applications.
+
+The LBM approximations of the velocity profile across the channel (more specifically, $u_x(\pos_j, 25000)$ where $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, i.e.
+$\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction) for the BGK collision operator using $m = 10^5$ and $m = 10^8$ are plotted with the analytical solution \eqref{eq:bingham} in \Fref{fig:bing-analyt-bgk-5} and \Fref{fig:bing-analyt-bgk-8}, respectively.
+Due to the smoothness of the LBM approximation in \Fref{fig:bing-analyt-bgk-5}, one can conclude that the error for the BGK model with $m = 10^5$ is not due to nonphysical oscillations, but instead the inaccuracy of the Papastasiou approximation with a lower stress growth exponent.
+In contrast, the LBM approximation in \Fref{fig:bing-analyt-bgk-8} is not smooth, which suggests that the error for the BGK model with $m = 10^8$ is due to nonphysical oscillations.
 
@@
 
-\Fref{fig:feq-vs-f_bgk} compares particle distributions to their respective quasiequilbiriums.
-It can be seen that in the BGK, $m = 10^8$ solution, the noise due to nonlinear constitutive relation pollutes the quasiequilibrium values.
-Because the noise makes its way into the quasiequilibrium values, entropic median filtering does not help dampen the noise but instead contracts the particle distributions closer to the quasiequilibrium noise--explaining why the median filtered results were less accurate for Bingham plastic Poiseuille flow.
-In order to ensure that this phenomena was a side effect of all entropic filtering and not just entropic median filtering with $n_s = 2.7$, entropic filtering was tested with median filtering and Ehrenfests' regularization for $n_s = 2.7$ and for $\Delta S$ thresholds of $\delta = 10^{-4}$ and $\delta = 10^{-5}$.
-All filtering schemes suffered from approximately the same level of noise.
+In order to better understand why median filtering did not eliminate the nonphysical oscillations in high yield stress flows, but instead exacerbates the problem, it is necessary to investigate what is happening at the particle distribution scale.
+\Fref{fig:feq-vs-f_bgk} compares particle distributions to quasiequilibrium.
+More specifically, \Fref{fig:feq-vs-f_bgk} compares $f_i(\pos_j, 25000)$ to $f_i^{eq}(\pos_j, 25000)$ where $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, and $i = 5, 8$.
+It can be seen that in the BGK, $m = 10^8$ solution, that nonphysical oscillations pollute the quasiequilibrium profile as well.
+These nonphysical oscillations likely originate at the lattice scale due to the sharp gradient in the macroscopic velocity, $\mvel$, near the walls, and the sharp gradient in $\mvel$ is a result of a sharp gradient in $\mu_{app}$, namely the sharp gradient of the constitutive relationship.
+Because the nonphysical oscillations make their way into the quasiequilibrium values, entropic median filtering does not help dampen the oscillations but instead contracts the particle distributions closer to the quasiequilibrium oscillations, which explains why the median filtered results were less accurate for Bingham plastic Poiseuille flow.
+In order to ensure that this phenomena was a side effect of all entropic filtering, and not just entropic median filtering with $\theta = 1.0 \times 10^{-6}$ and $n_s = 2.7$, optimization was used to find the value of $\theta$ that minimized error for both median filtering and Ehrenfests' regularization.
+The optimization problem was defined as follows:
+\begin{align*}
+  \min_\theta \hspace{0.2in} & f(\theta)\\
+\text{such that} \hspace{0.2in} & \theta \in [10^{-10}, 2.0]
+\end{align*}
+\noindent where $f(\theta) \equiv $ the relative $L_2$ error between the LBM approximation using $m = 10^8$ and entropic filtering with a $\Delta S$ threshold of $\theta$ (used in the criteria defined in \eqref{eq:ds-threshold}).
+The bounds on $\theta$, namely $\theta \in [10^{-10}, 2]$ represent $\theta$ such that 100\% of lattice sites are filtered for every time step ($\theta = 10^{-10}$) and $\theta$ such that no lattice sites are filtered for any time step ($\theta = 2.0$).
+The optimization problem was solved for the cases of using median filtering and Ehrenfests' regularization using both Brent's method and the Golden Section search.
+For all four cases, the optimal solution was $\theta = 2.0$ and the cost function, $f(2.0)$, was equal to the relative $L_2$ error for the BGK collision operator, $m = 10^8$ and no entropic filtering.
+The results of the optimization suggest that entropic filtering, at best does not affect the accuracy of LBM, and often adversly affects the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization is used, and regardless of what the $\Delta S$ threshold, $\theta$, is set to.
+Thus, entropic filtering is not effective for increasing the accuracy of Bingham plastic Poiseuille flow.
 
@@
 \begin{tabulary}{\linewidth}{c c}
-\includegraphics[width=2.1in]{figs/poise-bingham/bgk-5/feq-vs-f_5.png}
+\includegraphics[width=2.75in]{figs/poise-bingham/bgk-5/feq-vs-f_5.png}
 &
-\includegraphics[width=2.1in]{figs/poise-bingham/bgk-5/feq-vs-f_8.png}
-\\
-\includegraphics[width=2.1in]{figs/poise-bingham/bgk-8/feq-vs-f_5.png}
+\includegraphics[width=2.75in]{figs/poise-bingham/bgk-5/feq-vs-f_8.png}
+\\
+\includegraphics[width=2.75in]{figs/poise-bingham/bgk-8/feq-vs-f_5.png}
 &
-\includegraphics[width=2.1in]{figs/poise-bingham/bgk-8/feq-vs-f_8.png}
+\includegraphics[width=2.75in]{figs/poise-bingham/bgk-8/feq-vs-f_8.png}
 \end{tabulary}
@@
 
-On average, the LBM model using the MRT collision operator was 2.9 times slower than the fastest model for the given material parameters.
-However, the increased computational time can be justified for the MRT collision operator because it did not suffer from the same level of noise as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
+On average, the LBM model using the MRT collision operator took 2.9 times more computation time than the BGK collision operator with $m = 10^5$.
+However, the increased computational time can be justified for the MRT collision operator because MRT solution did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
 when $m = 10^8$.
 The MRT collision operator was therefore the most accurate solution for all cases.
-\Fref{fig:feq-vs-f_mrt} compares particle distributions to quasiequilibrium for the MRT model; in contrast to the BGK with $m = 10^8$, there is no noise in the quasiequilibrium distributions.
+\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilbirium profiles, $f_i^{eq}$ for the MRT model; in contrast to the BGK with $m = 10^8$, there are no nonphysical oscillations in the quasiequilibrium distributions.
 
@@
     \begin{tabulary}{\linewidth}{c c}
-        \includegraphics[width=2.1in]{figs/poise-bingham/mrt/feq-vs-f_5.png}
+        \includegraphics[width=2.75in]{figs/poise-bingham/mrt/feq-vs-f_5.png}
         &
-        \includegraphics[width=2.1in]{figs/poise-bingham/mrt/feq-vs-f_8.png}
+        \includegraphics[width=2.75in]{figs/poise-bingham/mrt/feq-vs-f_8.png}
     \end{tabulary}
@@
 
-A natural question would be to ask ``if the strength of the MRT collision operator is damping out the ghost modes, then is the noise for the BGK with $m = 10^8$ (with and without median filtering) indeed due to the ghost modes?''.
-\Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment with respect to time for each of the collision schemes and was calculated using:
+Another important question to ask is, if the strength of the MRT collision operator is damping out the ghost modes, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of the ghost modes?.
+\Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment, $\boldsymbol{\epsilon}^{neq}$, with respect to time for each of the collision schemes, which was calculated as:
 \begin{equation}
@@
 \end{equation}
-\noindent where $\epsilon_j = 4f_0(\pos_j) - 2f_1(\pos_j) - 2f_2(\pos_j) - 2f_3(\pos_j) - 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) + f_7(\pos_j) + f_8(\pos_j)$ is related to the square of the energy, and $\pos_j = (16, j)$ for $j = 1, 2, ..., 64$, i.e.
+\noindent where $||\hdots||_2$ is the Euclidean norm, $\epsilon_j = 4f_0(\pos_j) - 2f_1(\pos_j) - 2f_2(\pos_j) - 2f_3(\pos_j) - 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) + f_7(\pos_j) + f_8(\pos_j)$ is related to the square of the energy, and $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, i.e.
 $\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
-The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT or BGK with $m = 10^5$ collision schemes--which suggests that the numerical noise the BGK with $m = 10^8$ suffered from was in part due to the $\epsilon$ moment.
+The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ suffered from was in part due to the $\epsilon$ moment.
 
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon}
-    \caption{Evolution of relative $L_2$, $\epsilon^{neq}$ with time.
+    \caption{Evolution of relative $L_2$ norm, $\epsilon^{neq}$ with time.
 The norm of the $\epsilon^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $\epsilon$ moment.}
@@
 
-\Fref{fig:qx} shows a measure of the nonequilibrium $q_x$ moment, or energy flux in the x-direction, with respect to time for each of the collision schemes and was calculated using:
+\Fref{fig:qx} shows a measure of the nonequilibrium $q_x$ moment, $\mathbf{q_x}^{neq}$, or energy flux in the x-direction, with respect to time for each of the collision schemes, which was calculated as:
 \begin{equation}
@@
 The values measured for the BGK with $m = 10^8$ and median filtered were consistently greater than any other collision scheme.
-The $q_x$ moment was likely the cause of the significant error and noise in the BGK with $m = 10^8$ and median filtering solution.
-As expected, the $q_x^{neq}$ relative $L_2$ for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $q_x^{neq}$ relative $L_2$ was negligible for the MRT collision scheme.
-These results are consistent with the solution noise and error.
+The relative $L_2$ norm of the $\mathbf{q_x}^{neq}$ for the BGK collision operator with $m = 10^8$ and median filtering solution spent most of the simulation between approximately 50\% and 100\%, this suggests that the $q_x$ moment was likely the cause of the nonphysical oscillations and error in its solution.
+As expected (if the dominant source of error in the LBM approximation for Bingham plastic Poiseuille flow is indeed due to the $q_x$ moment), the $\mathbf{q_x^{neq}}$ relative $L_2$ norm for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $\mathbf{q_x^{neq}}$ relative $L_2$ was negligible for the MRT collision scheme.
+These results are consistent with solution error.
 
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/qx}
-    \caption{Evolution of relative $L_2$, $q_x^{neq}$ with time.
-The norm of the $q_x^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $q_x$ moment.}
+    \caption{Evolution of relative $L_2$ norm, $\mathbf{q_x}^{neq}$ with time.
+The norm of the $\mathbf{q_x^{neq}}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $q_x$ moment.}
     \label{fig:qx}
@@
 
-It is likely that the nonequilibrium $\epsilon$ and $q_x$ moment cause oscillations at the lattice level that have a cumulative effect in time.
-\Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time.
-\Fref{fig:qx-cumulative} suggests that the $q_x^{neq}$ dominates and is the primary source of oscillations.
-If an LBM collision scheme is to be developed for simulating high yield stress flows that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it should focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.
+It is likely that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time, i.e.
+osciallations build in amplitude with time.
+\Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time, which was calculated as:
+\begin{equation}
+  \int_0^{T} \frac{||\boldsymbol{\epsilon}(t) - \boldsymbol{\epsilon}^{eq}(t)||_2}{||\boldsymbol{\epsilon}^{eq}(t)||_2} dt,
+\end{equation}
+\noindent and,
+\begin{equation}
+  \int_0^{T} \frac{||\mathbf{q_x}(t) - \mathbf{q_x}^{eq}(t)||_2}{||\mathbf{q_x}^{eq}(t)||_2} dt,
+\end{equation}
+\noindent where $T$ is the current time. 
+\Fref{fig:qx} and \Fref{fig:qx-cumulative} shows that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the ghost mode associated with the $q_x$ moment dominates and is the primary source of oscillations.
+It can be inferred that the $q_x$ moment was significant source of oscillations in this case because the Poiseuille flow was in the x-direction.
+If an LBM collision scheme is to be developed for simulating high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it should focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.
 
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon_cumulative}
-    \caption{Cumulative relative $L_2$, $\epsilon^{neq}$ with time.
+    \caption{Cumulative relative $L_2$ norm, $\epsilon^{neq}$ with time.
 Oscillations can have a positive effect on each other, building up.
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/qx_cumulative}
-    \caption{Cumulative relative $L_2$, $q_x^{neq}$ with time.
+    \caption{Cumulative relative $L_2$ norm, $\mathbf{q_x^{neq}}$ with time.
 Oscillations can have a positive effect on each other, building up.
-The cumulative relative $L_2$, $q_x^{neq}$ is a measure of how much oscillations due to the $q_x$ moment may have been building in time.}
+The cumulative relative $L_2$, $\mathbf{q_x^{neq}}$ is a measure of how much oscillations due to the $q_x$ moment may have been building in time.}
     \label{fig:qx-cumulative}
@@
 
-The power-law simulations were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a flow consistency index of $k = 0.2$.
-The flow behavior index was varied between the four different values $n = [0.5, 0.75, 1.25, 1.50]$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $[0.05, 1.995]$.
+The simulations of the Poiseuille flow with the power-law constitutive relationship were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a flow consistency index of $k = 0.2$.
+The flow behavior index, $n$, was varied between the four different values: $[0.5, 0.75, 1.25, 1.50]$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $[0.05, 1.995]$.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
-The analytical solution for power-law fluid flow is given by:
+The analytical solution for the velocity profile of Poiseuille flow with the power-law constitutive relationship is given by:
 \begin{equation} \label{eq:analytical-power-law}
@@
 
-The results are presented in \Fref{tab:poise-power-law}.
-In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
-The Reynold's number was computed by $Re = \frac{\rho U^{2-n} H^n}{k}$, where $U$ is the maximum velocity given by the analytical solution and $H = 2h$ is the total height of the channel.
+The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-power-law}.
+The Reynold's number was computed by $Re = \frac{\rho U^{2-n} H^n}{k}$.
 
@@
 
-There are some interesting things to note from the power-law Poiseuille flow results.
-Although for $n = 0.5, 0.75, 1.25$ there is still the same efficiency for accuracy trade off when it comes to deciding between the BGK and MRT collision operators, for $n = 1.5$ the BGK operator with median filtering was the fastest, most accurate, and only solution that remained stable.
-In fact, besides for the case of $n = 0.5$, i.e. rather extreme shear-thinning, the BGK with median filtering was both fast and accurate.
-The BGK-BRT had a moderate improvement of accuracy and efficiency--with respect to the regular BGK collision scheme--for flows where the constitutive relationship was only somewhat nonlinear, i.e.
+Although for $n = 0.5, 0.75, 1.25$ the MRT collision operator was more accurate and required more computation time than the BGK collision operator, for $n = 1.5$ the BGK operator with median filtering required the least computation time and was the most accurate with respect to the analytical solution.
+In fact, besides for the case of $n = 0.5$, i.e. relatively extreme shear-thinning, the BGK with median filtering was both computationally efficient and accurate.
+The BGK-BRT had a moderate improvement of accuracy and efficiency--with respect to the regular BGK collision scheme--for flow where the constitutive relationship was only somewhat nonlinear, i.e.
 for $n = 0.75, 1.25$.
-Note that none of the collision schemes were sufficiently accurate for the case of $n = 0.5$, and to improve this accuracy it would be necessary for a modeler to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
+Note that none of the collision schemes were sufficiently accurate (error greater than 15\%) for the case of $n = 0.5$, and to improve this accuracy it would be necessary to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
 
@@
 
-Lid-driven cavity flow is another good benchmark for two-dimensional fluid flow because it is sufficiently simple in terms of boundary conditions and implementation, while also complex in the flows and vortices it produces.
 The lid-driven benchmark was chosen for this numerical study because there are many results available in literature in which to compare with and because the vorticity of the flow coupled with the nonlinear constitutive equations should result in a challenge in terms of stability.
 Lid-driven cavity flow is generally characterized by a square cavity where a fluid velocity is prescribed tangential to the upper boundary and the remaining boundaries have a no-slip boundary condition.
-Lid-driven cavity flow is depicted in \Fref{fig:lid-driven}.
+The example of lid-driven cavity flow utilized herein is presented schematically in \Fref{fig:lid-driven}.
 
@@
 \includegraphics{figs/lid-driven}
-\caption{Schematic of lid-driven cavity flow; a velocity is prescribed tangential to the top boundary and no-slip is enforced at the remaining boundaries.}
+\caption{Schematic of lid-driven cavity flow; a velocity is prescribed tangential to the top boundary and no-slip is enforced at the remaining boundaries. $L = 100$}
 \label{fig:lid-driven}
@@
 
-The lid-driven cavity simulations presented in the section were all simulated on a coarse, $100 \times 100$ lattice; a coarse lattice was chosen in order to highlight concerns with stability and accuracy.
-The simulations were run for either 50000 timesteps, or until convergence was met.
+The lid-driven cavity simulations presented in the section were all simulated on a relatively coarse, $100 \times 100$ lattice (other studies simulating lid-driven cavity flow of a non-Newtonian fluid tend to use lattice sizes on the order of $256 \times 256$ and $512 \times 512$~\cite{chen2014simulations,li2014simulation}, this study follows that of~\cite{brownlee2008nonequilibrium}).
+A coarse lattice was chosen in order to highlight concerns with stability and accuracy.
+The simulations were run for either 50000 timesteps or until convergence was met.
 Convergence was defined by,
@@
 \noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current timestep.
-In reference to computational time, each of the simulations in this section were run on a single core of either an Intel I7-860 Quad-Core 2.80GHz processor, or an Intel Xeon v2 2.60 GHz processor.
-The lid velocity was prescribed at $u_o = 0.1$.
-For the MRT relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$.
+The lid velocity was prescribed as $u_o = 0.1$.
+For the MRT relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$ as these values have been successfully applied to simulating lid-driven cavity flow of non-Newtonian fluids in the past~\cite{chen2014simulations,li2014simulation}.
 All coordinate values presented in this section (used to describe the location of the center of vortices) are given normalized with respect to the length of the cavity side, i.e. as $(x / L, y / L)$.
-Note that all results presented as ``-'' indicate that the tests were numerically unstable to the extent that the degrees of freedom all over the domain were ``NaN''.
+Note that all results presented as ``-'' indicate that the tests were numerically unstable to the extent that the degrees of freedom all over the domain were nonsensical.
 
@@
 
-For the Bingham plastic numerical tests the Reynold's number was varied, $Re = 100, 1000, 5000, 10000$, and the Bingham number was varied, $Bn = 1, 10, 100$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
+For the Bingham plastic numerical tests the Reynold's number was varied:$[100, 1000, 5000, 10000]$, and the Bingham number was varied: $[1, 10, 100]$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
 The collision schemes that were tested were: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
@@
 
-Just as before with the Bingham plastic Poiseuille flow, if a modeler is using the BGK collision operator and entropic filtering is not available, a stress growth exponent of $m = 10^5$ yields faster, more accurate results (assuming the literature values are a good measure of accuracy) than using a larger stress growth exponent.
+Just as before with the Bingham plastic Poiseuille flow, if a modeler is using the BGK collision operator and entropic filtering is not available, a stress growth exponent of $m = 10^5$ yields faster, more accurate results than using a larger stress growth exponent.
 A smaller stress growth exponent is also more effective at producing accurate results in the BGK collision scheme than placing bounds on the relaxation time (BGK-BRT).
-However, although the BGK with $m = 10^5$ was the fastest model in all cases, it was also--like the other two BGK models without median filtering--unstable for $Re \ge 5000$.
-
-For flows with $Re \ge 5000$, the BGK with $m = 10^8$ and median filtering produced numerically stable results, but did not always seem to be consistent with literature values.
+However, although the BGK with $m = 10^5$ was the fastest model in all cases, it was also like the other two BGK models without median filtering in that it was unstable for $Re \ge 5000$.
+
+For flows with $Re \ge 5000$, the BGK with $m = 10^8$ and median filtering produced numerically stable results.
+However, in general, the BGK with $m = 10^8$ and median filtering was consistently different from literature values. 
 The apparent inaccuracy with regards to BGK with $m = 10^8$ and median filtering is probably due to how the numerical stability is enhanced--namely through artificial, nonphysical dissipation.
-In contrast to the nonphysical nature in which median filtering enhances stability, the stability enhancement used by MRT does not directly affect the macroscopic hydrodynamics of interest, so it again provides a stable, and typically, the most accurate solution.
+In contrast to the nonphysical nature in which median filtering enhances stability, the stability enhancement used by MRT does not directly affect the macroscopic hydrodynamics of interest, so it again provides a stable, and in most of the cases examined herein, the most accurate solution.
 % MG: verify this sentence is correct
@@
 
-For the sake of completeness, the MRT collision operator was also tested with median filtering applied.
-There does not seem to be any reason to recommend the MRT collision operator with median filtering in practice because the MRT collision operator is stable all on its own and the artificial dissipation introduced by median filtering has the potential to adversely affect the accuracy of the results.
-
-In summary, 
+In summary for the lid-driven cavity flow of a Bingham plastic fluid, 
 \begin{itemize}
-    \item for low Reynold's number flows when computational efficiency is important, the BGK with $m = 10^5$ is recommended,
-    \item for high Reynold's number flows when computational efficiency is important, the BGK with $m = 10^8$ and median filtering is recommended,
-    \item and for high Reynold's number flows when accuracy is more important than computational efficiency, the MRT with $m = 10^8$ is recommended.
+    \item for low Reynold's number flows the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
+    \item for high Reynold's number flows the BGK collision operator requires entropic filtering to remain stable,
+    \item and the MRT collision operator with $m = 10^8$ produces solutions with high accuracy and stability for all of the cases examined herein, but at an increased (approximately 5-10 times more) computational cost.
 \end{itemize}
@@
 
-For the power-law numerical tests the Reynold's number was varied, $Re = 100, 1000, 5000, 1000$, and the flow behavior index was varied, $n = 0.5, 1.5$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
+For the power-law numerical tests the Reynold's number was varied: $[100, 1000, 5000, 10000]$, and the flow behavior index, $n$, was varied: $[0.5, 1.5]$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
 The collision schemes that were tested were: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT with $m = 10^8$ and (5) MRT with $m = 10^8$ and median filtering.
 \Fref{tab:lid-powerlaw05} and \Fref{tab:lid-powerlaw15} compare the center location of the main vortex to literature values taken from~\citet{li2014simulation}.
-The main vortex location is determined by calculating the stream function using Simpson's rule and finding where it attains a maximum.
 
@@
 
-For the shear-thinning results, $n = 0.5$, the BGK and BGK-BRT collision schemes were unstable when $Re \ge 1000$.
-Although there still appears to be somewhat of a trade off between computational efficiency and accuracy in regards to the BGK with median filtering and MRT collision schemes, it does not seem to be nearly as drastic as when simulating Bingham plastic lid-driven flow; e.g.
-neither collision scheme consistently agrees with the literature values; and the MRT collision operator is only 2-3 times slower than the BGK with median filtering.
-
-For the shear-thickening results, $n = 1.5$, there are no issues of stability.
-What is interesting is that all of the collision schemes produce similar results for each of the cases (with the exception of the high Reynold's number case, $Re = 10000$), and the BGK-BRT is consistently the fastest solution.
-The reason the BGK-BRT may be the fastest is that because the collision frequency, $\omega$, is bounded, and consequently $\mu_{app}$ is bounded, meaning the solution to the constitutive equation may be converging to a bound with less iterations than the other collision schemes.
-
-In summary, 
-\begin{itemize}
-    \item for shear-thinning flows, either the BGK with median filtering or MRT collision schemes are recommended for stability reasons,
-    \item for shear-thickening flows, there is much less concern for numerical instability.
-\end{itemize}
-
-\section{Conclusion}
+For the shear-thinning results, $n = 0.5$, at the lowest Reynold's number considered, $Re = 100$, all of the collision schemes produced results that agreed well with the literature value.
+However, the solutions became unstable at higher Reynold's numbers and entropic filtering was necessary to produce a stable soltuion when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
+Although there still appears to be somewhat of a tradeoff between computational efficiency and accuracy in regards to the BGK with median filtering and MRT with median filtering, it does not seem to be nearly as drastic as when simulating Bingham plastic lid-driven flow; e.g.
+neither collision scheme consistently agrees with the literature values; and the MRT with median filtering is only 2-3 times slower than the BGK with median filtering.
+A good example of this tradeoff is the shear-thinning results for $Re = 5000$: the location of the main vortex for the BGK with median filtering differed from literature by more than 10\% in both the $x$ and $y$ directions but the MRT with median filtering has approximately twice the computational cost.
+
+For the shear-thickening results, $n = 1.5$, there were no issues of stability.
+What is interesting is that all of the collision schemes produce similar results for each of the cases, with the difference in the location of the main vortex being no greater than 1\% in either the $x$ or $y$ direction for any two collision schemes (with the exception of the high Reynold's number case, $Re = 10000$).
+The BGK-BRT consistently needed the least amount of computation time.
+The reason the BGK-BRT may be the most efficient is that because the collision frequency, $\omega$, is bounded, and consequently $\mu_{app}$ is bounded, meaning the solution to the constitutive equation may be converging to a bound with less iterations than the other collision schemes.
+
+\section{Conclusions}
 
@@
 \begin{itemize}
-  \item For high yield stress fluids in Poiseuille-type flows, only the MRT collision operator did not suffer from numerical noise.
+  \item For high yield stress fluids in Poiseuille-type flows, only the MRT collision operator did not suffer from nonphysical oscillations.
   \begin{itemize}
-      \item The noise appeared to be due to oscillations arising from high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
+      \item The oscillations appeared to be due to high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
       \item If a collision scheme is to be developed for high yield stress, Poiseuille-type flows that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, then it should focus on dampening the nonequilibrium energy flux moment.
@@
   \item In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
-  \item To summarize plastic, lid-driven cavity flows,
+  \item To summarize Bingham plastic, lid-driven cavity flows,
     \begin{itemize}
-  \item for low Reynold's number flows when computational efficiency is important, the BGK with $m = 10^5$ is recommended,
-  \item for high Reynold's number flows when computational efficiency is important, the BGK with $m = 10^8$ and median filtering is recommended,
-  \item and for high Reynold's number flows when accuracy is more important than computational efficiency, the MRT with $m = 10^8$ is recommended.
+  \item for low Reynold's number flows the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
+    \item for high Reynold's number flows the BGK collision operator requires entropic filtering to remain stable.
     \end{itemize}
@@
     \begin{itemize}
-  \item for shear-thinning either the BGK with median filtering or MRT collision schemes are recommended for stability reasons,
+      \item for shear-thinning fluids at high Reynold's numbers ($Re \ge 5000$) median filtering is necessary to produce a stable solution,
   \item for shear-thickening, lid-driven cavity flows, there is much less concern for numerical instability.

### 2016-07-26 figs/poise-bingham/mrt/process5.jl

@@ -1,0 +1,20 @@
+using Gadfly
+
+include("/home/clementine/Shared/Dev/lbxflow/inc/LBXFlow.jl");
+using LBXFlow
+
+sim = LBXFlow.loadsim_jld("bak.jld")[2];
+const i = 16;
+const ni, nj = size(sim.msm.rho);
+const xs = linspace(-0.5, 0.5, nj);
+for k=1:sim.lat.n
+  feq = zeros(nj);
+  for j=1:nj
+    feq[j] = LBXFlow.feq_incomp(sim.lat, sim.msm.rho[i, j], sim.msm.u[:, i, j], 
+                                k);
+  end
+  draw(PNG("feq-$k.png", 5.5inch, 3.inch), 
+       plot(x=xs, y=feq, Geom.line, Guide.XLabel("y / H"),
+            Guide.YLabel("f<sub>$k</sub><sup>eq</sup>")
+            ));
+end

### 2016-07-27 src/algorithm/streaming.tex
wrote some basic algorithms

@@
 \begin{algorithmic}[1]
-\Procedure{streaming}{$f$}
-Initialize temporary 3D array the size of lattice, $f_temp$
+\Procedure{streaming}{$f$} \Comment{$f$ is a 3D Array of particle distributions}
+\Statex \Comment{$f[k, i, j] \equiv f_k(\pos_{ij}, t)$}
+\State Initialize temporary 3D array the size of lattice, $f_{temp}$
+\Statex
+\For{each node $(i, j)$}
+  \For{each discrete velocity vector, $\pvel_k$}
+    \If {$(i + \pvel_{k1} \Delta t, j + \pvel_{k2} \Delta t)$ is in bounds and is not a solid wall}
+      \State $f_{temp}[k, i + \pvel_{k1} \Delta t, j + \pvel_{k2} \Delta t] = f[k, i, j]$
+    \EndIf
+  \EndFor
+\EndFor
+\Statex
+\State \textbf{copy} $f_{temp}$ to $f$
+\Statex
 \EndProcedure

### 2016-07-27 src/algorithm/update-cell-states.tex
wrote some basic algorithms

\begin{algorithm}
\caption{Update Cell States and Mass Redistribution} \label{algo:update-cell-states}
\begin{algorithmic}[1]
\Procedure{UpdateStates}{}

\Statex
\State Initialize (empty) data structure, $\iset_f$ \Comment{$\iset$ cells to be transitioned to $\fset$}
\State Initialize (empty) data structure, $\iset_g$ \Comment{$\iset$ cells to be transitioned to $\gset$}
\State Initialize variable that represents global excess mass, $M_{gex} = 0$

\Statex
\ForAll{cells $\pos \in \iset$} \Comment{Mark interface cells to transition}
  \If{$M(\pos, t) < -\delta_{trans}$}
    \State Add $\pos$ to $\iset_f$
  \ElsIf{$M(\pos, t) > \rho(\pos, t) + \delta_{trans}$}
    \State Add $\pos$ to $\iset_g$
  \EndIf
\EndFor

\Statex
\ForAll{cells $\pos \in \iset_f$}
\Statex \Comment{Prepare neighborhoods of $\iset$ cells that are to be transitioned to $\fset$ cells}
  \State Calculate the average density, $\rho_{avg}$, of the neighborhood
  \State Calculate the average velocity, $\mathbf{u}_{avg}$, of the neighborhood
  \ForAll{cells $\pos_n$ in the neighborhood of $\pos$}
    \If{$\pos_n \in \gset$}
      \State Transition $\pos_n$ from a $\gset$ state to $\iset$ state
      \State $f_i(\pos_n, t) = f_i^{eq}(\rho_{avg}, \mathbf{u}_{avg})$
    \ElsIf{$\pos_n \in \iset_g$}
      \State Remove $\pos_n$ from $\iset_g$ \Comment{Ensure interface is continuous}
    \EndIf
  \EndFor
\EndFor

\Statex
\ForAll{cells $\pos \in \iset_g$}
\Statex \Comment{Prepare neighborhoods of $\iset$ cells that are to be transitioned to $\gset$ cells}
  \ForAll{cells $\pos_n$ in the neighborhood of $\pos$}
    \If{$\pos_n \in \fset$}
      \State Transition $\pos_n$ from a $\fset$ state to $\iset$ state
    \EndIf
  \EndFor
\EndFor

\algstore{ucsalg1}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\begin{algorithmic}[1]
\algrestore{ucsalg1}
\ForAll{cells $\pos \in \iset_f$} 
\Statex \Comment{Transition from $\iset$ to $\fset$ and redistribute excess mass}
  \State Let $M_{ex} = M(\pos, t) - \rho(\pos, t)$
  \State Approximate the unit normal from the interface toward empty space, \indent \hspace{0.3in} $\mathbf{n} = \frac{1}{2} \begin{Bmatrix}
\epsilon_{i-1, j} - \epsilon_{i+1, j} \\
\epsilon_{i, j-1} - \epsilon_{i, j+1}
\end{Bmatrix}$ \Comment{Central difference of fluid fraction}
  \State Initialize $v_{sum} = 0$
  \State Initialize (empty) data structure, $\iset_r$
  \ForAll{cells $\pos_n$ in the neighborhood of $\pos$}
    \If{$\pos_n \in \iset$}
      \State Add $\pos_n$ to $\iset_r$
      \State Let $v_i = \begin{cases}
					\mathbf{n} \cdot \pvel_i, & \mathbf{n} \cdot \pvel_i > 0 \\
					0, & \text{otherwise}
				\end{cases}$
      \State $v_{sum} = v_{sum} + v_i$
    \EndIf
  \EndFor
  \If{$v_{sum} > 0$}
    \ForAll{cells $\pos_j \in \iset_r$}
      \State $M(\pos_j, t) = M(\pos_j, t) + M_{ex} \frac{v_i}{v_{sum}}$
    \EndFor
  \ElsIf{$v_{sum} = 0$ and $\iset_r \neq \emptyset$}
    \State Let $n_r$ be the number of cells in $\iset_r$
    \ForAll{cells $\pos_j \in \iset_r$}
      \State $M(\pos_j, t) = M(\pos_j, t) + \frac{M_{ex}}{n_r}$
    \EndFor
  \Else
    \State $M_{gex} = M_{gex} + M_{ex}$
  \EndIf
  \State Transition $\pos$ from an $\iset$ state to $\fset$ state
  \State Set $M(\pos, t) = \rho(\pos, t)$
\EndFor

\algstore{ucsalg2}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\begin{algorithmic}[1]
\algrestore{ucsalg2}
\ForAll{cells $\pos \in \iset_g$}
\Statex \Comment{Transition from $\iset$ to $\gset$ and redistribute excess mass}
  \State Let $M_{ex} = M(\pos, t)$
  \State Approximate the unit normal from the interface toward empty space, \indent \hspace{0.3in} $\mathbf{n} = \frac{1}{2} \begin{Bmatrix}
\epsilon_{i-1, j} - \epsilon_{i+1, j} \\
\epsilon_{i, j-1} - \epsilon_{i, j+1}
\end{Bmatrix}$ \Comment{Central difference of fluid fraction}
  \State Initialize $v_{sum} = 0$
  \State Initialize (empty) data structure, $\iset_r$
  \ForAll{cells $\pos_n$ in the neighborhood of $\pos$}
    \If{$\pos_n \in \iset$}
      \State Add $\pos_n$ to $\iset_r$
      \State Let $v_i = \begin{cases}
					-\mathbf{n} \cdot \pvel_i, & \mathbf{n} \cdot \pvel_i < 0 \\
					0, & \text{otherwise}
				\end{cases}$
      \State $v_{sum} = v_{sum} + v_i$
    \EndIf
  \EndFor
  \If{$v_{sum} > 0$}
    \ForAll{cells $\pos_j \in \iset_r$}
      \State $M(\pos_j, t) = M(\pos_j, t) + M_{ex} \frac{v_i}{v_{sum}}$
    \EndFor
  \ElsIf{$v_{sum} = 0$ and $\iset_r \neq \emptyset$}
    \State Let $n_r$ be the number of cells in $\iset_r$
    \ForAll{cells $\pos_j \in \iset_r$}
      \State $M(\pos_j, t) = M(\pos_j, t) + \frac{M_{ex}}{n_r}$
    \EndFor
  \Else
    \State $M_{gex} = M_{gex} + M_{ex}$
  \EndIf
  \State Transition $\pos$ from an $\iset$ state to $\gset$ state
  \State Set $M(\pos, t) = 0$
\EndFor

\Statex
\State Let $n_i$ be the number of cells in $\iset$
\ForAll{cells $\pos \in \iset$}
\Statex \Comment{Redistribute remaining excess mass evenly among interface cells}
  \State $M(\pos, t) = M(\pos, t) + \frac{M_{gex}}{n_i}$
\EndFor

\EndProcedure
\end{algorithmic}
\end{algorithm}

### 2016-07-28 thesis.tex
wrote some basic algorithms
finished first round of revisions

@@
 \usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
+\usepackage{algpseudocode}
 
@@
 \listoffigures
+\listofalgorithms
 
@@
 When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
-Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
-An important yet vulnerable barrier is the cement annulus.
-Creating the cement annulus, a process known as primary cementing, is difficult to perform optimally.
+%Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
+An important yet vulnerable barrier is the cement annulus and is created through a process known as primary cementing.
 Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
-Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
+%Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
 Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
@@
 
-Of critical importance to drilling operations of all kinds is that barriers are placed to prevent gasses and fluids from migrating from one geological zone to another (i.e., zonal isolation is achieved).
+Of critical importance to drilling operations of all kinds is that barriers are placed to prevent gasses and fluids from migrating from one geological zone to another (i.e., zonal isolation is achieved)~\cite{economides19901,jackson1993effect,goodwin1992cement,dusseault2014towards,li2016theory}.
 If zonal isolation is not achieved, gasses and fluids may migrate through the well to the surface and up into the atmosphere--potentially causing pollution, or at the very least, contributing to an increase in greenhouse gasses--or, worse yet, formation gasses and fluids can migrate into aquifers, potentially harming wild life and people.
@@
 There are various mechanisms by which this cement annulus can fail.
-According to previous studies~\cite{Bit02,Che85,Lev79,Ste88,Tal93,Zul12}, the most prevalent failure mechanisms of the cement annulus are:
+According to previous studies~\cite{Bit02,Che85,Lev79,Ste88,Tal93,Zul12,li2016theory}, the most prevalent failure mechanisms of the cement annulus are:
 \begin{itemize}
@@
 \item Poor fill of the annulus.
-If cement slurry does not fill the annulus then if drilling mud is left behind it can dry, crack, and provide a weak path for fluids and gasses under pressure to push through, or if a dry hole is is cemented and not filled properly, fluids and gasses can travel through channels and voids left in the annular space.
+If cement slurry does not fill the annulus and if drilling mud is left behind the mud can dry, crack, and provide a weak path for fluids and gasses under pressure to push through, or if a dry hole is cemented and not filled properly, fluids and gasses can travel through channels and voids left in the annular space.
 \item Gas channeling through the annulus during cement curing.
@@
 When the column begins to support itself, the hydrostatic pressure of the cement column--against the rock formation--begins to drop.
-If the hydrostatic pressure of the cement column drops below the pressure of a formation gas before the cement annulus has developed adequate strength to resist it, the formation gas can penetrate int the cement annulus and form channels, or degrade the cement annulus in other ways.
+If the hydrostatic pressure of the cement column drops below the pressure of a formation gas before the cement annulus has developed adequate strength to resist it, the formation gas can penetrate into the cement annulus and form channels, or degrade the cement annulus in other ways.
 \end{itemize}
@@
 In order to better understand, and thus improve primary cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
-For example,~\citet{Bei77} developed a mathematical model to ``describe the miscible displacement of drilling muds by cement slurries under laminar flow conditions''.
-Among their assumptions were that the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the y-coordinate.
+For example,~\citet{Bei77} developed a mathematical model to describe the displacement of drilling mud by the the cement slurry during primary cementing.
+Among their assumptions were that the flow as laminar, the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the distance from the center of the annulus to either the rock formation or steel casing surface.
 The model was then used to explore the effect of various parameters on the displacement efficiency of the spacer, such as the density ratio between the mud and the spacer, the viscosity ratio between the mud and the spacer, the displacement rate, the displaced phase yield stress, and the displacing phase yield stress.
@@
 \citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
-The model was more general than the model developed by~\citeauthor{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
-Annulus eccentricity occurs often in wellbores and can sometimes be extreme as it is difficult to keep long strings of casing centered in an imperfectly drilled hole.
+The model was more general than the model developed by~\citet{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
 The bulk fluid motion of the spacer and drilling mud during primary cementing were investigated.
 The intended use of the model was not to make general statements about what properties of the spacer and drilling mud were desirable, but to instead be used iteratively during the design process for the cement job.
-Results showed that the model met its intended purpose, and in particular, was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
+Results showed that the model was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
 \citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
-The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citeauthor{Bei77} or~\citeauthor{Bit02}.
+The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citet{Bei77} or~\citet{Bit02}.
 The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
-\citeauthor{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
+\citet{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
 The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
 \citet{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
-\citeauthor{zhao2016lattice} used the model to investigate the displacement interface shape in a horizontal, eccentric annulus; and in order to simplify the model from a 3D to 2D flow, used the classical Hele-Shaw approach.
-It was found, in general, that a greater density ratio between the displacing phase, i.e.
-cement slurry, and displaced phase, i.e.
-drilling mud, and lower eccentricity, results in better displacement in a horizontal, eccentric annulus.
+\citet{zhao2016lattice} used the model to investigate the shape of the interface between the displacing phase (i.e.
+cement slurry) and displaced phase (i.e.
+drilling mud) in a horizontal, eccentric annulus.
+It was found, in general, that a greater density ratio between the displacing phase and displaced phase and lower eccentricity results in better displacement in a horizontal, eccentric annulus.
 
@@
 
-The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features (i.e., the purpose of the current work is an extension of the work present in~\citet{grasinger2015simulation}).
-The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation, among other benefits.
+The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features.
+The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation.
 LBM also has other advantages as an approach for simulating primary cementing:
@@
 \item LBM is easily written in parallel.
-As hardware architectures shift from one or a few cores to, especially in the case of scientific computing, several CPUs, it becomes increasingly important that code for simulating complex and computationally expense phenomena be written in parallel.
+As hardware architectures shift from one or a few cores to several CPUs, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
 \end{itemize}
@@
 
-For the sake of computational simplicity and computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
-In addition, the collision operator used was the BGK collision operator.
-For more details on the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.
+For the sake of computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
+Idealizing the annulus as two-dimensional limits the modeling approach in that annulus eccentricity and other three-dimensional phenomena could not be considered.
+A two-dimensional representation of the annulus was deemed sufficient as the focus of this study is to establish the Lattice Boltzmann Method as a suitable method for simulating primary cementing in an annulus with an imperfect rock formation surface.
+In addition to using a D2Q9 lattice, the BGK collision operator used was in the current chapter.
+For more details on the fundamentals of the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.
 
@@
 
-Cement slurry is best modeled using a non-Newtonian constitutive relationship.
-A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
+Cement slurry is typically modeled using a non-Newtonian constitutive relationship.
+%A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
+As is commonly utilized for modeling cement and concrete fluid dynamics~\cite{ferraris2001fresh,ferraris1998testing,tattersall2003workability,roussel2007computational,roussel2015numerical,svec2012free}, the Bingham plastic constitutive relationship was utilized herein.
 Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
 The Bingham plastic constitutive model is difficult to work with numerically.
-Often, a smooth approximation is used instead--the Papastasiou approximation.
+Therefore, a smooth approximation, specifically the Papastasiou approximation, is often used~\cite{papanastasiou1987flows,chen2014simulations,tang2011bingham,Cha11,wang2008lattice,phillips2011lattice} and is used herein.
 Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
@@
 The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
-The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice,thurey2005interactive}.
+The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice} and \citet{thurey2005interactive}.
 
@@
 Consider at each node a cell of size $\text{d}x \times \text{d}x$ where $\text{d}x$ is the distance between nodes and the cell is centered at the node.
-In order to track where the primary fluid is and where it is not, i.e.
-which cells contain the primary fluid, a variable is introduced, $\epsilon$, which describes what fraction of the cell is covered by the primary fluid.
-Cells that have a fluid fraction of $\epsilon = 0$ contain no primary fluid, i.e.
-are empty; cells that have a fluid fraction of $\epsilon = 1$ are filled; and cells that have a fluid fraction $0 < \epsilon < 1$ are neither completely filled nor are they completely empty.
+In order to track where the primary fluid is and where it is not (i.e.
+which cells contain the primary fluid), a variable is introduced, $\epsilon$, which describes what fraction of the cell is covered by the primary fluid.
+Cells that have a fluid fraction of $\epsilon = 0$ contain no primary fluid (i.e.
+are empty); cells that have a fluid fraction of $\epsilon = 1$ are filled; and cells that have a fluid fraction $0 < \epsilon < 1$ are neither completely filled nor are they completely empty.
 The fluid fraction concept is similar to how fluids are tracked in Volume of Fluid (VOF) methods.
@@
 
-\newcommand{\opidx}{{\bar{i}}}
+\newcommand{\opidx}{{\bar{\i}}}
 \newcommand{\fset}{\mathcal{F}}
@@
 An interface is partially covered, and there is always a layer of interface cells between fluid and gas cells so that a fluid cell and a gas cell are never neighbors.
-The change in mass is calculated across fluid and interface cells for each discrete velocity direction as follows:
+The change in mass is calculated for fluid and interface cells (i.e. $\pos \in \fset \cup \iset$) for each discrete velocity direction as follows:
 \begin{equation} \label{eq:dM}
 \Delta M_i(\pos, t) = \begin{cases}
-f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t), & \pos \in \fset \text{ or } \pos + \pvel_i \in \fset \\
-\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \in \iset \\
-0, & \pos + \pvel_i \in \gset
+f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t), & \pos \in \fset \text{ or } \pos + \pvel_i \Delta t \in \fset \\
+%\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
+\frac{\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t}{2} (f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
+0, & \pos + \pvel_i \Delta t \in \gset
 \end{cases}
 \end{equation}
-\noindent where $\pvel_\opidx$ represents the discrete velocity vector that is opposite in direction to $\pvel_i$, i.e. $\pvel_\opidx = - \pvel_i$.
+\noindent where $\pvel_\opidx$ represents the discrete velocity vector that is opposite in direction to $\pvel_i$ (i.e. $\pvel_\opidx = - \pvel_i$).
 \eqref{eq:dM} states that when the border of the two cells is covered with fluid, the change in mass for a discrete velocity direction is equal to the difference of the particle distribution entering the cell from the opposite direction and particle distribution leaving the cell in that direction.
-When the border of the two cells are not covered with fluid, i.e.
-when both cells are interface cells, the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the cell border that is covered in fluid.
-The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M(\pos, t)$.
+When the border of the two cells are not covered with fluid (i.e.
+when both cells are interface cells) the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the cell border that is covered in fluid.
+The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M_i(\pos, t)$.
 Finally, note that $\Delta M_i(\pos, t) = -\Delta M_\opidx(\pos + \pvel_i \Delta t, t)$ for all $\pos \in \domain$ where $\domain$ is the domain.
-As a consequence, it is clear that $\sum_{\pos \in \domain}\Delta M(\pos, t) = 0$, i.e. that mass is conserved for the mass transfer step.
+As a consequence, it is clear that $\sum_{\pos \in \domain}\Delta M(\pos, t) = 0$ (i.e. that mass is conserved for the mass transfer step).
 
@@
 \textbf{Current State} & \textbf{Condition} & \textbf{Transition} \\
-\hline &  & \\
+\hline & \\
 $(\pos, t) \in \iset$ & $M(\pos, t) < -\tconst$ & $(\pos, t + \Delta t) \rightarrow \gset$, $M(\pos, t + \Delta t) = 0$ \\
                       & $M(\pos, t) > \tconst + \rho(\pos, t)$ & $(\pos, t + \Delta t) \rightarrow \fset$, $M(\pos, t + \Delta t) = \rho(\pos, t + \Delta t)$ \\
-                      &\\
+                      & &\\
 $(\pos, t) \in \fset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \gset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
-&\\
+& & \\
 $(\pos, t) \in \gset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \fset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
@@
 \centering
-\includegraphics{figs/cell-transition}
+\includegraphics[width=\linewidth]{figs/cell-transition}
 \caption{Example of cell state transitions.
@@
 
-The rules for updating cell states in \Fref{tab:cell-transition-rules} are fairly intuitive, however, there are some bookkeeping nuances that must be considered in order to ensure mass is properly conserved and interface continuity is achieved; e.g.
-note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells.
+The rules for updating cell states in \Fref{tab:cell-transition-rules} are fairly intuitive, however, there are some details that must be considered in order to ensure mass is properly conserved and interface continuity is achieved (e.g.
+note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells).
 The algorithm for updating cell states follows that which was presented in~\citet{thurey2005interactive}.
-It is presented here for the sake of completeness and for outlining minor modifications.
-\begin{enumerate}
-\item Initialize a variable that represents the global excess mass, e.g. $M_{gex} = 0$.
-\item For all cells $\pos \in \iset$, store cells that should be transitioned to $\fset$ in one data structure, $\iset_f$, and store cells that should be transitioned to $\gset$ in another, $\iset_g$.
-\item \label{enum:prep-fnbrs} Loop over cells that will be transitioned from interface to fluid cells, i.e. loop over $\iset_f$, and prepare their neighborhoods for transition:
-	\begin{enumerate}
-	\item Calculate the average density, $\rho_{avg}$, and velocity, $\mathbf{u}_{avg}$, of each neighborhood.
-	\item For all cells in the neighborhood of the new fluid cell:
-		\begin{enumerate}
-			\item If the cell is a gas cell, transition to interface state and set $f_i = f_i^{eq}(\rho_{avg}, \mathbf{u}_{avg})$, $\rho = \rho_{avg}$, and $\mathbf{u} = \mathbf{u}_{avg}$.
-			\item If the cell is an interface cell in $\iset_g$, remove it from $\iset_g$.
-		\end{enumerate}
-	\end{enumerate}
-\item \label{enum:prep-gnbrs} Loop over cells that will be transitioned from interface to gas cells, i.e. loop over $\iset_g$, and prepare their neighborhoods for transition:
-	\begin{enumerate}
-	\item For all cells in the neighborhood of the new gas cell:
-		\begin{enumerate}
-			\item If the cell is a fluid cell, transition to interface state.
-		\end{enumerate}
-	\end{enumerate}
-\item \label{enum:rloop-fluid} Loop over cells that will be transitioned from interface to fluid cells, i.e. loop over $\pos \in \iset_f$, and:
-\begin{enumerate}
-	\item Let $M_{ex} = M - \rho$.
-	\item Calculate the unit norm outward from the interface, i.e. toward empty space, $\mathbf{n}$.
-	Initialize $v_{sum} = 0$.
-	\item For all cells in the neighborhood,
-	\begin{enumerate}
-		\item if an interface cell, store in a data structure, $\iset_{r}$, and for the discrete velocity vector direction, $\pvel_i$, of the respective neighbor, let $v_i = \begin{cases}
-					\mathbf{n} \cdot \pvel_i, & \mathbf{n} \cdot \pvel_i > 0 \\
-					0, & \text{otherwise}
-				\end{cases}$.
-		\item $v_{sum} = v_{sum} + v_i$.
-	\end{enumerate}
-	\item If $v_{sum} \ge 0$, then for each cell, $\pos_j \in \iset_r$, redistribute the excess mass according to $M(\pos_j) = M(\pos_j) + M_{ex} \frac{v_i}{v_{sum}}$. \label{enum:redist-dop-f}
-	\item \label{enum:redist-neighbors-f} If $v_{sum} = 0$ and $\iset_r \neq \emptyset$, then for each cell, $\pos_j \in \iset_r$, redistributed the excess mass according to $M(\pos_j) = M(\pos_j) +  M_{ex} / n_r$ where $n_r$ is the number of cells in $\iset_r$.
-	\item Otherwise, increment the global excess mass, $M_{gex} = M_{gex} + M_{ex}$.
-	\item Lastly, transition $\pos$ from $\iset$ to $\fset$ and $M(\pos) = \rho(\pos)$.
-\end{enumerate}
-\item \label{enum:rloop-gas} Loop over cells that will be transitioned from interface to gas cells, i.e. loop over $\pos \in \iset_g$, and:
-\begin{enumerate}
-	\item Let $M_{ex} = M$.
-	\item Calculate the unit norm outward from the interface, i.e. toward empty space, $\mathbf{n}$.
-	Initialize $v_{sum} = 0$.
-	\item For all cells in the neighborhood,
-	\begin{enumerate}
-		\item if an interface cell, store in a data structure, $\iset_{r}$, and for the discrete velocity vector direction, $\pvel_i$, of the respective neighbor, let $v_i = \begin{cases}
-					-\mathbf{n} \cdot \pvel_i, & \mathbf{n} \cdot \pvel_i < 0 \\
-					0, & \text{otherwise}
-				\end{cases}$.
-		\item $v_{sum} = v_{sum} + v_i$.
-	\end{enumerate}
-	\item \label{enum:redist-dop-g} If $v_{sum} \ge 0$ and $\iset_r \neq \emptyset$, then for each cell, $\pos_j \in \iset_r$, redistribute the excess mass according to $M(\pos_j) = M(\pos_j) +  M_{ex} \frac{v_i}{v_{sum}}$.
-	\item \label{enum:redist-neighbors-g} If $v_{sum} = 0$ and $\iset \neq \emptyset$, then for each cell, $\pos_j \in \iset_r$, redistributed the excess mass according to $M(\pos_j) = M(\pos_j) + M_{ex} / n_r$ where $n_r$ is the number of cells in $\iset_r$.
-	\item Otherwise, increment the global excess mass, $M_{gex} = M_{gex} + M_{ex}$.
-	\item Lastly, transition $\pos$ from $\iset$ to $\gset$ and $M(\pos) = 0$.
-\end{enumerate}
-\item \label{enum:final-redist} The global excess mass, i.e. the mass that could not be redistributed by other means, is redistributed according to:
-\begin{enumerate}
-	\item \label{enum:final-rloop} for all $\pos \in \iset$, $M(\pos) = M(\pos) + M_{gex} / n_i$ where $n_i$ is equal to the number of cells in the interface set, $\iset$.
-\end{enumerate}
-\end{enumerate}
-The algorithm, as presented, ensures both mass conservation and continuity of the free-surface.
-Moreover, when mass redistribution inevitably occurs, it is first attempted to be redistributed and weighted in the direction in which the interface is propagating, e.g.
-as in Step \ref{enum:rloop-fluid}.\ref{enum:redist-dop-f} or \ref{enum:rloop-gas}.\ref{enum:redist-dop-g}.
-If there are no interface cells in the direction of interface propagation, it is instead redistributed to neighboring interface cells, e.g.
-Step \ref{enum:rloop-fluid}.\ref{enum:redist-neighbors-f} or \ref{enum:rloop-gas}.\ref{enum:redist-neighbors-g}.
-On the off-chance that there are no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells, e.g.
-Step \ref{enum:final-redist}.
+It is presented here as \Fref{algo:update-cell-states} for the sake of completeness and for outlining minor modifications.
+
+\input{src/algorithm/update-cell-states.tex}
+
+\Fref{algo:update-cell-states}, as presented, ensures both mass conservation and continuity of the free-surface.
+Moreover, when mass redistribution inevitably occurs, it is first attempted to be redistributed and weighted in the direction in which the interface is propagating (e.g.
+line \algref{algo:update-cell-states}{line:redist-dop-f} or line \algref{line:redist-dop-g}.
+If there are no interface cells in the direction of interface propagation, it is instead redistributed to neighboring interface cells (e.g.
+line \algref{line:redist-neighbors-f} or line \algref{line:redist-neighbors-g}).
+If there is no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells (e.g.
+line \algref{line:final-redist}).
 The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...
@@
 
-Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and a still more subtle consideration, how particle distributions that are missing from the streaming step are reconstructed.
+Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and how particle distributions that are missing from the streaming step are reconstructed.
 The degrees of freedom, $f_i$, are not tracked for gas cells.
@@
 
-The missing particle distributions need reconstructed.
-Also momentum needs to be conserved at the interface.
+The missing particle distributions need to be reconstructed.
+Also, momentum needs to be conserved at the interface.
 When the particle distributions are reconstructed, they will be reconstructed in such a way that momentum is conserved between the primary fluid and the secondary fluid.
@@
 
-\subsection{Resulting Algorithm}
-
-The details of the free-surface algorithm have not been presented in chronological order, but instead in a logical order.
-For reasons of clarity, the steps of the algorithm are presented in order:
-\begin{enumerate}
-\item Mass transfer (\Fref{sec:mass-transfer})
-\item Stream
-\item Reconstruct particle distributions (\Fref{sec:bc-at-fs})
-\item Collide
-\item Enforce boundary conditions
-\item Calculate macroscopic variables
-\item Update cell states (\Fref{sec:cell-updates})
-\end{enumerate}
+\subsection{Resulting Algorithm for Simulating Free-Surface Flow using the Lattice Boltzmann Method}
+
+For clarity, the complete algorithm used in the current work for simulating free-surface flow using the Lattice Boltzmann Method is presented in \Fref{algo:free-surface-complete}.
+
+\input{src/algorithm/free-surface-complete.tex}
 
@@
 
-\section{Primary Cementing} \label{sec:results}
+\section{Numerical Study of Primary Cementing} \label{sec:results}
 
 The free-surface LBM model was implemented and used to simulate primary cementing in a dry annulus, i.e. an annulus without drilling mud.
-The focus of the current study was the performance of different cement slurry mixes and characteristic flows when used to cement different imperfect wellbore geometries.
+The focus of the current study was the performance of different characteristic flow behaviors (as could potentially relate to different cement slurry mixes) when used to cement different imperfect wellbore geometries.
 Because the focus in regards to the wellbore is on geometrical imperfections, the entire wellbore annulus is not simulated, but instead a small section around a geometrical imperfection is considered.
-An example is shown in \Fref{fig:wellbore-cementing-schematic}.
+A schematic of an example primary cementing simulation, including boundary conditions, is shown in \Fref{fig:wellbore-cementing-schematic}.
 
@@
 \caption{Example simulation used in primary cementing study.
-The black rectangles on the left boundary represent notches protruding from the rock formation surface. The contours are of the primary fluid (cement slurry) mass.}
+The black rectangles on the left boundary represent notches protruding from the rock formation surface. The contours are of the primary fluid (cement slurry) mass.
+The figure on the right is originally from~\cite{cementing2015web}}
 \label{fig:wellbore-cementing-schematic}
@@
 
-The simulations were carried out on a $\nodesi \times \nodesj$ lattice, which corresponds to an annulus width of \anw \hspace{0.05cm} and a sectional height of \anh.
+%The simulations were carried out on a $\nodesi \times \nodesj$ lattice, which corresponds to an annulus width of \anw \hspace{0.05cm} and a sectional height of \anh.
+The simulations were carried out on a $\nodesi \times \nodesj$ lattice.
 For the mass inlet, a~\citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
@@
 This corresponded to a constant mass flow, $\dot{M}_{in} = M_0 u_0$, at inlet.
-For the outlet, mass was allowed to transfer out of the domain during the mass transfer step such that $f_\opidx(\pos + \pvel_i \Delta t, t) = 0$ in \eqref{eq:dM} for all $\pos + \pvel_i$ that lie outside of the domain on the other side of the mass outlet, i.e.
-the boundary condition at the outlet is such that there is no backflow.
-All simulations were executed for \nsteps time steps.
-
-Both the rock formation geometries and cement slurry flow properties were parameterized.
+For the outlet, mass was allowed to transfer out of the domain during the mass transfer step such that $f_\opidx(\pos + \pvel_i \Delta t, t) = 0$ in \eqref{eq:dM} for all $\pos + \pvel_i \Delta t$ that lie outside of the domain on the other side of the mass outlet (i.e.
+the boundary condition at the outlet is such that there is no backflow).
+All simulations were executed for $\nsteps$ time steps.
+
 The different imperfect geometries that were considered at the rock formation surface were: flow over a cavity that is 30\% of the annulus width, ``cav $30 \times 1$''; flow over a cavity that is 60\% of the annulus width, ``cav $60 \times 1$''; flow over a square obstacle with sides that are 30\% of the annulus width, ``obs $30 \times 1$''; flow over two square obstacles with sides that are 30\% of the annulus width, ``obs $30 \times 2$''; flow over a square obstacle with sides that are 60\% of the annulus width, ``obs $60 \times 1$''; flow over two square obstacles with sides that are 60\% of the annulus width, ``obs $60 \times 2$''; and flow over a rectangular obstacle with width that is 60\% of the annulus width and length that is four times the width, ``obs $60 \times L$''.
-The wellbore geometries are depicted in \Fref{fig:wellbore-geometries}.
+The wellbore geometries that were considered herein are depicted in \Fref{fig:wellbore-geometries}.
 
@@
 
-In addition to wellbore geometry, different cement slurry material properties and cement slurry flow rates were parameterized, which are characterized by the dimensionless Reynold's number, $Re = \frac{\rho U L}{\mu_p}$, and Bingham number, $Bn = \frac{\tau_y L}{\mu_p U}$, of the flow where $L \equiv $ the width of the annulus.
-The goal of the study was to investigate which Reynold's numbers and Bingham numbers were preferable for each geometry, and which Reynold's numbers and Bingham numbers performed well for all geometries.
-The dimensionless numbers were parameterized as follows, $(Re, Bn) = (1.00, 4.00)$,  $(2.50, 0.00)$, $(2.50, 1.00)$, $(2.50, 10.0)$, $(2.50, 25.0)$, $(5.00, 5.00)$, and $(12.5, 2.0)$.
-\Fref{fig:percent-filled} shows the percentage of the \anh \hspace{0.05cm} section wellbore filled after cementing.
+In addition to wellbore geometry, different cement slurry material properties and cement slurry flow rates were considered, which are characterized by the dimensionless Reynold's number, $Re = \frac{\rho U L}{\mu_p}$, and Bingham number, $Bn = \frac{\tau_y L}{\mu_p U}$, of the flow, where $L \equiv $ the width of the annulus.
+A particular goal of the study was to investigate which Reynold's numbers and Bingham numbers were preferable for each geometry, and which Reynold's numbers and Bingham numbers performed well for all geometries.
+The dimensionless numbers were defined as follows, $(Re, Bn) = (1.00, 4.00)$,  $(2.50, 0.00)$, $(2.50, 1.00)$, $(2.50, 10.0)$, $(2.50, 25.0)$, $(5.00, 5.00)$, and $(12.5, 2.0)$.
+\Fref{fig:percent-filled} shows the percentage of the wellbore annulus section filled at the end of the cement simulation duration.
+The percent fill was calculated as:
+\begin{equation}
+  \frac{\sum_{\pos \in \domain} \epsilon(\pos, 16000)}{\sum_{\pos \in \domain} 1.0}.
+\end{equation}
 
@@
 \includegraphics[width=\linewidth]{figs/percent_filled}
-\caption{Wellbore geometries considered in the study.}
+\caption{Percent of voids in wellbore at the end of simulation for each cement slurry flow through each wellbore geometry.}
 \label{fig:percent-filled}
@@
 
-\Fref{fig:percent-filled} shows that there are some general statements to be made about what properties of the cement slurry flow are preferable.
-For example, low yield stress, i.e. low Bingham number flows, had good fill for every wellbore geometry considered.
-In addition, the good performance of the $Re = 5.00, Bn = 5.00$ and $Re = 12.5, Bn = 2.0$ slurry flows suggests that a if the ratio of $Bn / Re$ is low, that the slurry flow will perform well for cementing a dry hole.
-As $Bn / Re$ increases, it appears that, in certain cases and wellbore geometries, the slurry flow tends to perform worse.
-What is interesting to note is that for slurry flows that performed poorly in some cases, there was a lack of consistency across wellbore geometries in terms of which slurry flow performed worse and which slurry flows performed better, which suggests that generalities cannot always be made and that computational methods are necessary for understanding and designing slurry mixes for real-life wellbores.
-
-\section{Conclusion} \label{sec:conclusion}
-
-An LBM model for simulating the non-Newtonnian and free-surface flows that occurs while cementing dry annuli in oil and gas wellbores was implemented and presented.
-It was then used to study how different cement slurry flows, parameterized by different Reynold's numbers and Bingham numbers, would perform in different wellbore geometries.
-It was found that in general, it is preferable that the yield stress and $Bn / Re$ ratio of the slurry flow be low.
+Based upon \Fref{fig:percent-filled}, some general statements can be made about what properties of the cement slurry flow are preferable.
+For example, low yield stress, and consquently low Bingham number flows ($Bn \le 1.0$), approximately completely filled the annulus for every wellbore geometry considered.
+In addition, the negligible percentage of voids for the ($Re = 5.00, Bn = 5.00$) and ($Re = 12.5, Bn = 2.0$) slurry flows suggests that a if the ratio of $Bn / Re$ is low, that the slurry flow will result in a low percentage of voids for cementing a dry hole as is desired.
+As $Bn / Re$ increases, it appears that, in certain cases and wellbore geometries, the slurry flow tends to result in a larger percentage of voids (approximately 10\%-40\%).
+What is interesting to note is that for slurry flows that resulted in a percentage of voids greater than 3\%, there was a lack of consistency across wellbore geometries in terms of which slurry flow resulted in more voids and which slurry flow resulted in less, which suggests that generalities cannot always be made and that computational methods are necessary for understanding and designing slurry mixes for real-life wellbores.
+
+\section{Conclusions} \label{sec:conclusion}
+
+An LBM model for simulating non-Newtonnian and free-surface flows to represent cementing dry annuli in oil and gas wellbores was presented.
+The model was then used to study how different cement slurry flows, defined by different Reynold's numbers and Bingham numbers, would perform in different wellbore geometries.
+The results showed that in general, it is preferable that the yield stress and $Bn / Re$ ratio of the slurry flow be low.
 However, beyond these considerations, it is not clear what slurry flow may be preferable for a given rock formation geometry.
 The results suggest that computational methods are necessary for (a) better understanding the mechanics of primary cementing and (b) for designing a cement mix for a given wellbore.
-It is a goal of the current work to identify a computational method for efficiently simulating cement slurry flow through a realistic wellbore so that optimization methods may be used to determine the optimal cement slurry flow for primary cementing.
 The lattice Boltzmann method extended for simulating non-Newtonian and free-surface flows as presented in the current work has several properties that make it a good candidate for such a design framework:
@@
 \item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components.
-\item LBM is easily written in parallel and as hardware architectures shifts sequential to parallel computing, it becomes increasingly important that code for simulating complex and computationally expense phenomena be written in parallel.
+\item LBM is easily written in parallel and as hardware architectures shifts sequential to parallel computing, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
 \item The LBM no-slip boundary condition works well with complex geometries, such as those that occur in realistic rock formation surfaces in drilled wellbores.

### 2016-07-28 thesis.tex
minor edits after first round of revisions

@@
 \end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
-Determining the apparent viscosity and strain-rate of a flow will often require an iterative solution, a general Picard-type algorithm is:
+Determining the apparent viscosity and strain-rate distribution of a flow over time will often require an iterative solution at every time step, a general Picard-type algorithm for such a process can be given as:
 
@@
 \label{eq:rho} \rho(\pos, t) =& \sum_{i} f_i(\pos, t), \\
-\label{eq:mom} \mmom(\pos, t) = \rho(\pos, t) \mvel(\pos, t) =& \sum_{i} \pvel_i(\pos, t) f_i(\pos, t)
+\label{eq:mom} \mmom(\pos, t) = \rho(\pos, t) \mvel(\pos, t) =& \sum_{i} \pvel_i f_i(\pos, t)
 \end{align}
@@
 
-\subsection{Collision Operator}
+\subsection{Collision Operator} \label{sec:colop}
 
@@
 To illustrate this, consider a flow in which there is a sharp spatial gradient in either $\rho$ or $\mvel$ at $\pos$.
-As $f_i^{eq}$ depends on both $\rho$ and $\mvel$ \eqref(eq:feq), it may be the case that $|f_i^{eq}(\pos, t) - f_i^{eq}(\pos + \pvel_i \Delta t, t + \Delta t)| >> 0$, i.e.
+As $f_i^{eq}$ depends on both $\rho$ and $\mvel$ \eqref{eq:feq}, it may be the case that $|f_i^{eq}(\pos, t) - f_i^{eq}(\pos + \pvel_i \Delta t, t + \Delta t)| >> 0$, i.e.
 there may be a large difference in the quasiequilibrium for the $i^{th}$ discrete velocity at $(\pos, t)$ and $(\pos + \pvel_i \Delta t, t + \Delta t)$.
-In this case, if $f_i$ is ``near'' to quasiequilibrium at $\pos$ it will be ``far'' from quasiequilibrium after the streaming step when it moves to the node at $\pos + \pvel_i \Delta$.
-Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \mvel \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \mvel \Delta t,t + \Delta t)$ but on the ``other side''.
+In this case, if $f_i$ is ``near'' to quasiequilibrium at $\pos$ it will be ``far'' from quasiequilibrium after the streaming step when it moves to the node at $\pos + \pvel_i \Delta t$.
+Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
 Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations such as those depicted in \Fref{fig:overrelaxation}.
@@
 
-\subsection{Boundary Conditions}
+\subsection{Boundary Conditions} \label{sec:bcs}
 
@@
 %\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
-\frac{\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t}{2} (f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
+\frac{\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t}{2} (f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t)), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
 0, & \pos + \pvel_i \Delta t \in \gset
@@
 Moreover, when mass redistribution inevitably occurs, it is first attempted to be redistributed and weighted in the direction in which the interface is propagating (e.g.
-line \algref{algo:update-cell-states}{line:redist-dop-f} or line \algref{line:redist-dop-g}.
+line \algref{algo:update-cell-states}{line:redist-dop-f} or line \algref{algo:update-cell-states}{line:redist-dop-g}.
 If there are no interface cells in the direction of interface propagation, it is instead redistributed to neighboring interface cells (e.g.
-line \algref{line:redist-neighbors-f} or line \algref{line:redist-neighbors-g}).
+line \algref{algo:update-cell-states}{line:redist-neighbors-f} or line \algref{algo:update-cell-states}{line:redist-neighbors-g}).
 If there is no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells (e.g.
-line \algref{line:final-redist}).
+line \algref{algo:update-cell-states}{line:final-redist}).
 The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...
@@
 The black rectangles on the left boundary represent notches protruding from the rock formation surface. The contours are of the primary fluid (cement slurry) mass.
-The figure on the right is originally from~\cite{cementing2015web}}
+The schematic on the right is courtesy of~\cite{cementing2015web}}.
 \label{fig:wellbore-cementing-schematic}
@@
 For the mass inlet, a~\citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
-$M(\pos_i, t) = M_0 = 1.0$ for all $\pos_j$ along the inlet.
+$M(\pos_j, t) = M_0 = 1.0$ for all $\pos_j$ along the inlet.
 This corresponded to a constant mass flow, $\dot{M}_{in} = M_0 u_0$, at inlet.

### 2016-07-29 thesis.tex
formatting and main abstract edits

@@
 
-\title{Evaluating the Capabilities of Lattice Boltzmann Method for non-Newtonian and Free-surface Flows towards Applications in Wellbore Cementing}
+\title{Evaluating the Capabilities of the Lattice Boltzmann Method for non-Newtonian and Free-surface Flow towards Applications in Wellbore Cementing}
 % The optional argument is the version of the title that will appear in Acrobat Reader's Document Info dialog box.
 \author{Matthew Grasinger}
-\degree{BS in Civil and Environmental Engineering, University of Pittsburgh, Pittsburgh, PA 2013}
+\degree{BS in Civil Engineering, University of Pittsburgh, Pittsburgh, PA 2013}
 \date{July 21, 2016}
@@
 % be treated by pittetd as the committee chair (thesis/dissertation advisor).
-\committeemember{John C. Brigham, PhD, Senior Lecturer, School of Engineering and Computing Sciences at Durham University}
+\committeemember{John C. Brigham, PhD, Associate Professor, Department of Civil and Environmental Engineering}
 \committeemember{Julie M. Vandenbossche, PhD, Associate Professor, Department of Civil and Environmental Engineering}
@@
 An important yet vulnerable barrier is the cement annulus.
-Creating the cement annulus, a process known as primary cementing, is difficult to perform optimally.
-Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
-Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
-Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
+Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for cementing all wells.
 The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
@@
 Despite the advantages of LBM, there are considerations that must be made, as with all computational methods, in regards to the accuracy and numerical stability of the solution.
-Issues with accuracy and numerical stability are especially prevalent in non-Newtonian flows because of the nonlinear constitutive relationship.
-\Fref{chap:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LBM models in simulating non-Newtonian flows.
+Issues with accuracy and numerical stability are especially relevant in non-Newtonian flows because of the nonlinear constitutive relationship between shear stress and strain-rate.
+\Fref{chap:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LB methods in simulating non-Newtonian flows.
+The accuracy and computational time are presented in \Fref{sec:numerical-study} for various LB methods applied to two different benchmark problems, Poiseuille flow and lid-driven cavity flow, with two different non-Newtonian constitutive models, the Bingham plastic constitutive relationship and a power-law constitutive relationship.
 
@@
 The extended LBM model was used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
-More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
+More specifically, the study involved defining different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
 The study is followed by conclusions and a discussion of future work.
@@
 
-\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows} \label{chap:lbm-stability}
+\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Fluid Flow} \label{chap:lbm-stability}
 
@@
 
-The Lattice Boltzmann method (LBM) is a method based on computational statistical mechanics that is well-suited for complex flows such as non-Newtonian, free surface, and multiphase multicomponent flows.
-Non-Newtonian flows are the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten metals and plastics, blood flows, etc., are best modeled as non-Newtonian fluids.
+The Lattice Boltzmann method (LBM) is a method based on computational statistical mechanics that is well-suited for complex flow such as non-Newtonian, free surface, and multiphase multicomponent flow.
+Non-Newtonian flow is the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten metals and plastics, blood flows, etc., are best modeled as non-Newtonian fluids.
 LBM is typically applied to simulate flow through a series of time seps, each consisting of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
@@
 constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
-In this paper, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
+In this chapter, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
 The investigation was carried out as a numerical study looking for qualitative, yet practical, results; it included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluid flows.
@@
 
-\section{Numerical Study}
+\section{Numerical Study} \label{sec:numerical-study}
 

### 2016-07-30 thesis.tex
started working on final conclusions

@@
 Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for cementing all wells.
-The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
+The lattice Boltzmann method (LBM) is a promising technique for simulating primary wellbore cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
 
 Despite the advantages of LBM, there are considerations that must be made, as with all computational methods, in regards to the accuracy and numerical stability of the solution.
-Issues with accuracy and numerical stability are especially relevant in non-Newtonian flows because of the nonlinear constitutive relationship between shear stress and strain-rate.
-\Fref{chap:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LB methods in simulating non-Newtonian flows.
+Issues with accuracy and numerical stability are especially relevant in the flow of non-Newtonian fluids because of the nonlinear constitutive relationship between shear stress and strain-rate.
+\Fref{chap:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LB methods in simulating non-Newtonian fluid flow.
 The accuracy and computational time are presented in \Fref{sec:numerical-study} for various LB methods applied to two different benchmark problems, Poiseuille flow and lid-driven cavity flow, with two different non-Newtonian constitutive models, the Bingham plastic constitutive relationship and a power-law constitutive relationship.
 
-Once the ground-work has been laid for simulating non-Newtonian flows using LBM, the algorithm was extended in \Fref{chap:lbm-cement} to incorporate simulation of free-surface flows.
+Once the ground-work has been laid for simulating non-Newtonian flows using LBM, the algorithm was extended in \Fref{chap:lbm-cement} to incorporate simulation of free-surface flow.
 The extended LBM model was used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
@@
 The Lattice Boltzmann method (LBM) is a method based on computational statistical mechanics that is well-suited for complex flow such as non-Newtonian, free surface, and multiphase multicomponent flow.
-Non-Newtonian flow is the primary focus of this paper, as many practical engineering problems such as the flow of cement slurries and concretes, the filling of molds by molten metals and plastics, blood flows, etc., are best modeled as non-Newtonian fluids.
+Non-Newtonian flow is the primary focus of this Chapter, as many practical engineering problems such as the flow of cement slurry and concrete, the filling of molds by molten metals and plastics, blood flow, etc., are best modeled as non-Newtonian fluids.
 LBM is typically applied to simulate flow through a series of time seps, each consisting of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
@@
 constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
-In this chapter, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
-The investigation was carried out as a numerical study looking for qualitative, yet practical, results; it included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluid flows.
+In this Chapter, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
+The investigation was carried out as a numerical study looking for qualitative, yet practical, results; it included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
 Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
-The test results are followed by recommendations for choice collision scheme given the modeler's priorities and problem type.
+The test results are followed by recommendations for choice collision scheme given priority and problem type.
 
@@
 
-Analytical solutions rarely exist for even the simplest non-Newtonian flows because of the complexity that a nonlinear constitutive relationship entails.
-It is generally more practical to approximate non-Newtonian flows using numerical methods. %MG: ask John about the following pargraph...
+Analytical solutions rarely exist for even the simplest non-Newtonian fluid flows because of the complexity that a nonlinear constitutive relationship entails.
+It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods.
 However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
@@
 This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
-Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flows over some other numerical methods.
-
-The lattice Boltzmann method has been studied and successively applied to modeling various non-Newtonian flows.
-For example, ~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flows.
-The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flows which shows the feasibility of using LBM models for yield-stress flows.
-LBM models for power-law fluid flows~\citet{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\citet{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
+Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flow over some other numerical methods.
+
+The lattice Boltzmann method has been studied and successively applied to modeling various flows of non-Newtonian fluids.
+For example, ~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
+The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flow, which shows the feasibility of using LBM models for yield-stress fluids.
+LBM models for power-law fluid flow~\citet{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\citet{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
 %MG: does citing this way appear too dense?
@@
 %Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
-Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian flows because the nonlinear relationship between shear stress and strain-rate can lead to highly nonlinear fluctuations.
+Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian fluids because the nonlinear relationship between shear stress and strain-rate can lead to highly nonlinear fluctuations.
 %MG: be more precise than 'fluctuations'?
-Various schemes and strategies for incorporating the physics of non-Newtonian flows with LBM and yet maintaining a stable numerical method have been developed and studied.
+Various schemes and strategies for incorporating the physics of non-Newtonian flow with LBM and yet maintaining a stable numerical method have been developed and studied.
 The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}, is to make the collision frequency, which is proportional to apparent viscosity, variable and dependent on the local strain rate.
@@
 The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach zero, and will not result in overrelaxation, and the collision frequency will not reach values much greater than one, and so there is no reason to bound the collision frequency in a way that is nonphysical.
-\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flows by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
+\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
 The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum, i.e.
@@
 \citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
-Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
+Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
 \citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
@@
 \begin{itemize}
-    \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flows?
-    \item Under what conditions, e.g. material parameters, physical problem, etc., for Bingham plastic fluid flows is the MRT collision operator necessary to maintain stability and/or accuracy?
-    \item Under what conditions, e.g. material parameters, physical problem, etc, for power-law fluid flows is the MRT collision operator necessary to maintain stability and/or accuracy?
+    \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flow?
+    \item Under what conditions, e.g. material parameters, physical problem, etc., for Bingham plastic fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
+    \item Under what conditions, e.g. material parameters, physical problem, etc, for power-law fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
     \item What are additional strategies for increasing stability and accuracy, and what are their associated computational costs?
@@
 \citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
-Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in non-Newtonian flows.
-
-The goal of this paper is to numerically study the implications of accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flows using LBM.
+Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in simulating the flow of non-Newtonian fluids.
+
+The goal of this Chapter is to numerically study the implications of accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flow using LBM.
 The intention of the study is to aid scientists and engineers in understanding which strategy is best suited to their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
@@
 Despite the utility of the BGK collision operator, it does have a few drawbacks.
-For example, in low viscosity flows the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
+For example, in low viscosity fluids the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
 It is well known that when large nonequilibrium distributions exist in the LBM approximation that overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
@@
 
-The challenge associated with high viscosity flows is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
+The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
 %MG: this needs cleaned up, I think
 
-Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flows because of the nonlinear constitutive relationship between shear stress and strain-rate.
+Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
 The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
 
-Due to the instabilities associated with the collision frequency being too high, e.g.
-approaching 2, or too low, e.g.
-approaching 0, a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flows is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
+Due to the instabilities associated with the collision frequency being too high (e.g.
+approaching 2) or too low (e.g.
+approaching 0), a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flow is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
 This simple methodology for increasing stability will henceforth be referred to bounded-relaxation time BGK, or BGK-BRT.
@@
 
-The MRT collision operator has a greater numerical stability than its BGK counterpart~\cite{lallemand2000theory,dellar2003incompressible,d2002multiple}, and because of the challenges associated with simulating non-Newtonian flow, the MRT collision operator has become popular for simulating non-Newtonian flows~\cite{chen2014simulations,fallah2012multiple,tang2011bingham,vikhansky2008lattice,chai2011multiple,grasinger2015simulation}.
+The MRT collision operator has a greater numerical stability than its BGK counterpart~\cite{lallemand2000theory,dellar2003incompressible,d2002multiple}, and because of the challenges associated with simulating non-Newtonian flow, the MRT collision operator has become popular for simulating non-Newtonian fluids~\cite{chen2014simulations,fallah2012multiple,tang2011bingham,vikhansky2008lattice,chai2011multiple,grasinger2015simulation}.
 The main drawback of the MRT collision operator is its computational expense.
@@
 It has been reported that MRT is approximately 15\% slower than BGK~\cite{d2002multiple}, but this was in the context of Newtonian flow.
-As will be shown later, for certain LBM implementations and non-Newtonian flows the increase in computational expense can be much greater.
+As will be shown later, for certain LBM implementations and non-Newtonian fluid flows the increase in computational expense can be much greater.
 
@@
 Although this achieves the desired result, namely damping out large nonequilibrium fluctuations, it does so in a way that is not smooth or gradual, but sharp.
-An alternative, the median filter, has been used successfully in conjunction with both the BGK and MRT collision schemes for simulating one-dimensional shock tubes and lid-driven cavity flows~\cite{brownlee2008nonequilibrium,gorban2014enhancement,packwood2009entropy}.
+An alternative, the median filter, has been used successfully in conjunction with both the BGK and MRT collision schemes for simulating one-dimensional shock tubes and lid-driven cavity flow~\cite{brownlee2008nonequilibrium,gorban2014enhancement,packwood2009entropy}.
 Median filtering is an effective noise reduction technique, often used in image processing, for ``speckle noise'' or ``salt and pepper noise''~\cite{brownlee2008nonequilibrium}.
@@
 
-The Bingham plastic constitutive model is popular for yield stress flows.
+The Bingham plastic constitutive model is popular for yield stress flow.
 A Bingham plastic does not flow, i.e.
@@
 
-A numerical study was carried out in order to investigate the suitability of different LBM collision schemes for simulating non-Newtonian flows in terms of accuracy, numerical stability and computational efficiency.
+A numerical study was carried out in order to investigate the suitability of different LBM collision schemes for simulating non-Newtonian flow in terms of accuracy, numerical stability and computational efficiency.
 For all simulations presented in this section, the apparent viscosity is approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the maximum number of iterations set at 15 and the convergence criteria:
@@
 In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
-For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$, which follows the recommendation of~\cite{lallemand2000theory} for reasons of stability, and has been successfully used to simulate non-Newtonian flows in the past~\cite{fallah2012multiple,grasinger2015simulation}.
+For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$, which follows the recommendation of~\cite{lallemand2000theory} for reasons of stability, and has been successfully used to simulate non-Newtonian flow in the past~\cite{fallah2012multiple,grasinger2015simulation}.
 
@@
 Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helps mitigate errors that occur as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
-In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress flows (relative errors of approximately 25-75\%).
+In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress fluids (relative errors of approximately 25-75\%).
 
@@
 This can be attributed to the fact that a smoother approximation of the Bingham plastic constitutive model would lead to a solution for $\mu_{app}$ converging with less iterations.
-The BGK collision operator using a stress growth exponent of $m = 10^5$ experienced relatively low error for lower yield stress flows, however, for $\tau_y = 12 \times 10^{-5}$ and $16 \times 10^{-5}$ the relative $L_\infty$ errors were 8.9\% and 19\%, respectively, which is larger than what would be considered acceptable for most engineering applications.
+The BGK collision operator using a stress growth exponent of $m = 10^5$ experienced relatively low error for lower yield stress fluids, however, for $\tau_y = 12 \times 10^{-5}$ and $16 \times 10^{-5}$ the relative $L_\infty$ errors were 8.9\% and 19\%, respectively, which is larger than what would be considered acceptable for most engineering applications.
 
@@
 
-In order to better understand why median filtering did not eliminate the nonphysical oscillations in high yield stress flows, but instead exacerbates the problem, it is necessary to investigate what is happening at the particle distribution scale.
+In order to better understand why median filtering did not eliminate the nonphysical oscillations in high yield stress fluids, but instead exacerbates the problem, it is necessary to investigate what is happening at the particle distribution scale.
 \Fref{fig:feq-vs-f_bgk} compares particle distributions to quasiequilibrium.
@@
 
-For flows with $Re \ge 5000$, the BGK with $m = 10^8$ and median filtering produced numerically stable results.
+For flow with $Re \ge 5000$, the BGK with $m = 10^8$ and median filtering produced numerically stable results.
 However, in general, the BGK with $m = 10^8$ and median filtering was consistently different from literature values. 
@@
 \begin{itemize}
-    \item for low Reynold's number flows the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
-    \item for high Reynold's number flows the BGK collision operator requires entropic filtering to remain stable,
+    \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
+    \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable,
     \item and the MRT collision operator with $m = 10^8$ produces solutions with high accuracy and stability for all of the cases examined herein, but at an increased (approximately 5-10 times more) computational cost.
@@
 
-\section{Conclusions}
-
-A numerical investigation into the accuracy, stability, and efficiency of LBM collision models when applied to non-Newtonian flows was presented.
-The numerical investigation included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluid flows.
-Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
+\section{Conclusions} \label{sec:numerical-study-conclusions}
+
+A numerical investigation into the accuracy, stability, and efficiency of LBM collision models when applied to non-Newtonian flow was presented.
+The numerical investigation included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
+Two different benchmark problems were chosen for the study: Poiseuille flow, and lid-driven square cavity flow.
 The results showed that:
 \begin{itemize}
-  \item For high yield stress fluids in Poiseuille-type flows, only the MRT collision operator did not suffer from nonphysical oscillations.
+  \item For high yield stress fluids in Poiseuille-type flow, only the MRT collision operator did not suffer from nonphysical oscillations.
   \begin{itemize}
       \item The oscillations appeared to be due to high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
-      \item If a collision scheme is to be developed for high yield stress, Poiseuille-type flows that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, then it should focus on dampening the nonequilibrium energy flux moment.
+      \item If a collision scheme is to be developed for high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, then it should focus on dampening the nonequilibrium energy flux moment.
     \end{itemize}
-  \item Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flows; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
+  \item Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
   \item In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
-  \item To summarize Bingham plastic, lid-driven cavity flows,
+  \item To summarize Bingham plastic, lid-driven cavity flow,
     \begin{itemize}
-  \item for low Reynold's number flows the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
-    \item for high Reynold's number flows the BGK collision operator requires entropic filtering to remain stable.
+  \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
+    \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable.
     \end{itemize}
-  \item To summarize power-law, lid-driven cavity flows,
+  \item To summarize power-law, lid-driven cavity flow,
     \begin{itemize}
       \item for shear-thinning fluids at high Reynold's numbers ($Re \ge 5000$) median filtering is necessary to produce a stable solution,
-  \item for shear-thickening, lid-driven cavity flows, there is much less concern for numerical instability.
+  \item for shear-thickening, lid-driven cavity flow, there is much less concern for numerical instability.
     \end{itemize}
@@
 Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
-The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
+The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flow, multiphase multicomponent flow, and flow through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
 In \Fref{sec:numerical-methods}, an algorithm for simulating non-Newtonian free-surface flow using LBM is presented.
@@
 Many fluids present in primary cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
-\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components.
+\item LBM is well-suited for simulating multiphase multicomponent flow and free-surface flow because it can handle complicated interface shapes between fluid phases and components.
 \item LBM is easily written in parallel.
@@
 Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
-The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flows.
+The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flow.
 For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.
@@
 
-An LBM model for simulating non-Newtonnian and free-surface flows to represent cementing dry annuli in oil and gas wellbores was presented.
+An LBM model for simulating non-Newtonnian and free-surface flow to represent cementing dry annuli in oil and gas wellbores was presented.
 The model was then used to study how different cement slurry flows, defined by different Reynold's numbers and Bingham numbers, would perform in different wellbore geometries.
@@
 The results suggest that computational methods are necessary for (a) better understanding the mechanics of primary cementing and (b) for designing a cement mix for a given wellbore.
-The lattice Boltzmann method extended for simulating non-Newtonian and free-surface flows as presented in the current work has several properties that make it a good candidate for such a design framework:
+The lattice Boltzmann method extended for simulating non-Newtonian and free-surface flow as presented in the current work has several properties that make it a good candidate for such a design framework:
 \begin{itemize}
 \item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
-\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components.
+\item LBM is well-suited for simulating multiphase multicomponent flow and free-surface flow because it can handle complicated interface shapes between fluid phases and components.
 \item LBM is easily written in parallel and as hardware architectures shifts sequential to parallel computing, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
@@
 
+\chapter{Conclusions}
+
+Various Lattice Boltzmann methods were investigated for their efficacy in simulating primary cementing.
+Although non-Newtonian fluid mechanics only represent a portion of the challenges associated with simulating primary cementing, the fluid components that can be present in primary cementing exhibit various non-Newtonian behaviors such as yield stress or Bingham plastic behavior (e.g.
+cement slurry), power-law behavior (e.g.
+drilling mud), among other non-Newtonian behaviors.
+In \Fref{chap:lbm-stability}, the accuracy, stability, and efficiency of Lattice Boltzmann models in simulating non-Newtonian fluid flow was investigated.
+Two different benchmark problems were chosen for the study: Poiseuille flow, and lid-driven square cavity flow, as Poiseuille flow is characterized by the development of velocity profiles and lid-driven cavity flow is characterized by the development of vortices.
+Conclusions from the study were presented in \Fref{sec:numerical-study-conclusions}.
+In the context of studying the accuracy, stability, and efficiency of Lattice Boltzmann models for simulating non-Newtonian flow, future work would include investigating the effect of the choice of time step size and simulation Mach number (as in~\cite{conrad2015accuracy}) on the convergence rate and accuracy of the results and LB methods presented in \Fref{sec:numerical-study}.
+In addition, possible future work could include investigating other physical problems such as planar sudden expansion flow~\cite{tang2011binghami,chai2011multiple,chen2014simulations}, reentrant corner flow~\cite{gabbanelli2005lattice},
+
 \bibliographystyle{abbrvnat}

### 2016-07-30 thesis.tex
last adjustments before submitting

@@
 \listoffigures
-\listofalgorithms
+%\listofalgorithms
 
@@
 I would like to thank the best advisor ever, Dr. John Brigham, for his patience and for all that he has taught me. 
-I would also like to thank the other members of my committee, Dr. Julie Vandenbossche, and Dr. Anthony Iannacchione, whom have both been great guides and teachers to me.
+I would also like to thank the other members of my committee, Dr. Julie Vandenbossche, and Dr. Anthony Iannacchione, both of whom have been great guides and teachers to me.
 Lastly, I would like to thank my fellow students, especially Rob, Jing, Mo, and Scott, for helping me to learn along the way and for the comradery that every graduate student needs.
@@
 
+\pagebreak
 The macroscopic variables of interest can be calculated from the particle distribution functions, $f(\pos, \pvel, t)$, by integrating moments of $f$ over velocity space.
@@
 The criteria described in \eqref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead can think in terms of what would be the maximum percentage of sites one would want artificial dissipation to be added to.
-A disadvantage of \eqref{eq:dissipation-critera} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artifical dissipation is unnecessary.
-By required that both \eqref{eq:ds-theshold} and \eqref{eq:dissipation-criteria} be met before artifical dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
+A disadvantage of \eqref{eq:dissipation-criteria} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artifical dissipation is unnecessary.
+By required that both \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} be met before artifical dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
 
@@
 In LBM this is a desirable way to introduce dissipation because it has the potential to reduce high frequency nonequilibrium fluctuations that might lead to numerical instability while retaining the lower frequency dynamics.
+\pagebreak
 To use the median filter one performs the collision step and then checks over the domain for lattice sites with $\Delta S$ that meet the criteria for artificial dissipation; sites that meet the criteria are updated as follows:
@@
 Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helps mitigate errors that occur as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
+\pagebreak
 In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress fluids (relative errors of approximately 25-75\%).
@@
     \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable.
-    \end{itemize}
+    \end{itemize} \pagebreak
   \item To summarize power-law, lid-driven cavity flow,
@@
 The primary fluid is modeled using a standard CFD method, such as LBM, and the interaction of the secondary fluid on the primary fluid is modeled as a boundary condition at the interface between the two fluids.
-The interface is what is referred to as the free-surface.
+The interface is what is referred to as the free-surface. \pagebreak
 The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
@@
 \begin{equation}
-  \frac{\sum_{\pos \in \domain} \epsilon(\pos, 16000)}{\sum_{\pos \in \domain} 1.0}.
+  \frac{\sum\limits_{\pos \in \domain} \epsilon(\pos, 16000)}{\sum\limits_{\pos \in \domain} 1.0}.
 \end{equation}
@@
 In addition, the negligible percentage of voids for the ($Re = 5.00, Bn = 5.00$) and ($Re = 12.5, Bn = 2.0$) slurry flows suggests that a if the ratio of $Bn / Re$ is low, that the slurry flow will result in a low percentage of voids for cementing a dry hole as is desired.
-As $Bn / Re$ increases, it appears that, in certain cases and wellbore geometries, the slurry flow tends to result in a larger percentage of voids (approximately 10\%-40\%).
+As $Bn / Re$ increases, it appears that, in certain cases and wellbore geometries, the slurry flow tends to result in a larger percentage of voids (approximately 10\%-40\%). \pagebreak
 What is interesting to note is that for slurry flows that resulted in a percentage of voids greater than 3\%, there was a lack of consistency across wellbore geometries in terms of which slurry flow resulted in more voids and which slurry flow resulted in less, which suggests that generalities cannot always be made and that computational methods are necessary for understanding and designing slurry mixes for real-life wellbores.
@@
 
-An LBM model for simulating non-Newtonnian and free-surface flow to represent cementing dry annuli in oil and gas wellbores was presented.
+An LBM model for simulating non-Newtonian and free-surface flow to represent cementing dry annuli in oil and gas wellbores was presented.
 The model was then used to study how different cement slurry flows, defined by different Reynold's numbers and Bingham numbers, would perform in different wellbore geometries.
@@
 
+Future work would include developing a more physically realistic model.
+Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
+Possible future work would be make the model more general so as to be capable of simulating multiphase multicomponent flow.
+In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.
+
 \chapter{Conclusions}
@@
 In the context of studying the accuracy, stability, and efficiency of Lattice Boltzmann models for simulating non-Newtonian flow, future work would include investigating the effect of the choice of time step size and simulation Mach number (as in~\cite{conrad2015accuracy}) on the convergence rate and accuracy of the results and LB methods presented in \Fref{sec:numerical-study}.
-In addition, possible future work could include investigating other physical problems such as planar sudden expansion flow~\cite{tang2011binghami,chai2011multiple,chen2014simulations}, reentrant corner flow~\cite{gabbanelli2005lattice},
-
-\bibliographystyle{abbrvnat}
-\bibliography{master}
+In addition, possible future work could include investigating other physical problems such as planar sudden expansion flow~\cite{tang2011binghami,chai2011multiple,chen2014simulations,wang2008lattice}, reentrant corner flow~\cite{gabbanelli2005lattice,yoshino2007numerical}, and flow around a cylinder~\cite{fallah2012multiple}.
+In the context of simulating wellbore cementing, future work would include studying the accuracy, stability, and efficiency of Lattice Boltzmann models for simulating the thixotropic and time-dependent behavior of cement slurry~\cite{roussel2005steady}.
+
+In \Fref{chap:lbm-cement}, the LBM algorithm for non-Newtonian flow is extended to simulate free surface flow.
+The extended LBM model is then used to study wellbore cementing.
+Future work for the wellbore cementing study would include developing a more physically realistic model.
+Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
+Possible future work would be make the model more general so as to be capable of simulating multiphase multicomponent flow.
+In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.
+
+%\bibliographystyle{abbrvnat}
+%\bibliography{master}
+
+% bullshit Pitt formatting
+\bibliographystyle{unsrt}
+\safebibliography{master}
 

### 2016-08-02 thesis.tex
formatting changes for the final submission. fixed bad line breaks

--- 2016-07-30 thesis.tex
@@
 
+\pagebreak
 Analytical solutions rarely exist for even the simplest non-Newtonian fluid flows because of the complexity that a nonlinear constitutive relationship entails.
@@
 The lattice Boltzmann method has been studied and successively applied to modeling various flows of non-Newtonian fluids.
-For example, ~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
+For example, ~\cite{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
 The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flow, which shows the feasibility of using LBM models for yield-stress fluids.
-LBM models for power-law fluid flow~\citet{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\citet{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
+LBM models for power-law fluid flow~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
 %MG: does citing this way appear too dense?
@@
 The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach zero, and will not result in overrelaxation, and the collision frequency will not reach values much greater than one, and so there is no reason to bound the collision frequency in a way that is nonphysical.
-\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
+\cite{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
 The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum, i.e.
@@
 
-To improve upon the stability of LBM models,~\citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
-\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
+To improve upon the stability of LBM models,~\cite{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
+\cite{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
+\pagebreak
 Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
-\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
+\cite{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
 However, what remains unclear is:
@@
 In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
-\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
+\cite{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
 Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in simulating the flow of non-Newtonian fluids.
@@
 
+\pagebreak
 The D2Q9 lattice was used in the current work (shown in \Fref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}. % MG: do we even need this citation?
@@
 approaching 2) or too low (e.g.
-approaching 0), a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flow is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
+approaching 0), a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flow is to simply put bounds on the values in which the collision frequency may attain (such as in~\cite{svec2011flow,svec2012free,gabbanelli2005lattice}).
 This simple methodology for increasing stability will henceforth be referred to bounded-relaxation time BGK, or BGK-BRT.
@@
 The collision is then performed in the moment space.
-There are a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
+There are a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space: \pagebreak
 \begin{enumerate}
@@
 The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
-It is common practice, and~\citet{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.
+It is common practice, and~\cite{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.
 
@@
 A potential drawback of defining a threshold \textit{a priori} is that in order to ensure the model still retains some physical integrity, only a small number of sites can have artificial dissipation added.
+\pagebreak
 If the threshold is too low, too many sites may have dissipation added.
@@
 For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
-For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
+For velocity or pressure boundary conditions, the method proposed by~\cite{zou1997pressure} can be used.
 The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
 
+\pagebreak
 \subsection{Applied Forces}
@@
 Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
-Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\citet{papanastasiou1987flows} is often used as:
+Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\cite{papanastasiou1987flows} is often used as:
 \begin{equation} \label{eq:papa}
@@
 A larger stress growth exponent results in a more accurate Papastasiou approximation of the true Bingham plastic constitutive model, however, it leads to more nonphysical oscillations that degrade the numerical solution for the BGK collision operator.
+\pagebreak
 Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helps mitigate errors that occur as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
@@
 $\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
+\pagebreak
 The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ suffered from was in part due to the $\epsilon$ moment.
@@
 The collision schemes that were tested were: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
-Tables \ref{tab:lid-bing1}--\ref{tab:lid-bing100} compare the center location of the main vortex to literature values taken from~\citet{syrakos2014performance}.
+Tables \ref{tab:lid-bing1}--\ref{tab:lid-bing100} compare the center location of the main vortex to literature values taken from~\cite{syrakos2014performance}.
 The main vortex location is determined by calculating the stream function using Simpson's rule and finding where it attains a maximum.
@@
 \eqref{eq:mrt-colop}, but instead calculating the strain-rate \eqref{eq:srtensor-mrt} for each iteration of the solution for $\mu_{app}$.
+\pagebreak
 
@@
 The collision schemes that were tested were: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT with $m = 10^8$ and (5) MRT with $m = 10^8$ and median filtering.
-\Fref{tab:lid-powerlaw05} and \Fref{tab:lid-powerlaw15} compare the center location of the main vortex to literature values taken from~\citet{li2014simulation}.
+\Fref{tab:lid-powerlaw05} and \Fref{tab:lid-powerlaw15} compare the center location of the main vortex to literature values taken from~\cite{li2014simulation}.
 
@@
 
+\pagebreak
 For the shear-thickening results, $n = 1.5$, there were no issues of stability.
@@
 In order to better understand, and thus improve primary cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
-For example,~\citet{Bei77} developed a mathematical model to describe the displacement of drilling mud by the the cement slurry during primary cementing.
+For example,~\cite{Bei77} developed a mathematical model to describe the displacement of drilling mud by the the cement slurry during primary cementing.
 Among their assumptions were that the flow as laminar, the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the distance from the center of the annulus to either the rock formation or steel casing surface.
@@
 It was concluded that displacement efficiency could be improved by increasing the density of the displacing phase, increasing the viscosity of the displacing phase, (in general) decreasing the displacement rate, decreasing the yield stress of the displaced phase, and increasing the yield stress of the displacing phase.
-\citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
-The model was more general than the model developed by~\citet{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
+\cite{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
+The model was more general than the model developed by~\cite{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
 The bulk fluid motion of the spacer and drilling mud during primary cementing were investigated.
@@
 Results showed that the model was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
-\citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
-The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citet{Bei77} or~\citet{Bit02}.
+\cite{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
+The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\cite{Bei77} or~\cite{Bit02}.
 The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
-\citet{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
+\cite{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
 The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
-\citet{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
-\citet{zhao2016lattice} used the model to investigate the shape of the interface between the displacing phase (i.e.
+\cite{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
+\cite{zhao2016lattice} used the model to investigate the shape of the interface between the displacing phase (i.e.
 cement slurry) and displaced phase (i.e.
@@
 The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features.
+\pagebreak
 The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation.
@@
 The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
-The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice} and \citet{thurey2005interactive}.
+The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\cite{korner2005lattice} and \cite{thurey2005interactive}.
 
@@
 note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells).
-The algorithm for updating cell states follows that which was presented in~\citet{thurey2005interactive}.
+The algorithm for updating cell states follows that which was presented in~\cite{thurey2005interactive}.
 It is presented here as \Fref{algo:update-cell-states} for the sake of completeness and for outlining minor modifications.
@@
 
+\pagebreak
 \Fref{algo:update-cell-states}, as presented, ensures both mass conservation and continuity of the free-surface.
@@
 line \algref{algo:update-cell-states}{line:redist-neighbors-f} or line \algref{algo:update-cell-states}{line:redist-neighbors-g}).
+\pagebreak
 If there is no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells (e.g.
@@
 The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...
-Lastly, note that the interface normal, $\mathbf{n}$, is calculated using a finite difference approximation,
+Lastly, note that the interface normal, $\mathbf{n}$, is calculated using a finite difference approximation:
 \begin{equation} \label{eq:n}
@@
 The simulations were carried out on a $\nodesi \times \nodesj$ lattice.
-For the mass inlet, a~\citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
+For the mass inlet, a~\cite{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
 $M(\pos_j, t) = M_0 = 1.0$ for all $\pos_j$ along the inlet.
@@
 A particular goal of the study was to investigate which Reynold's numbers and Bingham numbers were preferable for each geometry, and which Reynold's numbers and Bingham numbers performed well for all geometries.
+\pagebreak
 The dimensionless numbers were defined as follows, $(Re, Bn) = (1.00, 4.00)$,  $(2.50, 0.00)$, $(2.50, 1.00)$, $(2.50, 10.0)$, $(2.50, 25.0)$, $(5.00, 5.00)$, and $(12.5, 2.0)$.
@@
 
+\pagebreak Future work would include developing a more physically realistic model.
+Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
+Possible future work would be make the model more general so as to be capable of simulating multiphase multicomponent flow.
+In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.
+
 \chapter{Conclusions}
@@
 In the context of studying the accuracy, stability, and efficiency of Lattice Boltzmann models for simulating non-Newtonian flow, future work would include investigating the effect of the choice of time step size and simulation Mach number (as in~\cite{conrad2015accuracy}) on the convergence rate and accuracy of the results and LB methods presented in \Fref{sec:numerical-study}.
-In addition, possible future work could include investigating other physical problems such as planar sudden expansion flow~\cite{tang2011binghami,chai2011multiple,chen2014simulations}, reentrant corner flow~\cite{gabbanelli2005lattice},
-
-\bibliographystyle{abbrvnat}
-\bibliography{master}
+In addition, possible future work could include investigating other physical problems such as planar sudden expansion flow~\cite{tang2011bingham,chai2011multiple,chen2014simulations,wang2008lattice}, reentrant corner flow~\cite{gabbanelli2005lattice,yoshino2007numerical}, and flow around a cylinder~\cite{fallah2012multiple}.
+In the context of simulating wellbore cementing, future work would include studying the accuracy, stability, and efficiency of Lattice Boltzmann models for simulating the thixotropic and time-dependent behavior of cement slurry~\cite{roussel2005steady}.
+
+In \Fref{chap:lbm-cement}, the LBM algorithm for non-Newtonian flow is extended to simulate free surface flow.
+The extended LBM model is then used to study wellbore cementing.
+Future work for the wellbore cementing study would include developing a more physically realistic model.
+Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
+Possible future work would be make the model more general so as to be capable of simulating multiphase multicomponent flow.
+In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.
+
+%\bibliographystyle{abbrvnat}
+
+% bullshit Pitt formatting
+\bibliographystyle{unsrt}
+\safebibliography{master}
+%\bibliography{master}
 

### 2016-08-11 src/stability-paper.tex
breaking thesis into separate papers; addressed Robs comments

--- 2016-07-30 thesis.tex
@@
-% MG: replace \~\cite with ~\cite to ensure citations stay on same line
-% MG: need an overall consistency in 'flucuation' vs. 'oscillation' vs. 'noise'; use same word if you mean same thing.
-% MG: same note as above for 'local equilibrium' vs. 'quasiequilibrium'
-% MG: consider using collision frequency, \omega, instead of relaxation time, \tau, so that the relaxation time is not confused with shear stress
-
-\documentclass[pdftex,ms]{pittetd}
-
-\usepackage[pdftex]{graphicx}
-
-\usepackage[affil-it]{authblk}
-\usepackage{lineno}
-\usepackage{soul}
-\usepackage{color}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage[square,numbers,sort&compress]{natbib}
-\usepackage{enumitem}
-\setlistdepth{9}
-
-\newlist{myEnumerate}{enumerate}{9}
-\setlist[myEnumerate,1]{label=(\arabic*)}
-\setlist[myEnumerate,2]{label=(\Roman*)}
-\setlist[myEnumerate,3]{label=(\Alph*)}
-\setlist[myEnumerate,4]{label=(\roman*)}
-\setlist[myEnumerate,5]{label=(\alph*)}
-\setlist[myEnumerate,6]{label=(\arabic*)}
-\setlist[myEnumerate,7]{label=(\Roman*)}
-\setlist[myEnumerate,8]{label=(\Alph*)}
-\setlist[myEnumerate,9]{label=(\roman*)}
-
-%\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{mathtools}
-\usepackage{bm}
-\usepackage[plain]{fancyref}
-\usepackage{tabulary}
-\usepackage{pbox}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-
-\makeatletter
-\def\BState{\State\hskip-\ALG@thistlm}
-\def\mkfancyprefix#1#2{%
-  \@namedef{fancyref#1labelprefix}{#1}%
-  \begingroup\def\x{\endgroup\frefformat{plain}}%
-    \expandafter\x\csname fancyref#1labelprefix\endcsname
-      {\MakeLowercase{#2}\fancyrefdefaultspacing##1}%
-  \begingroup\def\x{\endgroup\Frefformat{plain}}%
-    \expandafter\x\csname fancyref#1labelprefix\endcsname
-      {#2\fancyrefdefaultspacing##1}%
-}
-\makeatother
-\mkfancyprefix{algo}{Algorithm}
-
-\patch{amsmatch}
-\patch{amsthm}
-
-\newcommand{\pos}{\mathbf{x}}
-\newcommand{\pvel}{\boldsymbol{\xi}}
-\newcommand{\mvel}{\mathbf{u}}
-\newcommand{\colop}{\Omega}
-\newcommand{\grad}{\boldsymbol{\nabla}}
-\newcommand{\mmom}{\mathbf{j}}
-\newcommand{\transM}{\mathbf{M}}
-\newcommand{\relaxM}{\mathbf{S}}
-
-\newcommand{\expnumber}[2]{{#1}\mathrm{e}{#2}}
-
-\title{Evaluating the Capabilities of the Lattice Boltzmann Method for non-Newtonian and Free-surface Flow towards Applications in Wellbore Cementing}
-% The optional argument is the version of the title that will appear in Acrobat Reader's Document Info dialog box.
-\author{Matthew Grasinger}
-\degree{BS in Civil Engineering, University of Pittsburgh, Pittsburgh, PA 2013}
-\date{July 21, 2016}
-\keywords{lattice Boltzmann method, non-Newtonian flow, free surface flow, primary cementing}
-\subject{Draft}
-\school{Swanson School of Engineering}
-\degreesought{Master of Science}
-
-\chapterfloats
-
-\begin{document}
-\maketitle
-
-% For the committee membership page, you have to provide the names and affiliations of the members. The first one will
-% be treated by pittetd as the committee chair (thesis/dissertation advisor).
-\committeemember{John C. Brigham, PhD, Associate Professor, Department of Civil and Environmental Engineering}
-\committeemember{Julie M. Vandenbossche, PhD, Associate Professor, Department of Civil and Environmental Engineering}
-\committeemember{Anthony T. Iannacchione, PhD, Associate Professor, Department of Civil and Environmental Engineering}
-
-\school{Swanson School of Engineering}
-\makecommittee
-\copyrightpage
-
-\begin{abstract}
-
-When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
-Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
-An important yet vulnerable barrier is the cement annulus.
-Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for cementing all wells.
-The lattice Boltzmann method (LBM) is a promising technique for simulating primary wellbore cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
-
-Despite the advantages of LBM, there are considerations that must be made, as with all computational methods, in regards to the accuracy and numerical stability of the solution.
-Issues with accuracy and numerical stability are especially relevant in the flow of non-Newtonian fluids because of the nonlinear constitutive relationship between shear stress and strain-rate.
-\Fref{chap:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LB methods in simulating non-Newtonian fluid flow.
-The accuracy and computational time are presented in \Fref{sec:numerical-study} for various LB methods applied to two different benchmark problems, Poiseuille flow and lid-driven cavity flow, with two different non-Newtonian constitutive models, the Bingham plastic constitutive relationship and a power-law constitutive relationship.
-
-Once the ground-work has been laid for simulating non-Newtonian flows using LBM, the algorithm was extended in \Fref{chap:lbm-cement} to incorporate simulation of free-surface flow.
-The extended LBM model was used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
-More specifically, the study involved defining different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
-The study is followed by conclusions and a discussion of future work.
-
-\end{abstract}
-
-\tableofcontents
-\listoftables
-\listoffigures
-%\listofalgorithms
-
-\preface
-I would like to thank my wife and best friend, Shelby, my parents, and my family.
-Their love, support, and example means everything to me.
-I would like to thank the best advisor ever, Dr. John Brigham, for his patience and for all that he has taught me. 
-I would also like to thank the other members of my committee, Dr. Julie Vandenbossche, and Dr. Anthony Iannacchione, both of whom have been great guides and teachers to me.
-Lastly, I would like to thank my fellow students, especially Rob, Jing, Mo, and Scott, for helping me to learn along the way and for the comradery that every graduate student needs.
-
-\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Fluid Flow} \label{chap:lbm-stability}
-
 \section*{Abstract}
@@
 
-\newcommand{\srtensor}{D_{\alpha\beta}}
-
 \subsection{Strain-rate Tensor}
@@
 All numerical values given in this section are in lattice units unless otherwise stated.
-
-\newcommand{\pgrad}{\frac{\partial p}{\partial x}}
-\newcommand{\unorm}{\mathbf{\hat{n}}}
 
@@
 % MG: should we mention: code is open source? input files used in this study are also freely available?
-
-\chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method} \label{chap:lbm-cement}
-
-\section*{Abstract}
-
-When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
-%Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
-An important yet vulnerable barrier is the cement annulus and is created through a process known as primary cementing.
-Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
-%Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
-Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
-The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flow, multiphase multicomponent flow, and flow through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
-In \Fref{sec:numerical-methods}, an algorithm for simulating non-Newtonian free-surface flow using LBM is presented.
-The algorithm was implemented and used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
-More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
-The study is followed by conclusions and a discussion of future work.
-
-\section{Introduction}
-
-% what is primary cementing?
-
-Of critical importance to drilling operations of all kinds is that barriers are placed to prevent gasses and fluids from migrating from one geological zone to another (i.e., zonal isolation is achieved)~\cite{economides19901,jackson1993effect,goodwin1992cement,dusseault2014towards,li2016theory}.
-If zonal isolation is not achieved, gasses and fluids may migrate through the well to the surface and up into the atmosphere--potentially causing pollution, or at the very least, contributing to an increase in greenhouse gasses--or, worse yet, formation gasses and fluids can migrate into aquifers, potentially harming wild life and people.
-Furthermore, it can be argued that the most important, yet vulnerable, barrier in terms of achieving zonal isolation of a wellbore is the cement annulus that is created between the casing and the formation.
-There are various mechanisms by which this cement annulus can fail.
-According to previous studies~\cite{Bit02,Che85,Lev79,Ste88,Tal93,Zul12,li2016theory}, the most prevalent failure mechanisms of the cement annulus are:
-\begin{itemize}
-\item Stresses developing in the cement annulus as a result of temperature gradients, moisture gradients, overburden pressure, and chemical shrinkage of the cement matrix.
-These stresses can cause cracking in the cement, or debonding at either the casing--cement or cement--formation interfaces.
-\item Poor fill of the annulus.
-If cement slurry does not fill the annulus and if drilling mud is left behind the mud can dry, crack, and provide a weak path for fluids and gasses under pressure to push through, or if a dry hole is cemented and not filled properly, fluids and gasses can travel through channels and voids left in the annular space.
-\item Gas channeling through the annulus during cement curing.
-Initially, after the annulus is poured, the cement column transmits its full hydrostatic pressure against the rock formation.
-As the cement annulus hydrates, it solidifies, becoming stiffer and more self-supporting.
-When the column begins to support itself, the hydrostatic pressure of the cement column--against the rock formation--begins to drop.
-If the hydrostatic pressure of the cement column drops below the pressure of a formation gas before the cement annulus has developed adequate strength to resist it, the formation gas can penetrate into the cement annulus and form channels, or degrade the cement annulus in other ways.
-\end{itemize}
-% what research has been done that is considered classical?
-
-% why do we need modelling?
-% Before the end of this paragraph--modeling is important because...
-
-In order to better understand, and thus improve primary cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
-For example,~\citet{Bei77} developed a mathematical model to describe the displacement of drilling mud by the the cement slurry during primary cementing.
-Among their assumptions were that the flow as laminar, the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the distance from the center of the annulus to either the rock formation or steel casing surface.
-The model was then used to explore the effect of various parameters on the displacement efficiency of the spacer, such as the density ratio between the mud and the spacer, the viscosity ratio between the mud and the spacer, the displacement rate, the displaced phase yield stress, and the displacing phase yield stress.
-It was concluded that displacement efficiency could be improved by increasing the density of the displacing phase, increasing the viscosity of the displacing phase, (in general) decreasing the displacement rate, decreasing the yield stress of the displaced phase, and increasing the yield stress of the displacing phase.
-\citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
-The model was more general than the model developed by~\citet{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
-The bulk fluid motion of the spacer and drilling mud during primary cementing were investigated.
-The intended use of the model was not to make general statements about what properties of the spacer and drilling mud were desirable, but to instead be used iteratively during the design process for the cement job.
-Results showed that the model was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
-\citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
-The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citet{Bei77} or~\citet{Bit02}.
-The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
-\citet{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
-The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
-\citet{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
-\citet{zhao2016lattice} used the model to investigate the shape of the interface between the displacing phase (i.e.
-cement slurry) and displaced phase (i.e.
-drilling mud) in a horizontal, eccentric annulus.
-It was found, in general, that a greater density ratio between the displacing phase and displaced phase and lower eccentricity results in better displacement in a horizontal, eccentric annulus.
-
-% talk about new LBM study
-
-Overall, past studies have tended to idealize the rock formation as a straight wall.
-Although this is a desirable simplification to make for capturing bulk fluid phenomena, it does not accurately describe what is happening locally at the surface of the rock formation.
-Flow near the surface of the formation, and in and around the imperfections of the formation geometry, could influence the bond of the cement annulus to the rock formation, and the fill of the cement slurry very near to the rock formation.
-If an adequate bond does not develop, debonding can occur as a result of thermal and mechanical stresses.
-Even if most of the annulus is filled by the cement slurry, a small gap at the cement--formation interface that stretches for a long enough span of the wellbore would still be susceptible to fluid and gas migration, i.e.
-loss of zonal isolation.
-One could even argue that the characteristic of the cement slurry flow is most important local to the cement--formation interface, and that considering imperfections in the rock formation is imperative to determining whether zonal isolation has been achieved.
-
-The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features.
-The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation.
-LBM also has other advantages as an approach for simulating primary cementing:
-\begin{itemize}
-\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
-Many fluids present in primary cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
-\item LBM is well-suited for simulating multiphase multicomponent flow and free-surface flow because it can handle complicated interface shapes between fluid phases and components.
-\item LBM is easily written in parallel.
-As hardware architectures shift from one or a few cores to several CPUs, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
-\end{itemize}
-The current work focuses on the problem of cementing a dry hole, i.e. an annulus that is not filled with drilling mud prior to cementing.
-In order to numerically investigate this problem, an LBM model for non-Newtonian and free-surface flow is developed.
-The details of the LBM model are outlined in \Fref{sec:numerical-methods}.
-What follows in \Fref{sec:results} is a preliminary study on how different characteristics of flow, e.g.
-cement slurry yield stress, cement slurry inlet velocity, etc., perform in regards to fill of the annulus for different wellbores with imperfect rock formation geometries.
-\Fref{sec:conclusion} outlines conclusions resulting from the numerical study and a discussion of future work.
-
-\section{Numerical Methods} \label{sec:numerical-methods}
-
-\subsection{Lattice Boltzmann Method}
-
-% We need to explain/justify why 2D vs. 3D
-
-For the sake of computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
-Idealizing the annulus as two-dimensional limits the modeling approach in that annulus eccentricity and other three-dimensional phenomena could not be considered.
-A two-dimensional representation of the annulus was deemed sufficient as the focus of this study is to establish the Lattice Boltzmann Method as a suitable method for simulating primary cementing in an annulus with an imperfect rock formation surface.
-In addition to using a D2Q9 lattice, the BGK collision operator used was in the current chapter.
-For more details on the fundamentals of the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.
-
-\subsection{Cement Slurry Constitutive Relationship}
-
-Cement slurry is typically modeled using a non-Newtonian constitutive relationship.
-%A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
-As is commonly utilized for modeling cement and concrete fluid dynamics~\cite{ferraris2001fresh,ferraris1998testing,tattersall2003workability,roussel2007computational,roussel2015numerical,svec2012free}, the Bingham plastic constitutive relationship was utilized herein.
-Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
-The Bingham plastic constitutive model is difficult to work with numerically.
-Therefore, a smooth approximation, specifically the Papastasiou approximation, is often used~\cite{papanastasiou1987flows,chen2014simulations,tang2011bingham,Cha11,wang2008lattice,phillips2011lattice} and is used herein.
-Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
-The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flow.
-For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.
-
-\subsection{Free Surface Flow}
-
-A free-surface flow is a flow in which two-fluids exist but the dynamics of one fluid does not need to be explicitly modeled.
-Typically, the phase of the primary fluid, or the fluid of interest, is a liquid; and the secondary phase that does not need to be explicitly modeled is a gas, often times air at atmospheric pressure.
-The primary fluid is modeled using a standard CFD method, such as LBM, and the interaction of the secondary fluid on the primary fluid is modeled as a boundary condition at the interface between the two fluids.
-The interface is what is referred to as the free-surface. \pagebreak
-The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
-The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice} and \citet{thurey2005interactive}.
-
-\subsection{Capturing the Free-Surface}
-
-Consider at each node a cell of size $\text{d}x \times \text{d}x$ where $\text{d}x$ is the distance between nodes and the cell is centered at the node.
-In order to track where the primary fluid is and where it is not (i.e.
-which cells contain the primary fluid), a variable is introduced, $\epsilon$, which describes what fraction of the cell is covered by the primary fluid.
-Cells that have a fluid fraction of $\epsilon = 0$ contain no primary fluid (i.e.
-are empty); cells that have a fluid fraction of $\epsilon = 1$ are filled; and cells that have a fluid fraction $0 < \epsilon < 1$ are neither completely filled nor are they completely empty.
-The fluid fraction concept is similar to how fluids are tracked in Volume of Fluid (VOF) methods.
-
-\subsubsection{Mass Transfer}
-
-It is important to note that the particle distributions, $f_0, ..., f_8$, cannot be used to directly determine the fluid fraction of a cell.
-Neither can the zeroth velocity moment of the particle distributions, $\rho = \sum_i f_i$.
-For the algorithm used in this work, $\epsilon(\pos, t) \neq \epsilon(\mathbf{f(\pos, t)}, \rho(\pos, t))$.
-Instead, each $f_i$ is related to the amount of primary fluid advection in each of the $\pvel_i$ discrete velocity directions.
-To track the fluid fraction and how it changes, another variable is introduced, $M(\pos, t)$, which denotes the (primary) fluid mass in each cell.
-The fluid mass is related to the fluid fraction by $\epsilon = M / \rho$.
-The change in mass, $\Delta M(\pos, t)$ is related to $\mathbf{f}(\pos, t)$ and particle distribution in the neighborhood of $\pos$.
-Given initial conditions on $M(\pos, 0)$ and a fixed time step size $\Delta t $, the fluid mass is determined by $M(\pos, t_n) = M(\pos, 0) + \Delta t \sum_{k=1}^n \Delta M(\pos, t_k)$.
-What remains to track the fluid mass is to develop an expression for $\Delta M(\pos, t)$.
-
-\newcommand{\opidx}{{\bar{\i}}}
-\newcommand{\fset}{\mathcal{F}}
-\newcommand{\iset}{\mathcal{I}}
-\newcommand{\gset}{\mathcal{G}}
-\newcommand{\domain}{\mathcal{X}}
-
-\subsubsection{Cell States} \label{sec:mass-transfer}
-
-Each of the cells in the domain are assigned one of three states: fluid, interface, or gas.
-The set of fluid, interface, and gas cells are denoted by $\fset$, $\iset$, and $\gset$, respectively.
-A fluid cell is a cell that is entirely covered by the primary fluid; and a gas cell is a cell that is completely empty, or does not contain any primary fluid.
-An interface is partially covered, and there is always a layer of interface cells between fluid and gas cells so that a fluid cell and a gas cell are never neighbors.
-The change in mass is calculated for fluid and interface cells (i.e. $\pos \in \fset \cup \iset$) for each discrete velocity direction as follows:
-\begin{equation} \label{eq:dM}
-\Delta M_i(\pos, t) = \begin{cases}
-f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t), & \pos \in \fset \text{ or } \pos + \pvel_i \Delta t \in \fset \\
-%\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
-\frac{\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t}{2} (f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t)), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
-0, & \pos + \pvel_i \Delta t \in \gset
-\end{cases}
-\end{equation}
-\noindent where $\pvel_\opidx$ represents the discrete velocity vector that is opposite in direction to $\pvel_i$ (i.e. $\pvel_\opidx = - \pvel_i$).
-\eqref{eq:dM} states that when the border of the two cells is covered with fluid, the change in mass for a discrete velocity direction is equal to the difference of the particle distribution entering the cell from the opposite direction and particle distribution leaving the cell in that direction.
-When the border of the two cells are not covered with fluid (i.e.
-when both cells are interface cells) the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the cell border that is covered in fluid.
-The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M_i(\pos, t)$.
-Finally, note that $\Delta M_i(\pos, t) = -\Delta M_\opidx(\pos + \pvel_i \Delta t, t)$ for all $\pos \in \domain$ where $\domain$ is the domain.
-As a consequence, it is clear that $\sum_{\pos \in \domain}\Delta M(\pos, t) = 0$ (i.e. that mass is conserved for the mass transfer step).
-
-\newcommand{\tconst}{\delta_{trans}}
-
-Tracking the transport of mass throughout the domain is an important step in determining what parts of the domain contain the primary fluid and what do not.
-However, for the free-surface to change shape and propagate, cells need to be able to transition state.
-Cell states change based on the rule set presented in \Fref{tab:cell-transition-rules} where $\tconst$ is a small, computational constant that keeps cells from transitioning back and forth between fluid and interface, or interface and gas states in subsequent time steps.
-To summarize \Fref{tab:cell-transition-rules}, gas cells cannot transition to fluid cells and vice versa, interface cells transition to fluid cells when their mass has exceeded $\rho$ by a computational constant, and gas cells when their mass is less than zero minus a computational constant.
-In addition, fluid cells and gas cells transition to the interface state in order to maintain continuity of the interface, i.e.
-to ensure there is always a layer of interface cells in between fluid cells and gas cells.
-An example of a cell transitioning state is depicted in \Fref{fig:cell-transition}.
-
-\begin{table}
-\caption{Cell state transition rules.}
-\begin{tabulary}{\linewidth}{l|L|L}
-\textbf{Current State} & \textbf{Condition} & \textbf{Transition} \\
-\hline & \\
-$(\pos, t) \in \iset$ & $M(\pos, t) < -\tconst$ & $(\pos, t + \Delta t) \rightarrow \gset$, $M(\pos, t + \Delta t) = 0$ \\
-                      & $M(\pos, t) > \tconst + \rho(\pos, t)$ & $(\pos, t + \Delta t) \rightarrow \fset$, $M(\pos, t + \Delta t) = \rho(\pos, t + \Delta t)$ \\
-                      & &\\
-$(\pos, t) \in \fset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \gset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
-& & \\
-$(\pos, t) \in \gset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \fset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
-\end{tabulary}
-\label{tab:cell-transition-rules}
-\end{table}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{figs/cell-transition}
-\caption{Example of cell state transitions.
-The interface cell highlighted in red on the left is transitioned to a gas cell as it has emptied.
-On the right, the three cells highlighted in red were transitioned from fluid to interface cells in order to keep the interface continuous.}
-\label{fig:cell-transition}
-\end{figure}
-
-\subsubsection{Updating Cell States and Mass Redistribution} \label{sec:cell-updates}
-
-The rules for updating cell states in \Fref{tab:cell-transition-rules} are fairly intuitive, however, there are some details that must be considered in order to ensure mass is properly conserved and interface continuity is achieved (e.g.
-note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells).
-The algorithm for updating cell states follows that which was presented in~\citet{thurey2005interactive}.
-It is presented here as \Fref{algo:update-cell-states} for the sake of completeness and for outlining minor modifications.
-
-\input{src/algorithm/update-cell-states.tex}
-
-\Fref{algo:update-cell-states}, as presented, ensures both mass conservation and continuity of the free-surface.
-Moreover, when mass redistribution inevitably occurs, it is first attempted to be redistributed and weighted in the direction in which the interface is propagating (e.g.
-line \algref{algo:update-cell-states}{line:redist-dop-f} or line \algref{algo:update-cell-states}{line:redist-dop-g}.
-If there are no interface cells in the direction of interface propagation, it is instead redistributed to neighboring interface cells (e.g.
-line \algref{algo:update-cell-states}{line:redist-neighbors-f} or line \algref{algo:update-cell-states}{line:redist-neighbors-g}).
-If there is no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells (e.g.
-line \algref{algo:update-cell-states}{line:final-redist}).
-The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...
-Lastly, note that the interface normal, $\mathbf{n}$, is calculated using a finite difference approximation,
-\begin{equation} \label{eq:n}
-\mathbf{n}_{ij} = \frac{1}{2} \begin{Bmatrix}
-\epsilon_{i-1, j} - \epsilon_{i+1, j} \\
-\epsilon_{i, j-1} - \epsilon_{i, j+1}
-\end{Bmatrix}
-\end{equation}
-\noindent where $i$ is the index of the node in the x-direction and $j$ is the index of the node in the y-direction.
-
-\subsection{Boundary Conditions at the Free-Surface} \label{sec:bc-at-fs}
-
-Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and how particle distributions that are missing from the streaming step are reconstructed.
-The degrees of freedom, $f_i$, are not tracked for gas cells.
-After the streaming step, interface cells will be missing particle distributions that would have streamed from gas cells--an issue that is depicted in \Fref{fig:ps}.
-
-\begin{figure}
-\centering
-\includegraphics{figs/ps}
-\caption{Example of a scenario in which particle distributions are missing post-streaming.}
-\label{fig:ps}
-\end{figure}
-
-The missing particle distributions need to be reconstructed.
-Also, momentum needs to be conserved at the interface.
-When the particle distributions are reconstructed, they will be reconstructed in such a way that momentum is conserved between the primary fluid and the secondary fluid.
-In order to perform the reconstruction, there are a few assumptions that need to be made:
-\begin{enumerate}
-\item The velocity of the secondary fluid is equal to the velocity of the primary fluid at the free-surface, i.e. there is no-slip between the two-fluids.
-\item The secondary fluid is at equilibrium and at constant pressure, $p_G$.
-\item There is a force balance for opposite lattice directions.
-\end{enumerate}
-Because particle distributions are a measure of momentum transport, the pressure of the secondary fluid can be converted to the particle distribution scale and the missing particle distribution can be solved for algebraically.
-Consider \Fref{fig:fbi}.
-Balancing the opposite lattice directions results in:
-\begin{eqnarray}
-f_i + f_\opidx &=& f^{eq}_\opidx(\rho_G, \mathbf{u}) + f_i^{eq}(\rho_G, \mathbf{u}), \\
-\label{eq:reconst} f_\opidx &=& f_\opidx^{eq}(\rho_G, \mathbf{u}) + f_i^{eq}(\rho_G, \mathbf{u}) - f_i.
-\end{eqnarray}
-Missing distribution functions can be reconstructed according to \eqref{eq:reconst}.
-In order for conservation of momentum to be satisfied along the free-surface, distribution functions coming from the direction of the interface normal, $\mathbf{n}$, must also be reconstructed.
-Thus, distribution functions are reconstructed if coming from a gas cell, or if $\mathbf{n} \cdot \pvel_\opidx > 0$.
-
-\begin{figure}
-\centering
-\includegraphics{figs/force-bal-at-interface}
-\caption{Particle distributions at free-surface for a pair of opposing lattice directions after gas pressure, $\rho_G$, is in the particle distribution form.}
-\label{fig:fbi}
-\end{figure}
-
-\subsection{Resulting Algorithm for Simulating Free-Surface Flow using the Lattice Boltzmann Method}
-
-For clarity, the complete algorithm used in the current work for simulating free-surface flow using the Lattice Boltzmann Method is presented in \Fref{algo:free-surface-complete}.
-
-\input{src/algorithm/free-surface-complete.tex}
-
-\newcommand{\nodesi}{50}
-\newcommand{\nodesj}{500}
-\newcommand{\anw}{$1\frac{1}{4}$"}
-\newcommand{\anh}{1'}
-\newcommand{\nsteps}{16000}
-
-\section{Numerical Study of Primary Cementing} \label{sec:results}
-
-The free-surface LBM model was implemented and used to simulate primary cementing in a dry annulus, i.e. an annulus without drilling mud.
-The focus of the current study was the performance of different characteristic flow behaviors (as could potentially relate to different cement slurry mixes) when used to cement different imperfect wellbore geometries.
-Because the focus in regards to the wellbore is on geometrical imperfections, the entire wellbore annulus is not simulated, but instead a small section around a geometrical imperfection is considered.
-A schematic of an example primary cementing simulation, including boundary conditions, is shown in \Fref{fig:wellbore-cementing-schematic}.
-
-\begin{figure}
-\centering
-\includegraphics{figs/wellbore-cementing-schematic}
-\caption{Example simulation used in primary cementing study.
-The black rectangles on the left boundary represent notches protruding from the rock formation surface. The contours are of the primary fluid (cement slurry) mass.
-The schematic on the right is courtesy of~\cite{cementing2015web}}.
-\label{fig:wellbore-cementing-schematic}
-\end{figure}
-
-%The simulations were carried out on a $\nodesi \times \nodesj$ lattice, which corresponds to an annulus width of \anw \hspace{0.05cm} and a sectional height of \anh.
-The simulations were carried out on a $\nodesi \times \nodesj$ lattice.
-For the mass inlet, a~\citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
-$M(\pos_j, t) = M_0 = 1.0$ for all $\pos_j$ along the inlet.
-This corresponded to a constant mass flow, $\dot{M}_{in} = M_0 u_0$, at inlet.
-For the outlet, mass was allowed to transfer out of the domain during the mass transfer step such that $f_\opidx(\pos + \pvel_i \Delta t, t) = 0$ in \eqref{eq:dM} for all $\pos + \pvel_i \Delta t$ that lie outside of the domain on the other side of the mass outlet (i.e.
-the boundary condition at the outlet is such that there is no backflow).
-All simulations were executed for $\nsteps$ time steps.
-
-The different imperfect geometries that were considered at the rock formation surface were: flow over a cavity that is 30\% of the annulus width, ``cav $30 \times 1$''; flow over a cavity that is 60\% of the annulus width, ``cav $60 \times 1$''; flow over a square obstacle with sides that are 30\% of the annulus width, ``obs $30 \times 1$''; flow over two square obstacles with sides that are 30\% of the annulus width, ``obs $30 \times 2$''; flow over a square obstacle with sides that are 60\% of the annulus width, ``obs $60 \times 1$''; flow over two square obstacles with sides that are 60\% of the annulus width, ``obs $60 \times 2$''; and flow over a rectangular obstacle with width that is 60\% of the annulus width and length that is four times the width, ``obs $60 \times L$''.
-The wellbore geometries that were considered herein are depicted in \Fref{fig:wellbore-geometries}.
-
-\begin{figure}
-\includegraphics[width=\linewidth]{figs/wellbore-geometries}
-\caption{Wellbore geometries considered in the study.}
-\label{fig:wellbore-geometries}
-\end{figure}
-
-In addition to wellbore geometry, different cement slurry material properties and cement slurry flow rates were considered, which are characterized by the dimensionless Reynold's number, $Re = \frac{\rho U L}{\mu_p}$, and Bingham number, $Bn = \frac{\tau_y L}{\mu_p U}$, of the flow, where $L \equiv $ the width of the annulus.
-A particular goal of the study was to investigate which Reynold's numbers and Bingham numbers were preferable for each geometry, and which Reynold's numbers and Bingham numbers performed well for all geometries.
-The dimensionless numbers were defined as follows, $(Re, Bn) = (1.00, 4.00)$,  $(2.50, 0.00)$, $(2.50, 1.00)$, $(2.50, 10.0)$, $(2.50, 25.0)$, $(5.00, 5.00)$, and $(12.5, 2.0)$.
-\Fref{fig:percent-filled} shows the percentage of the wellbore annulus section filled at the end of the cement simulation duration.
-The percent fill was calculated as:
-\begin{equation}
-  \frac{\sum\limits_{\pos \in \domain} \epsilon(\pos, 16000)}{\sum\limits_{\pos \in \domain} 1.0}.
-\end{equation}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{figs/percent_filled}
-\caption{Percent of voids in wellbore at the end of simulation for each cement slurry flow through each wellbore geometry.}
-\label{fig:percent-filled}
-\end{figure}
-
-Based upon \Fref{fig:percent-filled}, some general statements can be made about what properties of the cement slurry flow are preferable.
-For example, low yield stress, and consquently low Bingham number flows ($Bn \le 1.0$), approximately completely filled the annulus for every wellbore geometry considered.
-In addition, the negligible percentage of voids for the ($Re = 5.00, Bn = 5.00$) and ($Re = 12.5, Bn = 2.0$) slurry flows suggests that a if the ratio of $Bn / Re$ is low, that the slurry flow will result in a low percentage of voids for cementing a dry hole as is desired.
-As $Bn / Re$ increases, it appears that, in certain cases and wellbore geometries, the slurry flow tends to result in a larger percentage of voids (approximately 10\%-40\%). \pagebreak
-What is interesting to note is that for slurry flows that resulted in a percentage of voids greater than 3\%, there was a lack of consistency across wellbore geometries in terms of which slurry flow resulted in more voids and which slurry flow resulted in less, which suggests that generalities cannot always be made and that computational methods are necessary for understanding and designing slurry mixes for real-life wellbores.
-
-\section{Conclusions} \label{sec:conclusion}
-
-An LBM model for simulating non-Newtonian and free-surface flow to represent cementing dry annuli in oil and gas wellbores was presented.
-The model was then used to study how different cement slurry flows, defined by different Reynold's numbers and Bingham numbers, would perform in different wellbore geometries.
-The results showed that in general, it is preferable that the yield stress and $Bn / Re$ ratio of the slurry flow be low.
-However, beyond these considerations, it is not clear what slurry flow may be preferable for a given rock formation geometry.
-The results suggest that computational methods are necessary for (a) better understanding the mechanics of primary cementing and (b) for designing a cement mix for a given wellbore.
-The lattice Boltzmann method extended for simulating non-Newtonian and free-surface flow as presented in the current work has several properties that make it a good candidate for such a design framework:
-\begin{itemize}
-\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
-\item LBM is well-suited for simulating multiphase multicomponent flow and free-surface flow because it can handle complicated interface shapes between fluid phases and components.
-\item LBM is easily written in parallel and as hardware architectures shifts sequential to parallel computing, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
-\item The LBM no-slip boundary condition works well with complex geometries, such as those that occur in realistic rock formation surfaces in drilled wellbores.
-\end{itemize}
-
-Future work would include developing a more physically realistic model.
-Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
-Possible future work would be make the model more general so as to be capable of simulating multiphase multicomponent flow.
-In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.
-
-\chapter{Conclusions}
-
-Various Lattice Boltzmann methods were investigated for their efficacy in simulating primary cementing.
-Although non-Newtonian fluid mechanics only represent a portion of the challenges associated with simulating primary cementing, the fluid components that can be present in primary cementing exhibit various non-Newtonian behaviors such as yield stress or Bingham plastic behavior (e.g.
-cement slurry), power-law behavior (e.g.
-drilling mud), among other non-Newtonian behaviors.
-In \Fref{chap:lbm-stability}, the accuracy, stability, and efficiency of Lattice Boltzmann models in simulating non-Newtonian fluid flow was investigated.
-Two different benchmark problems were chosen for the study: Poiseuille flow, and lid-driven square cavity flow, as Poiseuille flow is characterized by the development of velocity profiles and lid-driven cavity flow is characterized by the development of vortices.
-Conclusions from the study were presented in \Fref{sec:numerical-study-conclusions}.
-In the context of studying the accuracy, stability, and efficiency of Lattice Boltzmann models for simulating non-Newtonian flow, future work would include investigating the effect of the choice of time step size and simulation Mach number (as in~\cite{conrad2015accuracy}) on the convergence rate and accuracy of the results and LB methods presented in \Fref{sec:numerical-study}.
-In addition, possible future work could include investigating other physical problems such as planar sudden expansion flow~\cite{tang2011binghami,chai2011multiple,chen2014simulations,wang2008lattice}, reentrant corner flow~\cite{gabbanelli2005lattice,yoshino2007numerical}, and flow around a cylinder~\cite{fallah2012multiple}.
-In the context of simulating wellbore cementing, future work would include studying the accuracy, stability, and efficiency of Lattice Boltzmann models for simulating the thixotropic and time-dependent behavior of cement slurry~\cite{roussel2005steady}.
-
-In \Fref{chap:lbm-cement}, the LBM algorithm for non-Newtonian flow is extended to simulate free surface flow.
-The extended LBM model is then used to study wellbore cementing.
-Future work for the wellbore cementing study would include developing a more physically realistic model.
-Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
-Possible future work would be make the model more general so as to be capable of simulating multiphase multicomponent flow.
-In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.
-
-%\bibliographystyle{abbrvnat}
-%\bibliography{master}
-
-% bullshit Pitt formatting
-\bibliographystyle{unsrt}
-\safebibliography{master}
-
-\end{document}

### 2016-08-11 thesis.tex
breaking thesis into separate papers; addressed Robs comments

--- 2016-07-30 thesis.tex
@@
 
-\newcommand{\pos}{\mathbf{x}}
-\newcommand{\pvel}{\boldsymbol{\xi}}
-\newcommand{\mvel}{\mathbf{u}}
-\newcommand{\colop}{\Omega}
-\newcommand{\grad}{\boldsymbol{\nabla}}
-\newcommand{\mmom}{\mathbf{j}}
-\newcommand{\transM}{\mathbf{M}}
-\newcommand{\relaxM}{\mathbf{S}}
-
-\newcommand{\expnumber}[2]{{#1}\mathrm{e}{#2}}
+\input{src/notation.tex}
 
@@
 \keywords{lattice Boltzmann method, non-Newtonian flow, free surface flow, primary cementing}
-\subject{Draft}
+\subject{Final}
 \school{Swanson School of Engineering}
@@
 
-\section*{Abstract}
-
-The Lattice Boltzmann method (LBM) is a method based on computational statistical mechanics that is well-suited for complex flow such as non-Newtonian, free surface, and multiphase multicomponent flow.
-Non-Newtonian flow is the primary focus of this Chapter, as many practical engineering problems such as the flow of cement slurry and concrete, the filling of molds by molten metals and plastics, blood flow, etc., are best modeled as non-Newtonian fluids.
-LBM is typically applied to simulate flow through a series of time seps, each consisting of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
-The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated, e.g.
-constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
-In this Chapter, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
-The investigation was carried out as a numerical study looking for qualitative, yet practical, results; it included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
-Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
-The test results are followed by recommendations for choice collision scheme given priority and problem type.
-
-\section{Introduction} %MG: this intro will potentially be long. Should it be broken up into subsections (as it is now)?
-
-The dynamic viscosity of a fluid is a measure of its resistance to shear deformation.
-For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
-These fluids are known as Newtonian fluids.
-Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable even at a constant temperature and is often a function of strain-rate.
-There are a number of fluids in science and engineering applications that can be classified as non-Newtonian: pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.
-
-Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized: yield stress fluids and power-law fluids.
-Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
-Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g.
-pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
-%MG: citation is for blood as non-Newtonian
-As a consequence, simulating yield stress fluids can help engineers develop better pastes and paints, understand the flow and deformation of mud and clay in geotechnical engineering, understand blood circulation, and better manufacture metal and plastic parts.
-
-There other idealized non-Newtonian behavir, power-law behavior, is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate.
-Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include polymer mixes and molten plastics.
-Shear-thickening fluids are also known as \emph{dilatants}, and some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
-
-\label{sec:lbm-for-nnf}
-
-Analytical solutions rarely exist for even the simplest non-Newtonian fluid flows because of the complexity that a nonlinear constitutive relationship entails.
-It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods.
-However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
-\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma},
-\end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
-Determining the apparent viscosity and strain-rate distribution of a flow over time will often require an iterative solution at every time step, a general Picard-type algorithm for such a process can be given as:
-
-\begin{enumerate}
-	\item Start with initial guess for the apparent viscosity, $\mu_k = \mu_0$.
-	\item \label{step:solve-for-flow} Solve for the flow using the current value of apparent viscosity, $\dot{\gamma}_k = \tau_k / \mu_k$.
-	\item Update the apparent viscosity, $\mu_{k+1} = \mu(\dot{\gamma}_k)$.
-	\item Return to Step \ref{step:solve-for-flow} until convergence is met.
-\end{enumerate}
-
-Numerical solutions work by discretizing the equations that govern the physics of interest.
-When the solution for fluid flow problems vary in space and time, a numerical approximation requires breaking the problem up into discrete locations and discrete time steps.
-The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which can become computationally expensive.
-The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the advantage that computing the strain rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
-This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
-Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flow over some other numerical methods.
-
-The lattice Boltzmann method has been studied and successively applied to modeling various flows of non-Newtonian fluids.
-For example, ~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
-The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flow, which shows the feasibility of using LBM models for yield-stress fluids.
-LBM models for power-law fluid flow~\citet{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\citet{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
-%MG: does citing this way appear too dense?
-
-%MG: probably include a discussion of the work done by Conrad et. al. 2015 here
-LBM does however, have its drawbacks.
-LBM can be considered as a type of finite-difference scheme for the continuous Boltzmann equation, and as such, has numerical properties in common with finite-difference schemes.
-One such consideration associated with this view is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
-%MG: are all relevant sources cited?
-%Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
-Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian fluids because the nonlinear relationship between shear stress and strain-rate can lead to highly nonlinear fluctuations.
-%MG: be more precise than 'fluctuations'?
-Various schemes and strategies for incorporating the physics of non-Newtonian flow with LBM and yet maintaining a stable numerical method have been developed and studied.
-The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}, is to make the collision frequency, which is proportional to apparent viscosity, variable and dependent on the local strain rate.
-A potential issue with the stability of the variable relaxation time approach is that as the collision frequency approaches $2$ the viscosity approaches zero and overrelaxation occurs; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
-%MG: verify this statement
-In order to ensure that the variable collision frequencies did not approach values leading to numerical instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable collision frequencies.
-Although bounding the collision frequency was shown to be effective in terms of stability, it is also nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the collision does not reflect the proper constitutive relationship of interest.
-Another scheme for incorporating non-Newtonian effects into LBM is to use a constant collision frequency, typically unity, and to instead incorporate the local shear-rate effect through equilibrium distribution functions.
-This means particle distribution functions will always relax toward equilibrium at the same rate, but that the definition of equilibrium is modified to represent the correct stress--strain-rate relationship.
-The equilibrium distribution function is derived for the specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.
-The equilibrium distribution functions for Bingham plastic fluids, and for power-law and Carreau fluids were derived, implemented, and verified in~\cite{wang2008lattice} and~\cite{yoshino2007numerical}, respectively.
-The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach zero, and will not result in overrelaxation, and the collision frequency will not reach values much greater than one, and so there is no reason to bound the collision frequency in a way that is nonphysical.
-\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
-The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum, i.e.
-as an external forcing term, that is dependent on local shear-rate.
-Although the constant-collision frequency strategies present interesting alternatives, the variable collision scheme is used in the present study because of its simplicity and generality, i.e.
-it can be easily fitted to any constitutive equation without for example, performing Chapman-Enskog multiscale expansion.
-
-To improve upon the stability of LBM models,~\citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
-\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
-Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
-\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
-However, what remains unclear is:
-\begin{itemize}
-    \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flow?
-    \item Under what conditions, e.g. material parameters, physical problem, etc., for Bingham plastic fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
-    \item Under what conditions, e.g. material parameters, physical problem, etc, for power-law fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
-    \item What are additional strategies for increasing stability and accuracy, and what are their associated computational costs?
-\end{itemize}
-In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
-\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
-Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in simulating the flow of non-Newtonian fluids.
-
-The goal of this Chapter is to numerically study the implications of accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flow using LBM.
-The intention of the study is to aid scientists and engineers in understanding which strategy is best suited to their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
-Advantages, such as LBM's potential to scale well in parallel, can be much less realized if the collision operator is too computationally expensive, or if numerical instabilities ensue.
-A numerical study can help to determine approximate numerical values, domains, and boundary conditions in which one LBM scheme may be more advantageous than another so that LBM may be used in a computationally efficient and stable manner.
-
-%MG: is the beginning of this methods section too much fluff now? Would our audiance consider it fluff?
-
-\section{Lattice Boltzmann Method for Simulation of Non-Newtonian Fluids} \label{sec:LBM}
-
-The Lattice Boltzmann method is a numerical approach that uses statistical mechanics to represent a variety of physical processes, such as fluid flow.
-More specifically, LBM can be thought of as a special finite difference discretization of the Boltzmann equation~\cite{chen1998lattice}.
-The length scale of LBM is unique in contrast to most common numerical methods, and is referred to as the mesoscale.
-In contrast to continuum based methods, LBM simulates the kinetics of microscopic particles, and so it reaches a finer length scale than the macroscopic domain of continuum mechanics; and in contrast to molecular dynamics, discrete element method, and other particle scale approaches, LBM does not deal with a complete description of the degrees of freedom for each individual particle.
-LBM instead relies on a statistical description of particle distributions, making LBM, in general, more computationally efficient and requiring less memory than other particle methods.
-Thus, LBM can be seen as a compromise between continuum and particle methods, combining strengths from each.
-
-LBM has some advantages over other methods of CFD.
-For example, LBM is a computationally efficient approach for some CFD.
-This efficiency is a consequence of two distinct features of LBM: (1) the convective operator is linear, as opposed to the nonlinear convection terms that appear in continuum mechanics approaches; and (2) the fluid pressure is given by an equation of state.
-Solving for the fluid pressure in traditional method is more computationally expensive and requires special treatment such as iteration and/or relaxation~\cite{chen1998lattice}.
-
-\subsection{The Boltzmann Equation}
-The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
-The BE is given as:
-\begin{equation}
-\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop,
-\end{equation}
-\noindent where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, and $\colop = \colop(f)$ is the collision operator.
-The lattice part of LBM refers to the way in which the BE is discretized.
-The lattice discretizes the spatial domain with nodes that are connected to their neighbors through discrete lattice velocity vectors.
-The velocity vectors act as pathways for particle distributions to travel along.
-Each time step in LBM consists of two distinct actions:
-
-\begin{itemize}
-\item Streaming: particle distributions propagate to their neighbors along the lattice velocity vectors.
-The particles can only move along the vectors in their specified direction and can only move at a specific speed.
-\item Collision: particle distributions meet at a node and ``collide''.
-In LBM, collisions are not simulated in a realistic sense, meaning that each individual particle does not exist and glance off of, or interact with, one another.
-Instead the collision operator is formulated in such a way that particle distributions are relaxed toward equilibrium.
-What defines equilibrium depends on the mechanics of interest to be modeled.
-\end{itemize}
-
-The D2Q9 lattice was used in the current work (shown in \Fref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}. % MG: do we even need this citation?
-The lattice is two-dimensional with nine discrete velocities at each node.
-There is a stationary particle, there are four discrete velocities of magnitude 1, $\begin{Bmatrix}1 & 0\end{Bmatrix}^T$, $\begin{Bmatrix}0 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & 0\end{Bmatrix}^T$, $\begin{Bmatrix}0 & -1\end{Bmatrix}^T$, and there are four discrete velocities of magnitude $\sqrt{2}$, $\begin{Bmatrix}1 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & -1\end{Bmatrix}^T$, $\begin{Bmatrix}1 & -1\end{Bmatrix}^T$.
-The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as:
-\begin{equation}
-f_i(\pos + \pvel_i \Delta t, t + \Delta t) = f_i(\pos, t) + \colop_i(\pos, t),
-\end{equation}
-\noindent where, for D2Q9, $i = 0, 1, ..., 8$ is the index of the discrete velocity vector.
-\begin{figure}
-  \centering
-  \includegraphics[width=10cm]{figs/d2q9}
-  \caption{Schematic of D2Q9 lattice, each node is connected to its neighbor by one of eight discrete velocity vectors}
-  \label{fig:d2q9}
-\end{figure}
-
-\pagebreak
-The macroscopic variables of interest can be calculated from the particle distribution functions, $f(\pos, \pvel, t)$, by integrating moments of $f$ over velocity space.
-Due to the discrete nature of velocity in LBM, the integrals simply become summations.
-The mass density is given by the sum of the particle distributions and the momentum density is given by the first moment of the particle distributions over the velocity space:
-
-\begin{align}
-\label{eq:rho} \rho(\pos, t) =& \sum_{i} f_i(\pos, t), \\
-\label{eq:mom} \mmom(\pos, t) = \rho(\pos, t) \mvel(\pos, t) =& \sum_{i} \pvel_i f_i(\pos, t)
-\end{align}
-
-\noindent The fluid pressure is related to macroscopic density through an equation of state:
-
-\begin{equation}
-\label{eq:pres} p(\pos, t) = \rho(\pos, t) c_s^2,
-\end{equation}
-
-\noindent where $c_s$ is the lattice speed of sound ($c_s = \frac{1}{\sqrt{3}}$, for D2Q9).
-
-\subsection{Collision Operator} \label{sec:colop}
-
-\subsubsection{Bhatnagar-Gross-Krook (BGK)} \label{sec:bgk}
-
-The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision)~\cite{Cer90}.
-In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
-This equilibrium is determined by the macroscopic physical behavior of interest.
-In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq} = f_i^{eq}(\pos, t)$, is often given by:
-
-\begin{equation} \label{eq:feq}
-  f_i^{eq} = w_i \rho \left[1 + \frac{\pvel_i \cdot \mvel}{c_s^2} + \frac{(\pvel_i \cdot \mvel)^2}{4c_s^2} - \frac{\mvel^2}{2c_s^2} \right],
-\end{equation}
-
-\noindent where $\rho$ and $\mvel$ are dependent on $\pos$ and $t$, and $w_i$ is the weight in the $i^{th}$:%MG: maybe say something about where this comes from, e.g.
-feq is a Hermite polynomial expansion of the Maxwell-Boltzmann distribution
-
-\begin{equation} \label{eq:weights}
-w_i = \begin{cases}
-    \frac{4}{9}, & i = 0 \\
-    \frac{1}{9}, & i = 1, 2, 3, 4 \\
-    \frac{1}{36}, & i = 5, 6, 7, 8
-\end{cases}.
-\end{equation}
-
-Due to its simplicity and computational efficiency, the most common collision operator is the Bhatnagar--Gross--Krook (BGK) operator.
-BGK consists of a single relaxation time and is a linear relaxation of particle distributions toward equilibrium.
-The BGK collison operator for the $i^{th}$ discrete velocity is expressed as:
-
-\begin{equation} \label{eq:bgk}
-\Omega_i(\pos, t) = -\omega (f_i(\pos, t) - f_i^{eq}(\pos, t)),
-\end{equation}
-
-\noindent where $\omega$ is the collision frequency~\cite{Bha54}.
-The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis~\cite{wolf2000lattice}.
-For incompressible Newtonian flow, the collision frequency is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$; and from this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.
-The method for simulating non-Newtonian flow in the current work involves approximating a solution to the local value of apparent viscosity, $\mu_{app}(\pos, t)$, using \eqref{eq:non-newtonian} where $\tau$ and $\dot{\gamma}$ are dependent on $\pos$ and $t$, and then setting the value of the local collision frequency as follows:
-\begin{equation} \label{eq:colfeq-non-newtonian}
-\omega(\pos, t) = \frac{1}{\frac{\mu_{app}(\pos, t)}{c_s^2 \rho(\pos, t)} + \frac{1}{2}}
-\end{equation}
-
-Despite the utility of the BGK collision operator, it does have a few drawbacks.
-For example, in low viscosity fluids the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
-It is well known that when large nonequilibrium distributions exist in the LBM approximation that overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
-To illustrate this, consider a flow in which there is a sharp spatial gradient in either $\rho$ or $\mvel$ at $\pos$.
-As $f_i^{eq}$ depends on both $\rho$ and $\mvel$ \eqref{eq:feq}, it may be the case that $|f_i^{eq}(\pos, t) - f_i^{eq}(\pos + \pvel_i \Delta t, t + \Delta t)| >> 0$, i.e.
-there may be a large difference in the quasiequilibrium for the $i^{th}$ discrete velocity at $(\pos, t)$ and $(\pos + \pvel_i \Delta t, t + \Delta t)$.
-In this case, if $f_i$ is ``near'' to quasiequilibrium at $\pos$ it will be ``far'' from quasiequilibrium after the streaming step when it moves to the node at $\pos + \pvel_i \Delta t$.
-Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
-Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations such as those depicted in \Fref{fig:overrelaxation}.
-Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to diverge or ``pollute'' the system enough to make the results highly nonphysical and altogether useless~\cite{gorban2014enhancement}.
-
-\begin{figure}
-	\centering
-    \includegraphics[width=\linewidth]{figs/overrelaxation_tau-0513}
-    \caption{Example of oscillations that can occur as a result of overrelaxation; the collision frequency, $\omega = 1.95$ and $\mu_{app} \approx 0.0043$}
-    \label{fig:overrelaxation}
-\end{figure}
-
-The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
-%MG: this needs cleaned up, I think
-
-Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
-The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
-
-Due to the instabilities associated with the collision frequency being too high (e.g.
-approaching 2) or too low (e.g.
-approaching 0), a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flow is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
-This simple methodology for increasing stability will henceforth be referred to bounded-relaxation time BGK, or BGK-BRT.
-
-\subsubsection{Multiple-relaxation Time (MRT)}
-
-An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
-In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f} = \begin{Bmatrix}f_0 & f_1 & ...
-& f_8\end{Bmatrix}$, herein referred to as the ``moment space''.
-The collision is then performed in the moment space.
-There are a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
-\begin{enumerate}
-    \item Physical processes in fluids can be approximately described by coupling or interacting among modes, and the modes are directly related to the moments~\cite{lallemand2000theory}.
-    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor~\cite{dellar2003incompressible}.
-    Of the nine relaxation rates available, the three that correspond to the extra variables--often referred to as ``ghost variables'', and their associated modes as ``ghost modes''--can be tuned in order to dampen out their associated ghost modes, ensuring these modes do not dominate or cause numerical instabilities at the lattice scale.
-\end{enumerate}
-
-The MRT collision operator is given by:
-
-\begin{equation} \label{eq:mrt-colop}
-	\colop = - {\transM}^{-1} \relaxM \transM (\mathbf{f} - \mathbf{f}^{eq}),
-\end{equation}
-
-\noindent where $\transM$ is a transformation matrix that maps the particle distribution vector, $\mathbf{f}$, and quasiequilibrium distribution vector, $\mathbf{f}^{eq}$, from the particle distribution space into moment space.
-The result of mapping the vectors $\mathbf{f}$ and $\mathbf{f}^{eq}$ into moment space will be denoted by $\mathbf{m}$ and $\mathbf{m}^{eq}$, respectively.
-The relationships between $\mathbf{m}$, $\transM$ and $\mathbf{f}$ can be written as follows:
-
-\begin{equation} \label{eq:transformation-matrix}
-\mathbf{m} = \begin{bmatrix}
-\rho \\ e \\ \epsilon \\ j_x \\ q_x \\ j_y \\ q_y \\ p_{xx} \\ p_{xy}
-\end{bmatrix} = \begin{bmatrix}
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
--4 & -1 & -1 & -1 & -1 & 2 & 2 & 2 & 2 \\
-4 & -2 & -2 & -2 & -2 & 1 & 1 & 1 & 1 \\
-0 & 1 & 0 & -1 & 0 & 1 & -1 & -1 & 1 \\
-0 & -2 & 0 & 2 & 0 & 1 & -1 & -1 & 1 \\
-0 & 0 & 1 & 0 & -1 & 1 & 1 & -1 & -1 \\
-0 & 0 & -2 & 0 & 2 & 1 & 1 & -1 & -1 \\
-0 & 1 & -1 & 1 & -1 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 1 & -1 & 1 & -1
-\end{bmatrix} \begin{bmatrix}
-f_0 \\
-f_1 \\
-f_2 \\
-f_3 \\
-f_4 \\
-f_5 \\
-f_6 \\
-f_7 \\
-f_8
-\end{bmatrix} = \transM \mathbf{f}.
-\end{equation}
-
-\noindent where $\epsilon$ is related to the square of the energy $e$; $q_x$ and $q_y$ correspond to the energy fluxes in the x and y directions; and $p_{xx}$ and $p_{xy}$ correspond to the diagonal and off-diagonal component of the viscous stress tensor~\cite{lallemand2000theory}.
-% MG: verify this is correct
-The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i \in [0, 2], i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
-In the case when $s_0 = s_1 = ... = s_8 = \omega$ ($\omega$ is the collision frequency in the BGK sense), the MRT collision operator is equivalent to the BGK collision operator.
-The relaxation parameters $s_0$, $s_3$, and $s_5$ are all set to zero as mass and momentum should be conserved.
-The relaxation parameters $s_1$ and $s_7 = s_8$ are related to the bulk and shear viscosities, respectively.
-The relationship for the shear viscosity is given by:
-\begin{equation} \label{eq:mrt-s7}
-\nu = c_s^2 \Delta t (\frac{1}{s_7} - \frac{1}{2}),
-\end{equation}
-\noindent which is equivalent with the relationship to the collision frequency, $\omega$, in the BGK sense (when $\Delta t = 1$).
-The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
-It is common practice, and~\citet{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.
-
-The MRT collision operator has a greater numerical stability than its BGK counterpart~\cite{lallemand2000theory,dellar2003incompressible,d2002multiple}, and because of the challenges associated with simulating non-Newtonian flow, the MRT collision operator has become popular for simulating non-Newtonian fluids~\cite{chen2014simulations,fallah2012multiple,tang2011bingham,vikhansky2008lattice,chai2011multiple,grasinger2015simulation}.
-The main drawback of the MRT collision operator is its computational expense.
-Why MRT is more computationally expensive is clear when one considers that \eqref{eq:mrt-colop} requires multiple matrix multiplications and \eqref{eq:bgk} requires none.
-It has been reported that MRT is approximately 15\% slower than BGK~\cite{d2002multiple}, but this was in the context of Newtonian flow.
-As will be shown later, for certain LBM implementations and non-Newtonian fluid flows the increase in computational expense can be much greater.
-
-\subsection{Stability Enhancement through Artificial Dissipation: Entropic Filtering}
-
-To reduce nonequilibrium fluctuations in LBM, one can introduce artificial dissipation.
-The idea of artificial dissipation is to increase numerical stability, while sacrificing some physical accuracy.
-A model that has more physical justification but produces unstable and nonsensical results is much less useful than a model with some minor artificial features yet produces more stable results.
-%Lyapanov stability is the idea that solutions at equilibrium stay there.
-A practical goal then would be to use only the necessary amount of artificial dissipation in order to ensure a stable solution.
-From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation is necessary?'' and ``how much artificial dissipation does one introduce when it is necessary?''.
-
-To answer the first question, consider the discussion in \Fref{sec:bgk} on problems with overrelaxation and nonphysical oscillations.
-Nonphysical oscillations due to overrelaxation would be damped out more quickly if particle distributions ``far'' from quasiequilibrium were brought closer to quasiequilibrium.
-A particle distribution vector, $\mathbf{f}$, ``far'' from equilibrium would be a good candidate for artificial dissipation.
-There are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
-A metric that has been developed and used successfully for determining when artificial dissipation should be introduced at a lattice site is the so called relative nonequilibrium entropy~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy}.
-The relative nonequilibrium entropy, $\Delta S$, is given by:
-
-\begin{equation} \label{eq:rel-noneq-entropy}
-\Delta S = \sum_i f_i \ln(\frac{f_i}{f^{eq}_i}).
-\end{equation}
-
-\noindent A more computationally efficient approximation of $\Delta S$ can be achieved by instead using the second order Taylor expansion of \eqref{eq:rel-noneq-entropy}:
-
-\begin{equation} \label{eq:quadratic-entropy}
-\Delta S \approx \sum_i \frac{(f_i - f^{eq}_i)^2}{2f^{eq}_i}.
-\end{equation}
-
-\noindent Note that limiting nonequilibrium entropy in LBM is analogous to what flux limiters do in finite difference, finite volume, and finite element methods~\cite{brownlee2008nonequilibrium}.
-
-A criteria for introducing artificial dissipation that has been used successfully ~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium} is to define a threshold, $\theta$, such that dissipation is added when:
-\begin{equation} \label{eq:ds-threshold}
-  \Delta S(\pos, t) > \theta.
-\end{equation}
-A potential drawback of defining a threshold \textit{a priori} is that in order to ensure the model still retains some physical integrity, only a small number of sites can have artificial dissipation added.
-If the threshold is too low, too many sites may have dissipation added.
-If the threshold is too high, a stable solution may not be achieved.
-The threshold can be determined on a case-by-case basis through trial-and-error or by a preliminary analysis.
-In the current work, the criteria that is used for determining whether dissipation should be added is a combination of \eqref{eq:ds-threshold} and the following:
-
-\begin{equation} \label{eq:dissipation-criteria}
-  \Delta S(\pos, t) > \overline{\Delta S} + n_\sigma \cdot \sigma_{\Delta S},
-\end{equation}
-
-\noindent where $\overline{\Delta S}$ and $\sigma_{\Delta S}$ are the mean and standard deviation of $\Delta S$, respectively--both are calculated using values over the domain for the current timestep--and $n_\sigma$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S$ must be before dissipation is added.
-The number of standard deviations, $n_\sigma$, is chosen \textit{a priori}.
-The criteria described in \eqref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead can think in terms of what would be the maximum percentage of sites one would want artificial dissipation to be added to.
-A disadvantage of \eqref{eq:dissipation-criteria} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artifical dissipation is unnecessary.
-By required that both \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} be met before artifical dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
-
-Just as there are many ways to measure a lattice site's distance from quasiequilibrium, or from that measure define criteria for artificial dissipation, there are also many different ways of deciding how much dissipation to add.
-One method of adding dissipation is the so-called Ehrenfests' regularization~\cite{brownlee2006stabilization}, and involves setting a lattice site that is chosen for artificial dissipation to its quasiequilibrium state.
-Although this achieves the desired result, namely damping out large nonequilibrium fluctuations, it does so in a way that is not smooth or gradual, but sharp.
-An alternative, the median filter, has been used successfully in conjunction with both the BGK and MRT collision schemes for simulating one-dimensional shock tubes and lid-driven cavity flow~\cite{brownlee2008nonequilibrium,gorban2014enhancement,packwood2009entropy}.
-Median filtering is an effective noise reduction technique, often used in image processing, for ``speckle noise'' or ``salt and pepper noise''~\cite{brownlee2008nonequilibrium}.
-In other words, median filters are good at reducing high frequency noise while having a minimal effect on lower frequency noise.
-In LBM this is a desirable way to introduce dissipation because it has the potential to reduce high frequency nonequilibrium fluctuations that might lead to numerical instability while retaining the lower frequency dynamics.
-\pagebreak
-To use the median filter one performs the collision step and then checks over the domain for lattice sites with $\Delta S$ that meet the criteria for artificial dissipation; sites that meet the criteria are updated as follows:
-\begin{equation} \label{eq:median-filter}
-\mathbf{f} = \mathbf{f}^{eq} + \delta (\mathbf{f} - \mathbf{f}^{eq}),
-\end{equation}
-\noindent where $\delta = \sqrt{\Delta S_{med} / \Delta S}$ is the scaling coefficient, and $\Delta S_{med}$ is the median value of $\Delta S$ for the nearest neighbors of the lattice site.
-
-%MG: show equation for feq that maximizes entropy and give a brief word why we should care
-
-\subsection{Boundary Conditions} \label{sec:bcs}
-
-No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
-For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
-For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
-The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
-
-\subsection{Applied Forces}
-
-Incorporating external forces, such as gravity, pressure gradients, etc., is done by adding a source of particle distributions in the direction of the force.
-The increase in particle distributions leads to the desired macroscopic result--an increase in momentum.
-The LBE with external forces is:
-
-\begin{equation}
-f_i(\pos + \pvel_i \Delta t, t + \Delta t) = f_i(\pos, t) + \Omega_i(f) + \frac{w_i \Delta t}{c_s^2} \mathbf{F} \cdot \pvel_i
-\end{equation}
-
-\noindent where $\mathbf{F}$ is the body force vector.
-
-\newcommand{\srtensor}{D_{\alpha\beta}}
-
-\subsection{Strain-rate Tensor}
-
-In order to solve for the apparent viscosity, $\mu_{app}$, using the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf}, one must calculate the strain-rate, $\dot{\gamma}$.
-The strain-rate is given by the second invariant of the strain-rate tensor, $\srtensor$, i.e.:
-\begin{equation} \label{eq:second-invariant}
-\dot{\gamma} = \sqrt{2 \sum_{\alpha,\beta=1}^{2} \srtensor \srtensor}.
-\end{equation}
-
-\noindent When using the BGK collision scheme, the strain-rate tensor is determined by:
-\begin{equation} \label{eq:srtensor-bgk}
-D_{\alpha \beta} = -\frac{\omega}{2 \rho c_s^2} \sum_i \xi_{i \alpha} \xi_{i \beta} (f_i - f^{eq}_i),
-\end{equation} % MG: verify this
-
-\noindent and for the MRT collision scheme, the strain-rate tensor is determined by:
-\begin{equation} \label{eq:srtensor-mrt}
-D_{\alpha \beta} = -\frac{1}{2 \rho c_s^2 \Delta t} \sum_i \xi_{i \alpha} \xi_{i \beta} \sum_j (\transM^{-1} \relaxM \transM)_{ij}(f_i - f^{eq}_i).
-\end{equation} % MG: verify this
-
-\noindent Computing the strain-rate tensor by either \eqref{eq:srtensor-bgk} or \eqref{eq:srtensor-mrt} is second order accurate in space~\cite{kruger2009shear,kruger2010second}.
-
-Upon comparison of \eqref{eq:srtensor-bgk} and \eqref{eq:srtensor-mrt}, it is clear that calculating the strain-rate tensor for the MRT collision operator is more computationally expensive than it is for the BGK collision operator.
-The increase in computational expense is exacerbated by the fact that approximating the apparent viscosity, $\mu_{app}$, by the algorithm outlined in \Fref{sec:lbm-for-nnf} requires iteration and therefore multiple calculations of the strain-rate tensor.
-
-\subsection{Non-Newtonian Constitutive Equations}
-
-\subsubsection{Bingham Plastic} \label{sec:bp}
-
-The Bingham plastic constitutive model is popular for yield stress flow.
-A Bingham plastic does not flow, i.e.
-the strain-rate is zero, when the shear stress is below the yield stress, and behaves in an almost Newtonian manner when the shear stress is above the yield stress.
-The Bingham plastic relationship is described mathematically as:
-\begin{equation} \label{eq:bingham}
-\begin{cases}
-\tau = \tau_y + \mu_p \dot{\gamma}, & |\tau| \geq \tau_y \\
-\dot{\gamma} = 0, & |\tau| < \tau_y
-\end{cases}
-\end{equation}
-\noindent where $\tau$ is the shear stress, $\tau_y$ is the yield stress, and $\mu_p$ is the plastic viscosity~\cite{bingham1922fluidity}.
-
-Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
-Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\citet{papanastasiou1987flows} is often used as:
-\begin{equation} \label{eq:papa}
-\tau = \tau_y (1 - e^{-m |\dot{\gamma}|}) + \mu_p \dot{\gamma},
-\end{equation}
-\noindent where $m$ is the stress growth exponent.
-The larger the value of $m$, the closer the approximation is to the Bingham plastic model.
-
-Alternatively, the constitutive relationship can be interpreted through the apparent viscosity.
-Noting that $\mu_{app} = \frac{\tau}{\dot{\gamma}}$ and rearranging \eqref{eq:papa} results in the following expression for the apparent viscosity:
-\begin{equation} \label{eq:bing-mu-app}
-\mu_{app} = \frac{\tau_y}{\dot{\gamma}} (1 - e^{-m |\dot{\gamma}|}) + \mu_p.
-\end{equation}
-
-\subsubsection{Power-law}
-
-The power-law constitutive relationship is useful for modeling fluids that experience shear-thinning or shear-thickening.
-The power-law relationship is given by:
-\begin{equation} \label{eq:pl}
-\tau = k \dot{\gamma}^n,
-\end{equation}
-\noindent where $k$ is the flow consistency index and $n$ is the flow behavior index.
-When $n = 1$, \eqref{eq:pl} results in a Newtonian constitutive relationship with dynamic viscosity, $\mu = k$.
-A flow consistency index of $n < 1$ results in shear-thinning behavior, where as $n > 1$ results in shear-thickening.
-As with the Bingham plastic relationship, \eqref{eq:pl} can be modified to determine an apparent viscosity:
-\begin{equation} \label{eq:pl-mu-app}
-\mu_{app} = k \dot{\gamma}^{n-1}.
-\end{equation}
-
-\section{Numerical Study} \label{sec:numerical-study}
-
-A numerical study was carried out in order to investigate the suitability of different LBM collision schemes for simulating non-Newtonian flow in terms of accuracy, numerical stability and computational efficiency.
-For all simulations presented in this section, the apparent viscosity is approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the maximum number of iterations set at 15 and the convergence criteria:
-\begin{equation} \label{eq:mu-app-conv}
-\frac{\left|\mu_{app}^{k+1} - \mu_{app}^{k}\right|}{\mu_{app}^{k}} < 1.0 \times 10^{-6}.
-\end{equation}
-All BGK-BRT collision schemes use $\omega \in [0.05, 1.995]$ as the bounds on the collision frequency.
-All simulations with artificial dissipation use the median filter with $\theta = 1.0 \times 10^{-6}$ and $n_{\sigma} = 2.7$ where $\theta$ is the threshold that $\Delta S(\pos, t$ must exceed before dissipation is added \eqref{eq:ds-threshold} and $n_{\sigma}$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S(\pos, t)$ must be before dissipation is added \eqref{eq:dissipation-criteria}.
-All numerical values given in this section are in lattice units unless otherwise stated.
-
-\newcommand{\pgrad}{\frac{\partial p}{\partial x}}
-\newcommand{\unorm}{\mathbf{\hat{n}}}
-
-\subsection{Poiseuille Flow}
-
-Poiseuille flow is a useful benchmark because analytical solutions exist for both Bingham plastic and power-law fluids.
-Poiseuille flow is driven by a constant pressure gradient, $\pgrad$, through a two-dimensional channel.
-A schematic is shown in \Fref{fig:poise-schematic}.
-No-slip boundary conditions are enforced at the top and bottom boundaries with a wall velocity of zero so that $\mvel \times \unorm = 0$ where $\unorm$ is the unit normal vector to the boundary. 
-Periodic boundary conditions are enforced at the left and right boundaries.
-The total height of the channel is denoted by $H$.
-The center of the channel is $y = 0$ and $y \in [-h, h]$ where $h = \frac{H}{2}$.
-
-\begin{figure}
-	\centering
-    \includegraphics[width=\linewidth]{figs/poise-schematic}
-    \caption{Schematic of Poiseuille flow; no-slip boundary conditions are enforced at the top and bottom boundaries, and periodic boundary conditions are enforced at the left and right boundaries.
-The size of the domain is $L = 32$  and $H = 64$.}
-    \label{fig:poise-schematic}
-\end{figure}
-
-Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps.
-The lattice size was chosen to be sufficiently fine for accuracy (although the length, $L = 32$, seems small, the effect is negligible as the east-west boundary conditions are periodic) and the number of time steps were set high enough to ensure a steady-state would be reached.
-In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
-For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$, which follows the recommendation of~\cite{lallemand2000theory} for reasons of stability, and has been successfully used to simulate non-Newtonian flow in the past~\cite{fallah2012multiple,grasinger2015simulation}.
-
-\subsubsection{Bingham Plastic Fluids} \label{sec:poise-bing}
-
-The Bingham plastic simulations were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a plastic viscosity of $\mu_p = 0.2$.
-The yield stress was varied between four different values $\tau_y = [4.0, 8.0, 12.0, 16.0] \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
-Recall that $m$ is the stress growth exponent for the Papanastasiou approximation.
-The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
-The analytical solution for Poiseuille flow of a Bingham plastic fluid is given by:
-\begin{equation} \label{eq:analytical-bingham}
-u_x(y) = \begin{cases}
-\frac{1}{2 \mu_p} \left(-\pgrad\right) \left[h^2 - y_{\tau}^2 \right] - \frac{\tau_y}{\mu_p} \left(h - y_{\tau}\right), & 0 \leq |y| \leq y_{\tau}, \\
-\frac{1}{2 \mu_p} \left(-\pgrad\right) \left[h^2 - y^2 \right] - \frac{\tau_y}{\mu_p} \left(h - |y|\right), & y_{\tau} < |y| \leq h,
-\end{cases}
-\end{equation}
-\noindent where $y_{\tau} = -\tau_y / \pgrad$ is the vertical location at which the fluid yields.
-
-The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing}.
-The Reynold's number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution.
-The Bingham number was computed by $Bn = \frac{\tau_y H}{\mu_p U}$.
-
-% MG: it might be better to normalize the time with respect to the fastest, BGK with no filtering
-
-\begin{table}
-\centering
-\caption{Bingham plastic Poiseuille flow}
-\vspace{0.5cm}
-\begin{tabulary}{\linewidth}{l l r r r r r r r r}
-\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
-\hline \\
-BGK & No & $10^5$ & 4.0 & 6.05 & 0.68 & 0.0062 & 0.0153 & 1857 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0204 & 0.0411 & 2329 \\
-& & & 12.0 & 3.04 & 4.04 & 0.0503 & 0.0891 & 3345 \\
-& & & 16.0 & 1.92 & 8.52 & 0.1161 & 0.1879 & 2029 \\
-\\
-BGK & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0109 & 0.0282 & 2831 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0330 & 0.0670 & 3509 \\
-& & & 12.0 & 3.04 & 4.04 & 0.0788 & 0.1570 & 4838 \\
-& & & 16.0 & 1.92 & 8.52 & 0.1991 & 0.3539 & 3790 \\
-\\
-BGK & Yes & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0100 & 0.0273 & 2903 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0361 & 0.0823 & 3567 \\
-& & & 12.0 & 3.04 & 4.04 & 0.2439 & 0.3832 & 4800 \\
-& & & 16.0 & 1.92 & 8.52 & 0.7533 & 1.0718 & 3507 \\
-\\
-MRT & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0018 & 0.0018 & 7908 \\
-& & & 12.0 & 3.04 & 4.04 & 0.0026 & 0.0026 & 7160 \\
-& & & 16.0 & 1.92 & 8.52 & 0.0041 & 0.0041 & 5559 \\
-\\
-\label{tab:poise-bing}
-\end{tabulary}
-\end{table}
-
-As has been reported previously, for the BGK collision operator, using a stress growth exponent of $m = 10^5$ is more accurate with respect to the analytical solution than using a stress growth exponent of $m = 10^8$~\cite{chen2014simulations}.
-A larger stress growth exponent results in a more accurate Papastasiou approximation of the true Bingham plastic constitutive model, however, it leads to more nonphysical oscillations that degrade the numerical solution for the BGK collision operator.
-Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helps mitigate errors that occur as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
-\pagebreak
-In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress fluids (relative errors of approximately 25-75\%).
-
-In general, the BGK collision operator using a stress growth exponent of $m = 10^5$ had the lowest computational time.
-This can be attributed to the fact that a smoother approximation of the Bingham plastic constitutive model would lead to a solution for $\mu_{app}$ converging with less iterations.
-The BGK collision operator using a stress growth exponent of $m = 10^5$ experienced relatively low error for lower yield stress fluids, however, for $\tau_y = 12 \times 10^{-5}$ and $16 \times 10^{-5}$ the relative $L_\infty$ errors were 8.9\% and 19\%, respectively, which is larger than what would be considered acceptable for most engineering applications.
-
-The LBM approximations of the velocity profile across the channel (more specifically, $u_x(\pos_j, 25000)$ where $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, i.e.
-$\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction) for the BGK collision operator using $m = 10^5$ and $m = 10^8$ are plotted with the analytical solution \eqref{eq:bingham} in \Fref{fig:bing-analyt-bgk-5} and \Fref{fig:bing-analyt-bgk-8}, respectively.
-Due to the smoothness of the LBM approximation in \Fref{fig:bing-analyt-bgk-5}, one can conclude that the error for the BGK model with $m = 10^5$ is not due to nonphysical oscillations, but instead the inaccuracy of the Papastasiou approximation with a lower stress growth exponent.
-In contrast, the LBM approximation in \Fref{fig:bing-analyt-bgk-8} is not smooth, which suggests that the error for the BGK model with $m = 10^8$ is due to nonphysical oscillations.
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{figs/poise-bingham/bgk-5/analytical-vs-approx.png}
-\caption{LBM approximation using BGK and $m = 10^5$ compared to the analytical solution for $\tau_y = 16 \times 10^{-5}$.}
-\label{fig:bing-analyt-bgk-5}
-\end{figure}
-
-\begin{figure}
-	\centering
-    \includegraphics[width=\textwidth]{figs/poise-bingham/bgk-8/analytical-vs-approx.png}
-    \caption{LBM approximation using BGK and $m = 10^8$ compared to the analytical solution for $\tau_y = 16 \times 10^{-5}$.}
-\label{fig:bing-analyt-bgk-8}
-\end{figure}
-
-In order to better understand why median filtering did not eliminate the nonphysical oscillations in high yield stress fluids, but instead exacerbates the problem, it is necessary to investigate what is happening at the particle distribution scale.
-\Fref{fig:feq-vs-f_bgk} compares particle distributions to quasiequilibrium.
-More specifically, \Fref{fig:feq-vs-f_bgk} compares $f_i(\pos_j, 25000)$ to $f_i^{eq}(\pos_j, 25000)$ where $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, and $i = 5, 8$.
-It can be seen that in the BGK, $m = 10^8$ solution, that nonphysical oscillations pollute the quasiequilibrium profile as well.
-These nonphysical oscillations likely originate at the lattice scale due to the sharp gradient in the macroscopic velocity, $\mvel$, near the walls, and the sharp gradient in $\mvel$ is a result of a sharp gradient in $\mu_{app}$, namely the sharp gradient of the constitutive relationship.
-Because the nonphysical oscillations make their way into the quasiequilibrium values, entropic median filtering does not help dampen the oscillations but instead contracts the particle distributions closer to the quasiequilibrium oscillations, which explains why the median filtered results were less accurate for Bingham plastic Poiseuille flow.
-In order to ensure that this phenomena was a side effect of all entropic filtering, and not just entropic median filtering with $\theta = 1.0 \times 10^{-6}$ and $n_s = 2.7$, optimization was used to find the value of $\theta$ that minimized error for both median filtering and Ehrenfests' regularization.
-The optimization problem was defined as follows:
-\begin{align*}
-  \min_\theta \hspace{0.2in} & f(\theta)\\
-\text{such that} \hspace{0.2in} & \theta \in [10^{-10}, 2.0]
-\end{align*}
-\noindent where $f(\theta) \equiv $ the relative $L_2$ error between the LBM approximation using $m = 10^8$ and entropic filtering with a $\Delta S$ threshold of $\theta$ (used in the criteria defined in \eqref{eq:ds-threshold}).
-The bounds on $\theta$, namely $\theta \in [10^{-10}, 2]$ represent $\theta$ such that 100\% of lattice sites are filtered for every time step ($\theta = 10^{-10}$) and $\theta$ such that no lattice sites are filtered for any time step ($\theta = 2.0$).
-The optimization problem was solved for the cases of using median filtering and Ehrenfests' regularization using both Brent's method and the Golden Section search.
-For all four cases, the optimal solution was $\theta = 2.0$ and the cost function, $f(2.0)$, was equal to the relative $L_2$ error for the BGK collision operator, $m = 10^8$ and no entropic filtering.
-The results of the optimization suggest that entropic filtering, at best does not affect the accuracy of LBM, and often adversly affects the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization is used, and regardless of what the $\Delta S$ threshold, $\theta$, is set to.
-Thus, entropic filtering is not effective for increasing the accuracy of Bingham plastic Poiseuille flow.
-
-%MG: reformat these images so that text is more readable
-\begin{figure}
-\centering
-\begin{tabulary}{\linewidth}{c c}
-\includegraphics[width=2.75in]{figs/poise-bingham/bgk-5/feq-vs-f_5.png}
-&
-\includegraphics[width=2.75in]{figs/poise-bingham/bgk-5/feq-vs-f_8.png}
-\\
-\includegraphics[width=2.75in]{figs/poise-bingham/bgk-8/feq-vs-f_5.png}
-&
-\includegraphics[width=2.75in]{figs/poise-bingham/bgk-8/feq-vs-f_8.png}
-\end{tabulary}
-\caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums.
-The top two plots are for the BGK with $m = 10^5$, the bottom two are for the BGK with $m = 10^8$.}
-\label{fig:feq-vs-f_bgk}
-\end{figure}
-
-On average, the LBM model using the MRT collision operator took 2.9 times more computation time than the BGK collision operator with $m = 10^5$.
-However, the increased computational time can be justified for the MRT collision operator because MRT solution did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
-when $m = 10^8$.
-The MRT collision operator was therefore the most accurate solution for all cases.
-\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilbirium profiles, $f_i^{eq}$ for the MRT model; in contrast to the BGK with $m = 10^8$, there are no nonphysical oscillations in the quasiequilibrium distributions.
-
-\begin{figure}
-	\centering
-    \begin{tabulary}{\linewidth}{c c}
-        \includegraphics[width=2.75in]{figs/poise-bingham/mrt/feq-vs-f_5.png}
-        &
-        \includegraphics[width=2.75in]{figs/poise-bingham/mrt/feq-vs-f_8.png}
-    \end{tabulary}
-    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
-    \label{fig:feq-vs-f_mrt}
-\end{figure}
-
-Another important question to ask is, if the strength of the MRT collision operator is damping out the ghost modes, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of the ghost modes?.
-\Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment, $\boldsymbol{\epsilon}^{neq}$, with respect to time for each of the collision schemes, which was calculated as:
-\begin{equation}
-\frac{||\boldsymbol{\epsilon} - \boldsymbol{\epsilon}^{eq}||_2}{||\boldsymbol{\epsilon}^{eq}||_2}
-\end{equation}
-\noindent where $||\hdots||_2$ is the Euclidean norm, $\epsilon_j = 4f_0(\pos_j) - 2f_1(\pos_j) - 2f_2(\pos_j) - 2f_3(\pos_j) - 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) + f_7(\pos_j) + f_8(\pos_j)$ is related to the square of the energy, and $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, i.e.
-$\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
-The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ suffered from was in part due to the $\epsilon$ moment.
-
-\begin{figure}
-	\centering
-    \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon}
-    \caption{Evolution of relative $L_2$ norm, $\epsilon^{neq}$ with time.
-The norm of the $\epsilon^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $\epsilon$ moment.}
-    \label{fig:epsilon}
-\end{figure}
-
-\Fref{fig:qx} shows a measure of the nonequilibrium $q_x$ moment, $\mathbf{q_x}^{neq}$, or energy flux in the x-direction, with respect to time for each of the collision schemes, which was calculated as:
-\begin{equation}
-    \frac{||\mathbf{q_x} - \mathbf{q_x}^{eq}||_2}{||\mathbf{q_x}^{eq}||_2}
-\end{equation}
-\noindent where $q_{x_j} = - 2f_1(\pos_j) + 2f_3(\pos_j) + f_5(\pos_j) - f_6(\pos_j) - f_7(\pos_j) + f_8(\pos_j)$ and $\pos_j$ were taken across the height of the channel in the same manner as with the $\epsilon$ moment.
-The values measured for the BGK with $m = 10^8$ and median filtered were consistently greater than any other collision scheme.
-The relative $L_2$ norm of the $\mathbf{q_x}^{neq}$ for the BGK collision operator with $m = 10^8$ and median filtering solution spent most of the simulation between approximately 50\% and 100\%, this suggests that the $q_x$ moment was likely the cause of the nonphysical oscillations and error in its solution.
-As expected (if the dominant source of error in the LBM approximation for Bingham plastic Poiseuille flow is indeed due to the $q_x$ moment), the $\mathbf{q_x^{neq}}$ relative $L_2$ norm for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $\mathbf{q_x^{neq}}$ relative $L_2$ was negligible for the MRT collision scheme.
-These results are consistent with solution error.
-
-\begin{figure}
-	\centering
-    \includegraphics[width=\linewidth]{figs/poise-bingham/qx}
-    \caption{Evolution of relative $L_2$ norm, $\mathbf{q_x}^{neq}$ with time.
-The norm of the $\mathbf{q_x^{neq}}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $q_x$ moment.}
-    \label{fig:qx}
-\end{figure}
-
-It is likely that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time, i.e.
-osciallations build in amplitude with time.
-\Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time, which was calculated as:
-\begin{equation}
-  \int_0^{T} \frac{||\boldsymbol{\epsilon}(t) - \boldsymbol{\epsilon}^{eq}(t)||_2}{||\boldsymbol{\epsilon}^{eq}(t)||_2} dt,
-\end{equation}
-\noindent and,
-\begin{equation}
-  \int_0^{T} \frac{||\mathbf{q_x}(t) - \mathbf{q_x}^{eq}(t)||_2}{||\mathbf{q_x}^{eq}(t)||_2} dt,
-\end{equation}
-\noindent where $T$ is the current time. 
-\Fref{fig:qx} and \Fref{fig:qx-cumulative} shows that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the ghost mode associated with the $q_x$ moment dominates and is the primary source of oscillations.
-It can be inferred that the $q_x$ moment was significant source of oscillations in this case because the Poiseuille flow was in the x-direction.
-If an LBM collision scheme is to be developed for simulating high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it should focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon_cumulative}
-    \caption{Cumulative relative $L_2$ norm, $\epsilon^{neq}$ with time.
-Oscillations can have a positive effect on each other, building up.
-The cumulative relative $L_2$, $\epsilon^{neq}$ is a measure of how much oscillations due to the $\epsilon$ moment may have been building in time.}
-    \label{fig:epsilon-cumulative}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{figs/poise-bingham/qx_cumulative}
-    \caption{Cumulative relative $L_2$ norm, $\mathbf{q_x^{neq}}$ with time.
-Oscillations can have a positive effect on each other, building up.
-The cumulative relative $L_2$, $\mathbf{q_x^{neq}}$ is a measure of how much oscillations due to the $q_x$ moment may have been building in time.}
-    \label{fig:qx-cumulative}
-\end{figure}
-
-\subsubsection{Power-law Fluids} \label{sec:poise-powerlaw}
-
-The simulations of the Poiseuille flow with the power-law constitutive relationship were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a flow consistency index of $k = 0.2$.
-The flow behavior index, $n$, was varied between the four different values: $[0.5, 0.75, 1.25, 1.50]$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $[0.05, 1.995]$.
-The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
-The analytical solution for the velocity profile of Poiseuille flow with the power-law constitutive relationship is given by:
-\begin{equation} \label{eq:analytical-power-law}
-  u_x(y) = \frac{n}{n+1}\left(-\frac{1}{k}\pgrad\right)^{1/n}\left[h^{\frac{n+1}{n}} - |y|^{\frac{n+1}{n}}\right].
-\end{equation}
-
-The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-power-law}.
-The Reynold's number was computed by $Re = \frac{\rho U^{2-n} H^n}{k}$.
-
-% MG: it might be better to normalize the time with respect to the fastest, BGK with no filtering
-
-\begin{table}
-\centering
-\caption{Power-law Poiseuille flow}
-\vspace{0.5cm}
-\begin{tabulary}{\linewidth}{l l r r r r r}
-\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $n$ & $Re$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
-\hline \\
-BGK & No & 0.50 & 0.0007 & 28.06 & 30.37 & 5232 \\
-& & 0.75 & 0.9125 & 0.0328 & 0.0569 & 2399 \\
-& & 1.25 & 423.2 & 0.0051 & 0.0055 & 2311 \\
-& & 1.50 & 2213 & 1.0 & 1.0 & 3622 \\
-\\
-BGK & Yes & 0.50 & 0.0007 & 26.72 & 28.78 & 5347 \\
-& & 0.75 & 0.9125 & 0.0328 & 0.0569 & 2385 \\
-& & 1.25 & 423.2 & 0.0051 & 0.0055 & 2275 \\
-& & 1.50 & 2213 & 0.0570 & 0.0600 & 2357 \\
-\\
-MRT & No & 0.50 & 0.0007 & 0.1758 & 0.1401 & 4906 \\
-& & 0.75 & 0.9125 & 0.0058 & 0.0045 & 4880 \\
-& & 1.25 & 423.2 & 0.0051 & 0.0055 & 4863 \\
-& & 1.50 & 2213 & 1.0 & 1.0 & 15992 \\
-\\
-BGK-BRT & No & 0.50 & 0.0007 & 11.73 & 12.56 & 4450 \\
-& & 0.75 & 0.9125 & 0.0320 & 0.0475 & 1805 \\
-& & 1.25 & 423.2 & 0.0051 & 0.0055 & 1905 \\
-& & 1.50 & 2213 & 1.0 & 1.0 & 8194 \\
-\\
-\label{tab:poise-power-law}
-\end{tabulary}
-\end{table}
-
-Although for $n = 0.5, 0.75, 1.25$ the MRT collision operator was more accurate and required more computation time than the BGK collision operator, for $n = 1.5$ the BGK operator with median filtering required the least computation time and was the most accurate with respect to the analytical solution.
-In fact, besides for the case of $n = 0.5$, i.e. relatively extreme shear-thinning, the BGK with median filtering was both computationally efficient and accurate.
-The BGK-BRT had a moderate improvement of accuracy and efficiency--with respect to the regular BGK collision scheme--for flow where the constitutive relationship was only somewhat nonlinear, i.e.
-for $n = 0.75, 1.25$.
-Note that none of the collision schemes were sufficiently accurate (error greater than 15\%) for the case of $n = 0.5$, and to improve this accuracy it would be necessary to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
-
-\subsection{Lid-driven Cavity Flow} % MG: would it be of any use to show streamline plots from some of these flows?
-
-The lid-driven benchmark was chosen for this numerical study because there are many results available in literature in which to compare with and because the vorticity of the flow coupled with the nonlinear constitutive equations should result in a challenge in terms of stability.
-Lid-driven cavity flow is generally characterized by a square cavity where a fluid velocity is prescribed tangential to the upper boundary and the remaining boundaries have a no-slip boundary condition.
-The example of lid-driven cavity flow utilized herein is presented schematically in \Fref{fig:lid-driven}.
-
-\begin{figure}
-\centering
-\includegraphics{figs/lid-driven}
-\caption{Schematic of lid-driven cavity flow; a velocity is prescribed tangential to the top boundary and no-slip is enforced at the remaining boundaries. $L = 100$}
-\label{fig:lid-driven}
-\end{figure}
-
-The lid-driven cavity simulations presented in the section were all simulated on a relatively coarse, $100 \times 100$ lattice (other studies simulating lid-driven cavity flow of a non-Newtonian fluid tend to use lattice sizes on the order of $256 \times 256$ and $512 \times 512$~\cite{chen2014simulations,li2014simulation}, this study follows that of~\cite{brownlee2008nonequilibrium}).
-A coarse lattice was chosen in order to highlight concerns with stability and accuracy.
-The simulations were run for either 50000 timesteps or until convergence was met.
-Convergence was defined by,
-\begin{equation} \label{eq:convergence}
-\sum_{m=1}^{100} \sum_{i, j} \frac{|u_{i, j}^k - u_{i, j}^{k-m}|}{|u_{i, j}^{k-m}|} < 1.0 \times 10^{-7},
-\end{equation}
-\noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current timestep.
-The lid velocity was prescribed as $u_o = 0.1$.
-For the MRT relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$ as these values have been successfully applied to simulating lid-driven cavity flow of non-Newtonian fluids in the past~\cite{chen2014simulations,li2014simulation}.
-All coordinate values presented in this section (used to describe the location of the center of vortices) are given normalized with respect to the length of the cavity side, i.e. as $(x / L, y / L)$.
-Note that all results presented as ``-'' indicate that the tests were numerically unstable to the extent that the degrees of freedom all over the domain were nonsensical.
-
-\subsubsection{Bingham Plastic Fluids}
-
-For the Bingham plastic numerical tests the Reynold's number was varied:$[100, 1000, 5000, 10000]$, and the Bingham number was varied: $[1, 10, 100]$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
-The collision schemes that were tested were: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
-Tables \ref{tab:lid-bing1}--\ref{tab:lid-bing100} compare the center location of the main vortex to literature values taken from~\citet{syrakos2014performance}.
-The main vortex location is determined by calculating the stream function using Simpson's rule and finding where it attains a maximum.
-
-\begin{table}
-\centering
-\caption{Bingham plastic, lid-driven cavity flow; $Bn = 1$.}
-\small
-\vspace{0.5cm}
-\begin{tabulary}{\linewidth}{r r l l r r r r}
-$Bn$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
-\hline \\
-1 & 100 & BGK     & No  & $10^5$ & (0.63, 0.79) & (0.63, 0.79) & 17685 \\
-  &     & BGK     & No  & $10^8$ &              & (0.63, 0.79) & 19398 \\
-  &     & BGK-BRT & No  & $10^8$ &              & (0.63, 0.79) & 22656 \\
-  &     & BGK     & Yes & $10^8$ &              & (0.63, 0.79) & 21535 \\
-  &     & MRT     & No  & $10^8$ &              & (0.63, 0.79) & 79040 \\
-  &     & MRT     & Yes & $10^8$ &              & (0.63, 0.79) & 85295 \\
-\\
-1 & 1000 & BGK     & No  & $10^5$ & (0.54, 0.57) & (0.54, 0.57) & 16109 \\
-  &      & BGK     & No  & $10^8$ &              & (0.54, 0.57) & 17035 \\
-  &      & BGK-BRT & No  & $10^8$ &              & (0.54, 0.57) & 16879 \\
-  &      & BGK     & Yes & $10^8$ &              & (0.54, 0.57) & 20170 \\
-  &      & MRT     & No  & $10^8$ &              & (0.54, 0.57) & 46048 \\
-  &      & MRT     & Yes & $10^8$ &              & (0.54, 0.57) & 55818 \\
-\\
-1 & 5000 & BGK     & No  & $10^5$ & (0.52, 0.53) & - & - \\
-  &      & BGK     & No  & $10^8$ &              & - & - \\
-  &      & BGK-BRT & No  & $10^8$ &              & - & - \\
-  &      & BGK     & Yes & $10^8$ &              & (0.54, 0.53) & 17248 \\
-  &      & MRT     & No  & $10^8$ &              & (0.51, 0.55) & 50572 \\
-  &      & MRT     & Yes & $10^8$ &              & (0.52, 0.53) & 54225 \\
-\\
-1 & 10000 & BGK     & No  & $10^5$ & N/A          & - & - \\
-  &       & BGK     & No  & $10^8$ &              & - & - \\
-  &       & BGK-BRT & No  & $10^8$ &              & - & - \\
-  &       & BGK     & Yes & $10^8$ &              & (0.56, 0.60) & 18186 \\
-  &       & MRT     & No  & $10^8$ &              & (0.48, 0.48) & 43246 \\
-  &       & MRT     & Yes & $10^8$ &              & (0.46, 0.54) & 50864 \\
-\\
-\end{tabulary}
-\label{tab:lid-bing1}
-\end{table}
-\begin{table}
-\centering
-    \caption{Bingham plastic, lid-driven cavity flow; $Bn = 10$.}
-    \small
-    \vspace{0.5cm}
-\begin{tabulary}{\linewidth}{r r l l r r r r}
-    $Bn$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
-    \hline \\
-10 & 100 & BGK     & No  & $10^5$ & (0.53, 0.87) & (0.54, 0.87) & 25217 \\
-   &     & BGK     & No  & $10^8$ &              & (0.55, 0.87) & 38285 \\
-   &     & BGK-BRT & No  & $10^8$ &              & (0.55, 0.87) & 29539 \\
-   &     & BGK     & Yes & $10^8$ &              & (0.54, 0.87) & 35817 \\
-   &     & MRT     & No  & $10^8$ &              & (0.53, 0.88) & 143741 \\
-   &     & MRT     & Yes & $10^8$ &              & (0.54, 0.88) & 135059 \\
-\\
-10 & 1000 & BGK     & No  & $10^5$ & (0.80, 0.85) & (0.78, 0.83) & 11525 \\
-   &      & BGK     & No  & $10^8$ &              & (0.78, 0.83) & 26671 \\
-   &      & BGK-BRT & No  & $10^8$ &              & (0.78, 0.83) & 19248 \\
-   &      & BGK     & Yes & $10^8$ &              & (0.79, 0.84) & 34143 \\
-   &      & MRT     & No  & $10^8$ &              & (0.79, 0.84) & 105136 \\
-   &      & MRT     & Yes & $10^8$ &              & (0.79, 0.84) & 111942 \\
-\\
-10 & 5000 & BGK     & No  & $10^5$ & (0.60, 0.55) & - & - \\
-   &      & BGK     & No  & $10^8$ &              & - & - \\
-   &      & BGK-BRT & No  & $10^8$ &              & - & - \\
-   &      & BGK     & Yes & $10^8$ &              & (0.52, 0.55) & 34638 \\
-   &      & MRT     & No  & $10^8$ &              & (0.55, 0.55) & 111274 \\
-   &      & MRT     & Yes & $10^8$ &              & (0.55, 0.53) & 121685 \\
-\\
-10 & 10000 & BGK     & No  & $10^5$ & N/A          & - & - \\
-   &       & BGK     & No  & $10^8$ &              & - & - \\
-   &       & BGK-BRT & No  & $10^8$ &              & - & - \\
-   &       & BGK     & Yes & $10^8$ &              & (0.49, 0.54) & 19351 \\
-   &       & MRT     & No  & $10^8$ &              & (0.53, 0.54) & 69249 \\
-   &       & MRT     & Yes & $10^8$ &              & (0.53, 0.53) & 69054 \\
-\\
-\end{tabulary}
-\label{tab:lid-bing10}
-\end{table}
-\begin{table}
-\centering
-    \caption{Bingham plastic, lid-driven cavity flow; $Bn = 100$.}
-    \small
-    \vspace{0.5cm}
-    \begin{tabulary}{\linewidth}{r r l l r r r r}
-        $Bn$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
-        \hline \\
-100 & 100 & BGK     & No  & $10^5$ & (0.51, 0.95) & (0.51, 0.95) & 13237 \\
-    &     & BGK     & No  & $10^8$ &              & (0.54, 0.96) & 26782 \\
-    &     & BGK-BRT & No  & $10^8$ &              & (0.53, 0.96) & 33095 \\
-    &     & BGK     & Yes & $10^8$ &              & (0.58, 0.96) & 34430 \\
-    &     & MRT     & No  & $10^8$ &              & (0.49, 0.95) & 161998 \\
-    &     & MRT     & Yes & $10^8$ &              & (0.54, 0.96) & 162085 \\
-\\
-100 & 1000 & BGK     & No  & $10^5$ & (0.53, 0.95) & (0.54, 0.95) & 42233 \\
-    &      & BGK     & No  & $10^8$ &              & (0.64, 0.95) & 47364 \\
-    &      & BGK-BRT & No  & $10^8$ &              & (0.60, 0.92) & 46289 \\
-    &      & BGK     & Yes & $10^8$ &              & (0.79, 0.95) & 48225 \\
-    &      & MRT     & No  & $10^8$ &              & (0.54, 0.95) & 190119 \\
-    &      & MRT     & Yes & $10^8$ &              & (0.55, 0.95) & 188242 \\
-\\
-100 & 5000 & BGK     & No  & $10^5$ & (0.93, 0.97) & - & - \\
-    &      & BGK     & No  & $10^8$ &              & - & - \\
-    &      & BGK-BRT & No  & $10^8$ &              & - & - \\
-    &      & BGK     & Yes & $10^8$ &              & (0.91, 0.95) & 57354 \\
-    &      & MRT     & No  & $10^8$ &              & (0.92, 0.96) & 163318 \\
-    &      & MRT     & Yes & $10^8$ &              & (0.93, 0.95) & 159523 \\
-\\
-100 & 10000 & BGK     & No  & $10^5$ & (0.92, 0.94) & - & - \\
-    &       & BGK     & No  & $10^8$ &              & - & - \\
-    &       & BGK-BRT & No  & $10^8$ &              & - & - \\
-    &       & BGK     & Yes & $10^8$ &              & (0.84, 0.88) & 32080 \\
-    &       & MRT     & No  & $10^8$ &              & (0.89, 0.91) & 152707 \\
-    &       & MRT     & Yes & $10^8$ &              & (0.89, 0.91) & 145722 \\
-\\
-\end{tabulary}
-\label{tab:lid-bing100}
-\end{table}
-
-Just as before with the Bingham plastic Poiseuille flow, if a modeler is using the BGK collision operator and entropic filtering is not available, a stress growth exponent of $m = 10^5$ yields faster, more accurate results than using a larger stress growth exponent.
-A smaller stress growth exponent is also more effective at producing accurate results in the BGK collision scheme than placing bounds on the relaxation time (BGK-BRT).
-However, although the BGK with $m = 10^5$ was the fastest model in all cases, it was also like the other two BGK models without median filtering in that it was unstable for $Re \ge 5000$.
-
-For flow with $Re \ge 5000$, the BGK with $m = 10^8$ and median filtering produced numerically stable results.
-However, in general, the BGK with $m = 10^8$ and median filtering was consistently different from literature values. 
-The apparent inaccuracy with regards to BGK with $m = 10^8$ and median filtering is probably due to how the numerical stability is enhanced--namely through artificial, nonphysical dissipation.
-In contrast to the nonphysical nature in which median filtering enhances stability, the stability enhancement used by MRT does not directly affect the macroscopic hydrodynamics of interest, so it again provides a stable, and in most of the cases examined herein, the most accurate solution.
-% MG: verify this sentence is correct
-The apparent superiority in terms of stability and accuracy of the MRT collision operator still comes at a price though.
-The MRT collision operator was, in general, 5-10 times slower than any of the BGK collision schemes.
-The extreme increase in computational expense is probably not due to the collision process itself, i.e.
-\eqref{eq:mrt-colop}, but instead calculating the strain-rate \eqref{eq:srtensor-mrt} for each iteration of the solution for $\mu_{app}$.
-
-In summary for the lid-driven cavity flow of a Bingham plastic fluid, 
-\begin{itemize}
-    \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
-    \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable,
-    \item and the MRT collision operator with $m = 10^8$ produces solutions with high accuracy and stability for all of the cases examined herein, but at an increased (approximately 5-10 times more) computational cost.
-\end{itemize}
-
-\subsubsection{Power-law Fluids}
-
-For the power-law numerical tests the Reynold's number was varied: $[100, 1000, 5000, 10000]$, and the flow behavior index, $n$, was varied: $[0.5, 1.5]$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
-The collision schemes that were tested were: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT with $m = 10^8$ and (5) MRT with $m = 10^8$ and median filtering.
-\Fref{tab:lid-powerlaw05} and \Fref{tab:lid-powerlaw15} compare the center location of the main vortex to literature values taken from~\citet{li2014simulation}.
-
-%MG: two values still need filled in in these tables
-\begin{table}
-\centering
-\caption{Power-law, lid-driven cavity flow; $n = 0.5$.}
-\small
-\vspace{0.5cm}
-\begin{tabulary}{\linewidth}{r r l l r r r}
-$n$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
-\hline \\
-0.5 & 100 & BGK     & No  & (0.72, 0.78) & (0.71, 0.77) & 21429 \\
-    &     & BGK-BRT & No  &              & (0.71, 0.77) & 14547 \\
-    &     & BGK     & Yes &              & (0.72, 0.78) & 33221 \\
-    &     & MRT     & No  &              & (0.71, 0.77) & 77287 \\
-    &     & MRT     & Yes &              & (0.71, 0.77) & 110403 \\
-\\
-0.5 & 1000 & BGK     & No  & (0.58, 0.55) & - & - \\
-    &      & BGK-BRT & No  &              & - & - \\
-    &      & BGK     & Yes &              & (0.53, 0.59) & 24935 \\
-    &      & MRT     & No  &              & (0.53, 0.54) & 72102 \\
-    &      & MRT     & Yes &              & (0.53, 0.54) & 78300 \\
-\\
-0.5 & 5000 & BGK     & No  & (0.53, 0.52) & - & - \\
-    &      & BGK-BRT & No  &              & - & - \\
-    &      & BGK     & Yes &              & (0.63, 0.68) & 23517 \\
-    &      & MRT     & No  &              & - & - \\
-    &      & MRT     & Yes &              & (0.51, 0.58) & 54225 \\
-\\
-0.5 & 10000 & BGK     & No  & (0.53, 0.52) & - & - \\
-    &       & BGK-BRT & No  &              & - & - \\
-    &       & BGK     & Yes &              & (0.50, 0.55) & 22603 \\
-    &       & MRT     & No  &              & - & - \\
-    &       & MRT     & Yes &              & (0.50, 0.54) & 63892 \\
-\\
-\end{tabulary}
-\label{tab:lid-powerlaw05}
-\end{table}
-
-\begin{table}
-\centering
-\caption{Power-law, lid-driven cavity flow; $n = 1.5$.}
-\small
-\vspace{0.5cm}
-\begin{tabulary}{\linewidth}{r r l l r r r}
-$n$ & $Re$ & \pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & \pbox{20cm}{Vortex \\ Center \\ (literature)} &  \pbox{20cm}{Vortex \\ Center \\ (LBM)} & \pbox{20cm}{Time \\ (sec)} \\
-\hline \\
-1.5 & 100 & BGK     & No  & (0.56, 0.73) & (0.57, 0.73) & 10897 \\
-    &     & BGK-BRT & No  &              & (0.57, 0.73) & 6239 \\
-    &     & BGK     & Yes &              & (0.56, 0.73) & 11847 \\
-    &     & MRT     & No  &              & (0.56, 0.73) & 39017 \\
-    &     & MRT     & Yes &              & (0.56, 0.73) & 96033 \\
-\\
-  1.5 & 1000 & BGK   & No  & (0.55, 0.64) & (0.54, 0.61) & 19764 \\
-    &      & BGK-BRT & No  &                & (0.54, 0.61) & 11949 \\
-    &      & BGK     & Yes &                & (0.54, 0.61) & 24516 \\
-    &      & MRT     & No  &                & (0.54, 0.61) & 58714 \\
-    &      & MRT     & Yes &                & (0.54, 0.61) & 69884 \\
-\\
-  1.5 & 5000 & BGK     & No  & (0.53, 0.61) & (0.52, 0.57) & 20147 \\
-    &      & BGK-BRT & No  &              & (0.52, 0.57) & 12116 \\
-    &      & BGK     & Yes &              & (0.52, 0.57) & 23140 \\
-    &      & MRT     & No  &              & (0.53, 0.57) & 60584 \\
-    &      & MRT     & Yes &              & (0.52, 0.58) & 58519 \\
-\\
-  1.5 & 10000 & BGK     & No  & (0.51, 0.55) & (0.53, 0.55) & 21570 \\
-    &       & BGK-BRT & No  &              & (0.53, 0.55) & 12841 \\
-    &       & BGK     & Yes &              & (0.53, 0.55) & 21570 \\
-    &       & MRT     & No  &              & (0.49, 0.56) & 61453 \\
-    &       & MRT     & Yes &              & (0.49, 0.55) & 67546 \\
-\\
-\end{tabulary}
-\label{tab:lid-powerlaw15}
-\end{table}
-
-For the shear-thinning results, $n = 0.5$, at the lowest Reynold's number considered, $Re = 100$, all of the collision schemes produced results that agreed well with the literature value.
-However, the solutions became unstable at higher Reynold's numbers and entropic filtering was necessary to produce a stable soltuion when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
-Although there still appears to be somewhat of a tradeoff between computational efficiency and accuracy in regards to the BGK with median filtering and MRT with median filtering, it does not seem to be nearly as drastic as when simulating Bingham plastic lid-driven flow; e.g.
-neither collision scheme consistently agrees with the literature values; and the MRT with median filtering is only 2-3 times slower than the BGK with median filtering.
-A good example of this tradeoff is the shear-thinning results for $Re = 5000$: the location of the main vortex for the BGK with median filtering differed from literature by more than 10\% in both the $x$ and $y$ directions but the MRT with median filtering has approximately twice the computational cost.
-
-For the shear-thickening results, $n = 1.5$, there were no issues of stability.
-What is interesting is that all of the collision schemes produce similar results for each of the cases, with the difference in the location of the main vortex being no greater than 1\% in either the $x$ or $y$ direction for any two collision schemes (with the exception of the high Reynold's number case, $Re = 10000$).
-The BGK-BRT consistently needed the least amount of computation time.
-The reason the BGK-BRT may be the most efficient is that because the collision frequency, $\omega$, is bounded, and consequently $\mu_{app}$ is bounded, meaning the solution to the constitutive equation may be converging to a bound with less iterations than the other collision schemes.
-
-\section{Conclusions} \label{sec:numerical-study-conclusions}
-
-A numerical investigation into the accuracy, stability, and efficiency of LBM collision models when applied to non-Newtonian flow was presented.
-The numerical investigation included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
-Two different benchmark problems were chosen for the study: Poiseuille flow, and lid-driven square cavity flow.
-The results showed that:
-\begin{itemize}
-  \item For high yield stress fluids in Poiseuille-type flow, only the MRT collision operator did not suffer from nonphysical oscillations.
-  \begin{itemize}
-      \item The oscillations appeared to be due to high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
-      \item If a collision scheme is to be developed for high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, then it should focus on dampening the nonequilibrium energy flux moment.
-    \end{itemize}
-  \item Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
-  \item In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
-  \item To summarize Bingham plastic, lid-driven cavity flow,
-    \begin{itemize}
-  \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
-    \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable.
-    \end{itemize} \pagebreak
-  \item To summarize power-law, lid-driven cavity flow,
-    \begin{itemize}
-      \item for shear-thinning fluids at high Reynold's numbers ($Re \ge 5000$) median filtering is necessary to produce a stable solution,
-  \item for shear-thickening, lid-driven cavity flow, there is much less concern for numerical instability.
-    \end{itemize}
-\end{itemize}
-
-% MG: should we mention: code is open source? input files used in this study are also freely available?
+\input{src/stability-paper.tex}
 
@@
 
-\section*{Abstract}
-
-When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
-%Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
-An important yet vulnerable barrier is the cement annulus and is created through a process known as primary cementing.
-Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
-%Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
-Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
-The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flow, multiphase multicomponent flow, and flow through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
-In \Fref{sec:numerical-methods}, an algorithm for simulating non-Newtonian free-surface flow using LBM is presented.
-The algorithm was implemented and used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
-More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
-The study is followed by conclusions and a discussion of future work.
-
-\section{Introduction}
-
-% what is primary cementing?
-
-Of critical importance to drilling operations of all kinds is that barriers are placed to prevent gasses and fluids from migrating from one geological zone to another (i.e., zonal isolation is achieved)~\cite{economides19901,jackson1993effect,goodwin1992cement,dusseault2014towards,li2016theory}.
-If zonal isolation is not achieved, gasses and fluids may migrate through the well to the surface and up into the atmosphere--potentially causing pollution, or at the very least, contributing to an increase in greenhouse gasses--or, worse yet, formation gasses and fluids can migrate into aquifers, potentially harming wild life and people.
-Furthermore, it can be argued that the most important, yet vulnerable, barrier in terms of achieving zonal isolation of a wellbore is the cement annulus that is created between the casing and the formation.
-There are various mechanisms by which this cement annulus can fail.
-According to previous studies~\cite{Bit02,Che85,Lev79,Ste88,Tal93,Zul12,li2016theory}, the most prevalent failure mechanisms of the cement annulus are:
-\begin{itemize}
-\item Stresses developing in the cement annulus as a result of temperature gradients, moisture gradients, overburden pressure, and chemical shrinkage of the cement matrix.
-These stresses can cause cracking in the cement, or debonding at either the casing--cement or cement--formation interfaces.
-\item Poor fill of the annulus.
-If cement slurry does not fill the annulus and if drilling mud is left behind the mud can dry, crack, and provide a weak path for fluids and gasses under pressure to push through, or if a dry hole is cemented and not filled properly, fluids and gasses can travel through channels and voids left in the annular space.
-\item Gas channeling through the annulus during cement curing.
-Initially, after the annulus is poured, the cement column transmits its full hydrostatic pressure against the rock formation.
-As the cement annulus hydrates, it solidifies, becoming stiffer and more self-supporting.
-When the column begins to support itself, the hydrostatic pressure of the cement column--against the rock formation--begins to drop.
-If the hydrostatic pressure of the cement column drops below the pressure of a formation gas before the cement annulus has developed adequate strength to resist it, the formation gas can penetrate into the cement annulus and form channels, or degrade the cement annulus in other ways.
-\end{itemize}
-% what research has been done that is considered classical?
-
-% why do we need modelling?
-% Before the end of this paragraph--modeling is important because...
-
-In order to better understand, and thus improve primary cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
-For example,~\citet{Bei77} developed a mathematical model to describe the displacement of drilling mud by the the cement slurry during primary cementing.
-Among their assumptions were that the flow as laminar, the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the distance from the center of the annulus to either the rock formation or steel casing surface.
-The model was then used to explore the effect of various parameters on the displacement efficiency of the spacer, such as the density ratio between the mud and the spacer, the viscosity ratio between the mud and the spacer, the displacement rate, the displaced phase yield stress, and the displacing phase yield stress.
-It was concluded that displacement efficiency could be improved by increasing the density of the displacing phase, increasing the viscosity of the displacing phase, (in general) decreasing the displacement rate, decreasing the yield stress of the displaced phase, and increasing the yield stress of the displacing phase.
-\citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
-The model was more general than the model developed by~\citet{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
-The bulk fluid motion of the spacer and drilling mud during primary cementing were investigated.
-The intended use of the model was not to make general statements about what properties of the spacer and drilling mud were desirable, but to instead be used iteratively during the design process for the cement job.
-Results showed that the model was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
-\citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
-The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citet{Bei77} or~\citet{Bit02}.
-The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
-\citet{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
-The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
-\citet{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
-\citet{zhao2016lattice} used the model to investigate the shape of the interface between the displacing phase (i.e.
-cement slurry) and displaced phase (i.e.
-drilling mud) in a horizontal, eccentric annulus.
-It was found, in general, that a greater density ratio between the displacing phase and displaced phase and lower eccentricity results in better displacement in a horizontal, eccentric annulus.
-
-% talk about new LBM study
-
-Overall, past studies have tended to idealize the rock formation as a straight wall.
-Although this is a desirable simplification to make for capturing bulk fluid phenomena, it does not accurately describe what is happening locally at the surface of the rock formation.
-Flow near the surface of the formation, and in and around the imperfections of the formation geometry, could influence the bond of the cement annulus to the rock formation, and the fill of the cement slurry very near to the rock formation.
-If an adequate bond does not develop, debonding can occur as a result of thermal and mechanical stresses.
-Even if most of the annulus is filled by the cement slurry, a small gap at the cement--formation interface that stretches for a long enough span of the wellbore would still be susceptible to fluid and gas migration, i.e.
-loss of zonal isolation.
-One could even argue that the characteristic of the cement slurry flow is most important local to the cement--formation interface, and that considering imperfections in the rock formation is imperative to determining whether zonal isolation has been achieved.
-
-The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features.
-The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation.
-LBM also has other advantages as an approach for simulating primary cementing:
-\begin{itemize}
-\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
-Many fluids present in primary cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
-\item LBM is well-suited for simulating multiphase multicomponent flow and free-surface flow because it can handle complicated interface shapes between fluid phases and components.
-\item LBM is easily written in parallel.
-As hardware architectures shift from one or a few cores to several CPUs, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
-\end{itemize}
-The current work focuses on the problem of cementing a dry hole, i.e. an annulus that is not filled with drilling mud prior to cementing.
-In order to numerically investigate this problem, an LBM model for non-Newtonian and free-surface flow is developed.
-The details of the LBM model are outlined in \Fref{sec:numerical-methods}.
-What follows in \Fref{sec:results} is a preliminary study on how different characteristics of flow, e.g.
-cement slurry yield stress, cement slurry inlet velocity, etc., perform in regards to fill of the annulus for different wellbores with imperfect rock formation geometries.
-\Fref{sec:conclusion} outlines conclusions resulting from the numerical study and a discussion of future work.
-
-\section{Numerical Methods} \label{sec:numerical-methods}
-
-\subsection{Lattice Boltzmann Method}
-
-% We need to explain/justify why 2D vs. 3D
-
-For the sake of computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
-Idealizing the annulus as two-dimensional limits the modeling approach in that annulus eccentricity and other three-dimensional phenomena could not be considered.
-A two-dimensional representation of the annulus was deemed sufficient as the focus of this study is to establish the Lattice Boltzmann Method as a suitable method for simulating primary cementing in an annulus with an imperfect rock formation surface.
-In addition to using a D2Q9 lattice, the BGK collision operator used was in the current chapter.
-For more details on the fundamentals of the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.
-
-\subsection{Cement Slurry Constitutive Relationship}
-
-Cement slurry is typically modeled using a non-Newtonian constitutive relationship.
-%A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
-As is commonly utilized for modeling cement and concrete fluid dynamics~\cite{ferraris2001fresh,ferraris1998testing,tattersall2003workability,roussel2007computational,roussel2015numerical,svec2012free}, the Bingham plastic constitutive relationship was utilized herein.
-Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
-The Bingham plastic constitutive model is difficult to work with numerically.
-Therefore, a smooth approximation, specifically the Papastasiou approximation, is often used~\cite{papanastasiou1987flows,chen2014simulations,tang2011bingham,Cha11,wang2008lattice,phillips2011lattice} and is used herein.
-Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
-The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flow.
-For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.
-
-\subsection{Free Surface Flow}
-
-A free-surface flow is a flow in which two-fluids exist but the dynamics of one fluid does not need to be explicitly modeled.
-Typically, the phase of the primary fluid, or the fluid of interest, is a liquid; and the secondary phase that does not need to be explicitly modeled is a gas, often times air at atmospheric pressure.
-The primary fluid is modeled using a standard CFD method, such as LBM, and the interaction of the secondary fluid on the primary fluid is modeled as a boundary condition at the interface between the two fluids.
-The interface is what is referred to as the free-surface. \pagebreak
-The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
-The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice} and \citet{thurey2005interactive}.
-
-\subsection{Capturing the Free-Surface}
-
-Consider at each node a cell of size $\text{d}x \times \text{d}x$ where $\text{d}x$ is the distance between nodes and the cell is centered at the node.
-In order to track where the primary fluid is and where it is not (i.e.
-which cells contain the primary fluid), a variable is introduced, $\epsilon$, which describes what fraction of the cell is covered by the primary fluid.
-Cells that have a fluid fraction of $\epsilon = 0$ contain no primary fluid (i.e.
-are empty); cells that have a fluid fraction of $\epsilon = 1$ are filled; and cells that have a fluid fraction $0 < \epsilon < 1$ are neither completely filled nor are they completely empty.
-The fluid fraction concept is similar to how fluids are tracked in Volume of Fluid (VOF) methods.
-
-\subsubsection{Mass Transfer}
-
-It is important to note that the particle distributions, $f_0, ..., f_8$, cannot be used to directly determine the fluid fraction of a cell.
-Neither can the zeroth velocity moment of the particle distributions, $\rho = \sum_i f_i$.
-For the algorithm used in this work, $\epsilon(\pos, t) \neq \epsilon(\mathbf{f(\pos, t)}, \rho(\pos, t))$.
-Instead, each $f_i$ is related to the amount of primary fluid advection in each of the $\pvel_i$ discrete velocity directions.
-To track the fluid fraction and how it changes, another variable is introduced, $M(\pos, t)$, which denotes the (primary) fluid mass in each cell.
-The fluid mass is related to the fluid fraction by $\epsilon = M / \rho$.
-The change in mass, $\Delta M(\pos, t)$ is related to $\mathbf{f}(\pos, t)$ and particle distribution in the neighborhood of $\pos$.
-Given initial conditions on $M(\pos, 0)$ and a fixed time step size $\Delta t $, the fluid mass is determined by $M(\pos, t_n) = M(\pos, 0) + \Delta t \sum_{k=1}^n \Delta M(\pos, t_k)$.
-What remains to track the fluid mass is to develop an expression for $\Delta M(\pos, t)$.
-
-\newcommand{\opidx}{{\bar{\i}}}
-\newcommand{\fset}{\mathcal{F}}
-\newcommand{\iset}{\mathcal{I}}
-\newcommand{\gset}{\mathcal{G}}
-\newcommand{\domain}{\mathcal{X}}
-
-\subsubsection{Cell States} \label{sec:mass-transfer}
-
-Each of the cells in the domain are assigned one of three states: fluid, interface, or gas.
-The set of fluid, interface, and gas cells are denoted by $\fset$, $\iset$, and $\gset$, respectively.
-A fluid cell is a cell that is entirely covered by the primary fluid; and a gas cell is a cell that is completely empty, or does not contain any primary fluid.
-An interface is partially covered, and there is always a layer of interface cells between fluid and gas cells so that a fluid cell and a gas cell are never neighbors.
-The change in mass is calculated for fluid and interface cells (i.e. $\pos \in \fset \cup \iset$) for each discrete velocity direction as follows:
-\begin{equation} \label{eq:dM}
-\Delta M_i(\pos, t) = \begin{cases}
-f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t), & \pos \in \fset \text{ or } \pos + \pvel_i \Delta t \in \fset \\
-%\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
-\frac{\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t}{2} (f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t)), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
-0, & \pos + \pvel_i \Delta t \in \gset
-\end{cases}
-\end{equation}
-\noindent where $\pvel_\opidx$ represents the discrete velocity vector that is opposite in direction to $\pvel_i$ (i.e. $\pvel_\opidx = - \pvel_i$).
-\eqref{eq:dM} states that when the border of the two cells is covered with fluid, the change in mass for a discrete velocity direction is equal to the difference of the particle distribution entering the cell from the opposite direction and particle distribution leaving the cell in that direction.
-When the border of the two cells are not covered with fluid (i.e.
-when both cells are interface cells) the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the cell border that is covered in fluid.
-The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M_i(\pos, t)$.
-Finally, note that $\Delta M_i(\pos, t) = -\Delta M_\opidx(\pos + \pvel_i \Delta t, t)$ for all $\pos \in \domain$ where $\domain$ is the domain.
-As a consequence, it is clear that $\sum_{\pos \in \domain}\Delta M(\pos, t) = 0$ (i.e. that mass is conserved for the mass transfer step).
-
-\newcommand{\tconst}{\delta_{trans}}
-
-Tracking the transport of mass throughout the domain is an important step in determining what parts of the domain contain the primary fluid and what do not.
-However, for the free-surface to change shape and propagate, cells need to be able to transition state.
-Cell states change based on the rule set presented in \Fref{tab:cell-transition-rules} where $\tconst$ is a small, computational constant that keeps cells from transitioning back and forth between fluid and interface, or interface and gas states in subsequent time steps.
-To summarize \Fref{tab:cell-transition-rules}, gas cells cannot transition to fluid cells and vice versa, interface cells transition to fluid cells when their mass has exceeded $\rho$ by a computational constant, and gas cells when their mass is less than zero minus a computational constant.
-In addition, fluid cells and gas cells transition to the interface state in order to maintain continuity of the interface, i.e.
-to ensure there is always a layer of interface cells in between fluid cells and gas cells.
-An example of a cell transitioning state is depicted in \Fref{fig:cell-transition}.
-
-\begin{table}
-\caption{Cell state transition rules.}
-\begin{tabulary}{\linewidth}{l|L|L}
-\textbf{Current State} & \textbf{Condition} & \textbf{Transition} \\
-\hline & \\
-$(\pos, t) \in \iset$ & $M(\pos, t) < -\tconst$ & $(\pos, t + \Delta t) \rightarrow \gset$, $M(\pos, t + \Delta t) = 0$ \\
-                      & $M(\pos, t) > \tconst + \rho(\pos, t)$ & $(\pos, t + \Delta t) \rightarrow \fset$, $M(\pos, t + \Delta t) = \rho(\pos, t + \Delta t)$ \\
-                      & &\\
-$(\pos, t) \in \fset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \gset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
-& & \\
-$(\pos, t) \in \gset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \fset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
-\end{tabulary}
-\label{tab:cell-transition-rules}
-\end{table}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{figs/cell-transition}
-\caption{Example of cell state transitions.
-The interface cell highlighted in red on the left is transitioned to a gas cell as it has emptied.
-On the right, the three cells highlighted in red were transitioned from fluid to interface cells in order to keep the interface continuous.}
-\label{fig:cell-transition}
-\end{figure}
-
-\subsubsection{Updating Cell States and Mass Redistribution} \label{sec:cell-updates}
-
-The rules for updating cell states in \Fref{tab:cell-transition-rules} are fairly intuitive, however, there are some details that must be considered in order to ensure mass is properly conserved and interface continuity is achieved (e.g.
-note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells).
-The algorithm for updating cell states follows that which was presented in~\citet{thurey2005interactive}.
-It is presented here as \Fref{algo:update-cell-states} for the sake of completeness and for outlining minor modifications.
-
-\input{src/algorithm/update-cell-states.tex}
-
-\Fref{algo:update-cell-states}, as presented, ensures both mass conservation and continuity of the free-surface.
-Moreover, when mass redistribution inevitably occurs, it is first attempted to be redistributed and weighted in the direction in which the interface is propagating (e.g.
-line \algref{algo:update-cell-states}{line:redist-dop-f} or line \algref{algo:update-cell-states}{line:redist-dop-g}.
-If there are no interface cells in the direction of interface propagation, it is instead redistributed to neighboring interface cells (e.g.
-line \algref{algo:update-cell-states}{line:redist-neighbors-f} or line \algref{algo:update-cell-states}{line:redist-neighbors-g}).
-If there is no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells (e.g.
-line \algref{algo:update-cell-states}{line:final-redist}).
-The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...
-Lastly, note that the interface normal, $\mathbf{n}$, is calculated using a finite difference approximation,
-\begin{equation} \label{eq:n}
-\mathbf{n}_{ij} = \frac{1}{2} \begin{Bmatrix}
-\epsilon_{i-1, j} - \epsilon_{i+1, j} \\
-\epsilon_{i, j-1} - \epsilon_{i, j+1}
-\end{Bmatrix}
-\end{equation}
-\noindent where $i$ is the index of the node in the x-direction and $j$ is the index of the node in the y-direction.
-
-\subsection{Boundary Conditions at the Free-Surface} \label{sec:bc-at-fs}
-
-Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and how particle distributions that are missing from the streaming step are reconstructed.
-The degrees of freedom, $f_i$, are not tracked for gas cells.
-After the streaming step, interface cells will be missing particle distributions that would have streamed from gas cells--an issue that is depicted in \Fref{fig:ps}.
-
-\begin{figure}
-\centering
-\includegraphics{figs/ps}
-\caption{Example of a scenario in which particle distributions are missing post-streaming.}
-\label{fig:ps}
-\end{figure}
-
-The missing particle distributions need to be reconstructed.
-Also, momentum needs to be conserved at the interface.
-When the particle distributions are reconstructed, they will be reconstructed in such a way that momentum is conserved between the primary fluid and the secondary fluid.
-In order to perform the reconstruction, there are a few assumptions that need to be made:
-\begin{enumerate}
-\item The velocity of the secondary fluid is equal to the velocity of the primary fluid at the free-surface, i.e. there is no-slip between the two-fluids.
-\item The secondary fluid is at equilibrium and at constant pressure, $p_G$.
-\item There is a force balance for opposite lattice directions.
-\end{enumerate}
-Because particle distributions are a measure of momentum transport, the pressure of the secondary fluid can be converted to the particle distribution scale and the missing particle distribution can be solved for algebraically.
-Consider \Fref{fig:fbi}.
-Balancing the opposite lattice directions results in:
-\begin{eqnarray}
-f_i + f_\opidx &=& f^{eq}_\opidx(\rho_G, \mathbf{u}) + f_i^{eq}(\rho_G, \mathbf{u}), \\
-\label{eq:reconst} f_\opidx &=& f_\opidx^{eq}(\rho_G, \mathbf{u}) + f_i^{eq}(\rho_G, \mathbf{u}) - f_i.
-\end{eqnarray}
-Missing distribution functions can be reconstructed according to \eqref{eq:reconst}.
-In order for conservation of momentum to be satisfied along the free-surface, distribution functions coming from the direction of the interface normal, $\mathbf{n}$, must also be reconstructed.
-Thus, distribution functions are reconstructed if coming from a gas cell, or if $\mathbf{n} \cdot \pvel_\opidx > 0$.
-
-\begin{figure}
-\centering
-\includegraphics{figs/force-bal-at-interface}
-\caption{Particle distributions at free-surface for a pair of opposing lattice directions after gas pressure, $\rho_G$, is in the particle distribution form.}
-\label{fig:fbi}
-\end{figure}
-
-\subsection{Resulting Algorithm for Simulating Free-Surface Flow using the Lattice Boltzmann Method}
-
-For clarity, the complete algorithm used in the current work for simulating free-surface flow using the Lattice Boltzmann Method is presented in \Fref{algo:free-surface-complete}.
-
-\input{src/algorithm/free-surface-complete.tex}
-
-\newcommand{\nodesi}{50}
-\newcommand{\nodesj}{500}
-\newcommand{\anw}{$1\frac{1}{4}$"}
-\newcommand{\anh}{1'}
-\newcommand{\nsteps}{16000}
-
-\section{Numerical Study of Primary Cementing} \label{sec:results}
-
-The free-surface LBM model was implemented and used to simulate primary cementing in a dry annulus, i.e. an annulus without drilling mud.
-The focus of the current study was the performance of different characteristic flow behaviors (as could potentially relate to different cement slurry mixes) when used to cement different imperfect wellbore geometries.
-Because the focus in regards to the wellbore is on geometrical imperfections, the entire wellbore annulus is not simulated, but instead a small section around a geometrical imperfection is considered.
-A schematic of an example primary cementing simulation, including boundary conditions, is shown in \Fref{fig:wellbore-cementing-schematic}.
-
-\begin{figure}
-\centering
-\includegraphics{figs/wellbore-cementing-schematic}
-\caption{Example simulation used in primary cementing study.
-The black rectangles on the left boundary represent notches protruding from the rock formation surface. The contours are of the primary fluid (cement slurry) mass.
-The schematic on the right is courtesy of~\cite{cementing2015web}}.
-\label{fig:wellbore-cementing-schematic}
-\end{figure}
-
-%The simulations were carried out on a $\nodesi \times \nodesj$ lattice, which corresponds to an annulus width of \anw \hspace{0.05cm} and a sectional height of \anh.
-The simulations were carried out on a $\nodesi \times \nodesj$ lattice.
-For the mass inlet, a~\citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e.
-$M(\pos_j, t) = M_0 = 1.0$ for all $\pos_j$ along the inlet.
-This corresponded to a constant mass flow, $\dot{M}_{in} = M_0 u_0$, at inlet.
-For the outlet, mass was allowed to transfer out of the domain during the mass transfer step such that $f_\opidx(\pos + \pvel_i \Delta t, t) = 0$ in \eqref{eq:dM} for all $\pos + \pvel_i \Delta t$ that lie outside of the domain on the other side of the mass outlet (i.e.
-the boundary condition at the outlet is such that there is no backflow).
-All simulations were executed for $\nsteps$ time steps.
-
-The different imperfect geometries that were considered at the rock formation surface were: flow over a cavity that is 30\% of the annulus width, ``cav $30 \times 1$''; flow over a cavity that is 60\% of the annulus width, ``cav $60 \times 1$''; flow over a square obstacle with sides that are 30\% of the annulus width, ``obs $30 \times 1$''; flow over two square obstacles with sides that are 30\% of the annulus width, ``obs $30 \times 2$''; flow over a square obstacle with sides that are 60\% of the annulus width, ``obs $60 \times 1$''; flow over two square obstacles with sides that are 60\% of the annulus width, ``obs $60 \times 2$''; and flow over a rectangular obstacle with width that is 60\% of the annulus width and length that is four times the width, ``obs $60 \times L$''.
-The wellbore geometries that were considered herein are depicted in \Fref{fig:wellbore-geometries}.
-
-\begin{figure}
-\includegraphics[width=\linewidth]{figs/wellbore-geometries}
-\caption{Wellbore geometries considered in the study.}
-\label{fig:wellbore-geometries}
-\end{figure}
-
-In addition to wellbore geometry, different cement slurry material properties and cement slurry flow rates were considered, which are characterized by the dimensionless Reynold's number, $Re = \frac{\rho U L}{\mu_p}$, and Bingham number, $Bn = \frac{\tau_y L}{\mu_p U}$, of the flow, where $L \equiv $ the width of the annulus.
-A particular goal of the study was to investigate which Reynold's numbers and Bingham numbers were preferable for each geometry, and which Reynold's numbers and Bingham numbers performed well for all geometries.
-The dimensionless numbers were defined as follows, $(Re, Bn) = (1.00, 4.00)$,  $(2.50, 0.00)$, $(2.50, 1.00)$, $(2.50, 10.0)$, $(2.50, 25.0)$, $(5.00, 5.00)$, and $(12.5, 2.0)$.
-\Fref{fig:percent-filled} shows the percentage of the wellbore annulus section filled at the end of the cement simulation duration.
-The percent fill was calculated as:
-\begin{equation}
-  \frac{\sum\limits_{\pos \in \domain} \epsilon(\pos, 16000)}{\sum\limits_{\pos \in \domain} 1.0}.
-\end{equation}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{figs/percent_filled}
-\caption{Percent of voids in wellbore at the end of simulation for each cement slurry flow through each wellbore geometry.}
-\label{fig:percent-filled}
-\end{figure}
-
-Based upon \Fref{fig:percent-filled}, some general statements can be made about what properties of the cement slurry flow are preferable.
-For example, low yield stress, and consquently low Bingham number flows ($Bn \le 1.0$), approximately completely filled the annulus for every wellbore geometry considered.
-In addition, the negligible percentage of voids for the ($Re = 5.00, Bn = 5.00$) and ($Re = 12.5, Bn = 2.0$) slurry flows suggests that a if the ratio of $Bn / Re$ is low, that the slurry flow will result in a low percentage of voids for cementing a dry hole as is desired.
-As $Bn / Re$ increases, it appears that, in certain cases and wellbore geometries, the slurry flow tends to result in a larger percentage of voids (approximately 10\%-40\%). \pagebreak
-What is interesting to note is that for slurry flows that resulted in a percentage of voids greater than 3\%, there was a lack of consistency across wellbore geometries in terms of which slurry flow resulted in more voids and which slurry flow resulted in less, which suggests that generalities cannot always be made and that computational methods are necessary for understanding and designing slurry mixes for real-life wellbores.
-
-\section{Conclusions} \label{sec:conclusion}
-
-An LBM model for simulating non-Newtonian and free-surface flow to represent cementing dry annuli in oil and gas wellbores was presented.
-The model was then used to study how different cement slurry flows, defined by different Reynold's numbers and Bingham numbers, would perform in different wellbore geometries.
-The results showed that in general, it is preferable that the yield stress and $Bn / Re$ ratio of the slurry flow be low.
-However, beyond these considerations, it is not clear what slurry flow may be preferable for a given rock formation geometry.
-The results suggest that computational methods are necessary for (a) better understanding the mechanics of primary cementing and (b) for designing a cement mix for a given wellbore.
-The lattice Boltzmann method extended for simulating non-Newtonian and free-surface flow as presented in the current work has several properties that make it a good candidate for such a design framework:
-\begin{itemize}
-\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
-\item LBM is well-suited for simulating multiphase multicomponent flow and free-surface flow because it can handle complicated interface shapes between fluid phases and components.
-\item LBM is easily written in parallel and as hardware architectures shifts sequential to parallel computing, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
-\item The LBM no-slip boundary condition works well with complex geometries, such as those that occur in realistic rock formation surfaces in drilled wellbores.
-\end{itemize}
-
-Future work would include developing a more physically realistic model.
-Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
-Possible future work would be make the model more general so as to be capable of simulating multiphase multicomponent flow.
-In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.
+\input{src/cementing-paper.tex}
 
@@
 Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
-Possible future work would be make the model more general so as to be capable of simulating multiphase multicomponent flow.
+Possible future work would be to make the model more general so as to be capable of simulating multiphase multicomponent flow.
 In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.
 
-%\bibliographystyle{abbrvnat}
-%\bibliography{master}
-
-% bullshit Pitt formatting
-\bibliographystyle{unsrt}
-\safebibliography{master}
+\bibliographystyle{abbrvnat}
+\bibliography{master}
 

### 2016-09-02 src/stability-paper.tex
revised figures and addressed comments for publication

@@
 
-The Lattice Boltzmann method (LBM) is a method based on computational statistical mechanics that is well-suited for complex flow such as non-Newtonian, free surface, and multiphase multicomponent flow.
-Non-Newtonian flow is the primary focus of this Chapter, as many practical engineering problems such as the flow of cement slurry and concrete, the filling of molds by molten metals and plastics, blood flow, etc., are best modeled as non-Newtonian fluids.
-LBM is typically applied to simulate flow through a series of time seps, each consisting of streaming particle distributions to neighboring nodes, and collisions of particle distributions at each node through a collision operator.
-The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated, e.g.
-constitutive laws, interfacial dynamics, etc., and it has implications on numerical stability and computational efficiency.
-In this Chapter, various collision operators and methods for stability enhancement were examined for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
-The investigation was carried out as a numerical study looking for qualitative, yet practical, results; it included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
+The Lattice Boltzmann method (LBM) is a numerical method based on computational statistical mechanics that is well-suited for approximating complex flow behaviors such as non-Newtonian, free surface, and multiphase multicomponent flow.
+LBM is typically applied to simulate flow through a series of time steps, each consisting of streaming particle distributions to neighboring nodes and collisions of particle distributions at each node through a collision operator.
+The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated (e.g.
+constitutive laws, interfacial dynamics, etc.) and it has implications on numerical stability and computational efficiency.
+This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
+The investigation was carried out as a numerical study looking for qualitative, yet practical, results, including testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
 Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
@@
 
-The dynamic viscosity of a fluid is a measure of its resistance to shear deformation.
-For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
-These fluids are known as Newtonian fluids.
-Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable even at a constant temperature and is often a function of strain-rate.
-There are a number of fluids in science and engineering applications that can be classified as non-Newtonian: pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.
-
+There are a number of fluids in science and engineering applications that can be classified as non-Newtonian, such as pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.
 Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized: yield stress fluids and power-law fluids.
 Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
-Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g.
-pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
+Yield stress flow has been found to be relevant in many applictions, such as flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
 %MG: citation is for blood as non-Newtonian
-As a consequence, simulating yield stress fluids can help engineers develop better pastes and paints, understand the flow and deformation of mud and clay in geotechnical engineering, understand blood circulation, and better manufacture metal and plastic parts.
-
-There other idealized non-Newtonian behavir, power-law behavior, is more commonly known as \emph{shear-thinning}--when the apparent viscosity decreases with increasing strain-rate--or \emph{shear-thickening}--when the apparent viscosity increases with increasing strain-rate.
-Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include polymer mixes and molten plastics.
-Shear-thickening fluids are also known as \emph{dilatants}, and some examples include quicksand, a cornstarch and water mixture, and a silica and polyethylene glycol mixture.
+
+Power-law behavior, is more commonly known as \emph{shear-thinning}, when the apparent viscosity decreases with increasing strain-rate, or \emph{shear-thickening}, when the apparent viscosity increases with increasing strain-rate.
+Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include tomato concentrate, clay and water mixtures, and molten plastics~\cite{irgens2014rheology}.
+Shear-thickening fluids are also known as \emph{dilatants}, and a well-known example is a cornstarch and water mixture.
 
@@
 Analytical solutions rarely exist for even the simplest non-Newtonian fluid flows because of the complexity that a nonlinear constitutive relationship entails.
-It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods.
-However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
-\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma},
-\end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
-Determining the apparent viscosity and strain-rate distribution of a flow over time will often require an iterative solution at every time step, a general Picard-type algorithm for such a process can be given as:
-
-\begin{enumerate}
-	\item Start with initial guess for the apparent viscosity, $\mu_k = \mu_0$.
-	\item \label{step:solve-for-flow} Solve for the flow using the current value of apparent viscosity, $\dot{\gamma}_k = \tau_k / \mu_k$.
-	\item Update the apparent viscosity, $\mu_{k+1} = \mu(\dot{\gamma}_k)$.
-	\item Return to Step \ref{step:solve-for-flow} until convergence is met.
-\end{enumerate}
-
-Numerical solutions work by discretizing the equations that govern the physics of interest.
-When the solution for fluid flow problems vary in space and time, a numerical approximation requires breaking the problem up into discrete locations and discrete time steps.
-The significance of approximating a solution by discretizing the governing equations is that the iterative solution for the constitutive equation must be solved at each discrete location for each discrete time step, which can become computationally expensive.
-The lattice Boltzmann method (LBM) is a numerical method for fluid flow that has the advantage that computing the strain rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
-This means that although an iterative solution is still required to determine the local strain rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
+It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods~\cite{syrakos2014performance,owens2002computational,chai2011multiple}.
+
+The lattice Boltzmann method (LBM) is a numerical method for simulating fluid flow that has the advantage (among others) that computing the strain-rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
+This means that although an iterative solution is still required to determine the local strain-rate and apparent viscosity, each iterative solution can be done in parallel, by a separate process, as they are independent of each other.
 Because hardware architectures have shifted from single, sequential processing systems to parallel processing systems, the local nature of the stress--strain-rate relationship in LBM gives it a distinct advantage for simulating non-Newtonian flow over some other numerical methods.
-
-The lattice Boltzmann method has been studied and successively applied to modeling various flows of non-Newtonian fluids.
-For example, ~\citet{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
+For example,~\cite{tang2011bingham,chai2011multiple,fallah2012multiple,chen2014simulations,vikhansky2008lattice,wang2008lattice} developed LBM models for simulating yield stress flow.
 The LBM model results agreed well when compared to analytical solutions for Bingham plastic Poiseuille flow and values from literature for lid-driven cavity flow, which shows the feasibility of using LBM models for yield-stress fluids.
-LBM models for power-law fluid flow~\citet{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\citet{ashrafizaadeh2009comparison}, have also been successfully developed and verified.
-%MG: does citing this way appear too dense?
+Examples of LBM models for power-law fluid flow include~\cite{wang2011lattice,wang2015localized,boyd2006second,chai2011multiple}, and blood flows using the K-L, Casson, and Carreau-Yasuda constitutive relationships~\cite{ashrafizaadeh2009comparison}, which have also been successfully verified with benchmarks.
 
@@
 LBM can be considered as a type of finite-difference scheme for the continuous Boltzmann equation, and as such, has numerical properties in common with finite-difference schemes.
-One such consideration associated with this view is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
+One such consideration is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
 %MG: are all relevant sources cited?
@@
 %MG: be more precise than 'fluctuations'?
-Various schemes and strategies for incorporating the physics of non-Newtonian flow with LBM and yet maintaining a stable numerical method have been developed and studied.
-The simplest approach for simulating a shear-rate dependent viscosity, used in~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}, is to make the collision frequency, which is proportional to apparent viscosity, variable and dependent on the local strain rate.
-A potential issue with the stability of the variable relaxation time approach is that as the collision frequency approaches $2$ the viscosity approaches zero and overrelaxation occurs; and alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method also degrades~\cite{latt2007hydrodynamic}.
-%MG: verify this statement
+Various strategies for incorporating the physics of non-Newtonian flow with LBM while maintaining a stable numerical method have been developed and studied.
+The simplest approach for simulating a shear-rate dependent viscosity is to make the collision frequency, which is proportional to apparent viscosity, variable and dependent on the local strain-rate~\cite{boyd2006second,chen2014simulations,fallah2012multiple,tang2011bingham,svec2011flow,svec2012free,zhao2016lattice}.
+A potential issue with the stability of the variable relaxation time approach is that as the collision frequency approaches $2$ the viscosity approaches zero and overrelaxation occurs.
+Alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method degrades~\cite{latt2007hydrodynamic}. %MG: verify this statement
 In order to ensure that the variable collision frequencies did not approach values leading to numerical instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable collision frequencies.
-Although bounding the collision frequency was shown to be effective in terms of stability, it is also nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the collision does not reflect the proper constitutive relationship of interest.
+Although bounding the collision frequency was shown to be effective in terms of stability, it is nonphysical and can lead to approximations that are inaccurate, not because of round-off error or numerical instability, but because the collision does not reflect the proper constitutive relationship of interest.
 Another scheme for incorporating non-Newtonian effects into LBM is to use a constant collision frequency, typically unity, and to instead incorporate the local shear-rate effect through equilibrium distribution functions.
 This means particle distribution functions will always relax toward equilibrium at the same rate, but that the definition of equilibrium is modified to represent the correct stress--strain-rate relationship.
-The equilibrium distribution function is derived for the specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.
+The equilibrium distribution function can be derived for the specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.
 The equilibrium distribution functions for Bingham plastic fluids, and for power-law and Carreau fluids were derived, implemented, and verified in~\cite{wang2008lattice} and~\cite{yoshino2007numerical}, respectively.
-The strategy of using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach zero, and will not result in overrelaxation, and the collision frequency will not reach values much greater than one, and so there is no reason to bound the collision frequency in a way that is nonphysical.
-\citet{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
-The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum, i.e.
-as an external forcing term, that is dependent on local shear-rate.
-Although the constant-collision frequency strategies present interesting alternatives, the variable collision scheme is used in the present study because of its simplicity and generality, i.e.
-it can be easily fitted to any constitutive equation without for example, performing Chapman-Enskog multiscale expansion.
-
-To improve upon the stability of LBM models,~\citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
-\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
-Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
-\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
+Using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach values that lead to overrelaxation (e.g.
+$2$) or underrelaxation (e.g.
+$\approx 0$).
+\cite{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
+The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum (i.e.
+as an external forcing term) that is dependent on local shear-rate.
+Although the constant-collision frequency strategies present interesting alternatives, the variable collision frequency scheme is used in the present study because of its simplicity and generality, such as the fact that a variable collision frequency scheme can be fit to any constitutive relationship without performing Chapman-Enskog multiscale expansion.
+
+%The Bhatnagar, Gross, Krook (BGK) collision operator~\cite{bhatnagar1954model}.
+The Multiple-relaxation-time (MRT) collision operator ~\cite{d1994generalized} is another approach that has been developed to improve upon the stability of LBM.
+The MRT collision takes place in moment space and allows each moment to relax at a different rate.
+\cite{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model for fluid flow, and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
+Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the cost may increase significantly for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain-rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
+\cite{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
 However, what remains unclear is:
@@
     \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flow?
-    \item Under what conditions, e.g. material parameters, physical problem, etc., for Bingham plastic fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
-    \item Under what conditions, e.g. material parameters, physical problem, etc, for power-law fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
+    \item Under what conditions (e.g. material parameters, physical problem, etc.) for Bingham plastic fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
+    \item Under what conditions (e.g. material parameters, physical problem, etc.) for power-law fluid flow is the MRT collision operator necessary to maintain stability and/or accuracy?
     \item What are additional strategies for increasing stability and accuracy, and what are their associated computational costs?
@@
 In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
-\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
-Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in simulating the flow of non-Newtonian fluids.
-
-The goal of this Chapter is to numerically study the implications of accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flow using LBM.
+\cite{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
+Stability enhancement through artificial dissipation and entropic filtering has shown promise, but to the authors' knowledge has not yet been tested for use in simulating the flow of non-Newtonian fluids.
+
+The goal of the present study is to numerically study the implications on accuracy, stability, and efficiency for some of the different strategies for simulating non-Newtonian flow using LBM.
 The intention of the study is to aid scientists and engineers in understanding which strategy is best suited to their priorities and applications of interest so as to maximize the advantages LBM has in simulating non-Newtonian flow.
@@
 A numerical study can help to determine approximate numerical values, domains, and boundary conditions in which one LBM scheme may be more advantageous than another so that LBM may be used in a computationally efficient and stable manner.
+In \Fref{sec:LBM}, an overview of the Lattice Boltzmann Method for simulating fluid flow is presented, along with the collision operators, stability enhancements, and constitutive relationships that were investigated.
+\Fref{sec:numerical-study} outlines the boundary conditions, material parameters, and LBM collision schemes for each simulation in the numerical study; and is followed by results and discussion.
+Finally, concluding remarks are given in \Fref{sec:numerical-study-conclusions}.
 
@@
 
-LBM has some advantages over other methods of CFD.
+As noted previously, LBM has some advantages over other methods of CFD.
 For example, LBM is a computationally efficient approach for some CFD.
@@
 \item Collision: particle distributions meet at a node and ``collide''.
-In LBM, collisions are not simulated in a realistic sense, meaning that each individual particle does not exist and glance off of, or interact with, one another.
-Instead the collision operator is formulated in such a way that particle distributions are relaxed toward equilibrium.
+In LBM, collisions are not simulated in a realistic sense, meaning that each individual particle does not exist and glance off of or interact with one another.
+Instead, the collision operator is formulated in such a way that particle distributions are relaxed toward equilibrium.
 What defines equilibrium depends on the mechanics of interest to be modeled.
@@
 
-\pagebreak
 The macroscopic variables of interest can be calculated from the particle distribution functions, $f(\pos, \pvel, t)$, by integrating moments of $f$ over velocity space.
@@
 
-\subsubsection{Bhatnagar-Gross-Krook (BGK)} \label{sec:bgk}
+\subsubsection{Bhatnagar-Gross-Krook} \label{sec:bgk}
 
@@
 In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
-This equilibrium is determined by the macroscopic physical behavior of interest.
+This equilibrium is determined by the macroscopic physical behavior of interest (e.g. diffusion, incompressible flow, compressible flow, etc.).
 In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq} = f_i^{eq}(\pos, t)$, is often given by:
@@
 
-\noindent where $\rho$ and $\mvel$ are dependent on $\pos$ and $t$, and $w_i$ is the weight in the $i^{th}$:%MG: maybe say something about where this comes from, e.g.
+\noindent where $\rho$ and $\mvel$ are dependent on $\pos$ and $t$, and $w_i$ is the weight in the $i$th:%MG: maybe say something about where this comes from, e.g.
 feq is a Hermite polynomial expansion of the Maxwell-Boltzmann distribution
@@
 For incompressible Newtonian flow, the collision frequency is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$; and from this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.
-The method for simulating non-Newtonian flow in the current work involves approximating a solution to the local value of apparent viscosity, $\mu_{app}(\pos, t)$, using \eqref{eq:non-newtonian} where $\tau$ and $\dot{\gamma}$ are dependent on $\pos$ and $t$, and then setting the value of the local collision frequency as follows:
+The method for simulating non-Newtonian flow in the current work involves approximating a solution to the local value of apparent viscosity, $\mu_{app}(\pos, t)$, where the shear stress, $\tau$, and strain-rate, $\dot{\gamma}$, are dependent on $\pos$ and $t$, and then setting the value of the local collision frequency as follows:
 \begin{equation} \label{eq:colfeq-non-newtonian}
@@
 Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
-Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations such as those depicted in \Fref{fig:overrelaxation}.
-Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to diverge or ``pollute'' the system enough to make the results highly nonphysical and altogether useless~\cite{gorban2014enhancement}.
-
-\begin{figure}
-	\centering
-    \includegraphics[width=\linewidth]{figs/overrelaxation_tau-0513}
-    \caption{Example of oscillations that can occur as a result of overrelaxation; the collision frequency, $\omega = 1.95$ and $\mu_{app} \approx 0.0043$}
-    \label{fig:overrelaxation}
-\end{figure}
+Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations.
+Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to diverge or ``pollute'' the system enough to make the results highly nonphysical~\cite{gorban2014enhancement}.
 
 The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
-%MG: this needs cleaned up, I think
-
 Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
-The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the apparent viscosity, $\mu(\dot{\gamma}(\pos))$, may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
+The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the local collision frequency may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
 
@@
 
-\subsubsection{Multiple-relaxation Time (MRT)}
+\subsubsection{Multiple-relaxation Time}
 
@@
 In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f} = \begin{Bmatrix}f_0 & f_1 & ...
-& f_8\end{Bmatrix}$, herein referred to as the ``moment space''.
+& f_8\end{Bmatrix}^T$, herein referred to as the ``moment space''.
 The collision is then performed in the moment space.
@@
 The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i \in [0, 2], i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
-In the case when $s_0 = s_1 = ... = s_8 = \omega$ ($\omega$ is the collision frequency in the BGK sense), the MRT collision operator is equivalent to the BGK collision operator.
+In the case when $s_0 = s_1 = ... = s_8 = \omega$, the MRT collision operator is equivalent to the BGK collision operator ($\omega$ is the collision frequency in the BGK sense).
 The relaxation parameters $s_0$, $s_3$, and $s_5$ are all set to zero as mass and momentum should be conserved.
@@
 
-To answer the first question, consider the discussion in \Fref{sec:bgk} on problems with overrelaxation and nonphysical oscillations.
+%To answer the first question, consider the discussion in \Fref{sec:bgk} on problems with overrelaxation and nonphysical oscillations.
 Nonphysical oscillations due to overrelaxation would be damped out more quickly if particle distributions ``far'' from quasiequilibrium were brought closer to quasiequilibrium.
-A particle distribution vector, $\mathbf{f}$, ``far'' from equilibrium would be a good candidate for artificial dissipation.
-There are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
+Thus, a particle distribution vector, $\mathbf{f}$, ``far'' from quasiequilibrium would be a good candidate for artificial dissipation.
+However, there are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
 A metric that has been developed and used successfully for determining when artificial dissipation should be introduced at a lattice site is the so called relative nonequilibrium entropy~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy}.
@@
 The number of standard deviations, $n_\sigma$, is chosen \textit{a priori}.
-The criteria described in \eqref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead can think in terms of what would be the maximum percentage of sites one would want artificial dissipation to be added to.
+The criteria described in \eqref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead considers the maximum percentage of sites one would want artificial dissipation to be added to.
 A disadvantage of \eqref{eq:dissipation-criteria} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artifical dissipation is unnecessary.
-By required that both \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} be met before artifical dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
-
-Just as there are many ways to measure a lattice site's distance from quasiequilibrium, or from that measure define criteria for artificial dissipation, there are also many different ways of deciding how much dissipation to add.
+By requiring that both \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} be met before artifical dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
+
+Just as there are many ways to measure a lattice site's distance from quasiequilibrium and define criteria for artificial dissipation, there are also many different ways of deciding how much dissipation to add.
 One method of adding dissipation is the so-called Ehrenfests' regularization~\cite{brownlee2006stabilization}, and involves setting a lattice site that is chosen for artificial dissipation to its quasiequilibrium state.
@@
 An alternative, the median filter, has been used successfully in conjunction with both the BGK and MRT collision schemes for simulating one-dimensional shock tubes and lid-driven cavity flow~\cite{brownlee2008nonequilibrium,gorban2014enhancement,packwood2009entropy}.
-Median filtering is an effective noise reduction technique, often used in image processing, for ``speckle noise'' or ``salt and pepper noise''~\cite{brownlee2008nonequilibrium}.
-In other words, median filters are good at reducing high frequency noise while having a minimal effect on lower frequency noise.
 In LBM this is a desirable way to introduce dissipation because it has the potential to reduce high frequency nonequilibrium fluctuations that might lead to numerical instability while retaining the lower frequency dynamics.
-\pagebreak
-To use the median filter one performs the collision step and then checks over the domain for lattice sites with $\Delta S$ that meet the criteria for artificial dissipation; sites that meet the criteria are updated as follows:
+To use the median filter one performs the collision step and then checks over the domain for lattice sites with $\Delta S$ that meet the criteria for artificial dissipation.
+Sites that meet the criteria are updated as follows:
 \begin{equation} \label{eq:median-filter}
@@
 
-\subsection{Boundary Conditions} \label{sec:bcs}
+\subsection{Boundary Conditions and Applied Forces} \label{sec:bcs}
 
@@
 
-\subsection{Applied Forces}
-
 Incorporating external forces, such as gravity, pressure gradients, etc., is done by adding a source of particle distributions in the direction of the force.
-The increase in particle distributions leads to the desired macroscopic result--an increase in momentum.
+The increase in particle distributions leads to the desired macroscopic result, which is an increase in momentum.
 The LBE with external forces is:
@@
 
-The Bingham plastic constitutive model is popular for yield stress flow.
 A Bingham plastic does not flow, i.e.
@@
 Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
-Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\citet{papanastasiou1987flows} is often used as:
+Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\citet{papanastasiou1987flows} was used herein, which is defined as:
 \begin{equation} \label{eq:papa}
@@
 The larger the value of $m$, the closer the approximation is to the Bingham plastic model.
-
 Alternatively, the constitutive relationship can be interpreted through the apparent viscosity.
@@
 
-The power-law constitutive relationship is useful for modeling fluids that experience shear-thinning or shear-thickening.
-The power-law relationship is given by:
+The power-law relationship between stress and strain-rate is given by:
 \begin{equation} \label{eq:pl}
@@
 \noindent where $k$ is the flow consistency index and $n$ is the flow behavior index.
-When $n = 1$, \eqref{eq:pl} results in a Newtonian constitutive relationship with dynamic viscosity, $\mu = k$.
-A flow consistency index of $n < 1$ results in shear-thinning behavior, where as $n > 1$ results in shear-thickening.
+When $n = 1$, \eqref{eq:pl} results in a Newtonian constitutive relationship with dynamic viscosity $\mu = k$.
+A flow consistency index of $n < 1$ results in shear-thinning behavior, whereas $n > 1$ results in shear-thickening.
 As with the Bingham plastic relationship, \eqref{eq:pl} can be modified to determine an apparent viscosity:
@@
 
-A numerical study was carried out in order to investigate the suitability of different LBM collision schemes for simulating non-Newtonian flow in terms of accuracy, numerical stability and computational efficiency.
-For all simulations presented in this section, the apparent viscosity is approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the maximum number of iterations set at 15 and the convergence criteria:
+A numerical study was carried out to investigate the suitability of the different LBM collision schemes discussed for simulating non-Newtonian flow in terms of accuracy, numerical stability and computational efficiency.
+For all simulations presented in this section, the apparent viscosity was approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the stopping criteria set as (whichever was first) a maximum number of iterations set at 15 and the convergence criteria:
 \begin{equation} \label{eq:mu-app-conv}
@@
 \end{equation}
-All BGK-BRT collision schemes use $\omega \in [0.05, 1.995]$ as the bounds on the collision frequency.
-All simulations with artificial dissipation use the median filter with $\theta = 1.0 \times 10^{-6}$ and $n_{\sigma} = 2.7$ where $\theta$ is the threshold that $\Delta S(\pos, t$ must exceed before dissipation is added \eqref{eq:ds-threshold} and $n_{\sigma}$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S(\pos, t)$ must be before dissipation is added \eqref{eq:dissipation-criteria}.
+All BGK-BRT collision schemes used $\omega \in [0.05, 1.995]$ as the bounds on the collision frequency.
+All simulations with artificial dissipation used the median filter with $\theta = 1.0 \times 10^{-6}$ and $n_{\sigma} = 2.7$ where $\theta$ is the threshold that $\Delta S(\pos, t)$ must exceed before dissipation is added \eqref{eq:ds-threshold} and $n_{\sigma}$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S(\pos, t)$ must be before dissipation is added \eqref{eq:dissipation-criteria}.
 All numerical values given in this section are in lattice units unless otherwise stated.
@@
 Poiseuille flow is a useful benchmark because analytical solutions exist for both Bingham plastic and power-law fluids.
-Poiseuille flow is driven by a constant pressure gradient, $\pgrad$, through a two-dimensional channel.
+Poiseuille flow is created with a two-dimensional channel flow that is driven by a constant pressure gradient, $\pgrad$, through the domain.
 A schematic is shown in \Fref{fig:poise-schematic}.
-No-slip boundary conditions are enforced at the top and bottom boundaries with a wall velocity of zero so that $\mvel \times \unorm = 0$ where $\unorm$ is the unit normal vector to the boundary. 
+No-slip boundary conditions are enforced at the top and bottom boundaries with a wall velocity of zero, so that $\mvel \times \unorm = 0$ where $\unorm$ is the unit normal vector to the boundary. 
 Periodic boundary conditions are enforced at the left and right boundaries.
@@
 Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps.
-The lattice size was chosen to be sufficiently fine for accuracy (although the length, $L = 32$, seems small, the effect is negligible as the east-west boundary conditions are periodic) and the number of time steps were set high enough to ensure a steady-state would be reached.
+The lattice size was chosen to be sufficiently fine for accuracy and the number of time steps were set high enough to ensure a steady-state would be reached.
 In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
@@
 The yield stress was varied between four different values $\tau_y = [4.0, 8.0, 12.0, 16.0] \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
-Recall that $m$ is the stress growth exponent for the Papanastasiou approximation.
+Recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer to the true Bingham plastic behavior.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
@@
 \noindent where $y_{\tau} = -\tau_y / \pgrad$ is the vertical location at which the fluid yields.
+(Note that the analytical solution is based on the exact Bingham plastic model, and not the Papanastasiou approximation.)
 
@@
 A larger stress growth exponent results in a more accurate Papastasiou approximation of the true Bingham plastic constitutive model, however, it leads to more nonphysical oscillations that degrade the numerical solution for the BGK collision operator.
-Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helps mitigate errors that occur as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
-\pagebreak
+Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helped to mitigate errors that occured as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
 In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress fluids (relative errors of approximately 25-75\%).
@@
 \centering
-\includegraphics[width=\textwidth]{figs/poise-bingham/bgk-5/analytical-vs-approx.png}
+\includegraphics[width=\linewidth]{figs/poise-bingham/bgk-5/analytical-vs-approx.png}
 \caption{LBM approximation using BGK and $m = 10^5$ compared to the analytical solution for $\tau_y = 16 \times 10^{-5}$.}
@@
 	\centering
-    \includegraphics[width=\textwidth]{figs/poise-bingham/bgk-8/analytical-vs-approx.png}
+    \includegraphics[width=\linewidth]{figs/poise-bingham/bgk-8/analytical-vs-approx.png}
     \caption{LBM approximation using BGK and $m = 10^8$ compared to the analytical solution for $\tau_y = 16 \times 10^{-5}$.}
@@
 
-In order to better understand why median filtering did not eliminate the nonphysical oscillations in high yield stress fluids, but instead exacerbates the problem, it is necessary to investigate what is happening at the particle distribution scale.
+In order to better understand why median filtering did not eliminate the nonphysical oscillations in high yield stress fluids, but instead exacerbated the problem, it is necessary to investigate what is happening at the particle distribution scale.
 \Fref{fig:feq-vs-f_bgk} compares particle distributions to quasiequilibrium.
@@
 For all four cases, the optimal solution was $\theta = 2.0$ and the cost function, $f(2.0)$, was equal to the relative $L_2$ error for the BGK collision operator, $m = 10^8$ and no entropic filtering.
-The results of the optimization suggest that entropic filtering, at best does not affect the accuracy of LBM, and often adversly affects the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization is used, and regardless of what the $\Delta S$ threshold, $\theta$, is set to.
-Thus, entropic filtering is not effective for increasing the accuracy of Bingham plastic Poiseuille flow.
+The results of the optimization show that entropic filtering, at best did not affect the accuracy of LBM, and often adversly affected the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization was used, and regardless of what the $\Delta S$ threshold, $\theta$, was set to.
+
+\newcommand{\figwid}{2.28in}
 
@@
 \begin{tabulary}{\linewidth}{c c}
-\includegraphics[width=2.75in]{figs/poise-bingham/bgk-5/feq-vs-f_5.png}
+\includegraphics[width=\figwid]{figs/poise-bingham/bgk-5/feq-vs-f_5.png}
 &
-\includegraphics[width=2.75in]{figs/poise-bingham/bgk-5/feq-vs-f_8.png}
-\\
-\includegraphics[width=2.75in]{figs/poise-bingham/bgk-8/feq-vs-f_5.png}
+\includegraphics[width=\figwid]{figs/poise-bingham/bgk-5/feq-vs-f_8.png}
+\\
+\includegraphics[width=\figwid]{figs/poise-bingham/bgk-8/feq-vs-f_5.png}
 &
-\includegraphics[width=2.75in]{figs/poise-bingham/bgk-8/feq-vs-f_8.png}
+\includegraphics[width=\figwid]{figs/poise-bingham/bgk-8/feq-vs-f_8.png}
 \end{tabulary}
@@
 
-On average, the LBM model using the MRT collision operator took 2.9 times more computation time than the BGK collision operator with $m = 10^5$.
-However, the increased computational time can be justified for the MRT collision operator because MRT solution did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
+On average, the LBM model using the MRT collision operator took 2.9 times more computing time than the BGK collision operator with $m = 10^5$.
+However, the increased computing time can be justified for the MRT collision operator because the MRT solution did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
 when $m = 10^8$.
 The MRT collision operator was therefore the most accurate solution for all cases.
-\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilbirium profiles, $f_i^{eq}$ for the MRT model; in contrast to the BGK with $m = 10^8$, there are no nonphysical oscillations in the quasiequilibrium distributions.
+\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilbirium profiles, $f_i^{eq}$ for the MRT model.
+In contrast to the BGK with $m = 10^8$, there were no significant nonphysical oscillations in the quasiequilibrium distributions.
 
@@
     \begin{tabulary}{\linewidth}{c c}
-        \includegraphics[width=2.75in]{figs/poise-bingham/mrt/feq-vs-f_5.png}
+        \includegraphics[width=\figwid]{figs/poise-bingham/mrt/feq-vs-f_5.png}
         &
-        \includegraphics[width=2.75in]{figs/poise-bingham/mrt/feq-vs-f_8.png}
+        \includegraphics[width=\figwid]{figs/poise-bingham/mrt/feq-vs-f_8.png}
     \end{tabulary}
@@
 
-Another important question to ask is, if the strength of the MRT collision operator is damping out the ghost modes, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of the ghost modes?.
+Another important question to ask is, if the strength of the MRT collision operator is damping out the ghost modes, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of the ghost modes?
 \Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment, $\boldsymbol{\epsilon}^{neq}$, with respect to time for each of the collision schemes, which was calculated as:
@@
 \end{equation}
-\noindent where $||\hdots||_2$ is the Euclidean norm, $\epsilon_j = 4f_0(\pos_j) - 2f_1(\pos_j) - 2f_2(\pos_j) - 2f_3(\pos_j) - 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) + f_7(\pos_j) + f_8(\pos_j)$ is related to the square of the energy, and $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, i.e.
+\noindent where $||.||_2$ is the Euclidean norm, $\epsilon_j = 4f_0(\pos_j) - 2f_1(\pos_j) - 2f_2(\pos_j) - 2f_3(\pos_j) - 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) + f_7(\pos_j) + f_8(\pos_j)$ is related to the square of the energy, and $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, i.e.
 $\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
-The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ suffered from was in part due to the $\epsilon$ moment.
+The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ displayed was in part due to the $\epsilon$ moment.
 
@@
 The values measured for the BGK with $m = 10^8$ and median filtered were consistently greater than any other collision scheme.
-The relative $L_2$ norm of the $\mathbf{q_x}^{neq}$ for the BGK collision operator with $m = 10^8$ and median filtering solution spent most of the simulation between approximately 50\% and 100\%, this suggests that the $q_x$ moment was likely the cause of the nonphysical oscillations and error in its solution.
-As expected (if the dominant source of error in the LBM approximation for Bingham plastic Poiseuille flow is indeed due to the $q_x$ moment), the $\mathbf{q_x^{neq}}$ relative $L_2$ norm for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $\mathbf{q_x^{neq}}$ relative $L_2$ was negligible for the MRT collision scheme.
-These results are consistent with solution error.
+The relative $L_2$ norm of the $\mathbf{q_x}^{neq}$ for the BGK collision operator with $m = 10^8$ and median filtering solution spent most of the simulation between approximately 50\% and 100\%, suggesting that the $q_x$ moment was likely the cause of the nonphysical oscillations and error in the solution.
+As expected (if the dominant source of error in the LBM approximation for Bingham plastic Poiseuille flow is indeed due to the $q_x$ moment), the $\mathbf{q_x^{neq}}$ relative $L_2$ norm for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $\mathbf{q_x^{neq}}$ relative $L_2$ norm was negligible for the MRT collision scheme.
 
@@
 
-It is likely that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time, i.e.
-osciallations build in amplitude with time.
-\Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time, which was calculated as:
+It is likely that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time (i.e.
+osciallations build in amplitude with time).
+\Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time, which was calculated, respectively, as:
 \begin{equation}
@@
 \Fref{fig:qx} and \Fref{fig:qx-cumulative} shows that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the ghost mode associated with the $q_x$ moment dominates and is the primary source of oscillations.
-It can be inferred that the $q_x$ moment was significant source of oscillations in this case because the Poiseuille flow was in the x-direction.
-If an LBM collision scheme is to be developed for simulating high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it should focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.
+It can be inferred that the $q_x$ moment was a significant source of oscillations in this case because the Poiseuille flow was in the x-direction.
+Overall, the results presented suggest that if an LBM collision scheme is to be developed for simulating high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it should focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.
 
@@
 The simulations of the Poiseuille flow with the power-law constitutive relationship were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a flow consistency index of $k = 0.2$.
-The flow behavior index, $n$, was varied between the four different values: $[0.5, 0.75, 1.25, 1.50]$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $[0.05, 1.995]$.
+The flow behavior index, $n$, was varied between the four different values (from shear-thinning to shear-thickening): $[0.5, 0.75, 1.25, 1.50]$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $[0.05, 1.995]$.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
@@
 The Reynold's number was computed by $Re = \frac{\rho U^{2-n} H^n}{k}$.
+Although for $n = 0.5, 0.75, 1.25$ the MRT collision operator was more accurate and required more computation time than the BGK collision operator, for $n = 1.5$ the BGK operator with median filtering required the least computation time and was the most accurate with respect to the analytical solution.
+In fact, besides the case of $n = 0.5$ (i.e. relatively extreme shear-thinning) the BGK with median filtering was both computationally efficient and accurate.
+The BGK-BRT had a moderate improvement of accuracy and efficiency--with respect to the regular BGK collision scheme--for flow where the constitutive relationship was only somewhat nonlinear, i.e.
+for $n = 0.75, 1.25$.
+Note that none of the collision schemes were sufficiently accurate (error greater than 15\%) for the case of $n = 0.5$, and to improve this accuracy it would be necessary to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
 
@@
 
-Although for $n = 0.5, 0.75, 1.25$ the MRT collision operator was more accurate and required more computation time than the BGK collision operator, for $n = 1.5$ the BGK operator with median filtering required the least computation time and was the most accurate with respect to the analytical solution.
-In fact, besides for the case of $n = 0.5$, i.e. relatively extreme shear-thinning, the BGK with median filtering was both computationally efficient and accurate.
-The BGK-BRT had a moderate improvement of accuracy and efficiency--with respect to the regular BGK collision scheme--for flow where the constitutive relationship was only somewhat nonlinear, i.e.
-for $n = 0.75, 1.25$.
-Note that none of the collision schemes were sufficiently accurate (error greater than 15\%) for the case of $n = 0.5$, and to improve this accuracy it would be necessary to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
-
 \subsection{Lid-driven Cavity Flow} % MG: would it be of any use to show streamline plots from some of these flows?
 
-The lid-driven benchmark was chosen for this numerical study because there are many results available in literature in which to compare with and because the vorticity of the flow coupled with the nonlinear constitutive equations should result in a challenge in terms of stability.
-Lid-driven cavity flow is generally characterized by a square cavity where a fluid velocity is prescribed tangential to the upper boundary and the remaining boundaries have a no-slip boundary condition.
+A lid-driven benchmark problem was chosen for this numerical study because there are many results available in literature in which to compare with and because the vorticity of the flow coupled with the nonlinear constitutive equations should result in a challenge in terms of stability.
+As was done herein, lid-driven cavity flow is generally characterized by a square cavity where a fluid velocity is prescribed tangential to the upper boundary and the remaining boundaries have a no-slip boundary condition.
 The example of lid-driven cavity flow utilized herein is presented schematically in \Fref{fig:lid-driven}.
@@
 
-The lid-driven cavity simulations presented in the section were all simulated on a relatively coarse, $100 \times 100$ lattice (other studies simulating lid-driven cavity flow of a non-Newtonian fluid tend to use lattice sizes on the order of $256 \times 256$ and $512 \times 512$~\cite{chen2014simulations,li2014simulation}, this study follows that of~\cite{brownlee2008nonequilibrium}).
+The lid-driven cavity simulations presented in the section were all simulated on a relatively coarse, $100 \times 100$ lattice, similar to the work presented in ~\cite{brownlee2008nonequilibrium}.
 A coarse lattice was chosen in order to highlight concerns with stability and accuracy.
 The simulations were run for either 50000 timesteps or until convergence was met.
-Convergence was defined by,
+Convergence was defined by:
 \begin{equation} \label{eq:convergence}
@@
 The lid velocity was prescribed as $u_o = 0.1$.
-For the MRT relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$ as these values have been successfully applied to simulating lid-driven cavity flow of non-Newtonian fluids in the past~\cite{chen2014simulations,li2014simulation}.
-All coordinate values presented in this section (used to describe the location of the center of vortices) are given normalized with respect to the length of the cavity side, i.e. as $(x / L, y / L)$.
-Note that all results presented as ``-'' indicate that the tests were numerically unstable to the extent that the degrees of freedom all over the domain were nonsensical.
+For the MRT relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$, as these values have been successfully applied to simulating lid-driven cavity flow of non-Newtonian fluids in the past~\cite{chen2014simulations,li2014simulation}.
+All coordinate values presented in this section (used to describe the location of the center of vortices) are given normalized with respect to the length of the cavity side (i.e. as $(x / L, y / L)$).
+Note that all results presented as ``-'' indicate that the tests were numerically unstable to the extent that the degrees of freedom all over the domain were diverging toward $-\infty$ or $\infty$.
 
@@
 
-For the Bingham plastic numerical tests the Reynold's number was varied:$[100, 1000, 5000, 10000]$, and the Bingham number was varied: $[1, 10, 100]$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
-The collision schemes that were tested were: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
+For the Bingham plastic numerical tests the Reynold's number was varied with the following values:$[100, 1000, 5000, 10000]$, and the Bingham number was varied with the following values: $[1, 10, 100]$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
+Six collision schemes were tested: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
 Tables \ref{tab:lid-bing1}--\ref{tab:lid-bing100} compare the center location of the main vortex to literature values taken from~\citet{syrakos2014performance}.
@@
 
-Just as before with the Bingham plastic Poiseuille flow, if a modeler is using the BGK collision operator and entropic filtering is not available, a stress growth exponent of $m = 10^5$ yields faster, more accurate results than using a larger stress growth exponent.
-A smaller stress growth exponent is also more effective at producing accurate results in the BGK collision scheme than placing bounds on the relaxation time (BGK-BRT).
-However, although the BGK with $m = 10^5$ was the fastest model in all cases, it was also like the other two BGK models without median filtering in that it was unstable for $Re \ge 5000$.
+Just as before with the Bingham plastic Poiseuille flow, if using the BGK collision operator and entropic filtering is not available, a stress growth exponent of $m = 10^5$ yields faster and more accurate results than using a larger stress growth exponent.
+A smaller stress growth exponent is also more effective at producing accurate results in the BGK collision scheme than placing bounds on the relaxation time (i.e. BGK-BRT).
+However, although the BGK with $m = 10^5$ was the fastest model in all cases, it was also similar to the other two BGK models without median filtering in that it was unstable for $Re \ge 5000$.
 
@@
 However, in general, the BGK with $m = 10^8$ and median filtering was consistently different from literature values. 
-The apparent inaccuracy with regards to BGK with $m = 10^8$ and median filtering is probably due to how the numerical stability is enhanced--namely through artificial, nonphysical dissipation.
+The apparent inaccuracy with regards to BGK with $m = 10^8$ and median filtering is probably due to how the numerical stability is enhanced through artificial, nonphysical dissipation.
 In contrast to the nonphysical nature in which median filtering enhances stability, the stability enhancement used by MRT does not directly affect the macroscopic hydrodynamics of interest, so it again provides a stable, and in most of the cases examined herein, the most accurate solution.
@@
 The MRT collision operator was, in general, 5-10 times slower than any of the BGK collision schemes.
-The extreme increase in computational expense is probably not due to the collision process itself, i.e.
-\eqref{eq:mrt-colop}, but instead calculating the strain-rate \eqref{eq:srtensor-mrt} for each iteration of the solution for $\mu_{app}$.
-
-In summary for the lid-driven cavity flow of a Bingham plastic fluid, 
+The extreme increase in computational expense is probably not due to the collision process itself (i.e.
+\eqref{eq:mrt-colop}) but instead due to calculating the strain-rate \eqref{eq:srtensor-mrt} for each iteration of the solution for the apparent viscosity, $\mu_{app}$.
+
+In summary, for the lid-driven cavity flow of a Bingham plastic fluid, 
 \begin{itemize}
-    \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
+    \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a relatively small amount of computing time,
     \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable,
-    \item and the MRT collision operator with $m = 10^8$ produces solutions with high accuracy and stability for all of the cases examined herein, but at an increased (approximately 5-10 times more) computational cost.
+    \item and the MRT collision operator with $m = 10^8$ produces solutions with high accuracy and stability for all of the cases examined herein, but at an increased (approximately 5-10 times more) computing cost.
 \end{itemize}
@@
 
-For the power-law numerical tests the Reynold's number was varied: $[100, 1000, 5000, 10000]$, and the flow behavior index, $n$, was varied: $[0.5, 1.5]$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
-The collision schemes that were tested were: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT with $m = 10^8$ and (5) MRT with $m = 10^8$ and median filtering.
+For the power-law numerical tests the Reynold's number was varied with the following values: $[100, 1000, 5000, 10000]$, and the flow behavior index, $n$, was varied with the following values: $[0.5, 1.5]$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
+Five collision schemes were tested: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT and (5) MRT with median filtering.
 \Fref{tab:lid-powerlaw05} and \Fref{tab:lid-powerlaw15} compare the center location of the main vortex to literature values taken from~\citet{li2014simulation}.
@@
 However, the solutions became unstable at higher Reynold's numbers and entropic filtering was necessary to produce a stable soltuion when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
-Although there still appears to be somewhat of a tradeoff between computational efficiency and accuracy in regards to the BGK with median filtering and MRT with median filtering, it does not seem to be nearly as drastic as when simulating Bingham plastic lid-driven flow; e.g.
-neither collision scheme consistently agrees with the literature values; and the MRT with median filtering is only 2-3 times slower than the BGK with median filtering.
-A good example of this tradeoff is the shear-thinning results for $Re = 5000$: the location of the main vortex for the BGK with median filtering differed from literature by more than 10\% in both the $x$ and $y$ directions but the MRT with median filtering has approximately twice the computational cost.
-
-For the shear-thickening results, $n = 1.5$, there were no issues of stability.
+Although there still appears to be a tradeoff between computational efficiency and accuracy in regards to the BGK with median filtering and MRT with median filtering, it does not seem to be nearly as significant as when simulating Bingham plastic lid-driven flow (e.g.
+neither collision scheme consistently agrees with the literature values and the MRT with median filtering is only 2-3 times slower than the BGK with median filtering).
+An example of this tradeoff is the shear-thinning results for $Re = 5000$: the location of the main vortex for the BGK with median filtering differed from the literature by more than 10\% in both the $x$ and $y$ directions, but the MRT with median filtering has approximately twice the computing cost.
+
+For the shear-thickening results, $n = 1.5$, there were no issues of instability.
 What is interesting is that all of the collision schemes produce similar results for each of the cases, with the difference in the location of the main vortex being no greater than 1\% in either the $x$ or $y$ direction for any two collision schemes (with the exception of the high Reynold's number case, $Re = 10000$).
-The BGK-BRT consistently needed the least amount of computation time.
-The reason the BGK-BRT may be the most efficient is that because the collision frequency, $\omega$, is bounded, and consequently $\mu_{app}$ is bounded, meaning the solution to the constitutive equation may be converging to a bound with less iterations than the other collision schemes.
+The BGK-BRT consistently needed the least amount of computing time.
+The reason the BGK-BRT may be the most efficient is because the collision frequency, $\omega$, is bounded, and consequently $\mu_{app}$ is bounded, meaning the solution to the constitutive equation may be converging with less iterations than the other collision schemes.
 
@@
     \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable.
-    \end{itemize} \pagebreak
+    \end{itemize}
   \item To summarize power-law, lid-driven cavity flow,

### 2016-09-02 figs/poise-bingham/process3.jl

@@ -1,0 +1,69 @@
+using Winston
+
+names = Dict("epsilon" => "ϵ", "qx" => "q_x", "qy" => "q_y");
+
+tests = ["bgk-5","bgk-8","bgk-fltr","mrt"];
+labels = ["BGK, m = 10^5","BGK, m = 10^8","BGK, filtered","MRT, m = 10^8"];
+#props = ["k-o","k--<","k-.>","k:^"];
+props = ["k","k","k","k"];
+linetypes = ["dotted","longdashed","solid","dotdotdotdashed"];
+#linetypes = ["k:","k-.","k-","k--"];
+
+for (name, var) in names
+  args1 = [];
+  args2 = [];
+  for (prop, d) in zip(props, tests)
+    #const x = (d in ["bgk-5", "mrt"] || (name == "qx" && d == "bgk-8")) ? collect(0:62) * 400.0 : collect(0:125) * 200.0;
+    const x = collect(0:250) * 100.0;
+    val    =   readdlm("$d/$name.dsv", ','); 
+    valeq  =   readdlm("$d/$(name)_eq.dsv", ',');
+
+    #=y = if d in ["bgk-5", "mrt"] || (name == "qx" && d == "bgk-8")
+          map(i -> norm(val[i, :] - valeq[i, :], 2) / norm(valeq[i, :], 2), 1:4:251);
+        else
+          map(i -> norm(val[i, :] - valeq[i, :], 2) / norm(valeq[i, :], 2), 1:2:251);
+        end=#
+    y = map(i -> norm(val[i, :] - valeq[i, :], 2) / norm(valeq[i, :], 2), 1:251);
+    y2 = zeros(251); y2[1] = 100 * y[1];
+    for i=2:251; y2[i] = 100 * y[i] + y2[i-1]; end
+    #args1 = vcat(args1, Vector[x, y], [prop]);
+    push!(args1, (x, y));
+    push!(args2, (x, y2));
+  end
+  #=
+  draw(PNG("$name.png", 5.5inch, 3inch),
+       plot(df, x=:x, y=:y, color=:Series,
+       Geom.line,
+       Guide.XLabel("Time (sec)"),
+       Guide.YLabel("Relative L<sub>2</sub>, $var<sup>neq</sup>")
+       ));
+  draw(PNG("$(name)_cumulative.png", 5.5inch, 3inch),
+       plot(cf, x=:x, y=:y, color=:Series,
+       Geom.line,
+       Guide.XLabel("Time (sec)"),
[…]

### 2016-09-02 src/conversions.jl

@@ -22,6 +22,9 @@ const GAL_TO_CFT     = 0.133681;
 
 Re(u, L, ν) = u * L / ν;
 Bn(τ, L, μ, u) = τ * L / (μ * u);
+Wt(τ, ρ, L) = τ / (ρ * L);
 
 @show Res = [Re(vel_ftps[1], 0.75 / FT_TO_IN, ν_sqftpsec[2]); Re(vel_ftps[2], 0.75 / FT_TO_IN, ν_sqftpsec[1])];
 @show Bns = [Bn(τ_lbpsqft[1], 0.75 / FT_TO_IN, μ_lbpftsec[2], vel_ftps[2]), Bn(τ_lbpsqft[2], 0.75 / FT_TO_IN, μ_lbpftsec[1], vel_ftps[1])];
+@show Wts = [Wt(τ_lbpsqft[1], 0.75 / FT_TO_IN, ρ_lbpcft[2]); Wt(τ_lbpsqft[2], 0.75 / FT_TO_IN, ρ_lbpcft[1])];
+@show iWts = reverse(map(wt -> 1 / wt, Wts));

### 2016-09-03 src/stability-paper.tex
changed plain square bracketed arrays to bmatrixs

@@
 The MRT collision takes place in moment space and allows each moment to relax at a different rate.
-\cite{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model for fluid flow, and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used BGK collision operator.
+\cite{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model for fluid flow, and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used collision operator.
 Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the cost may increase significantly for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain-rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
@@
 The Bingham plastic simulations were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a plastic viscosity of $\mu_p = 0.2$.
-The yield stress was varied between four different values $\tau_y = [4.0, 8.0, 12.0, 16.0] \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
+The yield stress was varied between four different values $\tau_y = \begin{bmatrix}4.0,&8.0,&12.0,&16.0\end{bmatrix} \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
 Recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer to the true Bingham plastic behavior.
@@
 Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helped to mitigate errors that occured as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
-In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress fluids (relative errors of approximately 25-75\%).
+In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress fluids (relative errors of approximately 25--75\%).
 
@@
   \min_\theta \hspace{0.2in} & f(\theta)\\
-\text{such that} \hspace{0.2in} & \theta \in [10^{-10}, 2.0]
+  \text{such that} \hspace{0.2in} & \theta \in \begin{bmatrix}10^{-10}, & 2.0\end{bmatrix}
 \end{align*}
 \noindent where $f(\theta) \equiv $ the relative $L_2$ error between the LBM approximation using $m = 10^8$ and entropic filtering with a $\Delta S$ threshold of $\theta$ (used in the criteria defined in \eqref{eq:ds-threshold}).
-The bounds on $\theta$, namely $\theta \in [10^{-10}, 2]$ represent $\theta$ such that 100\% of lattice sites are filtered for every time step ($\theta = 10^{-10}$) and $\theta$ such that no lattice sites are filtered for any time step ($\theta = 2.0$).
+The bounds on $\theta$, namely $\theta \in \begin{bmatrix}10^{-10}, & 2\end{bmatrix}$ represent $\theta$ such that 100\% of lattice sites are filtered for every time step ($\theta = 10^{-10}$) and $\theta$ such that no lattice sites are filtered for any time step ($\theta = 2.0$).
 The optimization problem was solved for the cases of using median filtering and Ehrenfests' regularization using both Brent's method and the Golden Section search.
@@
 The simulations of the Poiseuille flow with the power-law constitutive relationship were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a flow consistency index of $k = 0.2$.
-The flow behavior index, $n$, was varied between the four different values (from shear-thinning to shear-thickening): $[0.5, 0.75, 1.25, 1.50]$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $[0.05, 1.995]$.
+The flow behavior index, $n$, was varied between the four different values (from shear-thinning to shear-thickening): $\begin{bmatrix}0.5,&0.75,&1.25,& 1.50\end{bmatrix}$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $\begin{bmatrix}0.05,&1.995\end{bmatrix}$.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
@@
 
-For the Bingham plastic numerical tests the Reynold's number was varied with the following values:$[100, 1000, 5000, 10000]$, and the Bingham number was varied with the following values: $[1, 10, 100]$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
+For the Bingham plastic numerical tests the Reynold's number was varied with the following values:$\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the Bingham number was varied with the following values: $\begin{bmatrix}1,&10,&100\end{bmatrix}$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
 Six collision schemes were tested: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
@@
 
-For the power-law numerical tests the Reynold's number was varied with the following values: $[100, 1000, 5000, 10000]$, and the flow behavior index, $n$, was varied with the following values: $[0.5, 1.5]$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
+For the power-law numerical tests the Reynold's number was varied with the following values: $\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the flow behavior index, $n$, was varied with the following values: $\begin{bmatrix}0.5,&1.5\end{bmatrix}$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
 Five collision schemes were tested: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT and (5) MRT with median filtering.

### 2016-09-03 src/stability-paper.tex
restructured, moved picard type algorithm details and added acknowledgements

@@
 The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis~\cite{wolf2000lattice}.
-For incompressible Newtonian flow, the collision frequency is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$; and from this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.
+For incompressible Newtonian flow, the collision frequency is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$; and from this relationship it is clear that $\omega \in \begin{bmatrix}0.0,& 2.0\end{bmatrix}$, otherwise the viscosity would be negative.
 The method for simulating non-Newtonian flow in the current work involves approximating a solution to the local value of apparent viscosity, $\mu_{app}(\pos, t)$, where the shear stress, $\tau$, and strain-rate, $\dot{\gamma}$, are dependent on $\pos$ and $t$, and then setting the value of the local collision frequency as follows:
-\begin{equation} \label{eq:colfeq-non-newtonian}
+\begin{equation} \label{eq:colfreq-non-newtonian}
 \omega(\pos, t) = \frac{1}{\frac{\mu_{app}(\pos, t)}{c_s^2 \rho(\pos, t)} + \frac{1}{2}}
@@
 
-\subsubsection{Multiple-relaxation Time}
+\subsubsection{Multiple-relaxation Time} \label{sec:mrt}
 
@@
 
-In order to solve for the apparent viscosity, $\mu_{app}$, using the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf}, one must calculate the strain-rate, $\dot{\gamma}$.
+In the present work, material behavior is simulated through the collision frequency, $\omega$, when using the BGK collision operator (\Fref{sec:bgk}), or the $s_7$ and $s_8$ relaxation rates when using the MRT collision operator (\Fref{sec:mrt}).
+Both the collision frequency and the $s_7$ and $s_8$ relaxation rates are dependent on the apparent visocity, $\mu_{app}$, through \eqref{eq:colfreq-non-newtonian} and \eqref{eq:mrt-s7}, respectively.
+The apparent viscosity, in the context of non-Newtonian flow, is a function of the local strain-rate, $\dot{\gamma}$.
 The strain-rate is given by the second invariant of the strain-rate tensor, $\srtensor$, i.e.:
@@
 
-Upon comparison of \eqref{eq:srtensor-bgk} and \eqref{eq:srtensor-mrt}, it is clear that calculating the strain-rate tensor for the MRT collision operator is more computationally expensive than it is for the BGK collision operator.
-The increase in computational expense is exacerbated by the fact that approximating the apparent viscosity, $\mu_{app}$, by the algorithm outlined in \Fref{sec:lbm-for-nnf} requires iteration and therefore multiple calculations of the strain-rate tensor.
-
-\subsection{Non-Newtonian Constitutive Equations}
+It can be seen from \eqref{eq:srtensor-bgk} and \eqref{eq:srtensor-mrt} that the strain-rate tensor is a function of the collision frequency in the case of the BGK collision operator, or the relaxation matrix, $\relaxM$, in the case of the MRT collision operator.
+This interdependence between the relaxation parameters ($\omega$, $s_7$ and $s_8$) and the strain-rate is a direct result of a nonlinear, non-Newtonian constitutive relationship.
+The present work uses a Picard-type algorithm to approximate a solution for $\dot{\gamma}$ and $\mu_{app}$:
+
+\label{sec:picard}
+\begin{enumerate}
+	\item Start with initial guess for the apparent viscosity, $\mu_k = \mu_0$.
+	      Set $\omega_k$ using \eqref{eq:colfreq-non-newtonian} (, or set $s_7 = s_8$ using \eqref{eq:mrt-s7})
+	\item \label{step:solve-for-flow} Solve for the strain-rate, $\dot{\gamma}_k$, using \eqref{eq:srtensor-bgk} (, or using \eqref{eq:srtensor-mrt}).
+	\item Update the apparent viscosity, $\mu_{k+1} = f(\dot{\gamma}_k)$.
+	\item Return to Step \ref{step:solve-for-flow} until convergence is met.
+\end{enumerate}
+\noindent where $f$ is form of the constitutive relationship that maps the local strain-rate to a local apparent visocisty.
+More details on the functions that are used for $f$ are given in \Fref{sec:constit}.
+
+It is clear that calculating the strain-rate tensor for the MRT collision operator is more computationally expensive than it is for the BGK collision operator.
+The increase in the overall computational expense for simulating non-Newtonian flow using the MRT collision operator is exacerbated by the fact that approximating the apparent viscosity requires iteration and therefore multiple calculations of the strain-rate tensor for each discrete moment in time and for each discrete location in the domain.
+
+\subsection{Non-Newtonian Constitutive Equations} \label{sec:constit}
 
@@
 
-A Bingham plastic does not flow, i.e.
-the strain-rate is zero, when the shear stress is below the yield stress, and behaves in an almost Newtonian manner when the shear stress is above the yield stress.
+A Bingham plastic does not flow (i.e.
+the strain-rate is zero) when the shear stress is below the yield stress, and behaves in an almost Newtonian manner when the shear stress is above the yield stress.
 The Bingham plastic relationship is described mathematically as:
@@
 \begin{equation} \label{eq:bing-mu-app}
-\mu_{app} = \frac{\tau_y}{\dot{\gamma}} (1 - e^{-m |\dot{\gamma}|}) + \mu_p.
+  \mu_{app} = f_B(\dot{\gamma}) = \frac{\tau_y}{\dot{\gamma}} (1 - e^{-m |\dot{\gamma}|}) + \mu_p.
 \end{equation}
@@
 \begin{equation} \label{eq:pl-mu-app}
-\mu_{app} = k \dot{\gamma}^{n-1}.
+  \mu_{app} = f_{PL}(\dot{\gamma}) = k \dot{\gamma}^{n-1}.
 \end{equation}
@@
 A numerical study was carried out to investigate the suitability of the different LBM collision schemes discussed for simulating non-Newtonian flow in terms of accuracy, numerical stability and computational efficiency.
-For all simulations presented in this section, the apparent viscosity was approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:lbm-for-nnf} with the stopping criteria set as (whichever was first) a maximum number of iterations set at 15 and the convergence criteria:
+For all simulations presented in this section, the apparent viscosity was approximated using its corresponding constitutive relationship and the Picard-type algorithm outlined in \Fref{sec:picard} with the stopping criteria set as (whichever was first) a maximum number of iterations set at 15 and the convergence criteria:
 \begin{equation} \label{eq:mu-app-conv}
@@
 
-\newcommand{\figwid}{2.28in}
+\newcommand{\figwid}{0.48\linewidth}
 
@@
 Two different benchmark problems were chosen for the study: Poiseuille flow, and lid-driven square cavity flow.
-The results showed that:
-\begin{itemize}
-  \item For high yield stress fluids in Poiseuille-type flow, only the MRT collision operator did not suffer from nonphysical oscillations.
-  \begin{itemize}
-      \item The oscillations appeared to be due to high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
-      \item If a collision scheme is to be developed for high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, then it should focus on dampening the nonequilibrium energy flux moment.
-    \end{itemize}
-  \item Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
-  \item In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
-  \item To summarize Bingham plastic, lid-driven cavity flow,
+The results showed that for high yield stress fluids in Poiseuille-type flow, only the MRT collision operator did not suffer from nonphysical oscillations.
+The oscillations appeared to be due to high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
+If a collision scheme is to be developed for high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, then it should focus on dampening the nonequilibrium energy flux moment.
+
+Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
+In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
+
+To summarize Bingham plastic, lid-driven cavity flow,
     \begin{itemize}
@@
     \end{itemize}
-  \item To summarize power-law, lid-driven cavity flow,
+To summarize power-law, lid-driven cavity flow,
     \begin{itemize}
@@
     \end{itemize}
-\end{itemize}
 

### 2016-09-03 src/stability-paper.tex
added better conclusions at the end of the abstract and a simple analysis of gradients at the end of the poiseulle flow section

@@
 Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
-The test results are followed by recommendations for choice collision scheme given priority and problem type.
+The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behavior, but at an increased computing cost that was, in some cases, as much as five times greater.
+It was also shown that artificial dissipation can be an effective technique for enhancing numerical stability of non-Newtonian, lid-driven cavity flow, and is particularly effective for efficiently and accurately simulating the flow of shear-thinning fluids.
 
@@
 
+Before closing the discussion on the Poiseuille flow numerical study, it is worthwhile to consider what precisely is common between the higher-yield stress fluids and the shear-thinning fluids, as they were the cases that exhibited numerical instabilities and were simulated with the least amount of accuracy for some of the collision schemes considered.
+One such commonality can be seen when considering the effect of the yield stress, $\tau_y$, and the effect of the flow behavior index, $n$, on the gradient of the theoretical flow velocity.
+Taking the partial with respect to $y$ of the analytical solution for the flow velocity of a Bingham plastic, Poiseuille flow (Equation \eqref{eq:analytical-bingham}) results in:
+\begin{equation} \label{eq:gradu-bingham}
+  \frac{\partial u_x}{\partial y} = \begin{cases}
+    0, & 0 \leq |y| \leq y_{\tau}, \\
+    \frac{1}{\mu_p} \left( \pgrad y + \tau_y \frac{y}{|y|} \right), & y_{\tau} < |y| \leq h,
+  \end{cases}
+\end{equation}
+and taking the partial with respect to $y$ of the analytical solution for the flow velocity of a power-law fluid, Poiseuille flow (Equation \eqref{eq:analytical-power-law}) results in:
+\begin{equation} \label{eq:gradu-power-law}
+  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{k} \pgrad \right)^{1/n} \left( -y |y|^{1/n - 1} \right).
+\end{equation}
+It can be seen from \eqref{eq:gradu-bingham} and \eqref{eq:gradu-power-law} that $\left| \frac{\partial u_x}{\partial y} \right|$ increases with the yield stress for Bingham plastics, and increases with $\frac{1}{n}$ for power-law fluids, which suggests that one of the challenges associated with the accuracy and stability of simulating higher yield-stress fluids and more extreme shear-thinning fluids (e.g.
+$n \le 0.5$) in Poiseuille-type flow is a result of having a larger magnitude of the gradient of macroscopic flow.
+This conclusion is consistent with the discussion in \Fref{sec:bgk} with regards to one of the drawbacks of the BGK collision operator, namely that when there is a sharp spatial gradient in either $\rho$ or $\mvel$ that it can lead to nonphysical oscillations that can degrade the accuracy and stability of the solution.
+However, it does not explain why the MRT collision operator was unable to accurately simulate the case of a power-law fluid, Poiseuille flow, with $n = 0.5$.
+
 \subsection{Lid-driven Cavity Flow} % MG: would it be of any use to show streamline plots from some of these flows?

### 2016-09-03 src/stability-paper.tex
minor updates

@@
 Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
-The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behavior, but at an increased computing cost that was, in some cases, as much as five times greater.
+The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behaviors, but at an increased computing cost that was, in some cases, as much as five times greater.
 It was also shown that artificial dissipation can be an effective technique for enhancing numerical stability of non-Newtonian, lid-driven cavity flow, and is particularly effective for efficiently and accurately simulating the flow of shear-thinning fluids.
@@
 \noindent where $f$ is form of the constitutive relationship that maps the local strain-rate to a local apparent visocisty.
-More details on the functions that are used for $f$ are given in \Fref{sec:constit}.
+More detail on the functions that are used for $f$ is given in \Fref{sec:constit}.
 
@@
 \begin{equation} \label{eq:gradu-power-law}
-  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{k} \pgrad \right)^{1/n} \left( -y |y|^{1/n - 1} \right).
+  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{k} \pgrad \right)^{1/n} \left( -y |y|^{\frac{1}{n} - 1} \right).
 \end{equation}
@@
 $n \le 0.5$) in Poiseuille-type flow is a result of having a larger magnitude of the gradient of macroscopic flow.
-This conclusion is consistent with the discussion in \Fref{sec:bgk} with regards to one of the drawbacks of the BGK collision operator, namely that when there is a sharp spatial gradient in either $\rho$ or $\mvel$ that it can lead to nonphysical oscillations that can degrade the accuracy and stability of the solution.
+This conclusion is consistent with the discussion in \Fref{sec:bgk} with regards to one of the drawbacks of the BGK collision operator, namely that a sharp spatial gradient in either $\rho$ or $\mvel$ can lead to nonphysical oscillations that can degrade the accuracy and stability of the solution.
 However, it does not explain why the MRT collision operator was unable to accurately simulate the case of a power-law fluid, Poiseuille flow, with $n = 0.5$.

### 2016-09-07 src/stability-paper.tex
modified conversions script and minor edits of abstract

@@
 Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
-The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behaviors, but at an increased computing cost that was, in some cases, as much as five times greater.
+The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behaviors, but at an increased computing cost that was, in some cases, as much as five times greater than the BGK collision operator.
 It was also shown that artificial dissipation can be an effective technique for enhancing numerical stability of non-Newtonian, lid-driven cavity flow, and is particularly effective for efficiently and accurately simulating the flow of shear-thinning fluids.

### 2016-09-07 src/conversions.jl

@@ -22,9 +22,9 @@ const GAL_TO_CFT     = 0.133681;
 
 Re(u, L, ν) = u * L / ν;
 Bn(τ, L, μ, u) = τ * L / (μ * u);
-Wt(τ, ρ, L) = τ / (ρ * L);
+Wt(τ, ρ, L) = (ρ * L) / τ;
 
 @show Res = [Re(vel_ftps[1], 0.75 / FT_TO_IN, ν_sqftpsec[2]); Re(vel_ftps[2], 0.75 / FT_TO_IN, ν_sqftpsec[1])];
 @show Bns = [Bn(τ_lbpsqft[1], 0.75 / FT_TO_IN, μ_lbpftsec[2], vel_ftps[2]), Bn(τ_lbpsqft[2], 0.75 / FT_TO_IN, μ_lbpftsec[1], vel_ftps[1])];
-@show Wts = [Wt(τ_lbpsqft[1], 0.75 / FT_TO_IN, ρ_lbpcft[2]); Wt(τ_lbpsqft[2], 0.75 / FT_TO_IN, ρ_lbpcft[1])];
+@show Wts = [Wt(τ_lbpsqft[2], 0.75 / FT_TO_IN, ρ_lbpcft[1]); Wt(τ_lbpsqft[1], 0.75 / FT_TO_IN, ρ_lbpcft[2])];
 @show iWts = reverse(map(wt -> 1 / wt, Wts));

### 2016-09-10 src/stability-paper.tex
minor updates

@@
 It is generally more practical to approximate the flow of non-Newtonian fluids using numerical methods~\cite{syrakos2014performance,owens2002computational,chai2011multiple}.
-
 The lattice Boltzmann method (LBM) is a numerical method for simulating fluid flow that has the advantage (among others) that computing the strain-rate is second-order accurate in space and local to each node~\cite{kruger2010second}.
@@
 In \Fref{sec:LBM}, an overview of the Lattice Boltzmann Method for simulating fluid flow is presented, along with the collision operators, stability enhancements, and constitutive relationships that were investigated.
-\Fref{sec:numerical-study} outlines the boundary conditions, material parameters, and LBM collision schemes for each simulation in the numerical study; and is followed by results and discussion.
+\Fref{sec:numerical-study} outlines the boundary conditions, material parameters, and LBM collision schemes for each simulation in the numerical study, and is followed by results and discussion.
 Finally, concluding remarks are given in \Fref{sec:numerical-study-conclusions}.
@@
 \begin{enumerate}
-	\item Start with initial guess for the apparent viscosity, $\mu_k = \mu_0$.
-	      Set $\omega_k$ using \eqref{eq:colfreq-non-newtonian} (, or set $s_7 = s_8$ using \eqref{eq:mrt-s7})
-	\item \label{step:solve-for-flow} Solve for the strain-rate, $\dot{\gamma}_k$, using \eqref{eq:srtensor-bgk} (, or using \eqref{eq:srtensor-mrt}).
-	\item Update the apparent viscosity, $\mu_{k+1} = f(\dot{\gamma}_k)$.
-	\item Return to Step \ref{step:solve-for-flow} until convergence is met.
+  \item Start with initial guess for the apparent viscosity, $\mu_{app}^k = \mu_{app}^0$.
+	      Set $\omega^k$ using \eqref{eq:colfreq-non-newtonian} (or set $s_7^k = s_8^k$ using \eqref{eq:mrt-s7})
+	\item \label{step:solve-for-flow} Solve for the strain-rate, $\dot{\gamma}^k$, using \eqref{eq:srtensor-bgk} (or using \eqref{eq:srtensor-mrt}).
+  \item Update the apparent viscosity, $\mu_{app}^{k+1} = \mu_{app}(\dot{\gamma}^k)$.
+	\item Return to Step \ref{step:solve-for-flow} until the apparent viscosity estimate converges.
 \end{enumerate}
-\noindent where $f$ is form of the constitutive relationship that maps the local strain-rate to a local apparent visocisty.
-More detail on the functions that are used for $f$ is given in \Fref{sec:constit}.
+\noindent where $\mu_{app}(\dot{\gamma}^k)$ is form of the constitutive relationship that maps the local strain-rate to a local apparent visocisty.
+More detail on the functions that are used for $\mu_{app}(\dot{\gamma}^k)$ is given in \Fref{sec:constit}.
 
@@
 \begin{equation} \label{eq:bing-mu-app}
-  \mu_{app} = f_B(\dot{\gamma}) = \frac{\tau_y}{\dot{\gamma}} (1 - e^{-m |\dot{\gamma}|}) + \mu_p.
+  \mu_{app}(\dot{\gamma}) = \frac{\tau_y}{\dot{\gamma}} (1 - e^{-m |\dot{\gamma}|}) + \mu_p.
 \end{equation}
@@
 \begin{equation} \label{eq:pl-mu-app}
-  \mu_{app} = f_{PL}(\dot{\gamma}) = k \dot{\gamma}^{n-1}.
+  \mu_{app}(\dot{\gamma}) = k \dot{\gamma}^{n-1}.
 \end{equation}
@@
 One such commonality can be seen when considering the effect of the yield stress, $\tau_y$, and the effect of the flow behavior index, $n$, on the gradient of the theoretical flow velocity.
-Taking the partial with respect to $y$ of the analytical solution for the flow velocity of a Bingham plastic, Poiseuille flow (Equation \eqref{eq:analytical-bingham}) results in:
+Taking the partial derivative with respect to $y$ of the analytical solution for the flow velocity of a Bingham plastic, Poiseuille flow (Equation \eqref{eq:analytical-bingham}) results in:
 \begin{equation} \label{eq:gradu-bingham}
@@
 \end{equation}
-and taking the partial with respect to $y$ of the analytical solution for the flow velocity of a power-law fluid, Poiseuille flow (Equation \eqref{eq:analytical-power-law}) results in:
+and taking the partial derivative with respect to $y$ of the analytical solution for the flow velocity of a power-law fluid, Poiseuille flow (Equation \eqref{eq:analytical-power-law}) results in:
 \begin{equation} \label{eq:gradu-power-law}
@@
 
-Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow; however, if the filter is not carefully tuned by properly adjusted the threshold, $\delta$, or number of standard deviations, $n_s$, then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
+Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow. 
+However, if the filter is not carefully tuned by properly adjusting the threshold or number of standard deviations then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
 In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
 
-To summarize Bingham plastic, lid-driven cavity flow,
-    \begin{itemize}
-  \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a short computation time,
-    \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable.
-    \end{itemize}
-To summarize power-law, lid-driven cavity flow,
-    \begin{itemize}
-      \item for shear-thinning fluids at high Reynold's numbers ($Re \ge 5000$) median filtering is necessary to produce a stable solution,
-  \item for shear-thickening, lid-driven cavity flow, there is much less concern for numerical instability.
-    \end{itemize}
-
 % MG: should we mention: code is open source? input files used in this study are also freely available?

### 2016-09-10 src/stability-paper.tex
added some comments in regards to conrad's work

@@
 The equilibrium distribution functions for Bingham plastic fluids, and for power-law and Carreau fluids were derived, implemented, and verified in~\cite{wang2008lattice} and~\cite{yoshino2007numerical}, respectively.
-Using an equilibrium distribution function that incorporates the local shear-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach values that lead to overrelaxation (e.g.
+Using an equilibrium distribution function that incorporates the local strain-rate effect has the advantage that, because the collision frequency is constant (at unity), the collision frequency will not approach values that lead to overrelaxation (e.g.
 $2$) or underrelaxation (e.g.
@@
 \cite{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model for fluid flow, and concluded LBM with the MRT collision operator was more stable, but with increased computational expense than the commonly used collision operator.
-Note that although this increased computational expense was decided to not be significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the cost may increase significantly for non-Newtonian flow because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain-rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
+Note that although this increased computational expense was decided to not be significant for Newtonian flow ($\approx$ 10--20\%~\cite{lallemand2000theory}), the cost may increase significantly for non-Newtonian flow if an iterative solution is used for the nonlinear constitutive equation because it can require that certain expensive computations be performed multiple times per time step.
 \cite{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
@@
 \end{itemize}
-In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
+In regards to the last question, much work has been done recently to enhance stability and accuracy of LBM models beyond the MRT collision operator.
 \cite{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations.
@@
 
-A numerical investigation into the accuracy, stability, and efficiency of LBM collision models when applied to non-Newtonian flow was presented.
+A numerical investigation into the accuracy, stability, and efficiency of LBM collision models and stability enhancements when applied to non-Newtonian flow was presented.
 The numerical investigation included testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
 Two different benchmark problems were chosen for the study: Poiseuille flow, and lid-driven square cavity flow.
-The results showed that for high yield stress fluids in Poiseuille-type flow, only the MRT collision operator did not suffer from nonphysical oscillations.
+The results showed that for higher yield stress fluids in Poiseuille-type flow, only the MRT collision operator was able use an accurate approximation to true Bingham plastic behavior and not suffer from nonphysical oscillations.
 The oscillations appeared to be due to high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
-If a collision scheme is to be developed for high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, then it should focus on dampening the nonequilibrium energy flux moment.
-
-Median filtering can be an effective technique for enhancing stability, especially for high Reynold's number flow. 
-However, if the filter is not carefully tuned by properly adjusting the threshold or number of standard deviations then the physical intregrity, and consquently accuracy, of the model can be adversely impacted.
-In general, the MRT collision operator is much more computationally expensive than its BGK counterpart and is some times even orders of magnitude slower.
+If a collision scheme is to be developed for higher yield stress, Poiseuille-type flow that is (1) more accurate than the BGK collision operator with $m = 10^5$ and (2) more computationally efficient than the MRT collision operator, then it should focus on dampening the nonequilibrium $\epsilon$ moment and energy flux moment (in the direction of flow).
+
+The results of the study suggest that entropic filtering can be an effective technique for enhancing stability in non-Newtonian flow, especially for high Reynold's number flows and flows that tend to produce vortices.
+However, if the filter is not carefully tuned by properly adjusting the threshold or number of standard deviations then the physical integrity, and consequently accuracy, of the model can be adversely impacted.
+Also, in general, regardless of how the filter is tuned, entropic filtering may not be effective for Bingham plastic, Poiseuille-type flow, and in some cases has the potential to adversely affect the accuracy of the simulation.
+
+For nearly all of the cases considered in the current study, the MRT collision operator was more accurate, but also more computationally expensive than its BGK counterpart, and is some times even orders of magnitude slower.
+
+In regards to future work, \cite{conrad2015accuracy} showed that non-Newtonian Lattice Boltzmann simulations are not second order accurate for every simulation Mach number, and showed that a greater accuracy may be achieved by properly tuning the simulation Mach number.
+\cite{conrad2015accuracy} predicted optimal simulation Mach numbers for various shear-thinning and shear-thickening fluids. 
+Possbile future work would include extending the numerical investigation to consider the effect of the simulation Mach number on the accuracy, stability, and efficiency of the various collision operators and stability enhancements in simulating non-Newtonian flow.
 

### 2016-09-10 sigandnov.tex
wrote short significance and novelty paragraph as required for submission

\documentclass{article}

\begin{document}

\section*{Significance and novelty of this paper}

In this paper, it is shown that Lattice Boltzmann methods have certain advantages that can make them effective for simulating non-Newtonian flows, but that producing an accurate and stable solution is non-trivial depending on the physical problem and the non-Newtonian constitutive relationship of interest.
Past work has been done to access and improve both the accuracy and stability of Lattice Boltzmann methods, however, little has been done to compare their relative effectiveness for simulating non-Newtonian fluids in terms of accuracy, stability, and efficiency.
The significance of this work is that it aims to, through a numerical investigation, determine the strengths and weaknesses of various Lattice Boltzmann Methods when applied to simulating non-Newtonian fluid flow; and in addition, it aims to identify the mechanisms through which error and instability are introduced into the simulation so as to make recommendations about how new developments can be made for improving Lattice Boltzmann methods that are applied to non-Newtonian flow.
The novelty of this paper is considered to be: (1) a quantitative comparison of the computational cost for the BGK and MRT collision operators as applied to non-Newtonian flow, (2) the application of entropic filtering to non-Newtonian flow, and (3) an identification of some of the possible mechanisms that lead to error and instability in Lattice Boltzmann Methods applied to non-Newtonian flow that, to the author's knowledge, have not previously been discussed.

\end{document}

### 2017-02-27 src/stability-paper.tex
modified figure

@@
 
-\section{Introduction} %MG: this intro will potentially be long. Should it be broken up into subsections (as it is now)?
+\section{Introduction}
 
@@
 Yield stress flow has been found to be relevant in many applictions, such as flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
-%MG: citation is for blood as non-Newtonian
 
@@
 
-%MG: probably include a discussion of the work done by Conrad et. al. 2015 here
 LBM does however, have its drawbacks.
@@
 One such consideration is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}.
-%MG: are all relevant sources cited?
 %Much work has been done to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
 Stability concerns are just as prevalent, if not more prevalent, in simulating non-Newtonian fluids because the nonlinear relationship between shear stress and strain-rate can lead to highly nonlinear fluctuations.
-%MG: be more precise than 'fluctuations'?
 Various strategies for incorporating the physics of non-Newtonian flow with LBM while maintaining a stable numerical method have been developed and studied.
@@
 A potential issue with the stability of the variable relaxation time approach is that as the collision frequency approaches $2$ the viscosity approaches zero and overrelaxation occurs.
-Alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method degrades~\cite{latt2007hydrodynamic}. %MG: verify this statement
+Alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method degrades~\cite{latt2007hydrodynamic}. 
 In order to ensure that the variable collision frequencies did not approach values leading to numerical instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable collision frequencies.
@@
 
-%MG: is the beginning of this methods section too much fluff now? Would our audiance consider it fluff?
-
 \section{Lattice Boltzmann Method for Simulation of Non-Newtonian Fluids} \label{sec:LBM}
@@
 
-The D2Q9 lattice was used in the current work (shown in \Fref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}. % MG: do we even need this citation?
+The D2Q9 lattice was used in the current work (shown in \Fref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}.
 The lattice is two-dimensional with nine discrete velocities at each node.
@@
 The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision)~\cite{Cer90}.
-In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
+In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. 
 This equilibrium is determined by the macroscopic physical behavior of interest (e.g. diffusion, incompressible flow, compressible flow, etc.).
@@
 \noindent where $\epsilon$ is related to the square of the energy $e$; $q_x$ and $q_y$ correspond to the energy fluxes in the x and y directions; and $p_{xx}$ and $p_{xy}$ correspond to the diagonal and off-diagonal component of the viscous stress tensor~\cite{lallemand2000theory}.
-% MG: verify this is correct
 The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i \in [0, 2], i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
@@
 
-%MG: show equation for feq that maximizes entropy and give a brief word why we should care
-
 \subsection{Boundary Conditions and Applied Forces} \label{sec:bcs}
@@
 D_{\alpha \beta} = -\frac{\omega}{2 \rho c_s^2} \sum_i \xi_{i \alpha} \xi_{i \beta} (f_i - f^{eq}_i),
-\end{equation} % MG: verify this
+\end{equation}
 
@@
 D_{\alpha \beta} = -\frac{1}{2 \rho c_s^2 \Delta t} \sum_i \xi_{i \alpha} \xi_{i \beta} \sum_j (\transM^{-1} \relaxM \transM)_{ij}(f_i - f^{eq}_i).
-\end{equation} % MG: verify this
+\end{equation}
 
@@
 The Bingham number was computed by $Bn = \frac{\tau_y H}{\mu_p U}$.
-
-% MG: it might be better to normalize the time with respect to the fastest, BGK with no filtering
 

### 2017-03-02 response-to-reviewers.tex
copied template for cover letter to editor and response to reviewers. fixed grammatical and spelling errors in the text

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}

\begin{document}

\section{Response to reviewers}

Thank you very much for your assistance in reviewing this paper. We are sure that with your help, this has become a much better work. We hope that the attached edited copy of "Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow" will answer any questions previously raised, and will be satisfactory for publication.

The following is a list of all comments recieved, and how they have been addressed in the updated version:

\end{document}

### 2017-03-02 src/stability-paper.tex
copied template for cover letter to editor and response to reviewers. fixed grammatical and spelling errors in the text

@@
 Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
-Yield stress flow has been found to be relevant in many applictions, such as flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
+Yield stress flow has been found to be relevant in many applications, such as flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
 
@@
 BGK consists of a single relaxation time and is a linear relaxation of particle distributions toward equilibrium.
-The BGK collison operator for the $i^{th}$ discrete velocity is expressed as:
+The BGK collision operator for the $i^{th}$ discrete velocity is expressed as:
 
@@
 In this case, if $f_i$ is ``near'' to quasiequilibrium at $\pos$ it will be ``far'' from quasiequilibrium after the streaming step when it moves to the node at $\pos + \pvel_i \Delta t$.
-Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
+Overrelaxation will occur if $\nu \rightarrow 0$ because consequently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
 Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations.
@@
 approaching 2) or too low (e.g.
-approaching 0), a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flow is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
+approaching 0), a natural, albeit nonphysical, approach to using the BGK collision operator for non-Newtonian flow is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
 This simple methodology for increasing stability will henceforth be referred to bounded-relaxation time BGK, or BGK-BRT.
@@
 
-\noindent where $\overline{\Delta S}$ and $\sigma_{\Delta S}$ are the mean and standard deviation of $\Delta S$, respectively--both are calculated using values over the domain for the current timestep--and $n_\sigma$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S$ must be before dissipation is added.
+\noindent where $\overline{\Delta S}$ and $\sigma_{\Delta S}$ are the mean and standard deviation of $\Delta S$, respectively--both are calculated using values over the domain for the current time step--and $n_\sigma$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S$ must be before dissipation is added.
 The number of standard deviations, $n_\sigma$, is chosen \textit{a priori}.
 The criteria described in \eqref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead considers the maximum percentage of sites one would want artificial dissipation to be added to.
-A disadvantage of \eqref{eq:dissipation-criteria} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artifical dissipation is unnecessary.
-By requiring that both \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} be met before artifical dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
+A disadvantage of \eqref{eq:dissipation-criteria} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artificial dissipation is unnecessary.
+By requiring that both \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} be met before artificial dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
 
@@
 For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
-The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
+The particle distributions that are missing after the streaming step are solved for by assuming a bounce back of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
 
@@
 In the present work, material behavior is simulated through the collision frequency, $\omega$, when using the BGK collision operator (\Fref{sec:bgk}), or the $s_7$ and $s_8$ relaxation rates when using the MRT collision operator (\Fref{sec:mrt}).
-Both the collision frequency and the $s_7$ and $s_8$ relaxation rates are dependent on the apparent visocity, $\mu_{app}$, through \eqref{eq:colfreq-non-newtonian} and \eqref{eq:mrt-s7}, respectively.
+Both the collision frequency and the $s_7$ and $s_8$ relaxation rates are dependent on the apparent viscosity, $\mu_{app}$, through \eqref{eq:colfreq-non-newtonian} and \eqref{eq:mrt-s7}, respectively.
 The apparent viscosity, in the context of non-Newtonian flow, is a function of the local strain-rate, $\dot{\gamma}$.
@@
 
-Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps.
+Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 time steps.
 The lattice size was chosen to be sufficiently fine for accuracy and the number of time steps were set high enough to ensure a steady-state would be reached.
@@
 A larger stress growth exponent results in a more accurate Papastasiou approximation of the true Bingham plastic constitutive model, however, it leads to more nonphysical oscillations that degrade the numerical solution for the BGK collision operator.
-Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helped to mitigate errors that occured as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
+Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helped to mitigate errors that occurred as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
 In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress fluids (relative errors of approximately 25--75\%).
@@
 For all four cases, the optimal solution was $\theta = 2.0$ and the cost function, $f(2.0)$, was equal to the relative $L_2$ error for the BGK collision operator, $m = 10^8$ and no entropic filtering.
-The results of the optimization show that entropic filtering, at best did not affect the accuracy of LBM, and often adversly affected the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization was used, and regardless of what the $\Delta S$ threshold, $\theta$, was set to.
+The results of the optimization show that entropic filtering, at best did not affect the accuracy of LBM, and often adversely affected the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization was used, and regardless of what the $\Delta S$ threshold, $\theta$, was set to.
 
@@
 The MRT collision operator was therefore the most accurate solution for all cases.
-\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilbirium profiles, $f_i^{eq}$ for the MRT model.
+\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilibrium profiles, $f_i^{eq}$ for the MRT model.
 In contrast to the BGK with $m = 10^8$, there were no significant nonphysical oscillations in the quasiequilibrium distributions.
@@
 Note that none of the collision schemes were sufficiently accurate (error greater than 15\%) for the case of $n = 0.5$, and to improve this accuracy it would be necessary to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
-
-% MG: it might be better to normalize the time with respect to the fastest, BGK with no filtering
 
@@
 A coarse lattice was chosen in order to highlight concerns with stability and accuracy.
-The simulations were run for either 50000 timesteps or until convergence was met.
+The simulations were run for either 50000 time steps or until convergence was met.
 Convergence was defined by:
@@
 \end{equation}
-\noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current timestep.
+\noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current time step.
 The lid velocity was prescribed as $u_o = 0.1$.
@@
 In contrast to the nonphysical nature in which median filtering enhances stability, the stability enhancement used by MRT does not directly affect the macroscopic hydrodynamics of interest, so it again provides a stable, and in most of the cases examined herein, the most accurate solution.
-% MG: verify this sentence is correct
 The apparent superiority in terms of stability and accuracy of the MRT collision operator still comes at a price though.
@@
 For the shear-thinning results, $n = 0.5$, at the lowest Reynold's number considered, $Re = 100$, all of the collision schemes produced results that agreed well with the literature value.
-However, the solutions became unstable at higher Reynold's numbers and entropic filtering was necessary to produce a stable soltuion when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
+However, the solutions became unstable at higher Reynold's numbers and entropic filtering was necessary to produce a stable solution when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
 Although there still appears to be a tradeoff between computational efficiency and accuracy in regards to the BGK with median filtering and MRT with median filtering, it does not seem to be nearly as significant as when simulating Bingham plastic lid-driven flow (e.g.
@@
 \cite{conrad2015accuracy} predicted optimal simulation Mach numbers for various shear-thinning and shear-thickening fluids. 
-Possbile future work would include extending the numerical investigation to consider the effect of the simulation Mach number on the accuracy, stability, and efficiency of the various collision operators and stability enhancements in simulating non-Newtonian flow.
+Possible future work would include extending the numerical investigation to consider the effect of the simulation Mach number on the accuracy, stability, and efficiency of the various collision operators and stability enhancements in simulating non-Newtonian flow.
 

### 2017-03-03 src/stability-paper.tex
added paragraph for Bingham plastic grid convergence results

@@
 
+Lastly, simulations with the same collision schemes, Reynold's numbers and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
+The plastic viscosity was held constant ($\mu_p = 0.2$.), while the yield stress and pressure gradient were varied in order to match the Reynold's numbers and Bingham numbers in \Fref{tab:poise-bing}.
+The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing-2}.
+
+As would be expected from doubling the number of lattice sites, the cost for each corresponding combination of collision scheme, $Re$, and $Bn$ increased by a factor of approximately 2-2.5.
+The 2-2.5 times increase in computing time resulted in the BGK collision schemes on the finer lattice being comparable in computational cost to the MRT collision scheme on the coarser lattice.
+In regards to error, for the larger Bingham number flows ($Bn = 4.04, 8.52$) both the $L_2$ and $L_\infty$ errors of the BGK collision schemes were reduced in general, sometimes by as much as a factor of 4.
+None of the relative $L_2$ errors exceeded 10\%, in contrast to the BGK simulations run on the coarser lattice.
+However, despite the increase in accuracy, it should be noted that for $Bn = 8.52$ all of the BGK collision schemes had $L_\infty$ errors that exceeded 10\%.
+The MRT collision operator, even on the coarser lattice, was more accurate (for the case of $Bn = 8.52$) than all of the BGK collision schemes on the finer lattice.
+
+\begin{table}
+	\centering
+	\caption{Bingham plastic Poiseuille flow, $128 \times 32$ lattice}
+	\vspace{0.5cm}
+	\begin{tabulary}{\linewidth}{l l r r r r r r r r}
+		\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
+		\hline \\
+		BGK & No & $10^5$ & 1.0 & 6.05 & 0.68 & 0.0419 & 0.0450 & 4951 \\
+		              & & & 2.0 & 4.42 & 1.85 & 0.0300 & 0.0370 & 5209 \\
+	                  & & & 3.0 & 3.04 & 4.04 & 0.0206 & 0.0260 & 5296 \\
+		              & & & 4.0 & 1.92 & 8.52 & 0.0649 & 0.1016 & 5304 \\
+		\\
+		BGK & No & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0426 & 0.0470 & 6484 \\
+		              & & & 2.0 & 4.42 & 1.85 & 0.0306 & 0.0394 & 7111 \\
+		              & & & 3.0 & 3.04 & 4.04 & 0.0230 & 0.0438 & 7595 \\
+		              & & & 4.0 & 1.92 & 8.52 & 0.0963 & 0.1659 & 7791 \\
+		\\
+		BGK & Yes & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0427 & 0.0468 & 5576 \\
+		               & & & 2.0 & 4.42 & 1.85 & 0.0309 & 0.0397 & 7512 \\
+		               & & & 3.0 & 3.04 & 4.04 & 0.0236 & 0.0443 & 7856 \\
+		               & & & 4.0 & 1.92 & 8.52 & 0.0934 & 0.1624 & 7848 \\
+		\\
+		MRT & No & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0452 & 0.0484 & 13837 \\
+		              & & & 2.0 & 4.42 & 1.85 & 0.0404 & 0.0435 & 16515 \\
+                      & & & 3.0 & 3.04 & 4.04 & 0.0316 & 0.0338 & 17142 \\
+                      & & & 4.0 & 1.92 & 8.52 & 0.0200 & 0.0210 & 17853 \\
+		\\
+		\label{tab:poise-bing-2}
+	\end{tabulary}
+\end{table}
+
 \subsubsection{Power-law Fluids} \label{sec:poise-powerlaw}
@@
 \begin{equation} \label{eq:analytical-power-law}
-  u_x(y) = \frac{n}{n+1}\left(-\frac{1}{k}\pgrad\right)^{1/n}\left[h^{\frac{n+1}{n}} - |y|^{\frac{n+1}{n}}\right].
+  u_x(y) = \frac{n}{n+1}\left(-\frac{1}{2k}\pgrad\right)^{1/n}\left[h^{\frac{n+1}{n}} - |y|^{\frac{n+1}{n}}\right].
 \end{equation}
@@
 \hline \\
-BGK & No & 0.50 & 0.0007 & 28.06 & 30.37 & 5232 \\
+BGK & No & 0.50 & 0.0007 & 28.1 & 30.4 & 5232 \\
 & & 0.75 & 0.9125 & 0.0328 & 0.0569 & 2399 \\
-& & 1.25 & 423.2 & 0.0051 & 0.0055 & 2311 \\
-& & 1.50 & 2213 & 1.0 & 1.0 & 3622 \\
-\\
-BGK & Yes & 0.50 & 0.0007 & 26.72 & 28.78 & 5347 \\
+& & 1.25 & 423.2 & 0.00510 & 0.00547 & 2311 \\
+& & 1.50 & 2213 & 1.00 & 1.00 & 3622 \\
+\\
+BGK & Yes & 0.50 & 0.0007 & 26.7 & 28.8 & 5347 \\
 & & 0.75 & 0.9125 & 0.0328 & 0.0569 & 2385 \\
-& & 1.25 & 423.2 & 0.0051 & 0.0055 & 2275 \\
+& & 1.25 & 423.2 & 0.00510 & 0.00547 & 2275 \\
 & & 1.50 & 2213 & 0.0570 & 0.0600 & 2357 \\
 \\
-MRT & No & 0.50 & 0.0007 & 0.1758 & 0.1401 & 4906 \\
-& & 0.75 & 0.9125 & 0.0058 & 0.0045 & 4880 \\
-& & 1.25 & 423.2 & 0.0051 & 0.0055 & 4863 \\
-& & 1.50 & 2213 & 1.0 & 1.0 & 15992 \\
-\\
-BGK-BRT & No & 0.50 & 0.0007 & 11.73 & 12.56 & 4450 \\
+MRT & No & 0.50 & 0.0007 & 0.140 & 0.176 & 4906 \\
+& & 0.75 & 0.9125 & 0.00450 & 0.00579 & 4880 \\
+& & 1.25 & 423.2 & 0.00506 & 0.00548 & 4863 \\
+& & 1.50 & 2213 & 1.00 & 1.00 & 15992 \\
+\\
+BGK-BRT & No & 0.50 & 0.0007 & 11.7 & 12.6 & 4450 \\
 & & 0.75 & 0.9125 & 0.0320 & 0.0475 & 1805 \\
-& & 1.25 & 423.2 & 0.0051 & 0.0055 & 1905 \\
-& & 1.50 & 2213 & 1.0 & 1.0 & 8194 \\
+& & 1.25 & 423.2 & 0.00510 & 0.00547 & 1905 \\
+& & 1.50 & 2213 & 1.00 & 1.00 & 8194 \\
 \\
 \label{tab:poise-power-law}
+\end{tabulary}
+\end{table}
+
+In addition, simulations with the same collision schemes and Reynold's numbers were run in $128 \times 32$ lattices in order to investigate the effect of grid resolution.
+The flow consistency index was held constant ($k = 0.2$.), while the pressure gradient was varied in order to match the Reynold's numbers in \Fref{tab:poise-power-law}.
+The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-power-law-2}.
+
+\begin{table}
+	\centering
+	\caption{Power-law Poiseuille flow, $128 \times 32$ lattice}
+	\vspace{0.5cm}
+	\begin{tabulary}{\linewidth}{l l r r r r r}
+		\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $n$ & $Re$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
+		\hline \\
+		BGK & No & 0.50 & 0.0007 & 17.4 & 20.8 & 8477 \\
+		       & & 0.75 & 0.9125 & 0.0140 & 0.0288 & 4317 \\
+		       & & 1.25 & 423.2 & 0.479 & 0.502 & 4683 \\
+		       & & 1.50 & 2213 & 0.873 & 0.897 & 4457 \\
+		\\
+		BGK & Yes & 0.50 & 0.0007 & 17.6 & 21.0 & 8407 \\
+		        & & 0.75 & 0.9125 & 0.0140 & 0.0288 & 4738 \\
+		        & & 1.25 & 423.2 & 0.479 & 0.502 & 4714 \\
+		        & & 1.50 & 2213 & 0.874 & 0.898 & 4879 \\
+		\\
+		MRT & No & 0.50 & 0.0007 & 0.0675 & 0.0539 & 8338 \\
+		       & & 0.75 & 0.9125 & 0.00187 & 0.00145 & 9241 \\
+		       & & 1.25 & 423.2 & 0.479 & 0.502 & 10856 \\
+		       & & 1.50 & 2213 & 0.872 & 0.898 & 9965 \\
+		\\
+		BGK-BRT & No & 0.50 & 0.0007 & 16.8 & 18.4 & 8959 \\
+		           & & 0.75 & 0.9125 & 0.0129 & 0.0158 & 4516 \\
+		           & & 1.25 & 423.2 & 0.479 & 0.502 & 4308 \\
+		           & & 1.50 & 2213 & 0.873 & 0.897 & 8194 \\
+		\\
+		\label{tab:poise-power-law-2}
 	\end{tabulary}
@@
 \begin{equation} \label{eq:gradu-power-law}
-  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{k} \pgrad \right)^{1/n} \left( -y |y|^{\frac{1}{n} - 1} \right).
+  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{2k} \pgrad \right)^{1/n} \left( -y |y|^{\frac{1}{n} - 1} \right).
 \end{equation}
@@
 
-The lid-driven cavity simulations presented in the section were all simulated on a relatively coarse, $100 \times 100$ lattice, similar to the work presented in ~\cite{brownlee2008nonequilibrium}.
+Unless specified otherwise, the lid-driven cavity simulations presented in the section were all simulated on a relatively coarse, $100 \times 100$ lattice, similar to the work presented in ~\cite{brownlee2008nonequilibrium}.
 A coarse lattice was chosen in order to highlight concerns with stability and accuracy.
@@
 
-For the Bingham plastic numerical tests the Reynold's number was varied with the following values:$\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the Bingham number was varied with the following values: $\begin{bmatrix}1,&10,&100\end{bmatrix}$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
+For the Bingham plastic numerical tests the Reynold's number was varied with the following values: $\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the Bingham number was varied with the following values: $\begin{bmatrix}1,&10,&100\end{bmatrix}$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
 Six collision schemes were tested: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
@@
 A smaller stress growth exponent is also more effective at producing accurate results in the BGK collision scheme than placing bounds on the relaxation time (i.e. BGK-BRT).
-However, although the BGK with $m = 10^5$ was the fastest model in all cases, it was also similar to the other two BGK models without median filtering in that it was unstable for $Re \ge 5000$.
+However, although the BGK with $m = 10^5$ was the fastest model in all cases, it was also unstable for $Re \ge 5000$.
+The collision schemes that were unstable for $Re \ge 5000$ (BGK with $m = 10^5$, BGK with $m = 10^8$, and BGK-BRT with $m = 10^8$) were run with a finer $200 \times 200$ lattice to see if increased lattice resolution could increase stability without causing a significant increase computational efficiency.
+With the finer lattice, the BGK models without median filtering were numerically unstable and took 85\% to 308\% of the computational time of the corresponding $100 \times 100$ lattice MRT model.
+BGK tests conducted on a $300 \times 300$ lattice with $m = 10^8$ (without entropic filtering) for Bingham numbers of $\begin{bmatrix}1,&10\end{bmatrix}$ and Reynold’s numbers of $\begin{bmatrix}5000,&10000\end{bmatrix}$ also yielded unstable results.
 
@@
 The MRT collision operator was, in general, 5-10 times slower than any of the BGK collision schemes.
-The extreme increase in computational expense is probably not due to the collision process itself (i.e.
+The increase in computational expense is probably not due to the collision process itself (i.e.
 \eqref{eq:mrt-colop}) but instead due to calculating the strain-rate \eqref{eq:srtensor-mrt} for each iteration of the solution for the apparent viscosity, $\mu_{app}$.
@@
 
+Lastly, it should be noted that for instances in which the BGK collision operator
+
 \section{Conclusions} \label{sec:numerical-study-conclusions}
@@
 Possible future work would include extending the numerical investigation to consider the effect of the simulation Mach number on the accuracy, stability, and efficiency of the various collision operators and stability enhancements in simulating non-Newtonian flow.
-
-% MG: should we mention: code is open source? input files used in this study are also freely available?

### 2017-03-03 response-to-reviewers.tex
removed Poiseuille power law stuff, finished draft of reviewer responses

@@
-\documentclass[a4paper,10pt]{article}
-\usepackage[utf8]{inputenc}
+\documentclass{article}
 
@@
 
-\section{Response to reviewers}
+\section*{Response to reviewers}
 
@@
 
-The following is a list of all comments recieved, and how they have been addressed in the updated version:
+The following is a list of all comments received, and how they have been addressed in the updated version:
+
+  \subsection{Reviewer 1}
+	\begin{enumerate}
+		\item \emph{In Fig.
+2, the scale of the figure should be modified in order to avoid readers’ misleading.
+If the authors calculate only one case with L = 32 and H = 64, the computational domain becomes vertically long.}
+
+		Thank you. This is an excellent point. The figure has been updated to reflect the relative scale between $L$ and $H$.
+
+		\item \emph{In Table 2, all the results of the error norms, $L_{2}$ and $L_{\infty}$, for $n = 1.25$ are the same values in spite of different LBM schemes.
+The values should be changed to correct ones.}
+
+		%The results are actually the same for all of the BGK variants.
+		%This is because for $n = 1.25$ the nonequilibrium entropy threshold for median filtering was not exceeded, and hence it was effectively the same as the unfiltered BGK scheme.
+		%Similarly, the relaxation time was within the bounds for the BGK-BRT scheme, which again resulted in the scheme to being effectively the same as the BGK collision scheme without bounds.
+		%The error norms for the MRT collision scheme was different, but the difference could not be seen in the precision of the results presented in the previous draft.
+		%The $n = 1.25$ error norms have been updated with higher precision so as to show the difference.
+
+		The Table referenced in the above comment (previously Table 2), along with the power-law Poiseuille flow section, have been removed.
+We decided to remove the power-law Poiseuille flow section because we do not believe it added much value to the manuscript, and because we wanted to address grid resolution effects without lengthening the manuscript.
+
+		\item \emph{The authors carried out the simulations with fixed lattice resolution.
+However, the convergence rate should be examined by calculating the benchmark problem (e.g., Poiseuille flow) with different lattice resolutions.
+		What order is the accuracy of the schemes used in the present study?}
+
+		Considerations regarding grid resolution were added in the following locations:
+		\begin{itemize}
+			\item Table 2 at the end of Section 3.1.1 and the last paragraph of Section 3.1.1 address the significance of grid resolution for Bingham plastic Poiseuille flow.
+			\item The third to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
+		\end{itemize}
+
+		\item \emph{Several grammatical errors occur in the text. The authors should be checked again throughout the manuscript.}
+
+		The manuscript was checked carefully for spelling and grammatical errors. We appreciate you pointing this out and we hope that no more grammatical errors persist.
+
+	\end{enumerate}
+
+	\subsection{Reviewer 2}
+	  The present work examines the single-relaxation-time (SRT) and the multiple-relaxation-time (MRT) lattice Boltzmann method (LBM) for Bingham plastic and power-law fluids.
+Constitutive relations in these fluids are typically stiff presenting substantial numerical challenges when the relations are incorporated into the LBM collision models.
+Entropic filtering as additional artificial dissipation and MRT collision are considered as the means to stabilize otherwise oscillatory or unstable SRT LBM.
+Optimal value for the median filtering was suggested and computational costs with these approaches are compared in detail.
+The conclusions are rather trivial; MRT is the most accurate and stable whereas its computational cost can be significantly large compared with SRT due to non-linear iteration.
+Filtering does stabilize SRT but the accuracy deteriorates particularly when the filter threshold is not optimally chosen.
+This work can be useful if readers are interested in choosing a suitable LB model for simulating non-Newtonian flows.
+However, it lacks novelty.
+Although MRT collision operator is shown to be the best choice, no detailed analysis on the values of optimum relaxation rates for various non-Newtonian flows.
+
+	As for Poiseulle Flow part, SRT simulations deviate from the analytical solution in the core region (Figs.3,4) due to the difference in the constitutive relations used in LBM and the analytical solution.
+Thus, the reported errors in Table 1 include both constitutive relation error and truncation error.
+This does not seem fair.
+It would be helpful if the analytical solution or a benchmark solution from NS equations of Papanastasiou approximation is directly compared with LBM results.
+The authors attributed the nonphysical oscillations of SRT simulations to sharp gradients in the velocity near the walls, which is then again a result of a sharp gradient in the viscosity.
+However, from Figs.3\&4, it is difficult to observe any sign of sharp gradients in the velocity near the walls.
+In Fig.4 for higher m, the oscillations occur at the center of the computational domain and they are not symmetric.
+
+	Grid convergence and its effect on the computation time would strengthen the manuscript.
+It is possible that MRT might become faster as the grid resolution improves.
+A minor comment is that as the streamwise BC is periodic in Sec.3.1, one might not need L=32 grid.
+One or two grid points would be enough.
+
+	\begin{itemize}
+	\item \emph{As for Poiseulle Flow part, SRT simulations deviate from the analytical solution in the core region (Figs.3,4) due to the difference in the constitutive relations used in LBM and the analytical solution.
+Thus, the reported errors in Table 1 include both constitutive relation error and truncation error.
+This does not seem fair.
+It would be helpful if the analytical solution or a benchmark solution from NS equations of Papanastasiou approximation is directly compared with LBM results.}
+
+	This is a good point, and should be both addressed and justified in the manuscript.
+We chose to compare with the analytical solution for a Bingham plastic, instead of a solution or approximation for the Papanastasiou constitutive approximation, because the Bingham plastic constitutive behavior is the behavior of interest.
+In addition, if the error was calculated with respect to the solution or approximation for the Papanastasiou constitutive approximation, then collision schemes with different stress growth exponents would have their error computed with respect to different solutions.
+Thus, it would be difficult to compare the relative errors between collision schemes.
+Lastly, note that although the errors computed with the respect to the Bingham plastic analytical solution do contain both constitutive relation error and truncation error, it can be seen from the accuracy of the MRT results that the constitutive relation error is likely negligible compared to truncation error and error due to nonphysical oscillations.
+
+	The comment below equation 28 (page 17) has been expanded to include the above reasoning.
+
+	\item \emph{The authors attributed the nonphysical oscillations of SRT simulations to sharp gradients in the velocity near the walls, which is then again a result of a sharp gradient in the viscosity.
+However, from Figs.3\&4, it is difficult to observe any sign of sharp gradients in the velocity near the walls.
+In Fig.4 for higher m, the oscillations occur at the center of the computational domain and they are not symmetric.}
+
+	We believe that the oscillations are more pronounced at the center because the velocity is prescribed at the wall, but as particle distributions advect from the wall toward the center overrelaxation can occur.
+It would be interesting to explore the exact cause of the oscillations more in future work.
+
+	The oscillations are not symmetric because Fig.
+4 is merely a snapshot in time.
+They tend to move back and forth across the channel in subsequent time steps such that, on average (in time), the oscillations are symmetric.
+
+	\item \emph{Grid convergence and its effect on the computation time would strengthen the manuscript. It is possible that MRT might become faster as the grid resolution improves.}
+
+	As mentioned in the responses to Reviewer 1, considerations regarding grid resolution were added in the following locations:
+	\begin{itemize}
+		\item Table 2 at the end of Section 3.1.1 and the last paragraph of Section 3.1.1 address the significance of grid resolution for Bingham plastic Poiseuille flow.
+		\item The second to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
+	\end{itemize}
+
+	\end{itemize}
+
+Thank you again for your help. Please let us know of any further improvements that can be made to this paper.
 

### 2017-03-03 src/stability-paper.tex
removed Poiseuille power law stuff, finished draft of reviewer responses

@@
 
-\subsection{Poiseuille Flow}
-
-Poiseuille flow is a useful benchmark because analytical solutions exist for both Bingham plastic and power-law fluids.
+\subsection{Bingham Plastic, Poiseuille Flow}
+
+Poiseuille flow is a useful benchmark because an analytical solution exists for both Bingham plastic flow.
 Poiseuille flow is created with a two-dimensional channel flow that is driven by a constant pressure gradient, $\pgrad$, through the domain.
@@
 In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
-For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$, which follows the recommendation of~\cite{lallemand2000theory} for reasons of stability, and has been successfully used to simulate non-Newtonian flow in the past~\cite{fallah2012multiple,grasinger2015simulation}.
-
-\subsubsection{Bingham Plastic Fluids} \label{sec:poise-bing}
 
@@
 Recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer to the true Bingham plastic behavior.
+For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$, which follows the recommendation of~\cite{lallemand2000theory} for reasons of stability, and has been successfully used to simulate non-Newtonian flow in the past~\cite{fallah2012multiple,grasinger2015simulation}.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
@@
 \noindent where $y_{\tau} = -\tau_y / \pgrad$ is the vertical location at which the fluid yields.
-(Note that the analytical solution is based on the exact Bingham plastic model, and not the Papanastasiou approximation.)
+(Note that the analytical solution is based on the exact Bingham plastic model, and not the Papanastasiou approximation.
+The choice was made to compare with the analytical solution for a Bingham plastic, instead of an analytical solution or benchmark for the Papanastasiou constitutive approximation, because the Bingham plastic constitutive behavior is the behavior of interest.
+In addition, if the error was calculated with respect to the solution or approximation for the Papanastasiou constitutive approximation, then collision schemes with different stress growth exponents would have their error computed with respect to different solutions.
+Thus, it would be difficult to compare the relative errors between collision schemes.
+Lastly, note that although the errors computed with the respect to the Bingham plastic analytical solution do contain error in the constitutive relation, it can be seen from the accuracy of the MRT collision scheme, and others, that the constitutive relation error is likely negligible compared to other sources of error.)
 
@@
 
-As would be expected from doubling the number of lattice sites, the cost for each corresponding combination of collision scheme, $Re$, and $Bn$ increased by a factor of approximately 2-2.5.
-The 2-2.5 times increase in computing time resulted in the BGK collision schemes on the finer lattice being comparable in computational cost to the MRT collision scheme on the coarser lattice.
+As would be expected from doubling the number of lattice sites, the cost for each corresponding combination of collision scheme, $Re$, and $Bn$ increased by a factor of approximately 2--2.5.
+The 2--2.5 times increase in computing time resulted in the BGK collision schemes on the finer lattice being comparable in computational cost to the MRT collision scheme on the coarser lattice.
 In regards to error, for the larger Bingham number flows ($Bn = 4.04, 8.52$) both the $L_2$ and $L_\infty$ errors of the BGK collision schemes were reduced in general, sometimes by as much as a factor of 4.
@@
 \end{table}
-
-\subsubsection{Power-law Fluids} \label{sec:poise-powerlaw}
-
-The simulations of the Poiseuille flow with the power-law constitutive relationship were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a flow consistency index of $k = 0.2$.
-The flow behavior index, $n$, was varied between the four different values (from shear-thinning to shear-thickening): $\begin{bmatrix}0.5,&0.75,&1.25,& 1.50\end{bmatrix}$, and four different LBM schemes were used: (1) BGK, (2) BGK with median filtering, (3) MRT, and (4) BGK-BRT, with $\omega$ bounded within $\begin{bmatrix}0.05,&1.995\end{bmatrix}$.
-The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
-The analytical solution for the velocity profile of Poiseuille flow with the power-law constitutive relationship is given by:
-\begin{equation} \label{eq:analytical-power-law}
-  u_x(y) = \frac{n}{n+1}\left(-\frac{1}{2k}\pgrad\right)^{1/n}\left[h^{\frac{n+1}{n}} - |y|^{\frac{n+1}{n}}\right].
-\end{equation}
-
-The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-power-law}.
-The Reynold's number was computed by $Re = \frac{\rho U^{2-n} H^n}{k}$.
-Although for $n = 0.5, 0.75, 1.25$ the MRT collision operator was more accurate and required more computation time than the BGK collision operator, for $n = 1.5$ the BGK operator with median filtering required the least computation time and was the most accurate with respect to the analytical solution.
-In fact, besides the case of $n = 0.5$ (i.e. relatively extreme shear-thinning) the BGK with median filtering was both computationally efficient and accurate.
-The BGK-BRT had a moderate improvement of accuracy and efficiency--with respect to the regular BGK collision scheme--for flow where the constitutive relationship was only somewhat nonlinear, i.e.
-for $n = 0.75, 1.25$.
-Note that none of the collision schemes were sufficiently accurate (error greater than 15\%) for the case of $n = 0.5$, and to improve this accuracy it would be necessary to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
-
-\begin{table}
-\centering
-\caption{Power-law Poiseuille flow}
-\vspace{0.5cm}
-\begin{tabulary}{\linewidth}{l l r r r r r}
-\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $n$ & $Re$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
-\hline \\
-BGK & No & 0.50 & 0.0007 & 28.1 & 30.4 & 5232 \\
-& & 0.75 & 0.9125 & 0.0328 & 0.0569 & 2399 \\
-& & 1.25 & 423.2 & 0.00510 & 0.00547 & 2311 \\
-& & 1.50 & 2213 & 1.00 & 1.00 & 3622 \\
-\\
-BGK & Yes & 0.50 & 0.0007 & 26.7 & 28.8 & 5347 \\
-& & 0.75 & 0.9125 & 0.0328 & 0.0569 & 2385 \\
-& & 1.25 & 423.2 & 0.00510 & 0.00547 & 2275 \\
-& & 1.50 & 2213 & 0.0570 & 0.0600 & 2357 \\
-\\
-MRT & No & 0.50 & 0.0007 & 0.140 & 0.176 & 4906 \\
-& & 0.75 & 0.9125 & 0.00450 & 0.00579 & 4880 \\
-& & 1.25 & 423.2 & 0.00506 & 0.00548 & 4863 \\
-& & 1.50 & 2213 & 1.00 & 1.00 & 15992 \\
-\\
-BGK-BRT & No & 0.50 & 0.0007 & 11.7 & 12.6 & 4450 \\
-& & 0.75 & 0.9125 & 0.0320 & 0.0475 & 1805 \\
-& & 1.25 & 423.2 & 0.00510 & 0.00547 & 1905 \\
-& & 1.50 & 2213 & 1.00 & 1.00 & 8194 \\
-\\
-\label{tab:poise-power-law}
-\end{tabulary}
-\end{table}
-
-In addition, simulations with the same collision schemes and Reynold's numbers were run in $128 \times 32$ lattices in order to investigate the effect of grid resolution.
-The flow consistency index was held constant ($k = 0.2$.), while the pressure gradient was varied in order to match the Reynold's numbers in \Fref{tab:poise-power-law}.
-The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-power-law-2}.
-
-\begin{table}
-	\centering
-	\caption{Power-law Poiseuille flow, $128 \times 32$ lattice}
-	\vspace{0.5cm}
-	\begin{tabulary}{\linewidth}{l l r r r r r}
-		\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $n$ & $Re$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
-		\hline \\
-		BGK & No & 0.50 & 0.0007 & 17.4 & 20.8 & 8477 \\
-		       & & 0.75 & 0.9125 & 0.0140 & 0.0288 & 4317 \\
-		       & & 1.25 & 423.2 & 0.479 & 0.502 & 4683 \\
-		       & & 1.50 & 2213 & 0.873 & 0.897 & 4457 \\
-		\\
-		BGK & Yes & 0.50 & 0.0007 & 17.6 & 21.0 & 8407 \\
-		        & & 0.75 & 0.9125 & 0.0140 & 0.0288 & 4738 \\
-		        & & 1.25 & 423.2 & 0.479 & 0.502 & 4714 \\
-		        & & 1.50 & 2213 & 0.874 & 0.898 & 4879 \\
-		\\
-		MRT & No & 0.50 & 0.0007 & 0.0675 & 0.0539 & 8338 \\
-		       & & 0.75 & 0.9125 & 0.00187 & 0.00145 & 9241 \\
-		       & & 1.25 & 423.2 & 0.479 & 0.502 & 10856 \\
-		       & & 1.50 & 2213 & 0.872 & 0.898 & 9965 \\
-		\\
-		BGK-BRT & No & 0.50 & 0.0007 & 16.8 & 18.4 & 8959 \\
-		           & & 0.75 & 0.9125 & 0.0129 & 0.0158 & 4516 \\
-		           & & 1.25 & 423.2 & 0.479 & 0.502 & 4308 \\
-		           & & 1.50 & 2213 & 0.873 & 0.897 & 8194 \\
-		\\
-		\label{tab:poise-power-law-2}
-	\end{tabulary}
-\end{table}
-
-Before closing the discussion on the Poiseuille flow numerical study, it is worthwhile to consider what precisely is common between the higher-yield stress fluids and the shear-thinning fluids, as they were the cases that exhibited numerical instabilities and were simulated with the least amount of accuracy for some of the collision schemes considered.
-One such commonality can be seen when considering the effect of the yield stress, $\tau_y$, and the effect of the flow behavior index, $n$, on the gradient of the theoretical flow velocity.
-Taking the partial derivative with respect to $y$ of the analytical solution for the flow velocity of a Bingham plastic, Poiseuille flow (Equation \eqref{eq:analytical-bingham}) results in:
-\begin{equation} \label{eq:gradu-bingham}
-  \frac{\partial u_x}{\partial y} = \begin{cases}
-    0, & 0 \leq |y| \leq y_{\tau}, \\
-    \frac{1}{\mu_p} \left( \pgrad y + \tau_y \frac{y}{|y|} \right), & y_{\tau} < |y| \leq h,
-  \end{cases}
-\end{equation}
-and taking the partial derivative with respect to $y$ of the analytical solution for the flow velocity of a power-law fluid, Poiseuille flow (Equation \eqref{eq:analytical-power-law}) results in:
-\begin{equation} \label{eq:gradu-power-law}
-  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{2k} \pgrad \right)^{1/n} \left( -y |y|^{\frac{1}{n} - 1} \right).
-\end{equation}
-It can be seen from \eqref{eq:gradu-bingham} and \eqref{eq:gradu-power-law} that $\left| \frac{\partial u_x}{\partial y} \right|$ increases with the yield stress for Bingham plastics, and increases with $\frac{1}{n}$ for power-law fluids, which suggests that one of the challenges associated with the accuracy and stability of simulating higher yield-stress fluids and more extreme shear-thinning fluids (e.g.
-$n \le 0.5$) in Poiseuille-type flow is a result of having a larger magnitude of the gradient of macroscopic flow.
-This conclusion is consistent with the discussion in \Fref{sec:bgk} with regards to one of the drawbacks of the BGK collision operator, namely that a sharp spatial gradient in either $\rho$ or $\mvel$ can lead to nonphysical oscillations that can degrade the accuracy and stability of the solution.
-However, it does not explain why the MRT collision operator was unable to accurately simulate the case of a power-law fluid, Poiseuille flow, with $n = 0.5$.
 
@@
 
-Lastly, it should be noted that for instances in which the BGK collision operator
-
 \section{Conclusions} \label{sec:numerical-study-conclusions}

### 2017-03-03 response-to-reviewers.tex
added tables

--- 2017-03-02 response-to-reviewers.tex
@@
-\documentclass[a4paper,10pt]{article}
-\usepackage[utf8]{inputenc}
+\documentclass{article}
 
 \begin{document}
+	\section{Reviewer 1}
+	\begin{enumerate}
+		\item \emph{In Fig.
+2, the scale of the figure should be modified in order to avoid readers’ misleading.
+If the authors calculate only one case with L = 32 and H = 64, the computational domain becomes vertically long.}
 
-\section{Response to reviewers}
+		The figure has been updated to reflect the relative scale between $L$ and $H$.
 
-Thank you very much for your assistance in reviewing this paper.
-We are sure that with your help, this has become a much better work.
-We hope that the attached edited copy of "Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow" will answer any questions previously raised, and will be satisfactory for publication.
+		\item \emph{In Table 2, all the results of the error norms, $L_{2}$ and $L_{\infty}$, for $n = 1.25$ are the same values in spite of different LBM schemes.
+The values should be changed to correct ones.}
 
-The following is a list of all comments recieved, and how they have been addressed in the updated version:
+		The results are actually the same for all of the BGK variants.
+		This is because for $n = 1.25$ the nonequilibrium entropy threshold for median filtering was not exceeded, and hence it was effectively the same as the unfiltered BGK scheme.
+		Similarly, the relaxation time was within the bounds for the BGK-BRT scheme, which again resulted in the scheme to being effectively the same as the BGK collision scheme without bounds.
+		The error norms for the MRT collision scheme was different, but the difference could not be seen in the precision of the results presented in the previous draft.
+		The $n = 1.25$ error norms have been updated with higher precision so as to show the difference.
 
+		\item \emph{The authors carried out the simulations with fixed lattice resolution.
+However, the convergence rate should be examined by calculating the benchmark problem (e.g., Poiseuille flow) with different lattice resolutions.
+		What order is the accuracy of the schemes used in the present study?}
+
+		Considerations regarding grid resolution were added in the following locations:
+		\begin{itemize}
+			\item Table 2 at the end of Section 3.1.1 and the last paragraph of Section 3.1.1 address the significance of grid resolution for Bingham plastic Poiseuille flow.
+			\item Table 4 at the end of Section 3.1.2 and the last paragraph of Section 3.1.2 address the significance of grid resolution for Power-law Poiseuille flow. 
+			\item The third to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
+		\end{itemize}
+	\end{enumerate}
+
+	\section{Reviewer 2}
+	  The present work examines the single-relaxation-time (SRT) and the multiple-relaxation-time (MRT) lattice Boltzmann method (LBM) for Bingham plastic and power-law fluids.
+Constitutive relations in these fluids are typically stiff presenting substantial numerical challenges when the relations are incorporated into the LBM collision models.
+Entropic filtering as additional artificial dissipation and MRT collision are considered as the means to stabilize otherwise oscillatory or unstable SRT LBM.
+Optimal value for the median filtering was suggested and computational costs with these approaches are compared in detail.
+The conclusions are rather trivial; MRT is the most accurate and stable whereas its computational cost can be significantly large compared with SRT due to non-linear iteration.
+Filtering does stabilize SRT but the accuracy deteriorates particularly when the filter threshold is not optimally chosen.
+This work can be useful if readers are interested in choosing a suitable LB model for simulating non-Newtonian flows.
+However, it lacks novelty.
+Although MRT collision operator is shown to be the best choice, no detailed analysis on the values of optimum relaxation rates for various non-Newtonian flows.
+	As for Poiseulle Flow part, SRT simulations deviate from the analytical solution in the core region (Figs.3,4) due to the difference in the constitutive relations used in LBM and the analytical solution.
+Thus, the reported errors in Table 1 include both constitutive relation error and truncation error.
+This does not seem fair.
+It would be helpful if the analytical solution or a benchmark solution from NS equations of Papanastasiou approximation is directly compared with LBM results.
+The authors attributed the nonphysical oscillations of SRT simulations to sharp gradients in the velocity near the walls, which is then again a result of a sharp gradient in the viscosity.
+However, from Figs.3&4, it is difficult to observe any sign of sharp gradients in the velocity near the walls.
+In Fig.4 for higher m, the oscillations occur at the center of the computational domain and they are not symmetric.
+
+	Grid convergence and its effect on the computation time would strengthen the manuscript.
+It is possible that MRT might become faster as the grid resolution improves.
+A minor comment is that as the streamwise BC is periodic in Sec.3.1, one might not need L=32 grid.
+One or two grid points would be enough.
 \end{document}

### 2017-03-03 src/stability-paper.tex
added tables

--- 2017-03-03 src/stability-paper.tex
@@
 Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
-Yield stress flow has been found to be relevant in many applications, such as flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
+Yield stress flow has been found to be relevant in many applictions, such as flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
 
@@
 BGK consists of a single relaxation time and is a linear relaxation of particle distributions toward equilibrium.
-The BGK collision operator for the $i^{th}$ discrete velocity is expressed as:
+The BGK collison operator for the $i^{th}$ discrete velocity is expressed as:
 
@@
 In this case, if $f_i$ is ``near'' to quasiequilibrium at $\pos$ it will be ``far'' from quasiequilibrium after the streaming step when it moves to the node at $\pos + \pvel_i \Delta t$.
-Overrelaxation will occur if $\nu \rightarrow 0$ because consequently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
+Overrelaxation will occur if $\nu \rightarrow 0$ because consquently $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_i(\pos + \pvel_i \Delta t,t + \Delta t)$ still being ``far'' from $f^{eq}_i(\pos + \pvel_i \Delta t,t + \Delta t)$ but on the ``other side''.
 Overrelaxation in subsequent time steps (along the streaming trajectory of $f_i$) could result in nonphysical oscillations.
@@
 approaching 2) or too low (e.g.
-approaching 0), a natural, albeit nonphysical, approach to using the BGK collision operator for non-Newtonian flow is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
+approaching 0), a natural, albiet nonphysical, approach to using the BGK collision operator for non-Newtonian flow is to simply put bounds on the values in which the collision frequency may attain (such as in~\citet{svec2011flow,svec2012free,gabbanelli2005lattice}).
 This simple methodology for increasing stability will henceforth be referred to bounded-relaxation time BGK, or BGK-BRT.
@@
 
-\noindent where $\overline{\Delta S}$ and $\sigma_{\Delta S}$ are the mean and standard deviation of $\Delta S$, respectively--both are calculated using values over the domain for the current time step--and $n_\sigma$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S$ must be before dissipation is added.
+\noindent where $\overline{\Delta S}$ and $\sigma_{\Delta S}$ are the mean and standard deviation of $\Delta S$, respectively--both are calculated using values over the domain for the current timestep--and $n_\sigma$ is the number of standard deviations greater than $\overline{\Delta S}$ that $\Delta S$ must be before dissipation is added.
 The number of standard deviations, $n_\sigma$, is chosen \textit{a priori}.
 The criteria described in \eqref{eq:dissipation-criteria} has the advantage that one does not need to determine \textit{a priori} what constitutes ``far'' from quasiequilibrium, but instead considers the maximum percentage of sites one would want artificial dissipation to be added to.
-A disadvantage of \eqref{eq:dissipation-criteria} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artificial dissipation is unnecessary.
-By requiring that both \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} be met before artificial dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
+A disadvantage of \eqref{eq:dissipation-criteria} is that if $\overline{\Delta S}$ and $n_\sigma$ are both small then it is possible for dissipation to be added when $\Delta S(\pos, t)$ is low and artifical dissipation is unnecessary.
+By requiring that both \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} be met before artifical dissipation is added, there is the potential for \eqref{eq:ds-threshold} and \eqref{eq:dissipation-criteria} to compensate for the each other's disadvantage.
 
@@
 For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
-The particle distributions that are missing after the streaming step are solved for by assuming a bounce back of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
+The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.
 
@@
 In the present work, material behavior is simulated through the collision frequency, $\omega$, when using the BGK collision operator (\Fref{sec:bgk}), or the $s_7$ and $s_8$ relaxation rates when using the MRT collision operator (\Fref{sec:mrt}).
-Both the collision frequency and the $s_7$ and $s_8$ relaxation rates are dependent on the apparent viscosity, $\mu_{app}$, through \eqref{eq:colfreq-non-newtonian} and \eqref{eq:mrt-s7}, respectively.
+Both the collision frequency and the $s_7$ and $s_8$ relaxation rates are dependent on the apparent visocity, $\mu_{app}$, through \eqref{eq:colfreq-non-newtonian} and \eqref{eq:mrt-s7}, respectively.
 The apparent viscosity, in the context of non-Newtonian flow, is a function of the local strain-rate, $\dot{\gamma}$.
@@
 
-Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 time steps.
+Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 timesteps.
 The lattice size was chosen to be sufficiently fine for accuracy and the number of time steps were set high enough to ensure a steady-state would be reached.
@@
 A larger stress growth exponent results in a more accurate Papastasiou approximation of the true Bingham plastic constitutive model, however, it leads to more nonphysical oscillations that degrade the numerical solution for the BGK collision operator.
-Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helped to mitigate errors that occurred as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
+Upon inspection of \Fref{tab:poise-bing}, it does not appear that entropic median filtering helped to mitigate errors that occured as a result of using $m = 10^8$ for Bingham plastic Poiseuille flow.
 In fact, the median filter resulted in a less accurate solution in all cases other than the case with the lowest yield stress considered (i.e., the smoothest flow field), and rendered the solution altogether useless for the higher yield stress fluids (relative errors of approximately 25--75\%).
@@
 For all four cases, the optimal solution was $\theta = 2.0$ and the cost function, $f(2.0)$, was equal to the relative $L_2$ error for the BGK collision operator, $m = 10^8$ and no entropic filtering.
-The results of the optimization show that entropic filtering, at best did not affect the accuracy of LBM, and often adversely affected the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization was used, and regardless of what the $\Delta S$ threshold, $\theta$, was set to.
+The results of the optimization show that entropic filtering, at best did not affect the accuracy of LBM, and often adversly affected the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization was used, and regardless of what the $\Delta S$ threshold, $\theta$, was set to.
 
@@
 The MRT collision operator was therefore the most accurate solution for all cases.
-\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilibrium profiles, $f_i^{eq}$ for the MRT model.
+\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilbirium profiles, $f_i^{eq}$ for the MRT model.
 In contrast to the BGK with $m = 10^8$, there were no significant nonphysical oscillations in the quasiequilibrium distributions.
@@
 
-Lastly, simulations with the same collision schemes, Reynold's numbers and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
+Lastly, simulations with the same collision schemes, Reynold's numbers and Bingham numbers were run in $128 \times 32$ lattices in order to investigate the effect of grid resolution.
 The plastic viscosity was held constant ($\mu_p = 0.2$.), while the yield stress and pressure gradient were varied in order to match the Reynold's numbers and Bingham numbers in \Fref{tab:poise-bing}.
 The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing-2}.
-
-As would be expected from doubling the number of lattice sites, the cost for each corresponding combination of collision scheme, $Re$, and $Bn$ increased by a factor of approximately 2-2.5.
-The 2-2.5 times increase in computing time resulted in the BGK collision schemes on the finer lattice being comparable in computational cost to the MRT collision scheme on the coarser lattice.
-In regards to error, for the larger Bingham number flows ($Bn = 4.04, 8.52$) both the $L_2$ and $L_\infty$ errors of the BGK collision schemes were reduced in general, sometimes by as much as a factor of 4.
-None of the relative $L_2$ errors exceeded 10\%, in contrast to the BGK simulations run on the coarser lattice.
-However, despite the increase in accuracy, it should be noted that for $Bn = 8.52$ all of the BGK collision schemes had $L_\infty$ errors that exceeded 10\%.
-The MRT collision operator, even on the coarser lattice, was more accurate (for the case of $Bn = 8.52$) than all of the BGK collision schemes on the finer lattice.
 
@@
 \begin{equation} \label{eq:analytical-power-law}
-  u_x(y) = \frac{n}{n+1}\left(-\frac{1}{2k}\pgrad\right)^{1/n}\left[h^{\frac{n+1}{n}} - |y|^{\frac{n+1}{n}}\right].
+  u_x(y) = \frac{n}{n+1}\left(-\frac{1}{k}\pgrad\right)^{1/n}\left[h^{\frac{n+1}{n}} - |y|^{\frac{n+1}{n}}\right].
 \end{equation}
@@
 Note that none of the collision schemes were sufficiently accurate (error greater than 15\%) for the case of $n = 0.5$, and to improve this accuracy it would be necessary to use a different simulation Mach number, a finer grid, and/or a smaller pressure gradient.
+
+% MG: it might be better to normalize the time with respect to the fastest, BGK with no filtering
 
@@
 \begin{equation} \label{eq:gradu-power-law}
-  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{2k} \pgrad \right)^{1/n} \left( -y |y|^{\frac{1}{n} - 1} \right).
+  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{k} \pgrad \right)^{1/n} \left( -y |y|^{\frac{1}{n} - 1} \right).
 \end{equation}
@@
 A coarse lattice was chosen in order to highlight concerns with stability and accuracy.
-The simulations were run for either 50000 time steps or until convergence was met.
+The simulations were run for either 50000 timesteps or until convergence was met.
 Convergence was defined by:
@@
 \end{equation}
-\noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current time step.
+\noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current timestep.
 The lid velocity was prescribed as $u_o = 0.1$.
@@
 For the shear-thinning results, $n = 0.5$, at the lowest Reynold's number considered, $Re = 100$, all of the collision schemes produced results that agreed well with the literature value.
-However, the solutions became unstable at higher Reynold's numbers and entropic filtering was necessary to produce a stable solution when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
+However, the solutions became unstable at higher Reynold's numbers and entropic filtering was necessary to produce a stable soltuion when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
 Although there still appears to be a tradeoff between computational efficiency and accuracy in regards to the BGK with median filtering and MRT with median filtering, it does not seem to be nearly as significant as when simulating Bingham plastic lid-driven flow (e.g.
@@
 \cite{conrad2015accuracy} predicted optimal simulation Mach numbers for various shear-thinning and shear-thickening fluids. 
-Possible future work would include extending the numerical investigation to consider the effect of the simulation Mach number on the accuracy, stability, and efficiency of the various collision operators and stability enhancements in simulating non-Newtonian flow.
+Possbile future work would include extending the numerical investigation to consider the effect of the simulation Mach number on the accuracy, stability, and efficiency of the various collision operators and stability enhancements in simulating non-Newtonian flow.
+
+% MG: should we mention: code is open source? input files used in this study are also freely available?

### 2017-03-03 response-to-reviewers.tex

@@
 \begin{document}
-	\section{Reviewer 1}
+
+\section{Response to reviewers}
+
+Thank you very much for your assistance in reviewing this paper.
+We are sure that with your help, this has become a much better work.
+We hope that the attached edited copy of "Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow" will answer any questions previously raised, and will be satisfactory for publication.
+
+The following is a list of all comments recieved, and how they have been addressed in the updated version:
+
+  \subsection{Reviewer 1}
 	\begin{enumerate}
@@
 
-	\section{Reviewer 2}
+	\subsection{Reviewer 2}
 	  The present work examines the single-relaxation-time (SRT) and the multiple-relaxation-time (MRT) lattice Boltzmann method (LBM) for Bingham plastic and power-law fluids.

### 2017-03-03 cover-letter-to-editor.tex

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}

\begin{document}

Fernando Pinho
Universidade do Porto, Porto, Portugal

Dear Prof. Pinho:

Thank you very much for your assistance with our article titled "Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow" by Matthew Grasinger, Scott Overacker, and John C. Brigham for publication in the Journal of Non-Newtonian Fluid Mechanics. We are submitting to you the revised manuscript in which we have addressed the suggestions by the reviewers. In addition to the revised manuscript, we have included a document of our responses to the reviewers. Please let us know if any further modification or information is required.

All correspond should be addressed to:

John C. Brigham
blah blah blah

\end{document}

### 2017-03-03 src/stability-paper.tex

--- 2017-03-03 src/stability-paper.tex
@@
 Possible future work would include extending the numerical investigation to consider the effect of the simulation Mach number on the accuracy, stability, and efficiency of the various collision operators and stability enhancements in simulating non-Newtonian flow.
+
+% MG: should we mention: code is open source? input files used in this study are also freely available?

### 2017-03-03 response-to-reviewers.tex
Updated Bingham Poiseuille section with MRT-1 and MRT-2
corrected in address in the cover letter
updated figure legends, MRT -> MRT-1
addressed MRT-1, MRT-2 comment in response

--- 2017-03-03 response-to-reviewers.tex
@@
 \begin{document}
+
+	\pagenumbering{gobble}
 
@@
 We are sure that with your help, this has become a much better work.
-We hope that the attached edited copy of "Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow" will answer any questions previously raised, and will be satisfactory for publication.
+We hope that the attached edited copy of ``Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow'' will answer any questions previously raised, and will be satisfactory for publication.
 
@@
 
-  \subsection{Reviewer 1}
+  \subsection*{Reviewer 1}
 	\begin{enumerate}
@@
 		\begin{itemize}
-			\item Table 2 at the end of Section 3.1.1 and the last paragraph of Section 3.1.1 address the significance of grid resolution for Bingham plastic Poiseuille flow.
-			\item The third to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
+			\item Table 2 at the end of Section 3.1 and the last paragraph of Section 3.1 address the significance of grid resolution for Bingham plastic Poiseuille flow.
+			\item The second to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
 		\end{itemize}
@@
 
-	\subsection{Reviewer 2}
-	  The present work examines the single-relaxation-time (SRT) and the multiple-relaxation-time (MRT) lattice Boltzmann method (LBM) for Bingham plastic and power-law fluids.
-Constitutive relations in these fluids are typically stiff presenting substantial numerical challenges when the relations are incorporated into the LBM collision models.
-Entropic filtering as additional artificial dissipation and MRT collision are considered as the means to stabilize otherwise oscillatory or unstable SRT LBM.
-Optimal value for the median filtering was suggested and computational costs with these approaches are compared in detail.
-The conclusions are rather trivial; MRT is the most accurate and stable whereas its computational cost can be significantly large compared with SRT due to non-linear iteration.
-Filtering does stabilize SRT but the accuracy deteriorates particularly when the filter threshold is not optimally chosen.
-This work can be useful if readers are interested in choosing a suitable LB model for simulating non-Newtonian flows.
-However, it lacks novelty.
-Although MRT collision operator is shown to be the best choice, no detailed analysis on the values of optimum relaxation rates for various non-Newtonian flows.
-
-	As for Poiseulle Flow part, SRT simulations deviate from the analytical solution in the core region (Figs.3,4) due to the difference in the constitutive relations used in LBM and the analytical solution.
-Thus, the reported errors in Table 1 include both constitutive relation error and truncation error.
-This does not seem fair.
-It would be helpful if the analytical solution or a benchmark solution from NS equations of Papanastasiou approximation is directly compared with LBM results.
-The authors attributed the nonphysical oscillations of SRT simulations to sharp gradients in the velocity near the walls, which is then again a result of a sharp gradient in the viscosity.
-However, from Figs.3\&4, it is difficult to observe any sign of sharp gradients in the velocity near the walls.
-In Fig.4 for higher m, the oscillations occur at the center of the computational domain and they are not symmetric.
-
-	Grid convergence and its effect on the computation time would strengthen the manuscript.
-It is possible that MRT might become faster as the grid resolution improves.
-A minor comment is that as the streamwise BC is periodic in Sec.3.1, one might not need L=32 grid.
-One or two grid points would be enough.
+	\subsection*{Reviewer 2}
 
 	\begin{itemize}
-	\item \emph{As for Poiseulle Flow part, SRT simulations deviate from the analytical solution in the core region (Figs.3,4) due to the difference in the constitutive relations used in LBM and the analytical solution.
+
+	\item \emph{Although MRT collision operator is shown to be the best choice, no detailed analysis on the values of optimum relaxation rates for various non-Newtonian flows.}
+
+	Thank you for the suggestion.
+Another collision scheme was added to the Bingham plastic Poiseuille flow study so that the choices for the free relaxation parameters of $s_1 = s_2 = s_4 = s_6 = 1.1$ and $s_1 = 1.1, s_2 = 1.0, s_4 = s_6 = 1.2$ could be compared.
+The error and computational cost for each choice of relaxation parameters are shown in Table 1 and Table 2, and the two collision schemes are compared in the discussions that follow each table.
+
+	\item \emph{As for Poiseuille Flow part, SRT simulations deviate from the analytical solution in the core region (Figs.3,4) due to the difference in the constitutive relations used in LBM and the analytical solution.
 Thus, the reported errors in Table 1 include both constitutive relation error and truncation error.
@@
 They tend to move back and forth across the channel in subsequent time steps such that, on average (in time), the oscillations are symmetric.
+A sentence has been added to the end of the first paragraph of page 19 in order to clarify the apparent asymmetry.
 
@@
 
+\vspace{1in}
+\noindent Sincerely, \\
+\indent John C. Brigham \\
+\indent Associate Professor \\
+\indent School of Engineering and Computing Sciences
+
 \end{document}

### 2017-03-03 cover-letter-to-editor.tex
Updated Bingham Poiseuille section with MRT-1 and MRT-2
corrected in address in the cover letter
updated figure legends, MRT -> MRT-1
addressed MRT-1, MRT-2 comment in response

@@
-\documentclass[a4paper,10pt]{article}
-\usepackage[utf8]{inputenc}
-
+\documentclass{letter}
+\signature{John C. Brigham \\
+	\vspace{0.1in}
+	Associate Professor \\
+	School of Engineering \\ \hspace{0.1in} and Computing Sciences}
+\address{South Road\\
+	Durham\\
+	DH1 3LE \\ United Kingdom}
 \begin{document}
-
-Fernando Pinho
-Universidade do Porto, Porto, Portugal
-
-Dear Prof. Pinho:
-
-Thank you very much for your assistance with our article titled "Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow" by Matthew Grasinger, Scott Overacker, and John C.
+	\begin{letter}{Fernando Pinho \\ Universidade do Porto, Porto, Portugal}
+		\opening{Dear Prof. Pinho,}
+		Thank you very much for your assistance with our article titled ``Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow'' by Matthew Grasinger, Scott Overacker, and John C.
 Brigham for publication in the Journal of Non-Newtonian Fluid Mechanics.
@@
 
-John C. Brigham
-blah blah blah
+		John C. Brigham\\
+		South Road\\
+		Durham\\
+		DH1 3LE\\
+		john.brigham@durham.ac.uk
 
+		\vspace{1in}
+		\closing{Sincerely,}
+	\end{letter}
 \end{document}

### 2017-03-03 src/stability-paper.tex
Updated Bingham Poiseuille section with MRT-1 and MRT-2
corrected in address in the cover letter
updated figure legends, MRT -> MRT-1
addressed MRT-1, MRT-2 comment in response

--- 2017-03-03 src/stability-paper.tex
@@
 
-\subsection{Bingham Plastic, Poiseuille Flow}
+\subsection{Bingham Plastic, Poiseuille Flow} \label{sec:poise-bing}
 
@@
 The Bingham plastic simulations were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a plastic viscosity of $\mu_p = 0.2$.
-The yield stress was varied between four different values $\tau_y = \begin{bmatrix}4.0,&8.0,&12.0,&16.0\end{bmatrix} \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, and (4) MRT with $m = 10^8$.
-Recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer to the true Bingham plastic behavior.
-For the MRT relaxation matrix the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$, which follows the recommendation of~\cite{lallemand2000theory} for reasons of stability, and has been successfully used to simulate non-Newtonian flow in the past~\cite{fallah2012multiple,grasinger2015simulation}.
+The yield stress was varied between four different values $\tau_y = \begin{bmatrix}4.0,&8.0,&12.0,&16.0\end{bmatrix} \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, (4) MRT-1 with $m = 10^8$, and (5) MRT-2 with $m = 10^8$.
+The MRT-1 and MRT-2 schemes differ in their choice of free parameters for the relaxation matrix.
+For the MRT-1 relaxation matrix, the free parameters were set to $s_1 = s_2 = s_4 = s_6 = 1.1$, which follows the recommendation of~\cite{lallemand2000theory} for reasons of stability, and has been successfully used to simulate non-Newtonian flow in the past~\cite{fallah2012multiple,grasinger2015simulation}.
+For the MRT-2 relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$, as these values have been successfully applied to simulating lid-driven cavity flow of non-Newtonian fluids in the past~\cite{chen2014simulations,li2014simulation}.
+Additionally, in regards to the LBM schemes, recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer to the true Bingham plastic behavior.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
@@
 \\
-MRT & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
+MRT-1 & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
 & & & 8.0 & 4.42 & 1.85 & 0.0018 & 0.0018 & 7908 \\
@@
 & & & 16.0 & 1.92 & 8.52 & 0.0041 & 0.0041 & 5559 \\
+\\
+MRT-2 & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0012 & 0.0012 & 5284 \\
+& & & 8.0 & 4.42 & 1.85 & 0.0017 & 0.0017 & 5290 \\
+& & & 12.0 & 3.04 & 4.04 & 0.0024 & 0.0024 & 5244 \\
+& & & 16.0 & 1.92 & 8.52 & 0.0038 & 0.0038 & 4532 \\
 \\
@@
 In contrast, the LBM approximation in \Fref{fig:bing-analyt-bgk-8} is not smooth, which suggests that the error for the BGK model with $m = 10^8$ is due to nonphysical oscillations.
+(Note: the velocity for the BGK model with $m = 10^8$ may appear asymmetric, but that is because the flow does not reach a steady-state and \Fref{fig:bing-analyt-bgk-8} is merely a snapshot in time.
+The oscillations in the velocity tend to move back and forth across the channel in subsequent time steps such that, on average (in time), the oscillations are indeed symmetric.)
 
@@
 
-On average, the LBM model using the MRT collision operator took 2.9 times more computing time than the BGK collision operator with $m = 10^5$.
-However, the increased computing time can be justified for the MRT collision operator because the MRT solution did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
+On average, the LBM model using the MRT-1 collision operator took 2.9 times more computing time than the BGK collision operator with $m = 10^5$; and the model using MRT-2 took 2.1 times more computing time than the BGK collision operator with $m = 10^5$.
+However, the increased computing time can be justified for the MRT collision operator because the MRT solutions did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
 when $m = 10^8$.
-The MRT collision operator was therefore the most accurate solution for all cases.
-\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilibrium profiles, $f_i^{eq}$ for the MRT model.
+The MRT collision schemes were therefore the most accurate solutions for all cases; and because the difference in accuracy between the MRT-1 and MRT-2 collision schemes was negligible, the remainder of the discussion in this section will focus on comparing the BGK collision schemes and the MRT-1 collision scheme.
+\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilibrium profiles, $f_i^{eq}$ for the MRT-1 model.
 In contrast to the BGK with $m = 10^8$, there were no significant nonphysical oscillations in the quasiequilibrium distributions.
@@
     \end{tabulary}
-    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT with $m = 10^8$.}
+    \caption{Particle distributions in the 5 direction (left) and 8 direction (right) compared to their respective quasiequilibriums for the MRT-1 with $m = 10^8$.}
     \label{fig:feq-vs-f_mrt}
@@
 $\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
-The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ displayed was in part due to the $\epsilon$ moment.
+The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT-1 or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ displayed was in part due to the $\epsilon$ moment.
 
@@
 The relative $L_2$ norm of the $\mathbf{q_x}^{neq}$ for the BGK collision operator with $m = 10^8$ and median filtering solution spent most of the simulation between approximately 50\% and 100\%, suggesting that the $q_x$ moment was likely the cause of the nonphysical oscillations and error in the solution.
-As expected (if the dominant source of error in the LBM approximation for Bingham plastic Poiseuille flow is indeed due to the $q_x$ moment), the $\mathbf{q_x^{neq}}$ relative $L_2$ norm for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $\mathbf{q_x^{neq}}$ relative $L_2$ norm was negligible for the MRT collision scheme.
+As expected (if the dominant source of error in the LBM approximation for Bingham plastic Poiseuille flow is indeed due to the $q_x$ moment), the $\mathbf{q_x^{neq}}$ relative $L_2$ norm for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $\mathbf{q_x^{neq}}$ relative $L_2$ norm was negligible for the MRT-1 collision scheme.
 
@@
 As would be expected from doubling the number of lattice sites, the cost for each corresponding combination of collision scheme, $Re$, and $Bn$ increased by a factor of approximately 2--2.5.
-The 2--2.5 times increase in computing time resulted in the BGK collision schemes on the finer lattice being comparable in computational cost to the MRT collision scheme on the coarser lattice.
+The 2--2.5 times increase in computing time resulted in the BGK collision schemes on the finer lattice being comparable in computational cost to the MRT collision schemes on the coarser lattice.
 In regards to error, for the larger Bingham number flows ($Bn = 4.04, 8.52$) both the $L_2$ and $L_\infty$ errors of the BGK collision schemes were reduced in general, sometimes by as much as a factor of 4.
@@
 However, despite the increase in accuracy, it should be noted that for $Bn = 8.52$ all of the BGK collision schemes had $L_\infty$ errors that exceeded 10\%.
-The MRT collision operator, even on the coarser lattice, was more accurate (for the case of $Bn = 8.52$) than all of the BGK collision schemes on the finer lattice.
+The MRT collision schemes, even on the coarser lattice, were more accurate (for the case of $Bn = 8.52$) than all of the BGK collision schemes on the finer lattice.
+Again, the difference in the accuracies between the MRT-1 and MRT-2 collision schemes appears to be negligible.
+The difference in computational cost is also low, though the MRT-2 collision scheme took 4\% less computing time on average.
 
@@
 		\\
-		MRT & No & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0452 & 0.0484 & 13837 \\
+		MRT-1 & No & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0452 & 0.0484 & 13837 \\
 		              & & & 2.0 & 4.42 & 1.85 & 0.0404 & 0.0435 & 16515 \\
                       & & & 3.0 & 3.04 & 4.04 & 0.0316 & 0.0338 & 17142 \\
-                      & & & 4.0 & 1.92 & 8.52 & 0.0200 & 0.0210 & 17853 \\
+                      & & & 4.0 & 1.92 & 8.52 & 0.0197 & 0.0210 & 17853 \\
+		\\
+		MRT-2 & No & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0453 & 0.0485 & 14030 \\
+		& & & 2.0 & 4.42 & 1.85 & 0.0404 & 0.0435 & 15290 \\
+		& & & 3.0 & 3.04 & 4.04 & 0.0316 & 0.0338 & 16328 \\
+		& & & 4.0 & 1.92 & 8.52 & 0.0198 & 0.0210 & 16816 \\
 		\\
@@
 The lid velocity was prescribed as $u_o = 0.1$.
-For the MRT relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$, as these values have been successfully applied to simulating lid-driven cavity flow of non-Newtonian fluids in the past~\cite{chen2014simulations,li2014simulation}.
 All coordinate values presented in this section (used to describe the location of the center of vortices) are given normalized with respect to the length of the cavity side (i.e. as $(x / L, y / L)$).
@@
 Six collision schemes were tested: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
+For the MRT relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$ (same as MRT-2 in \Fref{sec:poise-bing}).
 Tables \ref{tab:lid-bing1}--\ref{tab:lid-bing100} compare the center location of the main vortex to literature values taken from~\citet{syrakos2014performance}.
@@
     \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable,
-    \item and the MRT collision operator with $m = 10^8$ produces solutions with high accuracy and stability for all of the cases examined herein, but at an increased (approximately 5-10 times more) computing cost.
+    \item and the MRT collision operator with $m = 10^8$ produces solutions with high accuracy and stability for all of the cases examined herein, but at an increased (approximately 5--10 times more) computing cost.
 \end{itemize}
@@
 
-For the power-law numerical tests the Reynold's number was varied with the following values: $\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the flow behavior index, $n$, was varied with the following values: $\begin{bmatrix}0.5,&1.5\end{bmatrix}$ ($Re$ was calculated the same as in \Fref{sec:poise-powerlaw}).
+For the power-law numerical tests the Reynold's number was varied with the following values: $\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the flow behavior index, $n$, was varied with the following values: $\begin{bmatrix}0.5,&1.5\end{bmatrix}$.
+The Reynold's number was calculated as $\rho \frac{U^{2-n} H^n}{k}$.
 Five collision schemes were tested: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT and (5) MRT with median filtering.

### 2017-03-05 response-to-reviewers.tex
resubmission draft

@@
 			\item Table 2 at the end of Section 3.1 and the last paragraph of Section 3.1 address the significance of grid resolution for Bingham plastic Poiseuille flow.
-			\item The second to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
+			\item The end of the third to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
 		\end{itemize}
@@
 		\item Table 2 at the end of Section 3.1.1 and the last paragraph of Section 3.1.1 address the significance of grid resolution for Bingham plastic Poiseuille flow.
-		\item The second to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
+		\item The end of the third to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
 	\end{itemize}
@@
 \noindent Sincerely, \\
-\indent John C. Brigham \\
+\indent John C. Brigham \vspace{0.1in}\\
 \indent Associate Professor \\

### 2017-12-20 manuscript-srcs/src/stability-paper.tex
version first submitted to computers and fluids, with mark-up from last round of revisions with nonnewtonian fluid mechanics

--- 2017-03-03 src/stability-paper.tex
@@
 constitutive laws, interfacial dynamics, etc.) and it has implications on numerical stability and computational efficiency.
-This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
+This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of their accuracy, numerical stability and computational efficiency.
 The investigation was carried out as a numerical study looking for qualitative, yet practical, results, including testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
@@
 The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behaviors, but at an increased computing cost that was, in some cases, as much as five times greater than the BGK collision operator.
-It was also shown that artificial dissipation can be an effective technique for enhancing numerical stability of non-Newtonian, lid-driven cavity flow, and is particularly effective for efficiently and accurately simulating the flow of shear-thinning fluids.
+It was also shown that artificial dissipation can be an effective technique for enhancing the numerical stability of non-Newtonian, lid-driven cavity flow simulations, and is particularly effective for efficiently and accurately simulating the flow of shear-thinning fluids.
 
@@
 Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
-Yield stress flow has been found to be relevant in many applications, such as flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
-
-Power-law behavior, is more commonly known as \emph{shear-thinning}, when the apparent viscosity decreases with increasing strain-rate, or \emph{shear-thickening}, when the apparent viscosity increases with increasing strain-rate.
+Yield stress flow has been found to be relevant in many applications, such as the flow of pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}.
+
+Power-law behavior is also known as \emph{shear-thinning} when the apparent viscosity decreases with increasing strain-rate, or \emph{shear-thickening} when the apparent viscosity increases with increasing strain-rate.
 Shear-thinning fluids are also known as \emph{pseudoplastics}, and some examples include tomato concentrate, clay and water mixtures, and molten plastics~\cite{irgens2014rheology}.
@@
 
-LBM does however, have its drawbacks.
+LBM does, however, have its drawbacks.
 LBM can be considered as a type of finite-difference scheme for the continuous Boltzmann equation, and as such, has numerical properties in common with finite-difference schemes.
@@
 This means particle distribution functions will always relax toward equilibrium at the same rate, but that the definition of equilibrium is modified to represent the correct stress--strain-rate relationship.
-The equilibrium distribution function can be derived for the specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.
+The equilibrium distribution function can be derived for a specific constitutive relationship of interest using the Chapman-Enskog multiscale expansion.
 The equilibrium distribution functions for Bingham plastic fluids, and for power-law and Carreau fluids were derived, implemented, and verified in~\cite{wang2008lattice} and~\cite{yoshino2007numerical}, respectively.
@@
 $\approx 0$).
-\cite{wang2011lattice} developed another constant-collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
+\cite{wang2011lattice} developed another constant collision frequency LBM scheme for non-Newtonian flow by splitting the effects of constitutive relationship into Newtonian and non-Newtonian parts.
 The Newtonian part was modeled in the usual way, namely scaling the collision frequency to achieve the macroscopic (Newtonian) viscosity, and the non-Newtonian part was modeled as a source of momentum (i.e.
 as an external forcing term) that is dependent on local shear-rate.
-Although the constant-collision frequency strategies present interesting alternatives, the variable collision frequency scheme is used in the present study because of its simplicity and generality, such as the fact that a variable collision frequency scheme can be fit to any constitutive relationship without performing Chapman-Enskog multiscale expansion.
+Although the constant collision frequency strategies present interesting alternatives, the variable collision frequency scheme is used in the present study because of its simplicity and generality, such as the fact that a variable collision frequency scheme can be fit to any constitutive relationship without performing Chapman-Enskog multiscale expansion.
 
 %The Bhatnagar, Gross, Krook (BGK) collision operator~\cite{bhatnagar1954model}.
-The Multiple-relaxation-time (MRT) collision operator ~\cite{d1994generalized} is another approach that has been developed to improve upon the stability of LBM.
+The Multiple-relaxation-time (MRT) collision operator ~\cite{d1994generalized} is another approach that has been developed to improve the stability of LBM.
 The MRT collision takes place in moment space and allows each moment to relax at a different rate.
@@
 More specifically, LBM can be thought of as a special finite difference discretization of the Boltzmann equation~\cite{chen1998lattice}.
-The length scale of LBM is unique in contrast to most common numerical methods, and is referred to as the mesoscale.
+The length scale of LBM is unique in contrast to most common numerical methods and is referred to as the mesoscale.
 In contrast to continuum based methods, LBM simulates the kinetics of microscopic particles, and so it reaches a finer length scale than the macroscopic domain of continuum mechanics; and in contrast to molecular dynamics, discrete element method, and other particle scale approaches, LBM does not deal with a complete description of the degrees of freedom for each individual particle.
@@
 Despite the utility of the BGK collision operator, it does have a few drawbacks.
-For example, in low viscosity fluids the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
+For example, in low viscosity fluids, the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
 It is well known that when large nonequilibrium distributions exist in the LBM approximation that overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
@@
 To reduce nonequilibrium fluctuations in LBM, one can introduce artificial dissipation.
-The idea of artificial dissipation is to increase numerical stability, while sacrificing some physical accuracy.
+The idea of artificial dissipation is to increase numerical stability while sacrificing some physical accuracy.
 A model that has more physical justification but produces unstable and nonsensical results is much less useful than a model with some minor artificial features yet produces more stable results.
@@
 However, there are many ways one can measure the distance between $\mathbf{f}$ and $\mathbf{f}^{eq}$; for example, a reasonable choice would be $||\mathbf{f} - \mathbf{f}^{eq}||_p$ for some $p$ norm.
-A metric that has been developed and used successfully for determining when artificial dissipation should be introduced at a lattice site is the so called relative nonequilibrium entropy~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy}.
+A metric that has been developed and used successfully for determining when artificial dissipation should be introduced at a lattice site is the so-called relative nonequilibrium entropy~\cite{gorban2014enhancement,brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy}.
 The relative nonequilibrium entropy, $\Delta S$, is given by:
@@
 
-\noindent A more computationally efficient approximation of $\Delta S$ can be achieved by instead using the second order Taylor expansion of \eqref{eq:rel-noneq-entropy}:
+\noindent A more computationally efficient approximation of $\Delta S$ can be achieved by instead using the second-order Taylor expansion of \eqref{eq:rel-noneq-entropy}:
 
@@
 Just as there are many ways to measure a lattice site's distance from quasiequilibrium and define criteria for artificial dissipation, there are also many different ways of deciding how much dissipation to add.
-One method of adding dissipation is the so-called Ehrenfests' regularization~\cite{brownlee2006stabilization}, and involves setting a lattice site that is chosen for artificial dissipation to its quasiequilibrium state.
-Although this achieves the desired result, namely damping out large nonequilibrium fluctuations, it does so in a way that is not smooth or gradual, but sharp.
+One method of adding dissipation is the so-called Ehrenfests' regularization~\cite{brownlee2006stabilization} and involves setting a lattice site that is chosen for artificial dissipation to its quasiequilibrium state.
+Although this achieves the desired result, namely dampening out large nonequilibrium fluctuations, it does so in a way that is not smooth or gradual, but sharp.
 An alternative, the median filter, has been used successfully in conjunction with both the BGK and MRT collision schemes for simulating one-dimensional shock tubes and lid-driven cavity flow~\cite{brownlee2008nonequilibrium,gorban2014enhancement,packwood2009entropy}.
-In LBM this is a desirable way to introduce dissipation because it has the potential to reduce high frequency nonequilibrium fluctuations that might lead to numerical instability while retaining the lower frequency dynamics.
+In LBM this is a desirable way to introduce dissipation because it has the potential to reduce high-frequency nonequilibrium fluctuations that might lead to numerical instability while retaining the lower frequency dynamics.
 To use the median filter one performs the collision step and then checks over the domain for lattice sites with $\Delta S$ that meet the criteria for artificial dissipation.
@@
 A Bingham plastic does not flow (i.e.
-the strain-rate is zero) when the shear stress is below the yield stress, and behaves in an almost Newtonian manner when the shear stress is above the yield stress.
+the strain-rate is zero) when the shear stress is below the yield stress and behaves in an almost Newtonian manner when the shear stress is above the yield stress.
 The Bingham plastic relationship is described mathematically as:
@@
 
-Poiseuille flow is a useful benchmark because an analytical solution exists for both Bingham plastic flow.
+Poiseuille flow is a useful benchmark because an analytical solution exists for Bingham plastic flow.
 Poiseuille flow is created with a two-dimensional channel flow that is driven by a constant pressure gradient, $\pgrad$, through the domain.
@@
 Unless otherwise stated, all of the Poiseuille flow simulations were computed on a $32 \times 64$ lattice for 25000 time steps.
-The lattice size was chosen to be sufficiently fine for accuracy and the number of time steps were set high enough to ensure a steady-state would be reached.
-In reference to computational time, each of the simulations in this section were run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
+The lattice size was chosen to be sufficiently fine for accuracy and the number of time steps was set high enough to ensure a steady state would be reached.
+In reference to computational time, each of the simulations in this section was run on a single core of an Intel I7-860 Quad-Core 2.80GHz processor.
 
 The Bingham plastic simulations were carried out with a pressure gradient of $\pgrad = 1.0 \times 10^{-5}$, and a plastic viscosity of $\mu_p = 0.2$.
-The yield stress was varied between four different values $\tau_y = \begin{bmatrix}4.0,&8.0,&12.0,&16.0\end{bmatrix} \times 10^{-5}$, and four different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, (4) MRT-1 with $m = 10^8$, and (5) MRT-2 with $m = 10^8$.
+The yield stress was varied between four different values $\tau_y = \begin{bmatrix}4.0,&8.0,&12.0,&16.0\end{bmatrix} \times 10^{-5}$, and five different LBM schemes were used: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK with $m = 10^8$ and the median filter, (4) MRT-1 with $m = 10^8$, and (5) MRT-2 with $m = 10^8$.
 The MRT-1 and MRT-2 schemes differ in their choice of free parameters for the relaxation matrix.
@@
 For the MRT-2 relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$, as these values have been successfully applied to simulating lid-driven cavity flow of non-Newtonian fluids in the past~\cite{chen2014simulations,li2014simulation}.
-Additionally, in regards to the LBM schemes, recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer to the true Bingham plastic behavior.
+Additionally, in regards to the LBM schemes, recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer the approximation is to the true Bingham plastic behavior.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
@@
 Because the nonphysical oscillations make their way into the quasiequilibrium values, entropic median filtering does not help dampen the oscillations but instead contracts the particle distributions closer to the quasiequilibrium oscillations, which explains why the median filtered results were less accurate for Bingham plastic Poiseuille flow.
-In order to ensure that this phenomena was a side effect of all entropic filtering, and not just entropic median filtering with $\theta = 1.0 \times 10^{-6}$ and $n_s = 2.7$, optimization was used to find the value of $\theta$ that minimized error for both median filtering and Ehrenfests' regularization.
+In order to ensure that this phenomenon was a side effect of all entropic filtering, and not just entropic median filtering with $\theta = 1.0 \times 10^{-6}$ and $n_s = 2.7$, optimization was used to find the value of $\theta$ that minimized error for both median filtering and Ehrenfests' regularization.
 The optimization problem was defined as follows:
@@
 For all four cases, the optimal solution was $\theta = 2.0$ and the cost function, $f(2.0)$, was equal to the relative $L_2$ error for the BGK collision operator, $m = 10^8$ and no entropic filtering.
-The results of the optimization show that entropic filtering, at best did not affect the accuracy of LBM, and often adversely affected the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization was used, and regardless of what the $\Delta S$ threshold, $\theta$, was set to.
+The results of the optimization show that entropic filtering, at best, did not affect the accuracy of LBM; and often adversely affected the accuracy of LBM in approximating Bingham plastic Poiseuille flow, regardless of whether median filtering or Ehrenfests' regularization was used, and regardless of what the $\Delta S$ threshold, $\theta$, was set to.
 
@@
 when $m = 10^8$.
-The MRT collision schemes were therefore the most accurate solutions for all cases; and because the difference in accuracy between the MRT-1 and MRT-2 collision schemes was negligible, the remainder of the discussion in this section will focus on comparing the BGK collision schemes and the MRT-1 collision scheme.
+The MRT collision schemes were, therefore, the most accurate solutions for all cases; and because the difference in accuracy between the MRT-1 and MRT-2 collision schemes was negligible, the remainder of the discussion in this section will focus on comparing the BGK collision schemes and the MRT-1 collision scheme.
 \Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilibrium profiles, $f_i^{eq}$ for the MRT-1 model.
@@
 It is likely that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time (i.e.
-osciallations build in amplitude with time).
+oscillations build in amplitude with time).
 \Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time, which was calculated, respectively, as:
@@
 \noindent where $T$ is the current time. 
-\Fref{fig:qx} and \Fref{fig:qx-cumulative} shows that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the ghost mode associated with the $q_x$ moment dominates and is the primary source of oscillations.
+\Fref{fig:qx} and \Fref{fig:qx-cumulative} show that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the ghost mode associated with the $q_x$ moment dominates and is the primary source of oscillations.
 It can be inferred that the $q_x$ moment was a significant source of oscillations in this case because the Poiseuille flow was in the x-direction.
@@
     \caption{Cumulative relative $L_2$ norm, $\epsilon^{neq}$ with time.
-Oscillations can have a positive effect on each other, building up.
+Oscillations can have a positive effect on each other.
 The cumulative relative $L_2$, $\epsilon^{neq}$ is a measure of how much oscillations due to the $\epsilon$ moment may have been building in time.}
@@
     \caption{Cumulative relative $L_2$ norm, $\mathbf{q_x^{neq}}$ with time.
-Oscillations can have a positive effect on each other, building up.
+Oscillations can have a positive effect on each other.
 The cumulative relative $L_2$, $\mathbf{q_x^{neq}}$ is a measure of how much oscillations due to the $q_x$ moment may have been building in time.}
@@
 
-Lastly, simulations with the same collision schemes, Reynold's numbers and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
+Lastly, simulations with the same collision schemes, Reynold's numbers, and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
 The plastic viscosity was held constant ($\mu_p = 0.2$.), while the yield stress and pressure gradient were varied in order to match the Reynold's numbers and Bingham numbers in \Fref{tab:poise-bing}.
@@
 Two different benchmark problems were chosen for the study: Poiseuille flow, and lid-driven square cavity flow.
-The results showed that for higher yield stress fluids in Poiseuille-type flow, only the MRT collision operator was able use an accurate approximation to true Bingham plastic behavior and not suffer from nonphysical oscillations.
+The results showed that for higher yield stress fluids in Poiseuille-type flow, only the MRT collision operator was able to use an accurate approximation to true Bingham plastic behavior and not suffer from nonphysical oscillations.
 The oscillations appeared to be due to high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
@@
 
-For nearly all of the cases considered in the current study, the MRT collision operator was more accurate, but also more computationally expensive than its BGK counterpart, and is some times even orders of magnitude slower.
-
-In regards to future work, \cite{conrad2015accuracy} showed that non-Newtonian Lattice Boltzmann simulations are not second order accurate for every simulation Mach number, and showed that a greater accuracy may be achieved by properly tuning the simulation Mach number.
+For nearly all of the cases considered in the current study, the MRT collision operator was more accurate, but also more computationally expensive than its BGK counterpart, and is sometimes even orders of magnitude slower.
+
+In regards to future work, \cite{conrad2015accuracy} showed that non-Newtonian Lattice Boltzmann simulations are not second-order accurate for every simulation Mach number, and showed that a greater accuracy may be achieved by properly tuning the simulation Mach number.
 \cite{conrad2015accuracy} predicted optimal simulation Mach numbers for various shear-thinning and shear-thickening fluids. 

### 2017-12-20 manuscript-srcs/figs/poise-bingham/process4.jl

@@ -1,0 +1,41 @@
+using DataFrames, Gadfly
+
+names = Dict("epsilon" => "ϵ", "qx" => "q<sub>x</sub>", "qy" => "q<sub>y</sub>");
+tests = Dict("bgk-5" => "BGK, m = 10<sup>5</sup>", 
+             "bgk-8" => "BGK, m = 10<sup>8</sup>",
+             "bgk-fltr" => "BGK, filtered",
+             "mrt" => "MRT, m = 10<sup>8</sup>");
+
+for (name, var) in names
+  df = DataFrame(x = Float64[], y = Float64[], Series = AbstractString[]);
+  cf = DataFrame(x = Float64[], y = Float64[], Series = AbstractString[]);
+  for (d, label)  in tests
+    #const x = (d in ["bgk-5", "mrt"] || (name == "qx" && d == "bgk-8")) ? collect(0:62) * 400.0 : collect(0:125) * 200.0;
+    const x = collect(0:250) * 100.0;
+    val    =   readdlm("$d/$name.dsv", ','); 
+    valeq  =   readdlm("$d/$(name)_eq.dsv", ',');
+
+    #=y = if d in ["bgk-5", "mrt"] || (name == "qx" && d == "bgk-8")
+          map(i -> norm(val[i, :] - valeq[i, :], 2) / norm(valeq[i, :], 2), 1:4:251);
+        else
+          map(i -> norm(val[i, :] - valeq[i, :], 2) / norm(valeq[i, :], 2), 1:2:251);
+        end=#
+    y = map(i -> norm(val[i, :] - valeq[i, :], 2), 1:251);
+    y2 = zeros(251); y2[1] = y[1] * 100;
+    for i=2:251; y2[i] = y[i] * 100 + y2[i-1];  end
+    df = vcat(df, DataFrame(x=x, y=y, Series=fill(label, length(x))));
+    cf = vcat(cf, DataFrame(x=x, y=y2, Series=fill(label, length(x))));
+  end
+  draw(PNG("$name-abs.png", 5.5inch, 3inch),
+       plot(df, x=:x, y=:y, color=:Series,
+       Geom.line,
+       Guide.XLabel("Time (sec)"),
+       Guide.YLabel("L<sub>2</sub>, $var<sup>neq</sup>")
+       ));
+  draw(PNG("$(name)-abs_cumulative.png", 5.5inch, 3inch),
[…]

### 2018-01-03 response-to-reviewers.tex
addressed reviewers comments

--- 2017-03-03 response-to-reviewers.tex
@@
 
-\section{Response to reviewers}
+	\pagenumbering{gobble}
+
+\section*{Response to reviewers}
 
@@
 We are sure that with your help, this has become a much better work.
-We hope that the attached edited copy of "Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow" will answer any questions previously raised, and will be satisfactory for publication.
+We hope that the attached edited copy of ``Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow'' will answer any questions previously raised, and will be satisfactory for publication.
 
-The following is a list of all comments recieved, and how they have been addressed in the updated version:
+The following is a list of all comments received, and how they have been addressed in the updated version:
 
-  \subsection{Reviewer 1}
+  \subsection*{Reviewer 1}
 	\begin{enumerate}
-		\item \emph{In Fig.
-2, the scale of the figure should be modified in order to avoid readers’ misleading.
-If the authors calculate only one case with L = 32 and H = 64, the computational domain becomes vertically long.}
+		\item first
 
-		The figure has been updated to reflect the relative scale between $L$ and $H$.
-
-		\item \emph{In Table 2, all the results of the error norms, $L_{2}$ and $L_{\infty}$, for $n = 1.25$ are the same values in spite of different LBM schemes.
-The values should be changed to correct ones.}
-
-		The results are actually the same for all of the BGK variants.
-		This is because for $n = 1.25$ the nonequilibrium entropy threshold for median filtering was not exceeded, and hence it was effectively the same as the unfiltered BGK scheme.
-		Similarly, the relaxation time was within the bounds for the BGK-BRT scheme, which again resulted in the scheme to being effectively the same as the BGK collision scheme without bounds.
-		The error norms for the MRT collision scheme was different, but the difference could not be seen in the precision of the results presented in the previous draft.
-		The $n = 1.25$ error norms have been updated with higher precision so as to show the difference.
-
-		\item \emph{The authors carried out the simulations with fixed lattice resolution.
-However, the convergence rate should be examined by calculating the benchmark problem (e.g., Poiseuille flow) with different lattice resolutions.
-		What order is the accuracy of the schemes used in the present study?}
-
-		Considerations regarding grid resolution were added in the following locations:
-		\begin{itemize}
-			\item Table 2 at the end of Section 3.1.1 and the last paragraph of Section 3.1.1 address the significance of grid resolution for Bingham plastic Poiseuille flow.
-			\item Table 4 at the end of Section 3.1.2 and the last paragraph of Section 3.1.2 address the significance of grid resolution for Power-law Poiseuille flow. 
-			\item The third to last paragraph of Section 3.2.1 addresses the significance of grid resolution for the Bingham plastic lid-driven cavity flow.
-		\end{itemize}
 	\end{enumerate}
 
-	\subsection{Reviewer 2}
-	  The present work examines the single-relaxation-time (SRT) and the multiple-relaxation-time (MRT) lattice Boltzmann method (LBM) for Bingham plastic and power-law fluids.
-Constitutive relations in these fluids are typically stiff presenting substantial numerical challenges when the relations are incorporated into the LBM collision models.
-Entropic filtering as additional artificial dissipation and MRT collision are considered as the means to stabilize otherwise oscillatory or unstable SRT LBM.
-Optimal value for the median filtering was suggested and computational costs with these approaches are compared in detail.
-The conclusions are rather trivial; MRT is the most accurate and stable whereas its computational cost can be significantly large compared with SRT due to non-linear iteration.
-Filtering does stabilize SRT but the accuracy deteriorates particularly when the filter threshold is not optimally chosen.
-This work can be useful if readers are interested in choosing a suitable LB model for simulating non-Newtonian flows.
-However, it lacks novelty.
-Although MRT collision operator is shown to be the best choice, no detailed analysis on the values of optimum relaxation rates for various non-Newtonian flows.
-	As for Poiseulle Flow part, SRT simulations deviate from the analytical solution in the core region (Figs.3,4) due to the difference in the constitutive relations used in LBM and the analytical solution.
-Thus, the reported errors in Table 1 include both constitutive relation error and truncation error.
-This does not seem fair.
-It would be helpful if the analytical solution or a benchmark solution from NS equations of Papanastasiou approximation is directly compared with LBM results.
-The authors attributed the nonphysical oscillations of SRT simulations to sharp gradients in the velocity near the walls, which is then again a result of a sharp gradient in the viscosity.
-However, from Figs.3&4, it is difficult to observe any sign of sharp gradients in the velocity near the walls.
-In Fig.4 for higher m, the oscillations occur at the center of the computational domain and they are not symmetric.
+	\subsection*{Reviewer 2}
 
-	Grid convergence and its effect on the computation time would strengthen the manuscript.
-It is possible that MRT might become faster as the grid resolution improves.
-A minor comment is that as the streamwise BC is periodic in Sec.3.1, one might not need L=32 grid.
-One or two grid points would be enough.
+	\begin{itemize}
+
+	\item \emph{Page 1: The reviewer agrees that artificial dissipation can be an effective technique
+		to stabilize a simulation and thus make a certain simulation possible. But the
+		reviewer does not agree that it is “effective ... for accurately simulating the flow ...”.
+		Especially for non-Newtonian fluids an exact representation of the fluid viscosity is
+		essential. By adding numerical dissipation, i.e. viscosity, this is foiled. The Reynolds
+		number and thus the flow physics will change and accuracy is decreased. Please
+		revise this wording to strengthen that fact.}
+
+	Thank you for helping clarify this point. The last sentence of the abstract has been edited in order to more accurately reflect the effect of artificial dissipation on accuracy and stability.
+
+	\item \emph{Page 5: Please mention that Eq. (1) is without the effects of external forces}
+
+	The sentence prior to the Boltzmann Equation (Eq. (1)) has been edited to clarify this point.
+
+	\item \emph{Page 7: ``Due to the discrete nature of velocity in LBM, the integrals simply become
+		summations'' Please be more precise, it is an intrinsic part of the discretization
+		procedure to replace continuous integrals (moments) by numerical quadrature, i.e.
+		Gauss-Hermite quadrature in LBM.}
+
+	The aforementioned statement was removed and replaced by explicitly noting that the summations are a Gaussian-type quadrature.
+	This was followed by relating the weights of the quadrature to the equilibrium distribution function.
+	Additionally, the following connections to numerical quadrature are made directly following Eq.
+(7) by: (1) mentioning how the weights in Eq.
+(7) are determined, (2) the relationship between the weights in Eq.
+(7) and the quadrature weights are given and (3) the reader is referred to some additional reading if they are interested in further detail regarding the numerical quadrature.
+	Thank you for this comment.
+	The changes mentioned above should help improve the clarity of this paper.
+
+	\item \emph{Page 4/15: You mention that computing the strain rate tensor is second order
+		accurate, but the reviewer couldn’t find any comment on the accuracy of LBM to
+		Navier-Stokes. This could be mentioned in the context with Chapman-Enskog
+		expansion (p. 8)}
+
+	\item \emph{Page 10: ``A model that has more physical justification but produces unstable and
+		nonsensical results is much less useful than a model with some minor artificial
+		features yet produces more stable results.'' The reviewer does not agree with this
+		statement in general. A stable method, producing ``always'' a result which might have
+		significant errors is a dangerous tool...}
+
+	The authors agree with this comment and, as such, the above sentenced was removed.
+
+	\item \emph{Page 12: Boundary conditions: Please explain here or when describing the setup for
+		the numerical calculations if half way or full way bounce back was used}
+
+	In Section 2.4 (Boundary Conditions and Applied Forces), it is now explicitly mentioned that the complete bounceback scheme is used for the no-slip boundary condition.
+	Additionally, it is made clear that, in general, boundary conditions and the incorporation of external forces in LBM is not unique.
+
+	\item \emph{Page 15: Setup of Bingham Poiseuille flow: What is the Mach number in the lattice
+		for the simulations?}
+
+	A column has been added to each table to show the Mach number of each simulation.
+
+	\item \emph{Page 15: An exemplary figure (e.g. for $\tau_y = 4.0*10^-5$) comparing the viscosity
+		according to the exact Bingham relationship and the approximations according to
+		Papanastasiou with $m=10^5$ and $m=10^8$, respectively, would allow a comparison at
+		first glance. Also the reader could see differences directly.}
+
+	This figure was added.
+	Thank you for the suggestion.
+
+	\item \emph{Page 22: Why don’t you compare the MRT calculations to the BGK,
+		$m=10^8$, cases, since $m=10^8$ was used for MRT as well?}
+
+	\item \emph{Page 23, 24: You draw some conclusions between the ``ghost'' modes ($\epsilon$ and $q$) and
+		nonphysical oscillations. For example you state (page 23): ``that the nonphysical
+		oscillations the BGK with $m = 10^8$ displayed was in part due to the $\epsilon$ moment''.
+		To the reviewer there is no direct proof visible for these conclusions. It might also be
+		the other way around; oscillations in the distributions cause the oscillations in the
+		moments. Please explain this in more detail.}
+
+	\item \emph{Page 23, 24: Did you investigate if there are oscillations in the moment e, related to
+		the bulk viscosity? It is well known that a high bulk viscosity is stabilizing and thus
+		producing some of the superior MRT stability properties.}
+
+	\item \emph{Page 25: ``...ghost mode associated with the q x moment dominates and is the
+		primary source of oscillations.'' Again, this is not obvious for the reviewer, please
+		prove this conclusion.}
+
+	\item \emph{Page 25: ``Overall, the results presented suggest that if an LBM collision scheme is to
+		be developed ... , it should focus on a means of dampening the nonequilibrium
+		energy flux moments, namely q x and q y''. Did you consider the two relaxation times
+		model (TRT) of Ginzburg et al.?}
+
+	\item \emph{Page 25/27: Is the lattice number (128x32) correct? Since you reduced the number
+		of nodes along the channel height from 64 to 32? What was the Mach number for
+		these simulations? The error of all MRT results is drastically increased in Table 2
+		compared to Table 1 what fits to the reduced lattice number in along the channel
+		height. Can you explain this, also in regard to the behavior of the BGK error?}
+
+	The lattice number was incorrect (it should be 32x128).
+	Thank you for pointing out this mistake.
+	The Mach number of these simulations were added to the table.
+	Additionally, a paragraph was added to the end of Section 3.1 that aims to explain why the MRT results are less accurate despite being simulated with a finer lattice resolution.
+
+	\item \emph{Page 27: Lid driven cavity: Please explain the setup of the simulation in more detail:
+		e.g. boundary conditions, lattice Mach number, ...}
+
+	\item \emph{Page 30: Some exemplary figures of the flow pattern in the cavity, e.g. streamlines or
+		contour plots, would give more insight, especially about smoothness or oscillations in
+		the solution}
+
+	\item The typos that you mentioned in your review have also been corrected.
+
+	\end{itemize}
+
+Thank you again for your help. Please let us know of any further improvements that can be made to this paper.
+
+\vspace{1in}
+\noindent Sincerely, \\
+\indent John C. Brigham \vspace{0.1in}\\
+\indent Associate Professor \\
+\indent School of Engineering and Computing Sciences
+
 \end{document}

### 2018-01-03 src/stability-paper.tex
addressed reviewers comments

@@
 LBM is typically applied to simulate flow through a series of time steps, each consisting of streaming particle distributions to neighboring nodes and collisions of particle distributions at each node through a collision operator.
-The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated (e.g.
+The collision operator is of interest because it, along with the equilibrium distribution function, determines the physics that are simulated (e.g.
 constitutive laws, interfacial dynamics, etc.) and it has implications on numerical stability and computational efficiency.
-This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
+This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of their accuracy, numerical stability and computational efficiency.
 The investigation was carried out as a numerical study looking for qualitative, yet practical, results, including testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
@@
 The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behaviors, but at an increased computing cost that was, in some cases, as much as five times greater than the BGK collision operator.
-It was also shown that artificial dissipation can be an effective technique for enhancing numerical stability of non-Newtonian, lid-driven cavity flow, and is particularly effective for efficiently and accurately simulating the flow of shear-thinning fluids.
+It was also shown that, although the cost of artificial dissipation is that it introduces error in the constitutive response of the fluid (and therefore, may not accurately capture the physics of the flow), it can be an effective technique for stabilizing the numerics of non-Newtonian, lid-driven cavity flow simulations, and is particularly effective for stabilizing shear-thinning fluids.
 
@@
 A potential issue with the stability of the variable relaxation time approach is that as the collision frequency approaches $2$ the viscosity approaches zero and overrelaxation occurs.
-Alternatively, if the relaxation time is much greater than one, the accuracy and stability of the method degrades~\cite{latt2007hydrodynamic}. 
+Alternatively, if the relaxation time is much smaller than one, the accuracy and stability of the method degrades~\cite{latt2007hydrodynamic}. 
 In order to ensure that the variable collision frequencies did not approach values leading to numerical instabilities,~\cite{svec2011flow,svec2012free,gabbanelli2005lattice} set upper and lower bounds on allowable collision frequencies.
@@
 The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
-The BE is given as:
+The BE (in the absence of external forces) is given as:
 \begin{equation}
@@
 \end{figure}
+Using Chapman-Enskog multiscale analysis, one can show that in the limit of small Mach number and small Knudsen number (i.e.
+the ratio of the mean free path to the characteristic length of the flow) the LBE recovers the Navier-Stokes equations.
 
 The macroscopic variables of interest can be calculated from the particle distribution functions, $f(\pos, \pvel, t)$, by integrating moments of $f$ over velocity space.
-Due to the discrete nature of velocity in LBM, the integrals simply become summations.
+Such moments can be computed using a Gaussian-type quadrature.
 The mass density is given by the sum of the particle distributions and the momentum density is given by the first moment of the particle distributions over the velocity space:
@@
 
-\noindent The fluid pressure is related to macroscopic density through an equation of state:
+\noindent where $f_i(\pos, t) \equiv W_i f(\pos, \pvel_i, t)$, $\pvel_i$ are the quadrature points and $W_i$ are the quadrature weights~\cite{he1997priori}.
+(Note: in practice, the $W_i$ are not used or computed directly.
+Instead, the weights are enforced through relaxation and the equilibrium distribution function.
+More detail will be given in \Fref{sec:bgk}.) 
+In addition, the fluid pressure is related to macroscopic density through an equation of state:
 
@@
 
-\noindent where $\rho$ and $\mvel$ are dependent on $\pos$ and $t$, and $w_i$ is the weight in the $i$th:%MG: maybe say something about where this comes from, e.g.
-feq is a Hermite polynomial expansion of the Maxwell-Boltzmann distribution
+\noindent where $\rho$ and $\mvel$ are dependent on $\pos$ and $t$, and $w_i$ is the weight in the $i$th velocity direction:
 
@@
 \end{equation}
+\noindent Note first that \eqref{eq:feq} is only valid for small Mach numbers.
+Secondly, the above weights are determined by ensuring that the lattice velocity moments up to fourth order are identical to the respective velocity moments over the Maxwell distribution~\cite{wolf2000lattice,he1997priori}.
+The quadrature weights discussed in the context of \eqref{eq:rho} and \eqref{eq:mom} satisfy the relation $W_i = 2 \pi c_s^2 \exp \left(\pvel_i^2/2c_s^2\right)w_i$.
+(For more detail on the quadrature in velocity space, see \citet{he1997priori}, Section 5.2 of \citet{wolf2000lattice} or pg. 336 of \citet{chen1998lattice}.)
 
@@
 BGK consists of a single relaxation time and is a linear relaxation of particle distributions toward equilibrium.
-The BGK collision operator for the $i^{th}$ discrete velocity is expressed as:
+The BGK collision operator for the $i$th discrete velocity is expressed as:
 
@@
 Despite the utility of the BGK collision operator, it does have a few drawbacks.
-For example, in low viscosity fluids the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
-It is well known that when large nonequilibrium distributions exist in the LBM approximation that overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
+For example, in low viscosity fluids, the BGK operator results in an overrelaxation of particle distributions toward quasiequilibrium.
+It is well known that when large nonequilibrium distributions exist in the LBM approximation overrelaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
 To illustrate this, consider a flow in which there is a sharp spatial gradient in either $\rho$ or $\mvel$ at $\pos$.
@@
 
-The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
+The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as it is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
 Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
@@
 An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
-In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f} = \begin{Bmatrix}f_0 & f_1 & ...
-& f_8\end{Bmatrix}^T$, herein referred to as the ``moment space''.
+In the LB-MRT scheme, one constructs a space based on the particle velocity moments of $\mathbf{f} = \begin{Bmatrix}f_0 & f_1 & ... & f_8\end{Bmatrix}^T$, herein referred to as the ``moment space''.
 The collision is then performed in the moment space.
@@
 \begin{equation} \label{eq:mrt-colop}
-	\colop = - {\transM}^{-1} \relaxM \transM (\mathbf{f} - \mathbf{f}^{eq}),
+	\colop = - {\transM}^{-1} \relaxM \transM \left(\mathbf{f} - \mathbf{f}^{eq}\right),
 \end{equation}
@@
 \begin{equation} \label{eq:mrt-s7}
-\nu = c_s^2 \Delta t (\frac{1}{s_7} - \frac{1}{2}),
+\nu = c_s^2 \Delta t \left(\frac{1}{s_7} - \frac{1}{2}\right),
 \end{equation}
@@
 To reduce nonequilibrium fluctuations in LBM, one can introduce artificial dissipation.
-The idea of artificial dissipation is to increase numerical stability, while sacrificing some physical accuracy.
-A model that has more physical justification but produces unstable and nonsensical results is much less useful than a model with some minor artificial features yet produces more stable results.
+The idea of artificial dissipation is to increase numerical stability while sacrificing some physical accuracy.
 %Lyapanov stability is the idea that solutions at equilibrium stay there.
@@
 \begin{equation} \label{eq:rel-noneq-entropy}
-\Delta S = \sum_i f_i \ln(\frac{f_i}{f^{eq}_i}).
-\end{equation}
-
-\noindent A more computationally efficient approximation of $\Delta S$ can be achieved by instead using the second order Taylor expansion of \eqref{eq:rel-noneq-entropy}:
+\Delta S = \sum_i f_i \ln\left(\frac{f_i}{f^{eq}_i}\right).
+\end{equation}
+
+\noindent A more computationally efficient approximation of $\Delta S$ can be achieved by instead using the second-order Taylor expansion of \eqref{eq:rel-noneq-entropy}:
 
@@
 
-No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
+No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as bouncing back at the walls in the opposite direction from which they stream.
 For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
+In the LBM, the implementation of boundary conditions and external forces is generally not unique.
+The implementation of the no-slip boundary condition used in the current work is referred to as the \emph{complete bounceback} scheme~\cite{inamuro1995non,he1997analytic}.
 For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
@@
 
-Incorporating external forces, such as gravity, pressure gradients, etc., is done by adding a source of particle distributions in the direction of the force.
+Incorporating external forces, such as gravity, pressure gradients, etc., can be accomplished by adding a source of particle distributions in the direction of the force.
 The increase in particle distributions leads to the desired macroscopic result, which is an increase in momentum.
@@
 For the MRT-2 relaxation matrix, the free parameters were set to $s_1 = 1.1$, $s_2 = 1.0$, and $s_4 = s_6 = 1.2$, as these values have been successfully applied to simulating lid-driven cavity flow of non-Newtonian fluids in the past~\cite{chen2014simulations,li2014simulation}.
-Additionally, in regards to the LBM schemes, recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer to the true Bingham plastic behavior.
+Additionally, in regards to the LBM schemes, recall that $m$ is the stress growth exponent for the Papanastasiou approximation, and the higher $m$ is the closer the approximation is to the true Bingham plastic behavior.
+For comparison, a plot of the apparent viscosity as a function of strain rate for true Bingham behavior, and approximate Bingham behavior with $m = 10^8$, and $m = 10^5$ is shown in \Fref{fig:viscosity}.
 The relative $L_2$ and relative $L_{\infty}$ errors with respect to the analytical solution were computed for each simulation.
@@
 
+\begin{figure}
+	\centering
+	\includegraphics[width=\linewidth]{figs/viscosity_yield-stress}
+	\caption{Apparent viscosity for a yield stress fluid (with $\tau_y = 4 \times 10^{-5}$) as a function of strain rate for a true Bingham plastic, a Papanastasiou approximation with $m = 10^8$, and a Papanastasiou approximation with $m = 10^5$.}
+	\label{fig:viscosity}
+\end{figure}
+
 The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing}.
@@
 The Bingham number was computed by $Bn = \frac{\tau_y H}{\mu_p U}$.
+Because the accuracy of LBM in recovering the Navier-Stokes equations depends on the Mach number being small, the Mach number of each simulation is also tabulated.
 
@@
 \begin{tabulary}{\linewidth}{l l r r r r r r r r}
-\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
+\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Ma$ & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
 \hline \\
-BGK & No & $10^5$ & 4.0 & 6.05 & 0.68 & 0.0062 & 0.0153 & 1857 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0204 & 0.0411 & 2329 \\
-& & & 12.0 & 3.04 & 4.04 & 0.0503 & 0.0891 & 3345 \\
-& & & 16.0 & 1.92 & 8.52 & 0.1161 & 0.1879 & 2029 \\
-\\
-BGK & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0109 & 0.0282 & 2831 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0330 & 0.0670 & 3509 \\
-& & & 12.0 & 3.04 & 4.04 & 0.0788 & 0.1570 & 4838 \\
-& & & 16.0 & 1.92 & 8.52 & 0.1991 & 0.3539 & 3790 \\
-\\
-BGK & Yes & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0100 & 0.0273 & 2903 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0361 & 0.0823 & 3567 \\
-& & & 12.0 & 3.04 & 4.04 & 0.2439 & 0.3832 & 4800 \\
-& & & 16.0 & 1.92 & 8.52 & 0.7533 & 1.0718 & 3507 \\
-\\
-MRT-1 & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0018 & 0.0018 & 7908 \\
-& & & 12.0 & 3.04 & 4.04 & 0.0026 & 0.0026 & 7160 \\
-& & & 16.0 & 1.92 & 8.52 & 0.0041 & 0.0041 & 5559 \\
-\\
-MRT-2 & No & $10^8$ & 4.0 & 6.05 & 0.68 & 0.0012 & 0.0012 & 5284 \\
-& & & 8.0 & 4.42 & 1.85 & 0.0017 & 0.0017 & 5290 \\
-& & & 12.0 & 3.04 & 4.04 & 0.0024 & 0.0024 & 5244 \\
-& & & 16.0 & 1.92 & 8.52 & 0.0038 & 0.0038 & 4532 \\
+BGK & No & $10^5$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0062 & 0.0153 & 1857 \\
+& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0204 & 0.0411 & 2329 \\
+& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0503 & 0.0891 & 3345 \\
+& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.1161 & 0.1879 & 2029 \\
+\\
+BGK & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0109 & 0.0282 & 2831 \\
+& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0330 & 0.0670 & 3509 \\
+& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0788 & 0.1570 & 4838 \\
+& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.1991 & 0.3539 & 3790 \\
+\\
+BGK & Yes & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0100 & 0.0273 & 2903 \\
+& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0361 & 0.0823 & 3567 \\
+& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.2439 & 0.3832 & 4800 \\
+& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.7533 & 1.0718 & 3507 \\
+\\
+MRT-1 & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
+& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0018 & 0.0018 & 7908 \\
+& & & 12.0 & 3.04 & 4.04 & 0.016 &  0.0026 & 0.0026 & 7160 \\
+& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.0041 & 0.0041 & 5559 \\
+\\
+MRT-2 & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0012 & 0.0012 & 5284 \\
+& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0017 & 0.0017 & 5290 \\
+& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0024 & 0.0024 & 5244 \\
+& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.0038 & 0.0038 & 4532 \\
 \\
@@
 In contrast, the LBM approximation in \Fref{fig:bing-analyt-bgk-8} is not smooth, which suggests that the error for the BGK model with $m = 10^8$ is due to nonphysical oscillations.
-(Note: the velocity for the BGK model with $m = 10^8$ may appear asymmetric, but that is because the flow does not reach a steady-state and \Fref{fig:bing-analyt-bgk-8} is merely a snapshot in time.
+(Note: the velocity for the BGK model with $m = 10^8$ is asymmetric, but that is because the flow does not reach a steady-state and \Fref{fig:bing-analyt-bgk-8} is merely a snapshot in time.
 The oscillations in the velocity tend to move back and forth across the channel in subsequent time steps such that, on average (in time), the oscillations are indeed symmetric.)
@@
 More specifically, \Fref{fig:feq-vs-f_bgk} compares $f_i(\pos_j, 25000)$ to $f_i^{eq}(\pos_j, 25000)$ where $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, and $i = 5, 8$.
-It can be seen that in the BGK, $m = 10^8$ solution, that nonphysical oscillations pollute the quasiequilibrium profile as well.
+It can be seen in the BGK, $m = 10^8$ solution, that nonphysical oscillations pollute the quasiequilibrium profile as well.
 These nonphysical oscillations likely originate at the lattice scale due to the sharp gradient in the macroscopic velocity, $\mvel$, near the walls, and the sharp gradient in $\mvel$ is a result of a sharp gradient in $\mu_{app}$, namely the sharp gradient of the constitutive relationship.
 Because the nonphysical oscillations make their way into the quasiequilibrium values, entropic median filtering does not help dampen the oscillations but instead contracts the particle distributions closer to the quasiequilibrium oscillations, which explains why the median filtered results were less accurate for Bingham plastic Poiseuille flow.
-In order to ensure that this phenomena was a side effect of all entropic filtering, and not just entropic median filtering with $\theta = 1.0 \times 10^{-6}$ and $n_s = 2.7$, optimization was used to find the value of $\theta$ that minimized error for both median filtering and Ehrenfests' regularization.
+In order to ensure that this phenomenon was a side effect of all entropic filtering, and not just entropic median filtering with $\theta = 1.0 \times 10^{-6}$ and $n_s = 2.7$, optimization was used to find the value of $\theta$ that minimized error for both median filtering and Ehrenfests' regularization.
 The optimization problem was defined as follows:
@@
 On average, the LBM model using the MRT-1 collision operator took 2.9 times more computing time than the BGK collision operator with $m = 10^5$; and the model using MRT-2 took 2.1 times more computing time than the BGK collision operator with $m = 10^5$.
+The increased computing for the MRT-1 and MRT-2 with respect to the BGK collision operator with $m = 10^8$ was approximately 2.2 and 1.7, respectively, for the smaller yield stresses ($\tau_y = 4 \times 10^-5, 8 \times 10^-5$).
+For larger yield stresses ($\tau_y = 12 \times 10^-5, 16 \times 10^-5$), the increased computing time was less, namely 1.5 and 1.1, respectively.
+The increased computational efficiency of the BGK collision operator with $m = 10^8$ with respect to its MRT counterparts was less for the larger yield stresses, likely because the nonphysical oscillations in the BGK solutions resulted in requiring more iterations for the constitutive relation to converge.
 However, the increased computing time can be justified for the MRT collision operator because the MRT solutions did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
 when $m = 10^8$.
-The MRT collision schemes were therefore the most accurate solutions for all cases; and because the difference in accuracy between the MRT-1 and MRT-2 collision schemes was negligible, the remainder of the discussion in this section will focus on comparing the BGK collision schemes and the MRT-1 collision scheme.
+The MRT collision schemes were, therefore, the most accurate solutions for all cases; and because the difference in accuracy between the MRT-1 and MRT-2 collision schemes was negligible, the remainder of the discussion in this section will focus on comparing the BGK collision schemes and the MRT-1 collision scheme.
 \Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilibrium profiles, $f_i^{eq}$ for the MRT-1 model.
@@
 
-Another important question to ask is, if the strength of the MRT collision operator is damping out the ghost modes, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of the ghost modes?
-\Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment, $\boldsymbol{\epsilon}^{neq}$, with respect to time for each of the collision schemes, which was calculated as:
+Recall that: (1) when the relaxation rates $s_0 = s_1 = ...
+= s_8 = \omega$, the MRT collision operator is equivalent to the BGK collision operator (where $\omega$ is the collision frequency in the BGK sense), and (2) that the relaxation rates $s_1, s_2, s_4$ and $s_6$ are chosen with flexibility and may be tuned in an effort to increase stability.
+Thus, an important question to ask is, if the strength of the MRT collision operator is damping out the nonequilibrium velocity moments with tunable relaxation parameters, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of these nonequilibrium moments and if so which ones?
+\Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment, $\epsilon^{neq}$, with respect to time for each of the collision schemes, which was calculated as:
 \begin{equation}
@@
 $\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
-The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT-1 or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ displayed was in part due to the $\epsilon$ moment.
+The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT-1 or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ displayed may in part be due to the $\epsilon$ moment.
 
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon}
-    \caption{Evolution of relative $L_2$ norm, $\epsilon^{neq}$ with time.
-The norm of the $\epsilon^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $\epsilon$ moment.}
+    \caption{Evolution of relative $L_2$ norm, $\boldsymbol{\epsilon}^{neq}$ with time.
+The norm of the $\boldsymbol{\epsilon}^{neq}$ across the height of the channel is an indicator of an increase in oscillations at the lattice level due to the $\epsilon$ moment.}
     \label{fig:epsilon}
@@
 
-\Fref{fig:qx} shows a measure of the nonequilibrium $q_x$ moment, $\mathbf{q_x}^{neq}$, or energy flux in the x-direction, with respect to time for each of the collision schemes, which was calculated as:
+\Fref{fig:qx} shows a measure of the nonequilibrium $q_x$ moment, $q_x^{neq}$, or energy flux in the x-direction, with respect to time for each of the collision schemes, which was calculated as:
 \begin{equation}
@@
 The values measured for the BGK with $m = 10^8$ and median filtered were consistently greater than any other collision scheme.
-The relative $L_2$ norm of the $\mathbf{q_x}^{neq}$ for the BGK collision operator with $m = 10^8$ and median filtering solution spent most of the simulation between approximately 50\% and 100\%, suggesting that the $q_x$ moment was likely the cause of the nonphysical oscillations and error in the solution.
+The relative $L_2$ norm of the $q_x^{neq}$ for the BGK collision operator with $m = 10^8$ and median filtering solution spent most of the simulation between approximately 50\% and 100\%, suggesting that the $q_x$ moment was also possibly the cause of nonphysical oscillations and error in the solution.
 As expected (if the dominant source of error in the LBM approximation for Bingham plastic Poiseuille flow is indeed due to the $q_x$ moment), the $\mathbf{q_x^{neq}}$ relative $L_2$ norm for the BGK with $m = 10^8$ was slightly greater than the BGK with $m = 10^5$, and the $\mathbf{q_x^{neq}}$ relative $L_2$ norm was negligible for the MRT-1 collision scheme.
+The nonequilibrium moments for the other tunable relaxation rates, namely the $e$ and $q_y$, were also calculated.
+However, their relative values were orders of magnitude lower than $\epsilon^{neq}$ and $q_x^{neq}$.
 
@@
 
-It is likely that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time (i.e.
-osciallations build in amplitude with time).
+It is also possible that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time (i.e.
+oscillations build in amplitude with time).
 \Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time, which was calculated, respectively, as:
@@
 \noindent where $T$ is the current time. 
-\Fref{fig:qx} and \Fref{fig:qx-cumulative} shows that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the ghost mode associated with the $q_x$ moment dominates and is the primary source of oscillations.
-It can be inferred that the $q_x$ moment was a significant source of oscillations in this case because the Poiseuille flow was in the x-direction.
-Overall, the results presented suggest that if an LBM collision scheme is to be developed for simulating high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it should focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.
+\Fref{fig:qx} and \Fref{fig:qx-cumulative} show that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the nonequilibrium $q_x$ moment may be the dominated source of oscillations.
+It can be inferred that the nonequilibrium $q_x$ moment was much larger than the nonequilibrium $q_y$ moment in this case because the Poiseuille flow was in the x-direction.
+Overall, the results presented suggest that if an LBM collision scheme is to be developed for simulating high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it may be important to focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.
 
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/epsilon_cumulative}
-    \caption{Cumulative relative $L_2$ norm, $\epsilon^{neq}$ with time.
-Oscillations can have a positive effect on each other, building up.
-The cumulative relative $L_2$, $\epsilon^{neq}$ is a measure of how much oscillations due to the $\epsilon$ moment may have been building in time.}
+    \caption{Cumulative relative $L_2$ norm, $\boldsymbol{\epsilon}^{neq}$ with time.
+Oscillations can have a positive effect on each other.
+The cumulative relative $L_2$, $\boldsymbol{\epsilon}^{neq}$ is a measure of how much oscillations due to the $\epsilon$ moment may have been building in time.}
     \label{fig:epsilon-cumulative}
@@
     \includegraphics[width=\linewidth]{figs/poise-bingham/qx_cumulative}
-    \caption{Cumulative relative $L_2$ norm, $\mathbf{q_x^{neq}}$ with time.
-Oscillations can have a positive effect on each other, building up.
-The cumulative relative $L_2$, $\mathbf{q_x^{neq}}$ is a measure of how much oscillations due to the $q_x$ moment may have been building in time.}
+    \caption{Cumulative relative $L_2$ norm, $\mathbf{q_x}^{neq}$ with time.
+Oscillations can have a positive effect on each other.
+The cumulative relative $L_2$, $\mathbf{q_x}^{neq}$ is a measure of how much oscillations due to the $q_x$ moment may have been building in time.}
     \label{fig:qx-cumulative}
@@
 
-Lastly, simulations with the same collision schemes, Reynold's numbers and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
+An alternative explanation is that the oscillations develop first in other velocity moments.
+The nonequilibrium moments associated with $\rho$, $j_x$, $j_y$, and $p_{xy}$ did not differ appreciably between collision operators.
+The relative $L_2$ norm, $\mathbf{p_{xx}}^{neq}$ is shown in \Fref{fig:pxx}.
+The values associated with the BGK collision operator with $m = 10^8$ and with the BGK collision operator with $m = 10^8$ and entropic filtering are again much larger in comparison to the BGK with $m = 10^5$ and MRT-1 collision operators.
+However, the relaxation rate associated with the $p_{xx}$ moment is given by the constitutive relationship and therefore is consistent regardless of collision operator.
+So although the nonphysical oscillations (in the BGK with $m = 10^8$ and BGK with $m = 10^8$ and entropic filtering collision operators) may be in part caused by the nonequilibrium $p_{xx}$ moment, it is unlikely that they originate there.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\linewidth]{figs/poise-bingham/pxx}
+	\caption{Evolution of relative $L_2$ norm, $\mathbf{p_{xx}}^{neq}$ with time.
+The dotted line, the dashed line, the solid line and the dashed dotted line represent the BGK collision operator with $m = 10^5$, the BGK collision operator with $m = 10^8$, the BGK collision operator with entropic filtering, and the MRT collision operator, respectively.}
+	\label{fig:pxx}
+\end{figure}
+
+Lastly, simulations with the same collision schemes, Reynold's numbers, and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
 The plastic viscosity was held constant ($\mu_p = 0.2$.), while the yield stress and pressure gradient were varied in order to match the Reynold's numbers and Bingham numbers in \Fref{tab:poise-bing}.
@@
 
+It is also interesting that, despite the finer grid resolution for the same characteristic flow, the MRT-1 and MRT-2 collision operators were less accurate on the $32 \times 128$ lattices than the $32 \times 64$ lattices.
+A possible explanation for this nonintuitive result is that, as mentioned in~\cite{conrad2015accuracy}, although non-Newtonian lattice Boltzmann simulations are second order accurate, the accuracy is not preserved for every simulation Mach number.
+Although the LBE recovers the Navier-Stokes equations in the limit of small Mach number, an additional artificial viscosity can be introduced for certain Mach numbers, and the optimal simulation Mach number depends on the constitutive parameters of the non-Newtonian fluid and the collision operator.
+Thus, although the error in the LBE with respect to the Navier-Stokes equations was reduced by lowering the Mach number, an increase in the error in the constitutive relation may have been introduced that caused the MRT-1 and MRT-2 $32 \times 128$ simulations to have a greater net error (with respect to the analytical solution) than their $32 \times 64$ counterparts.
+
 \begin{table}
 	\centering
-	\caption{Bingham plastic Poiseuille flow, $128 \times 32$ lattice}
+	\caption{Bingham plastic Poiseuille flow, $32 \times 128$ lattice}
 	\vspace{0.5cm}
 	\begin{tabulary}{\linewidth}{l l r r r r r r r r}
-		\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
+		\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Ma$ & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
 		\hline \\
-		BGK & No & $10^5$ & 1.0 & 6.05 & 0.68 & 0.0419 & 0.0450 & 4951 \\
-		              & & & 2.0 & 4.42 & 1.85 & 0.0300 & 0.0370 & 5209 \\
-	                  & & & 3.0 & 3.04 & 4.04 & 0.0206 & 0.0260 & 5296 \\
-		              & & & 4.0 & 1.92 & 8.52 & 0.0649 & 0.1016 & 5304 \\
-		\\
-		BGK & No & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0426 & 0.0470 & 6484 \\
-		              & & & 2.0 & 4.42 & 1.85 & 0.0306 & 0.0394 & 7111 \\
-		              & & & 3.0 & 3.04 & 4.04 & 0.0230 & 0.0438 & 7595 \\
-		              & & & 4.0 & 1.92 & 8.52 & 0.0963 & 0.1659 & 7791 \\
-		\\
-		BGK & Yes & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0427 & 0.0468 & 5576 \\
-		               & & & 2.0 & 4.42 & 1.85 & 0.0309 & 0.0397 & 7512 \\
-		               & & & 3.0 & 3.04 & 4.04 & 0.0236 & 0.0443 & 7856 \\
-		               & & & 4.0 & 1.92 & 8.52 & 0.0934 & 0.1624 & 7848 \\
-		\\
-		MRT-1 & No & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0452 & 0.0484 & 13837 \\
-		              & & & 2.0 & 4.42 & 1.85 & 0.0404 & 0.0435 & 16515 \\
-                      & & & 3.0 & 3.04 & 4.04 & 0.0316 & 0.0338 & 17142 \\
-                      & & & 4.0 & 1.92 & 8.52 & 0.0197 & 0.0210 & 17853 \\
-		\\
-		MRT-2 & No & $10^8$ & 1.0 & 6.05 & 0.68 & 0.0453 & 0.0485 & 14030 \\
-		& & & 2.0 & 4.42 & 1.85 & 0.0404 & 0.0435 & 15290 \\
-		& & & 3.0 & 3.04 & 4.04 & 0.0316 & 0.0338 & 16328 \\
-		& & & 4.0 & 1.92 & 8.52 & 0.0198 & 0.0210 & 16816 \\
+		BGK & No & $10^5$ & 1.0 & 0.016 & 6.05 & 0.68 & 0.0419 & 0.0450 & 4951 \\
+		              & & & 2.0 & 0.012 & 4.42 & 1.85 & 0.0300 & 0.0370 & 5209 \\
+	                  & & & 3.0 & 0.008 & 3.04 & 4.04 & 0.0206 & 0.0260 & 5296 \\
+		              & & & 4.0 & 0.005 & 1.92 & 8.52 & 0.0649 & 0.1016 & 5304 \\
+		\\
+		BGK & No & $10^8$ & 1.0 & 0.016 & 6.05 & 0.68 & 0.0426 & 0.0470 & 6484 \\
+		              & & & 2.0 & 0.012 & 4.42 & 1.85 & 0.0306 & 0.0394 & 7111 \\
+		              & & & 3.0 & 0.008 & 3.04 & 4.04 & 0.0230 & 0.0438 & 7595 \\
+		              & & & 4.0 & 0.005 & 1.92 & 8.52 & 0.0963 & 0.1659 & 7791 \\
+		\\
+		BGK & Yes & $10^8$ & 1.0 & 0.016 & 6.05 & 0.68 & 0.0427 & 0.0468 & 5576 \\
+		               & & & 2.0 & 0.012 & 4.42 & 1.85 & 0.0309 & 0.0397 & 7512 \\
+		               & & & 3.0 & 0.008 & 3.04 & 4.04 & 0.0236 & 0.0443 & 7856 \\
+		               & & & 4.0 & 0.005 & 1.92 & 8.52 & 0.0934 & 0.1624 & 7848 \\
+		\\
+		MRT-1 & No & $10^8$ & 1.0 & 0.016 & 6.05 & 0.68 & 0.0452 & 0.0484 & 13837 \\
+		              & & & 2.0 & 0.012 & 4.42 & 1.85 & 0.0404 & 0.0435 & 16515 \\
+                      & & & 3.0 & 0.008 & 3.04 & 4.04 & 0.0316 & 0.0338 & 17142 \\
+                      & & & 4.0 & 0.005 & 1.92 & 8.52 & 0.0197 & 0.0210 & 17853 \\
+		\\
+		MRT-2 & No & $10^8$ & 1.0 & 0.016 & 6.05 & 0.68 & 0.0453 & 0.0485 & 14030 \\
+					& & & 2.0 & 0.012 & 4.42 & 1.85 & 0.0404 & 0.0435 & 15290 \\
+					& & & 3.0 & 0.008 & 3.04 & 4.04 & 0.0316 & 0.0338 & 16328 \\
+					& & & 4.0 & 0.005 & 1.92 & 8.52 & 0.0198 & 0.0210 & 16816 \\
 		\\
@@
 \noindent where $i$ is the node index in the x-direction, $j$ is the node index in the y-direction, and $k$ is the current time step.
-The lid velocity was prescribed as $u_o = 0.1$.
+The lid velocity was prescribed as $u_o = 0.1$; thus, the simulation Mach number is $Ma = u_o / c_s = 0.1 \sqrt{3}$.
 All coordinate values presented in this section (used to describe the location of the center of vortices) are given normalized with respect to the length of the cavity side (i.e. as $(x / L, y / L)$).
@@
 
+In order to compare the overall flow pattern for each of the collision schemes, the streamline plots for select $Re$ and $Bn$ numbers are presented.
+\Fref{fig:lid-Bn10-Re100}-\ref{fig:lid-Bn100-Re100} show the streamlines for $Bn = 10$, $Re = 100$; $Bn = 10$, $Re = 1000$; and $Bn = 100$, $Re = 100$, respectively.
+Something to note in \Fref{fig:lid-Bn10-Re100} and \Fref{fig:lid-Bn10-Re1000} is that in both cases (1) the streamlines near the main vortex look similar for each collision operator and (2) the collision operators with a more accurate Bingham approximation (i.e.
+BGK with $m = 10^8$, BGK with $m = 10^8$ and entropic filtering, MRT with $m = 10^8$ and MRT with $m = 10^8$ and entropic filtering) resulted in a less smooth, more complex flow regime in the bottom corners of the cavity.
+Additionally, adding artificial dissipation did not seem to alter the characteristic flow pattern much for either the BGK with $m = 10^8$ or MRT with $m = 10^8$; however, there is a clear difference between the BGK with $m = 10^8$ and MRT with $m = 10^8$ collision operators.
+For the larger Bingham number simulations depicted in \Fref{fig:lid-Bn100-Re100}, one notices significant differences in the flow throughout the cavity, even near the main vortices.
+
+\begin{figure}
+	\centering
+	\begin{tabulary}{\linewidth}{c c}
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-5_Bn10_Re100/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-8_Bn10_Re100/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_svec-8_Bn10_Re100/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-filter_Bn10_Re100/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_mrt-8_Bn10_Re100/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_mrt-filter_Bn10_Re100/sl_step-000050000}
+	\end{tabulary}
+	\caption{Lid-driven solutions at steady-state or 50000 time steps for $Bn = 10$ and $Re = 100$.
+Top-left: BGK with $m = 10^5$; top-right: BGK with $m = 10^8$; middle-left: BGK-BRT with $m = 10^8$; middle-right: BGK with $m = 10^8$ and entropic filtering; bottom-left: MRT with $m = 10^8$; bottom-right: MRT with $m = 10^8$ and entropic filtering.}
+	\label{fig:lid-Bn10-Re100}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{tabulary}{\linewidth}{c c}
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-5_Bn10_Re1000/sl_step-000037145}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-8_Bn10_Re1000/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_svec-8_Bn10_Re1000/sl_step-000036192}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-filter_Bn10_Re1000/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_mrt-8_Bn10_Re1000/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_mrt-filter_Bn10_Re1000/sl_step-000050000}
+	\end{tabulary}
+	\caption{Lid-driven solutions at steady-state or 50000 time steps for $Bn = 10$ and $Re = 1000$.
+Top-left: BGK with $m = 10^5$; top-right: BGK with $m = 10^8$; middle-left: BGK-BRT with $m = 10^8$; middle-right: BGK with $m = 10^8$ and entropic filtering; bottom-left: MRT with $m = 10^8$; bottom-right: MRT with $m = 10^8$ and entropic filtering.}
+	\label{fig:lid-Bn10-Re1000}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{tabulary}{\linewidth}{c c}
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-5_Bn100_Re100/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-8_Bn100_Re100/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_svec-8_Bn100_Re100/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_bgk-filter_Bn100_Re100/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_mrt-8_Bn100_Re100/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/bingham_100_mrt-filter_Bn100_Re100/sl_step-000050000}
+	\end{tabulary}
+	\caption{Lid-driven solutions at steady-state or 50000 time steps for $Bn = 100$ and $Re = 100$.
+Top-left: BGK with $m = 10^5$; top-right: BGK with $m = 10^8$; middle-left: BGK-BRT with $m = 10^8$; middle-right: BGK with $m = 10^8$ and entropic filtering; bottom-left: MRT with $m = 10^8$; bottom-right: MRT with $m = 10^8$ and entropic filtering.}
+	\label{fig:lid-Bn100-Re100}
+\end{figure}
+
 In summary, for the lid-driven cavity flow of a Bingham plastic fluid, 
@@
 
+To compare overall flow patterns, the streamlines (for convergent collision schemes) for $n = 0.5, Re = 1000$, $n = 0.5, Re = 10000$, $n = 1.5, Re = 1000$, and $n = 1.5, Re = 10000$ are shown in Figures \ref{fig:lid-n05-Re1000}-\ref{fig:lid-n15-Re10000}, respectively.
+The differences in streamlines between collision schemes that converge for the same flow behavior index and Reynold's number are subtle.
+\Fref{fig:n05-Re1000} shows that each convergent collision scheme for $n = 0.5, Re = 1000$ forms a main vortex near the center of the cavity, and vortices in the top-right, bottom-left, and bottom-right of the cavity.
+There is a slight variation in the shape of the vortices when comparing the BGK with entropic filtering to the two MRT collision schemes.
+For the same flow behavior index but larger Reynolds, $Re = 10000$, \Fref{fig:n05-Re10000}, again, vortices appear in similar locations in the cavity between collision schemes, but with slightly different shape.
+Note, the MRT with entropic filtering solution appears to be more smooth than the BGK with entropic filtering solution.
+\Fref{fig:lid-n15-Re1000} shows almost no difference in streamlines between the various collision schemes for the flow behavior index $n = 1.5$ and $Re = 1000$.
+For the same flow behavior index but larger Reynolds number, $Re = 10000$, \Fref{fig:lid-n15-Re10000} the main difference is between the BGK collision schemes and MRT collision schemes where the former has a single vortex form in the bottom right of the cavity while the latter has two vortices form in the bottom right of the cavity.
+
+\begin{figure}
+	\centering
+	\begin{tabulary}{\linewidth}{c c}
+		\includegraphics[width=\figwid]{figs/lid/power_bgk-filter_100_n0.5_Re1000/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/power_mrt_100_n0.5_Re1000/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/power_mrt-filter_100_n0.5_Re1000/sl_step-000050000}
+	\end{tabulary}
+	\caption{Lid-driven solutions at steady-state or 50000 time steps for $n = 0.5$ and $Re = 1000$. Top-left: BGK with entropic filtering; top-right: MRT; bottom-left: MRT with entropic filtering.}
+	\label{fig:lid-n05-Re1000}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{tabulary}{\linewidth}{c c}
+		\includegraphics[width=\figwid]{figs/lid/power_bgk-filter_100_n0.5_Re10000/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/power_mrt-filter_100_n0.5_Re10000/sl_step-000050000}
+	\end{tabulary}
+	\caption{Lid-driven solutions at steady-state or 50000 time steps for $n = 0.5$ and $Re = 10000$. Left: BGK with entropic filtering; right: MRT with entropic filtering.}
+	\label{fig:lid-n05-Re10000}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{tabulary}{\linewidth}{c c}
+		\includegraphics[width=\figwid]{figs/lid/power_bgk_100_n1.5_Re1000/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/power_svec_100_n1.5_Re1000/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/power_bgk-filter_100_n1.5_Re1000/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/power_mrt_100_n1.5_Re1000/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/power_mrt-filter_100_n1.5_Re1000/sl_step-000050000}
+	\end{tabulary}
+	\caption{Lid-driven solutions at steady-state or 50000 time steps for $n = 1.5$ and $Re = 1000$.
+Top-left: BGK; top-right: BGK-BRT; middle-left: BGK with entropic filtering; middle-right: MRT; bottom-left: MRT with entropic filtering.}
+	\label{fig:lid-n15-Re1000}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\begin{tabulary}{\linewidth}{c c}
+		\includegraphics[width=\figwid]{figs/lid/power_bgk_100_n1.5_Re10000/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/power_svec_100_n1.5_Re10000/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/power_bgk-filter_100_n1.5_Re10000/sl_step-000050000}
+		&
+		\includegraphics[width=\figwid]{figs/lid/power_mrt_100_n1.5_Re10000/sl_step-000050000}
+		\\
+		\includegraphics[width=\figwid]{figs/lid/power_mrt-filter_100_n1.5_Re10000/sl_step-000050000}
+	\end{tabulary}
+	\caption{Lid-driven solutions at steady-state or 50000 time steps for $n = 1.5$ and $Re = 10000$.
+Top-left: BGK; top-right: BGK-BRT; middle-left: BGK with entropic filtering; middle-right: MRT; bottom-left: MRT with entropic filtering.}
+	\label{fig:lid-n15-Re10000}
+\end{figure}
+
 \section{Conclusions} \label{sec:numerical-study-conclusions}
@@
 Two different benchmark problems were chosen for the study: Poiseuille flow, and lid-driven square cavity flow.
-The results showed that for higher yield stress fluids in Poiseuille-type flow, only the MRT collision operator was able use an accurate approximation to true Bingham plastic behavior and not suffer from nonphysical oscillations.
+The results showed that for higher yield stress fluids in Poiseuille-type flow, only the MRT collision operator was able to use an accurate approximation to true Bingham plastic behavior and not suffer from nonphysical oscillations.
 The oscillations appeared to be due to high nonequilibrium values for the moment related to the square of the energy, $\epsilon$, and the energy flux moment in the direction of flow, $q_x$.
-If a collision scheme is to be developed for higher yield stress, Poiseuille-type flow that is (1) more accurate than the BGK collision operator with $m = 10^5$ and (2) more computationally efficient than the MRT collision operator, then it should focus on dampening the nonequilibrium $\epsilon$ moment and energy flux moment (in the direction of flow).
+If a collision scheme is to be developed for higher yield stress, Poiseuille-type flow that is (1) more accurate than the BGK collision operator with $m = 10^5$ and (2) more computationally efficient than the MRT collision operator, then focusing on dampening the nonequilibrium $\epsilon$ moment and energy flux moment (in the direction of flow) may be an effective strategy.
 
@@
 
-For nearly all of the cases considered in the current study, the MRT collision operator was more accurate, but also more computationally expensive than its BGK counterpart, and is some times even orders of magnitude slower.
-
-In regards to future work, \cite{conrad2015accuracy} showed that non-Newtonian Lattice Boltzmann simulations are not second order accurate for every simulation Mach number, and showed that a greater accuracy may be achieved by properly tuning the simulation Mach number.
+For nearly all of the cases considered in the current study, the MRT collision operator was more accurate, but also more computationally expensive than its BGK counterpart, and is sometimes even orders of magnitude slower.
+
+In regards to future work, \cite{conrad2015accuracy} showed that non-Newtonian Lattice Boltzmann simulations are not second-order accurate for every simulation Mach number, and showed that a greater accuracy may be achieved by properly tuning the simulation Mach number.
 \cite{conrad2015accuracy} predicted optimal simulation Mach numbers for various shear-thinning and shear-thickening fluids. 
 Possible future work would include extending the numerical investigation to consider the effect of the simulation Mach number on the accuracy, stability, and efficiency of the various collision operators and stability enhancements in simulating non-Newtonian flow.
+Additionally, investigating more LBM models, such as the Two Relaxation Times (TRT) model~\cite{ginzburg2010optimal} and the entropic lattice Boltzmann method (ELBM), and more stabilization techniques, such as a Viscosity Counteracting Approach~\cite{zhang2016improving} might be of interest.

### 2018-01-03 response-to-reviewers.tex
finished responses to reviewers

@@
 	\begin{enumerate}
-		\item first
+		\item \emph{The authors may test other non-Newtonian fluids more than the present Bingham plastics and power-law fluids.}
+
+		Thank you for your comment.
+		More complex non-Newtonian behavior would indeed to be interesting to investigate.
+		However, as some non-Newtonian constitutive models combine and/or modify Bingham plastic and power-law behavior (e.g.
+Herschel–Bulkley fluids) these two non-Newtonian fluids were seen as representing the building blocks, in a sense, of more complex non-Newtonian behavior.
+		While investigating more non-Newtonian fluids presents interesting future work, we believe that, in the interest of brevity, the current work should remain focused on Bingham plastics and power-law fluids.
+
+		\item \emph{How about the non-Newtonian flow in complex geometries,such as the porous media? It is more strict for the stability requirement.}
+
+		Thank you for the suggestion.
+		Similarly, although the authors agree that such an investigation would be interesting, in the interest of brevity we have chosen not to expand the current investigation beyond Poiseuille flow and lid-driven cavity flow.
+
+		\item \emph{An outlook for further models to improve the simulation stability of LBM.}
+
+		The discussion of future work at the end of the paper has been expanded to discuss more possible LBM models and stabilization techniques.
 
@@
 
+	In the second to last paragraph of page 6, it is mentioned that in the limit of small Mach number and small Knudsen number that the LBE recovers the Navier-Stokes equations.
+	Also, the last paragraph of Section 3.1 discusses some of the nuances of the simulation Mach number in the context of simulating non-Newtonian fluids.
+	In particular, it mentions that, although reducing the Mach number should increase the accuracy of the LBE with respect to the Navier-Stokes equations, it can introduce error in the constitutive relation.
+
 	\item \emph{Page 10: ``A model that has more physical justification but produces unstable and
@@
 
+	This is a good point.
+	The discussed has been expanded to compare the MRT calculations to the BGK, $m = 10^8$ cases.
+
 	\item \emph{Page 23, 24: You draw some conclusions between the ``ghost'' modes ($\epsilon$ and $q$) and
@@
 
+	The authors agree with the reviewer.
+	The evidence is inconclusive and as such one can only make loose inferences on the actual cause of oscillations.
+	The language in the discussion in pages 23-29 has been altered to reflect the fact that the results do not offer direct proof.
+	In addition, parts of the discussion has been expanded to better motivate and explain why the authors believe the nonequilibrium $\epsilon$ and $q_x$ moments may be the primary source of oscillations.
+
 	\item \emph{Page 23, 24: Did you investigate if there are oscillations in the moment e, related to
@@
 		producing some of the superior MRT stability properties.}
+
+	The nonequilibrium $e$ moment was indeed calculated and investigated.
+	It has been added to the discussion of the nonequilibrium moments.
 
@@
 
+	Again, the authors agree that direct proof cannot be found in the results.
+	The language has been changed to reflect this.
+	In addition, parts of the discussion has been expanded to better motivate and explain why the authors believe the nonequilibrium $\epsilon$ and $q_x$ moments may be the primary source of oscillations.
+
 	\item \emph{Page 25: ``Overall, the results presented suggest that if an LBM collision scheme is to
@@
 		model (TRT) of Ginzburg et al.?}
+
+	The two relaxation time model was not investigating. However, the discussion of possible future work has been expanded to included the TRT.
 
@@
 
+	The setup of the simulations is given in more detail.
+
 	\item \emph{Page 30: Some exemplary figures of the flow pattern in the cavity, e.g. streamlines or
@@
 		the solution}
+
+	Streamlines comparing the flow patterns of the various collision schemes for some of the more interesting non-Newtonian fluids and Reynolds numbers are now provided and accompanied by discussions of the results.
+	Thank you for the suggestion.
+	This should give a clearer picture of the similarities and differences between the solutions of the various collision schemes.
 

### 2018-01-04 src/stability-paper.tex
finished responses to reviewers
addressed Scott's comments

@@
 The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behaviors, but at an increased computing cost that was, in some cases, as much as five times greater than the BGK collision operator.
-It was also shown that, although the cost of artificial dissipation is that it introduces error in the constitutive response of the fluid (and therefore, may not accurately capture the physics of the flow), it can be an effective technique for stabilizing the numerics of non-Newtonian, lid-driven cavity flow simulations, and is particularly effective for stabilizing shear-thinning fluids.
+It was also shown that, although it introduces error in the constitutive response of the fluid (and therefore, may not accurately capture the physics of the flow), artificial dissipation can be an effective technique for stabilizing the numerics of non-Newtonian, lid-driven cavity flow simulations, and is particularly effective for stabilizing shear-thinning fluids.
 
@@
 \end{figure}
-Using Chapman-Enskog multiscale analysis, one can show that in the limit of small Mach number and small Knudsen number (i.e.
-the ratio of the mean free path to the characteristic length of the flow) the LBE recovers the Navier-Stokes equations.
+Using Chapman-Enskog multiscale analysis, one can show that, in the limit of small Mach number and small Knudsen number (i.e.
+the ratio of the mean free path to the characteristic length of the flow), the LBE recovers the Navier-Stokes equations.
 
@@
 
-The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as it is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
+The challenge associated with high viscosity fluids is that some distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in underrelaxation.
 Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
-The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the local collision frequency may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.
+The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the local collision frequency may result in overrelaxation in certain parts of the domain and underrelaxation in others.
 
@@
 %Lyapanov stability is the idea that solutions at equilibrium stay there.
-A practical goal then would be to use only the necessary amount of artificial dissipation in order to ensure a stable solution.
-From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation is necessary?'' and ``how much artificial dissipation does one introduce when it is necessary?''.
+A practical goal then would be to use only the necessary amount of artificial dissipation in order to ensure a stable solution within a reasonable amount of error in the constitutive relation.
+From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation should be introduced?'' and ``how much artificial dissipation should be introduced?''.
 
@@
 The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing}.
-The Reynold's number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution.
+The Reynolds number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution.
 The Bingham number was computed by $Bn = \frac{\tau_y H}{\mu_p U}$.
@@
 BGK & No & $10^5$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0062 & 0.0153 & 1857 \\
-& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0204 & 0.0411 & 2329 \\
-& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0503 & 0.0891 & 3345 \\
-& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.1161 & 0.1879 & 2029 \\
+& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0204 & 0.0411 & 2329 \\
+& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.0503 & 0.0891 & 3345 \\
+& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.1161 & 0.1879 & 2029 \\
 \\
 BGK & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0109 & 0.0282 & 2831 \\
-& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0330 & 0.0670 & 3509 \\
-& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0788 & 0.1570 & 4838 \\
-& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.1991 & 0.3539 & 3790 \\
+& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0330 & 0.0670 & 3509 \\
+& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.0788 & 0.1570 & 4838 \\
+& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.1991 & 0.3539 & 3790 \\
 \\
 BGK & Yes & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0100 & 0.0273 & 2903 \\
-& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0361 & 0.0823 & 3567 \\
-& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.2439 & 0.3832 & 4800 \\
-& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.7533 & 1.0718 & 3507 \\
+& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0361 & 0.0823 & 3567 \\
+& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.2439 & 0.3832 & 4800 \\
+& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.7533 & 1.0718 & 3507 \\
 \\
 MRT-1 & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
-& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0018 & 0.0018 & 7908 \\
-& & & 12.0 & 3.04 & 4.04 & 0.016 &  0.0026 & 0.0026 & 7160 \\
-& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.0041 & 0.0041 & 5559 \\
+& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0018 & 0.0018 & 7908 \\
+& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.0026 & 0.0026 & 7160 \\
+& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.0041 & 0.0041 & 5559 \\
 \\
 MRT-2 & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0012 & 0.0012 & 5284 \\
-& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0017 & 0.0017 & 5290 \\
-& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0024 & 0.0024 & 5244 \\
-& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.0038 & 0.0038 & 4532 \\
+& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0017 & 0.0017 & 5290 \\
+& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.0024 & 0.0024 & 5244 \\
+& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.0038 & 0.0038 & 4532 \\
 \\
@@
 On average, the LBM model using the MRT-1 collision operator took 2.9 times more computing time than the BGK collision operator with $m = 10^5$; and the model using MRT-2 took 2.1 times more computing time than the BGK collision operator with $m = 10^5$.
-The increased computing for the MRT-1 and MRT-2 with respect to the BGK collision operator with $m = 10^8$ was approximately 2.2 and 1.7, respectively, for the smaller yield stresses ($\tau_y = 4 \times 10^-5, 8 \times 10^-5$).
-For larger yield stresses ($\tau_y = 12 \times 10^-5, 16 \times 10^-5$), the increased computing time was less, namely 1.5 and 1.1, respectively.
-The increased computational efficiency of the BGK collision operator with $m = 10^8$ with respect to its MRT counterparts was less for the larger yield stresses, likely because the nonphysical oscillations in the BGK solutions resulted in requiring more iterations for the constitutive relation to converge.
+The increased computing time for the MRT-1 and MRT-2 with respect to the BGK collision operator with $m = 10^8$ was approximately 2.2 and 1.7, respectively, for the smaller yield stresses ($\tau_y = 4 \times 10^{-5}, 8 \times 10^{-5}$).
+For the larger yield stresses ($\tau_y = 12 \times 10^{-5}, 16 \times 10^{-5}$), the increased computing time was less, namely 1.5 and 1.1, respectively.
+The increased computational efficiency of the BGK collision operator with $m = 10^8$ with respect to its MRT counterparts was less for the larger yield stresses, likely because the nonphysical oscillations in the BGK solutions required more iterations for the constitutive relation to converge.
 However, the increased computing time can be justified for the MRT collision operator because the MRT solutions did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e.
@@
 = s_8 = \omega$, the MRT collision operator is equivalent to the BGK collision operator (where $\omega$ is the collision frequency in the BGK sense), and (2) that the relaxation rates $s_1, s_2, s_4$ and $s_6$ are chosen with flexibility and may be tuned in an effort to increase stability.
-Thus, an important question to ask is, if the strength of the MRT collision operator is damping out the nonequilibrium velocity moments with tunable relaxation parameters, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of these nonequilibrium moments and if so which ones?
+Thus, an important question to ask is, if the strength of the MRT collision operator is damping out the nonequilibrium velocity moments with tunable relaxation parameters, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of these nonequilibrium moments and, if so, which ones?
 \Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment, $\epsilon^{neq}$, with respect to time for each of the collision schemes, which was calculated as:
@@
 \noindent where $T$ is the current time. 
-\Fref{fig:qx} and \Fref{fig:qx-cumulative} show that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the nonequilibrium $q_x$ moment may be the dominated source of oscillations.
+\Fref{fig:qx} and \Fref{fig:qx-cumulative} show that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the nonequilibrium $q_x$ moment may be the dominant source of oscillations.
 It can be inferred that the nonequilibrium $q_x$ moment was much larger than the nonequilibrium $q_y$ moment in this case because the Poiseuille flow was in the x-direction.
@@
 
-Lastly, simulations with the same collision schemes, Reynold's numbers, and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
-The plastic viscosity was held constant ($\mu_p = 0.2$.), while the yield stress and pressure gradient were varied in order to match the Reynold's numbers and Bingham numbers in \Fref{tab:poise-bing}.
+Lastly, simulations with the same collision schemes, Reynolds numbers, and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
+The plastic viscosity was held constant ($\mu_p = 0.2$.), while the yield stress and pressure gradient were varied in order to match the Reynolds numbers and Bingham numbers in \Fref{tab:poise-bing}.
 The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing-2}.
@@
 
-It is also interesting that, despite the finer grid resolution for the same characteristic flow, the MRT-1 and MRT-2 collision operators were less accurate on the $32 \times 128$ lattices than the $32 \times 64$ lattices.
+It is also worth noting that, despite the finer grid resolution for the same characteristic flow, the MRT-1 and MRT-2 collision operators were less accurate on the $32 \times 128$ lattices than the $32 \times 64$ lattices.
 A possible explanation for this nonintuitive result is that, as mentioned in~\cite{conrad2015accuracy}, although non-Newtonian lattice Boltzmann simulations are second order accurate, the accuracy is not preserved for every simulation Mach number.
@@
 As was done herein, lid-driven cavity flow is generally characterized by a square cavity where a fluid velocity is prescribed tangential to the upper boundary and the remaining boundaries have a no-slip boundary condition.
+More specifically, the upper boundary is a velocity Zou and He~\cite{zou1997pressure} boundary condition and the remaining boundaries are complete bounceback boundaries.
 The example of lid-driven cavity flow utilized herein is presented schematically in \Fref{fig:lid-driven}.
@@
 
-For the Bingham plastic numerical tests the Reynold's number was varied with the following values: $\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the Bingham number was varied with the following values: $\begin{bmatrix}1,&10,&100\end{bmatrix}$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
+For the Bingham plastic numerical tests the Reynolds number was varied with the following values: $\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the Bingham number was varied with the following values: $\begin{bmatrix}1,&10,&100\end{bmatrix}$ ($Bn$ and $Re$ were calculated the same as in \Fref{sec:poise-bing}).
 Six collision schemes were tested: (1) BGK with $m = 10^5$, (2) BGK with $m = 10^8$, (3) BGK-BRT with $m = 10^8$, (4) BGK with $m = 10^8$ and median filtering, (5) MRT with $m = 10^8$ and (6) MRT with $m = 10^8$ and median filtering.
@@
 In order to compare the overall flow pattern for each of the collision schemes, the streamline plots for select $Re$ and $Bn$ numbers are presented.
-\Fref{fig:lid-Bn10-Re100}-\ref{fig:lid-Bn100-Re100} show the streamlines for $Bn = 10$, $Re = 100$; $Bn = 10$, $Re = 1000$; and $Bn = 100$, $Re = 100$, respectively.
+Figures \ref{fig:lid-Bn10-Re100}-\ref{fig:lid-Bn100-Re100} show the streamlines for $Bn = 10$, $Re = 100$; $Bn = 10$, $Re = 1000$; and $Bn = 100$, $Re = 100$, respectively.
 Something to note in \Fref{fig:lid-Bn10-Re100} and \Fref{fig:lid-Bn10-Re1000} is that in both cases (1) the streamlines near the main vortex look similar for each collision operator and (2) the collision operators with a more accurate Bingham approximation (i.e.
@@
 \begin{itemize}
-    \item for low Reynold's number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a relatively small amount of computing time,
-    \item for high Reynold's number flow the BGK collision operator requires entropic filtering to remain stable,
+    \item for low Reynolds number flow the BGK collision operator with $m = 10^5$ provides an accurate solution with a relatively small amount of computing time,
+    \item for high Reynolds number flow the BGK collision operator requires entropic filtering to remain stable,
     \item and the MRT collision operator with $m = 10^8$ produces solutions with high accuracy and stability for all of the cases examined herein, but at an increased (approximately 5--10 times more) computing cost.
@@
 
-For the power-law numerical tests the Reynold's number was varied with the following values: $\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the flow behavior index, $n$, was varied with the following values: $\begin{bmatrix}0.5,&1.5\end{bmatrix}$.
-The Reynold's number was calculated as $\rho \frac{U^{2-n} H^n}{k}$.
+For the power-law numerical tests the Reynolds number was varied with the following values: $\begin{bmatrix}100,&1000,&5000,&10000\end{bmatrix}$, and the flow behavior index, $n$, was varied with the following values: $\begin{bmatrix}0.5,&1.5\end{bmatrix}$.
+The Reynolds number was calculated as $\rho \frac{U^{2-n} H^n}{k}$.
 Five collision schemes were tested: (1) BGK, (2) BGK-BRT, (3) BGK with median filtering, (4) MRT and (5) MRT with median filtering.
@@
 
-For the shear-thinning results, $n = 0.5$, at the lowest Reynold's number considered, $Re = 100$, all of the collision schemes produced results that agreed well with the literature value.
-However, the solutions became unstable at higher Reynold's numbers and entropic filtering was necessary to produce a stable solution when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
+For the shear-thinning results, $n = 0.5$, at the lowest Reynolds number considered, $Re = 100$, all of the collision schemes produced results that agreed well with the literature value.
+However, the solutions became unstable at higher Reynolds numbers and entropic filtering was necessary to produce a stable solution when $Re \ge 1000$ for the BGK collision operator, and when $Re \ge 5000$ for the MRT collision operator.
 Although there still appears to be a tradeoff between computational efficiency and accuracy in regards to the BGK with median filtering and MRT with median filtering, it does not seem to be nearly as significant as when simulating Bingham plastic lid-driven flow (e.g.
@@
 For the shear-thickening results, $n = 1.5$, there were no issues of instability.
-What is interesting is that all of the collision schemes produce similar results for each of the cases, with the difference in the location of the main vortex being no greater than 1\% in either the $x$ or $y$ direction for any two collision schemes (with the exception of the high Reynold's number case, $Re = 10000$).
+What is interesting is that all of the collision schemes produce similar results for each of the cases, with the difference in the location of the main vortex being no greater than 1\% in either the $x$ or $y$ direction for any two collision schemes (with the exception of the high Reynolds number case, $Re = 10000$).
 The BGK-BRT consistently needed the least amount of computing time.
@@
 
-To compare overall flow patterns, the streamlines (for convergent collision schemes) for $n = 0.5, Re = 1000$, $n = 0.5, Re = 10000$, $n = 1.5, Re = 1000$, and $n = 1.5, Re = 10000$ are shown in Figures \ref{fig:lid-n05-Re1000}-\ref{fig:lid-n15-Re10000}, respectively.
-The differences in streamlines between collision schemes that converge for the same flow behavior index and Reynold's number are subtle.
-\Fref{fig:n05-Re1000} shows that each convergent collision scheme for $n = 0.5, Re = 1000$ forms a main vortex near the center of the cavity, and vortices in the top-right, bottom-left, and bottom-right of the cavity.
+To compare overall flow patterns, the streamlines (for convergent collision schemes) for $n = 0.5, Re = 1000$; $n = 0.5, Re = 10000$; $n = 1.5, Re = 1000$; and $n = 1.5, Re = 10000$ are shown in Figures \ref{fig:lid-n05-Re1000}-\ref{fig:lid-n15-Re10000}, respectively.
+The differences in streamlines between collision schemes that converge for the same flow behavior index and Reynolds number are subtle.
+\Fref{fig:lid-n05-Re1000} shows that each convergent collision scheme for $n = 0.5, Re = 1000$ forms a main vortex near the center of the cavity, and vortices in the bottom-left and bottom-right of the cavity.
 There is a slight variation in the shape of the vortices when comparing the BGK with entropic filtering to the two MRT collision schemes.
-For the same flow behavior index but larger Reynolds, $Re = 10000$, \Fref{fig:n05-Re10000}, again, vortices appear in similar locations in the cavity between collision schemes, but with slightly different shape.
-Note, the MRT with entropic filtering solution appears to be more smooth than the BGK with entropic filtering solution.
+For the same flow behavior index but larger Reynolds, $Re = 10000$, \Fref{fig:lid-n05-Re10000}, again, vortices appear in similar locations in the cavity between collision schemes, but with slightly different shape.
+Note that the MRT with entropic filtering solution appears to be more smooth than the BGK with entropic filtering solution.
 \Fref{fig:lid-n15-Re1000} shows almost no difference in streamlines between the various collision schemes for the flow behavior index $n = 1.5$ and $Re = 1000$.
@@
 
-The results of the study suggest that entropic filtering can be an effective technique for enhancing stability in non-Newtonian flow, especially for high Reynold's number flows and flows that tend to produce vortices.
+The results of the study suggest that entropic filtering can be an effective technique for enhancing stability in non-Newtonian flow, especially for high Reynolds number flows and flows that tend to produce vortices.
 However, if the filter is not carefully tuned by properly adjusting the threshold or number of standard deviations then the physical integrity, and consequently accuracy, of the model can be adversely impacted.

### 2018-02-12 response-to-reviewers.tex
addressed Scott's comments
submission accepted for publicaton

@@
 We hope that the attached edited copy of ``Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow'' will answer any questions previously raised, and will be satisfactory for publication.
+The response is followed by a copy of the manuscript with marked up changes for the convenience of any reviewer interested in seeing such a copy.
 
@@
 	This is a good point.
-	The discussed has been expanded to compare the MRT calculations to the BGK, $m = 10^8$ cases.
+	The discussion has been expanded to compare the MRT calculations to the BGK, $m = 10^8$ cases.
 
@@
 
-	The two relaxation time model was not investigating. However, the discussion of possible future work has been expanded to included the TRT.
+	The two relaxation time model was not investigated. However, the discussion of possible future work has been expanded to included the TRT.
 
@@
 
-\vspace{1in}
+\vspace{6in}
 \noindent Sincerely, \\

### 2018-02-12 manuscript-srcs/src/cementing-notes.tex
submission accepted for publicaton

- turbulent flow is good for displacement
- typical pump rates include 2--15 (2, 5, 10, 15) bbl/min
-- 4-6 is most common
- typical yield stress 10 lbf / 100 sqft = approximately 5 pascals
-- range from 2.5 to 30 pascals
- typical plastic viscosities are 10-100 cP
-- 1 Pa s = 10 P = 1000 cP

- standard casing sizes 4.5" -- 20" in outside diameter
- corresponding annuli areas: 12 -- 48 sqin 
                              0.08 -- 0.333 sqft
- 1 bbl = 5.6145833 cft
- 2 bbl/min   = 0.18715 cft / sec
  15 bbl/min  = 1.40365 cft / sec

- velocities range: 0.562 -- 17.5 ft/sec
- dynamic visco range: 0.000208854 -- 0.00208854 lbf s / ft^2
- slurry density range: 12.0 -- 18.0 lb/gal
                        90   -- 135 lb/cft
- kinematic visco range: 1.5e-6 -- 2.32e-5 ft s
- Re range:

### 2018-02-12 manuscript-srcs/src/cementing-paper.tex
submission accepted for publicaton

\section*{Abstract}

When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
%Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
An important yet vulnerable barrier is the cement annulus and is created through a process known as primary cementing.
Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
%Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flow, multiphase multicomponent flow, and flow through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
In \Fref{sec:numerical-methods}, an algorithm for simulating non-Newtonian free-surface flow using LBM is presented.
The algorithm was implemented and used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
The study is followed by conclusions and a discussion of future work.

\section{Introduction}

% what is primary cementing?

Of critical importance to drilling operations of all kinds is that barriers are placed to prevent gasses and fluids from migrating from one geological zone to another (i.e., zonal isolation is achieved)~\cite{economides19901,jackson1993effect,goodwin1992cement,dusseault2014towards,li2016theory}.
If zonal isolation is not achieved, gasses and fluids may migrate through the well to the surface and up into the atmosphere--potentially causing pollution, or at the very least, contributing to an increase in greenhouse gasses--or, worse yet, formation gasses and fluids can migrate into aquifers, potentially harming wild life and people.
Furthermore, it can be argued that the most important, yet vulnerable, barrier in terms of achieving zonal isolation of a wellbore is the cement annulus that is created between the casing and the formation.
There are various mechanisms by which this cement annulus can fail.
According to previous studies~\cite{Bit02,Che85,Lev79,Ste88,Tal93,Zul12,li2016theory}, the most prevalent failure mechanisms of the cement annulus are:
\begin{itemize}
\item Stresses developing in the cement annulus as a result of temperature gradients, moisture gradients, overburden pressure, and chemical shrinkage of the cement matrix.
These stresses can cause cracking in the cement, or debonding at either the casing--cement or cement--formation interfaces.
\item Poor fill of the annulus.
If cement slurry does not fill the annulus and if drilling mud is left behind the mud can dry, crack, and provide a weak path for fluids and gasses under pressure to push through, or if a dry hole is cemented and not filled properly, fluids and gasses can travel through channels and voids left in the annular space.
\item Gas channeling through the annulus during cement curing.
Initially, after the annulus is poured, the cement column transmits its full hydrostatic pressure against the rock formation.
As the cement annulus hydrates, it solidifies, becoming stiffer and more self-supporting.
When the column begins to support itself, the hydrostatic pressure of the cement column--against the rock formation--begins to drop.
If the hydrostatic pressure of the cement column drops below the pressure of a formation gas before the cement annulus has developed adequate strength to resist it, the formation gas can penetrate into the cement annulus and form channels, or degrade the cement annulus in other ways.
\end{itemize}
% what research has been done that is considered classical?

% why do we need modelling?
% Before the end of this paragraph--modeling is important because...

In order to better understand, and thus improve primary cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
For example,~\citet{Bei77} developed a mathematical model to describe the displacement of drilling mud by the the cement slurry during primary cementing.
Among their assumptions were that the flow as laminar, the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the distance from the center of the annulus to either the rock formation or steel casing surface.
The model was then used to explore the effect of various parameters on the displacement efficiency of the spacer, such as the density ratio between the mud and the spacer, the viscosity ratio between the mud and the spacer, the displacement rate, the displaced phase yield stress, and the displacing phase yield stress.
It was concluded that displacement efficiency could be improved by increasing the density of the displacing phase, increasing the viscosity of the displacing phase, (in general) decreasing the displacement rate, decreasing the yield stress of the displaced phase, and increasing the yield stress of the displacing phase.
\citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
The model was more general than the model developed by~\citet{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
The bulk fluid motion of the spacer and drilling mud during primary cementing were investigated.
The intended use of the model was not to make general statements about what properties of the spacer and drilling mud were desirable, but to instead be used iteratively during the design process for the cement job.
Results showed that the model was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
\citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citet{Bei77} or~\citet{Bit02}.
The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
\citet{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
\citet{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
\citet{zhao2016lattice} used the model to investigate the shape of the interface between the displacing phase (i.e. cement slurry) and displaced phase (i.e. drilling mud) in a horizontal, eccentric annulus.
It was found, in general, that a greater density ratio between the displacing phase and displaced phase and lower eccentricity results in better displacement in a horizontal, eccentric annulus.

% talk about new LBM study

Overall, past studies have tended to idealize the rock formation as a straight wall.
Although this is a desirable simplification to make for capturing bulk fluid phenomena, it does not accurately describe what is happening locally at the surface of the rock formation.
Flow near the surface of the formation, and in and around the imperfections of the formation geometry, could influence the bond of the cement annulus to the rock formation, and the fill of the cement slurry very near to the rock formation.
If an adequate bond does not develop, debonding can occur as a result of thermal and mechanical stresses.
Even if most of the annulus is filled by the cement slurry, a small gap at the cement--formation interface that stretches for a long enough span of the wellbore would still be susceptible to fluid and gas migration, i.e. loss of zonal isolation.
One could even argue that the characteristic of the cement slurry flow is most important local to the cement--formation interface, and that considering imperfections in the rock formation is imperative to determining whether zonal isolation has been achieved.

The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features.
The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation.
LBM also has other advantages as an approach for simulating primary cementing:
\begin{itemize}
\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
Many fluids present in primary cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
\item LBM is well-suited for simulating multiphase multicomponent flow and free-surface flow because it can handle complicated interface shapes between fluid phases and components.
\item LBM is easily written in parallel.
As hardware architectures shift from one or a few cores to several CPUs, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
\end{itemize}
The current work focuses on the problem of cementing a dry hole, i.e. an annulus that is not filled with drilling mud prior to cementing.
In order to numerically investigate this problem, an LBM model for non-Newtonian and free-surface flow is developed.
The details of the LBM model are outlined in \Fref{sec:numerical-methods}.
What follows in \Fref{sec:results} is a preliminary study on how different characteristics of flow, e.g. cement slurry yield stress, cement slurry inlet velocity, etc., perform in regards to fill of the annulus for different wellbores with imperfect rock formation geometries.
\Fref{sec:conclusion} outlines conclusions resulting from the numerical study and a discussion of future work.

\section{Numerical Methods} \label{sec:numerical-methods}

\subsection{Lattice Boltzmann Method}

% We need to explain/justify why 2D vs. 3D

For the sake of computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
Idealizing the annulus as two-dimensional limits the modeling approach in that annulus eccentricity and other three-dimensional phenomena could not be considered.
A two-dimensional representation of the annulus was deemed sufficient as the focus of this study is to establish the Lattice Boltzmann Method as a suitable method for simulating primary cementing in an annulus with an imperfect rock formation surface.
In addition to using a D2Q9 lattice, the BGK collision operator used was in the current chapter.
For more details on the fundamentals of the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.

\subsection{Cement Slurry Constitutive Relationship}

Cement slurry is typically modeled using a non-Newtonian constitutive relationship.
%A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
As is commonly utilized for modeling cement and concrete fluid dynamics~\cite{ferraris2001fresh,ferraris1998testing,tattersall2003workability,roussel2007computational,roussel2015numerical,svec2012free}, the Bingham plastic constitutive relationship was utilized herein.
Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
The Bingham plastic constitutive model is difficult to work with numerically.
Therefore, a smooth approximation, specifically the Papastasiou approximation, is often used~\cite{papanastasiou1987flows,chen2014simulations,tang2011bingham,Cha11,wang2008lattice,phillips2011lattice} and is used herein.
Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flow.
For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.

\subsection{Free Surface Flow}

A free-surface flow is a flow in which two-fluids exist but the dynamics of one fluid does not need to be explicitly modeled.
Typically, the phase of the primary fluid, or the fluid of interest, is a liquid; and the secondary phase that does not need to be explicitly modeled is a gas, often times air at atmospheric pressure.
The primary fluid is modeled using a standard CFD method, such as LBM, and the interaction of the secondary fluid on the primary fluid is modeled as a boundary condition at the interface between the two fluids.
The interface is what is referred to as the free-surface. \pagebreak
The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice} and \citet{thurey2005interactive}.

\subsection{Capturing the Free-Surface}

Consider at each node a cell of size $\text{d}x \times \text{d}x$, where $\text{d}x$ is the distance between nodes, and the cell is centered at the node.
In order to track where the primary fluid is and where it is not (i.e. which cells contain the primary fluid), a variable is introduced, $\epsilon$, which describes what fraction of the cell is covered by the primary fluid.
Cells that have a fluid fraction of $\epsilon = 0$ contain no primary fluid (i.e. are empty); cells that have a fluid fraction of $\epsilon = 1$ are filled; and cells that have a fluid fraction $0 < \epsilon < 1$ are neither completely filled nor are they completely empty.
The fluid fraction concept is similar to how fluids are tracked in Volume of Fluid (VOF) methods~\cite{nichols1981volume}.

\subsubsection{Mass Transfer}

It is important to note that the particle distributions, $f_0, ..., f_8$, cannot be used to directly determine the fluid fraction of a cell.
Neither can the zeroth velocity moment of the particle distributions, $\rho = \sum_i f_i$.
For the algorithm used in this work, $\epsilon(\pos, t) \neq \epsilon(\mathbf{f(\pos, t)}, \rho(\pos, t))$.
Instead, each $f_i$ is related to the amount of primary fluid advection in each of the discrete velocity directions, $\pvel_i$.
To track the fluid fraction and how it changes, another variable is introduced, $M(\pos, t)$, which denotes the (primary) fluid mass in each cell.
The fluid mass is related to the fluid fraction by $\epsilon = M / \rho$.
The change in mass, $\Delta M(\pos, t)$ is related to $\mathbf{f}(\pos, t)$ and particle distribution in the neighborhood of $\pos$.
Given initial conditions on $M(\pos, 0)$ and a fixed time step size $\Delta t $, the fluid mass is determined by $M(\pos, t_n) = M(\pos, 0) + \Delta t \sum_{k=1}^n \Delta M(\pos, t_k)$.
What remains to track the fluid mass is to develop an expression for $\Delta M(\pos, t)$.

\subsubsection{Cell States} \label{sec:mass-transfer}

Each of the cells in the domain are assigned one of three states: fluid, interface, or gas.
The set of fluid, interface, and gas cells are denoted by $\fset$, $\iset$, and $\gset$, respectively.
A fluid cell is a cell that is entirely covered by the primary fluid; and a gas cell is a cell that is completely empty, or does not contain any primary fluid.
An interface is partially covered, and there is always a layer of interface cells between fluid and gas cells so that a fluid cell and a gas cell are never neighbors.
The change in mass is calculated for fluid and interface cells (i.e. $\pos \in \fset \cup \iset$) for each discrete velocity direction as follows:
\begin{equation} \label{eq:dM}
\Delta M_i(\pos, t) = \begin{cases}
f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t), & \pos \in \fset \text{ or } \pos + \pvel_i \Delta t \in \fset \\
%\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
  \frac{\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)}{2} (f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t)), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
0, & \pos + \pvel_i \Delta t \in \gset
\end{cases}
\end{equation}
\noindent where $\pvel_\opidx$ represents the discrete velocity vector that is opposite in direction to $\pvel_i$ (i.e. $\pvel_\opidx = - \pvel_i$).
\eqref{eq:dM} states that when the border of the two cells is covered with fluid, the change in mass for a discrete velocity direction is equal to the difference of the particle distribution entering the cell from the opposite direction and particle distribution leaving the cell in that direction.
When the border of the two cells are not covered with fluid (i.e. when both cells are interface cells) the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the cell border that is covered in fluid.
The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M_i(\pos, t)$.
Finally, note that $\Delta M_i(\pos, t) = -\Delta M_\opidx(\pos + \pvel_i \Delta t, t)$ for all $\pos \in \domain$ where $\domain$ is the domain.
As a consequence, it is clear that $\sum_{\pos \in \domain}\Delta M(\pos, t) = 0$ (i.e. that mass is conserved for the mass transfer step).

Tracking the transport of mass throughout the domain is an important step in determining what parts of the domain contain the primary fluid and what do not.
However, for the free-surface to change shape and propagate, cells need to be able to transition state.
Cell states change based on the rule set presented in \Fref{tab:cell-transition-rules} where $\tconst$ is a small, computational constant that keeps cells from transitioning back and forth between fluid and interface, or interface and gas states in subsequent time steps.
To summarize \Fref{tab:cell-transition-rules}, gas cells cannot transition to fluid cells and vice versa, interface cells transition to fluid cells when their mass has exceeded $\rho$ by a computational constant, and gas cells when their mass is less than zero minus a computational constant.
In addition, fluid cells and gas cells transition to the interface state in order to maintain continuity of the interface, i.e. to ensure there is always a layer of interface cells in between fluid cells and gas cells.
An example of a cell transitioning state is depicted in \Fref{fig:cell-transition}.

\begin{table}
\caption{Cell state transition rules.}
\begin{tabulary}{\linewidth}{l|L|L}
\textbf{Current State} & \textbf{Condition} & \textbf{Transition} \\
\hline & \\
$(\pos, t) \in \iset$ & $M(\pos, t) < -\tconst$ & $(\pos, t + \Delta t) \rightarrow \gset$, $M(\pos, t + \Delta t) = 0$ \\
                      & $M(\pos, t) > \tconst + \rho(\pos, t)$ & $(\pos, t + \Delta t) \rightarrow \fset$, $M(\pos, t + \Delta t) = \rho(\pos, t + \Delta t)$ \\
                      & &\\
$(\pos, t) \in \fset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \gset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
& & \\
$(\pos, t) \in \gset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \fset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
\end{tabulary}
\label{tab:cell-transition-rules}
\end{table}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figs/cell-transition}
\caption{Example of cell state transitions.
The interface cell highlighted in red on the left is transitioned to a gas cell as it has emptied.
On the right, the three cells highlighted in red were transitioned from fluid to interface cells in order to keep the interface continuous.}
\label{fig:cell-transition}
\end{figure}

\subsubsection{Updating Cell States and Mass Redistribution} \label{sec:cell-updates}

The rules for updating cell states in \Fref{tab:cell-transition-rules} are fairly intuitive, however, there are some details that must be considered in order to ensure mass is properly conserved and interface continuity is achieved (e.g. note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells).
The algorithm for updating cell states follows that which was presented in~\citet{thurey2005interactive}.
It is presented here as \Fref{algo:update-cell-states} for the sake of completeness and for outlining minor modifications.

\input{src/algorithm/update-cell-states.tex}

\Fref{algo:update-cell-states}, as presented, ensures both mass conservation and continuity of the free-surface.
Moreover, when mass redistribution inevitably occurs, it is first attempted to be redistributed and weighted in the direction in which the interface is propagating (e.g. line \algref{algo:update-cell-states}{line:redist-dop-f} or line \algref{algo:update-cell-states}{line:redist-dop-g}.
If there are no interface cells in the direction of interface propagation, it is instead redistributed to neighboring interface cells (e.g. line \algref{algo:update-cell-states}{line:redist-neighbors-f} or line \algref{algo:update-cell-states}{line:redist-neighbors-g}).
If there is no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells (e.g. line \algref{algo:update-cell-states}{line:final-redist}).
The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...
Lastly, note that the interface normal, $\mathbf{n}$, is calculated using a finite difference approximation,
\begin{equation} \label{eq:n}
\mathbf{n}_{ij} = \frac{1}{2} \begin{Bmatrix}
\epsilon_{i-1, j} - \epsilon_{i+1, j} \\
\epsilon_{i, j-1} - \epsilon_{i, j+1}
\end{Bmatrix}
\end{equation}
\noindent where $i$ is the index of the node in the x-direction and $j$ is the index of the node in the y-direction.

\subsection{Boundary Conditions at the Free-Surface} \label{sec:bc-at-fs}

Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and how particle distributions that are missing from the streaming step are reconstructed.
The degrees of freedom, $f_i$, are not tracked for gas cells.
After the streaming step, interface cells will be missing particle distributions that would have streamed from gas cells--an issue that is depicted in \Fref{fig:ps}.

\begin{figure}
\centering
\includegraphics{figs/ps}
\caption{Example of a scenario in which particle distributions are missing post-streaming.}
\label{fig:ps}
\end{figure}

The missing particle distributions need to be reconstructed.
Also, momentum needs to be conserved at the interface.
When the particle distributions are reconstructed, they will be reconstructed in such a way that momentum is conserved between the primary fluid and the secondary fluid.
In order to perform the reconstruction, there are a few assumptions that need to be made:
\begin{enumerate}
\item The velocity of the secondary fluid is equal to the velocity of the primary fluid at the free-surface, i.e. there is no-slip between the two-fluids.
\item The secondary fluid is at equilibrium and at constant pressure, $p_G$.
\item There is a force balance for opposite lattice directions.
\end{enumerate}
Because particle distributions are a measure of momentum transport, the pressure of the secondary fluid can be converted to the particle distribution scale and the missing particle distribution can be solved for algebraically.
Consider \Fref{fig:fbi}.
Balancing the opposite lattice directions results in:
\begin{eqnarray}
f_i + f_\opidx &=& f^{eq}_\opidx(\rho_G, \mathbf{u}) + f_i^{eq}(\rho_G, \mathbf{u}), \\
\label{eq:reconst} f_\opidx &=& f_\opidx^{eq}(\rho_G, \mathbf{u}) + f_i^{eq}(\rho_G, \mathbf{u}) - f_i.
\end{eqnarray}
Missing distribution functions can be reconstructed according to \eqref{eq:reconst}.
In order for conservation of momentum to be satisfied along the free-surface, distribution functions coming from the direction of the interface normal, $\mathbf{n}$, must also be reconstructed.
Thus, distribution functions are reconstructed if coming from a gas cell, or if $\mathbf{n} \cdot \pvel_\opidx > 0$.

\begin{figure}
\centering
\includegraphics{figs/force-bal-at-interface}
\caption{Particle distributions at free-surface for a pair of opposing lattice directions after gas pressure, $\rho_G$, is in the particle distribution form.}
\label{fig:fbi}
\end{figure}

\subsection{Resulting Algorithm for Simulating Free-Surface Flow using the Lattice Boltzmann Method}

For clarity, the complete algorithm used in the current work for simulating free-surface flow using the Lattice Boltzmann Method is presented in \Fref{algo:free-surface-complete}.

\input{src/algorithm/free-surface-complete.tex}

\newcommand{\nodesi}{50}
\newcommand{\nodesj}{500}
\newcommand{\anw}{$1\frac{1}{4}$"}
\newcommand{\anh}{1'}
\newcommand{\nsteps}{16000}

\section{Numerical Study of Primary Cementing} \label{sec:results}

The free-surface LBM model was implemented and used to simulate primary cementing in a dry annulus, i.e. an annulus without drilling mud.
The focus of the current study was the performance of different characteristic flow behaviors (as could potentially relate to different cement slurry mixes) when used to cement different imperfect wellbore geometries.
Because the focus in regards to the wellbore is on geometrical imperfections, the entire wellbore annulus is not simulated, but instead a small section around a geometrical imperfection is considered.
A schematic of an example primary cementing simulation, including boundary conditions, is shown in \Fref{fig:wellbore-cementing-schematic}.

\begin{figure}
\centering
\includegraphics{figs/wellbore-cementing-schematic}
\caption{Example simulation used in primary cementing study.
The black rectangles on the left boundary represent notches protruding from the rock formation surface. The contours are of the primary fluid (cement slurry) mass.
The schematic on the right is courtesy of~\cite{cementing2015web}}.
\label{fig:wellbore-cementing-schematic}
\end{figure}

%The simulations were carried out on a $\nodesi \times \nodesj$ lattice, which corresponds to an annulus width of \anw \hspace{0.05cm} and a sectional height of \anh.
The simulations were carried out on a $\nodesi \times \nodesj$ lattice.
For the mass inlet, a~\citet{zou1997pressure} velocity boundary condition was used in conjuction with a constant mass boundary condition, i.e. $M(\pos_j, t) = M_0 = 1.0$ for all $\pos_j$ along the inlet.
This corresponded to a constant mass flow, $\dot{M}_{in} = M_0 u_0$, at inlet.
For the outlet, mass was allowed to transfer out of the domain during the mass transfer step such that $f_\opidx(\pos + \pvel_i \Delta t, t) = 0$ in \eqref{eq:dM} for all $\pos + \pvel_i \Delta t$ that lie outside of the domain on the other side of the mass outlet (i.e. the boundary condition at the outlet is such that there is no backflow).
All simulations were executed for $\nsteps$ time steps.

The different imperfect geometries that were considered at the rock formation surface were: flow over a cavity that is 30\% of the annulus width, ``cav $30 \times 1$''; flow over a cavity that is 60\% of the annulus width, ``cav $60 \times 1$''; flow over a square obstacle with sides that are 30\% of the annulus width, ``obs $30 \times 1$''; flow over two square obstacles with sides that are 30\% of the annulus width, ``obs $30 \times 2$''; flow over a square obstacle with sides that are 60\% of the annulus width, ``obs $60 \times 1$''; flow over two square obstacles with sides that are 60\% of the annulus width, ``obs $60 \times 2$''; and flow over a rectangular obstacle with width that is 60\% of the annulus width and length that is four times the width, ``obs $60 \times L$''.
The wellbore geometries that were considered herein are depicted in \Fref{fig:wellbore-geometries}.

\begin{figure}
\includegraphics[width=\linewidth]{figs/wellbore-geometries}
\caption{Wellbore geometries considered in the study.}
\label{fig:wellbore-geometries}
\end{figure}

In addition to wellbore geometry, different cement slurry material properties and cement slurry flow rates were considered, which are characterized by the dimensionless Reynold's number, $Re = \frac{\rho U L}{\mu_p}$, and Bingham number, $Bn = \frac{\tau_y L}{\mu_p U}$, of the flow, where $L \equiv $ the width of the annulus.
A particular goal of the study was to investigate which Reynold's numbers and Bingham numbers were preferable for each geometry, and which Reynold's numbers and Bingham numbers performed well for all geometries.
The dimensionless numbers were defined as follows, $(Re, Bn) = (1.00, 4.00)$,  $(2.50, 0.00)$, $(2.50, 1.00)$, $(2.50, 10.0)$, $(2.50, 25.0)$, $(5.00, 5.00)$, and $(12.5, 2.0)$.
\Fref{fig:percent-filled} shows the percentage of the wellbore annulus section filled at the end of the cement simulation duration.
The percent fill was calculated as:
\begin{equation}
  \frac{\sum\limits_{\pos \in \domain} \epsilon(\pos, 16000)}{\sum\limits_{\pos \in \domain} 1.0}.
\end{equation}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figs/percent_filled}
\caption{Percent of voids in wellbore at the end of simulation for each cement slurry flow through each wellbore geometry.}
\label{fig:percent-filled}
\end{figure}

Based upon \Fref{fig:percent-filled}, some general statements can be made about what properties of the cement slurry flow are preferable.
For example, low yield stress, and consquently low Bingham number flows ($Bn \le 1.0$), approximately completely filled the annulus for every wellbore geometry considered.
In addition, the negligible percentage of voids for the ($Re = 5.00, Bn = 5.00$) and ($Re = 12.5, Bn = 2.0$) slurry flows suggests that if the ratio of $Bn / Re$ is low, that the slurry flow will result in a low percentage of voids for cementing a dry hole as is desired.
As $Bn / Re$ increases, it appears that, in certain cases and wellbore geometries, the slurry flow tends to result in a larger percentage of voids (approximately 10\%-40\%). \pagebreak
What is interesting to note is that for slurry flows that resulted in a percentage of voids greater than 3\%, there was a lack of consistency across wellbore geometries in terms of which slurry flow resulted in more voids and which slurry flow resulted in less, which suggests that generalities cannot always be made and that computational methods are necessary for understanding and designing slurry mixes for real-life wellbores.

\section{Conclusions} \label{sec:conclusion}

An LBM model for simulating non-Newtonian and free-surface flow to represent cementing dry annuli in oil and gas wellbores was presented.
The model was then used to study how different cement slurry flows, defined by different Reynold's numbers and Bingham numbers, would perform in different wellbore geometries.
The results showed that in general, it is preferable that the yield stress and $Bn / Re$ ratio of the slurry flow be low.
However, beyond these considerations, it is not clear what slurry flow may be preferable for a given rock formation geometry.
The results suggest that computational methods are necessary for (a) better understanding the mechanics of primary cementing and (b) for designing a cement mix for a given wellbore.
The lattice Boltzmann method extended for simulating non-Newtonian and free-surface flow as presented in the current work has several properties that make it a good candidate for such a design framework:
\begin{itemize}
\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
\item LBM is well-suited for simulating multiphase multicomponent flow and free-surface flow because it can handle complicated interface shapes between fluid phases and components.
\item LBM is easily written in parallel and as hardware architectures shifts sequential to parallel computing, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
\item The LBM no-slip boundary condition works well with complex geometries, such as those that occur in realistic rock formation surfaces in drilled wellbores.
\end{itemize}

Future work would include developing a more physically realistic model.
Primary cementing can involve multiple phases and multiple components, namely the cement slurry, a spacer fluid, and drilling mud.
Possible future work would be to make the model more general so as to be capable of simulating multiphase multicomponent flow.
In addition, a more physically realistic model would simulate three-dimensional flow so as to be able to simulate eccentric annuli and three-dimensional rock formation imperfections.

### 2018-02-12 manuscript-srcs/src/bingham.jl

@@ -1,0 +1,16 @@
+using PyPlot;
+
+tau = 4.0e-5;
+mu = 0.2;
+
+bingham(x) = tau / x + mu;
+papa(x, m) = tau / x * (1 - exp(-m*abs(x))) + mu;
+
+xs = [1e-11; 5e-11; 1e-10; 5e-10; 1e-9; 1e-8; 1e-7; 1e-6; 5e-6; 1e-5; 5e-5; 1e-4; 5e-4; 1e-3; 5e-3; 1e-2; 5e-2; 1e-1; 1];
+
+loglog(xs, map(bingham, xs), "k-", xs, map(x -> papa(x, 1e8), xs), "ks",
+     xs, map(x -> papa(x, 1e5), xs), "ko");
+legend(["Bingham","\$m = 10^8\$","\$m = 10^5\$"]);
+xlabel("\$\\dot\{\\gamma\}\$");
+ylabel("\$\\mu_p\$");
+show();

### 2018-02-12 manuscript-srcs/src/conversions.jl

@@ -1,0 +1,30 @@
+const cP_TO_LBPFTSEC = 0.00067196897514;
+const FT_TO_IN       = 12.0;
+const BLL_TO_CFT     = 5.6145833;
+const MIN_TO_SEC     = 60.0; 
+const GAL_TO_CFT     = 0.133681;
+
+@show pump_rates_bm = [2; 15];
+@show pump_rates_cfts = pump_rates_bm * BLL_TO_CFT / MIN_TO_SEC;
+
+@show annuli_dia_in = [4.5; 20.0];
+@show annuli_area_sqin = map(d -> π/4 * ((d + 1.5)^2 - d^2), annuli_dia_in);
+@show annuli_area_sqft = annuli_area_sqin / (FT_TO_IN^2);
+
+@show vel_ftps = [pump_rates_cfts[1] / annuli_area_sqft[2]; pump_rates_cfts[2] / annuli_area_sqft[1]];
+
+@show τ_lbpsqft  = [10.0; 100.0];
+@show μ_cP       = [10.0; 100.0];
+@show μ_lbpftsec = μ_cP * cP_TO_LBPFTSEC;
+@show ρ_lbpgal   = [12.0; 18.0];
+@show ρ_lbpcft   = ρ_lbpgal / GAL_TO_CFT;
+@show ν_sqftpsec = [μ_lbpftsec[1] / ρ_lbpcft[2]; μ_lbpftsec[2] / ρ_lbpcft[1]];
+
+Re(u, L, ν) = u * L / ν;
+Bn(τ, L, μ, u) = τ * L / (μ * u);
+Wt(τ, ρ, L) = (ρ * L) / τ;
+
+@show Res = [Re(vel_ftps[1], 0.75 / FT_TO_IN, ν_sqftpsec[2]); Re(vel_ftps[2], 0.75 / FT_TO_IN, ν_sqftpsec[1])];
+@show Bns = [Bn(τ_lbpsqft[1], 0.75 / FT_TO_IN, μ_lbpftsec[2], vel_ftps[2]), Bn(τ_lbpsqft[2], 0.75 / FT_TO_IN, μ_lbpftsec[1], vel_ftps[1])];
+@show Wts = [Wt(τ_lbpsqft[2], 0.75 / FT_TO_IN, ρ_lbpcft[1]); Wt(τ_lbpsqft[1], 0.75 / FT_TO_IN, ρ_lbpcft[2])];
+@show iWts = reverse(map(wt -> 1 / wt, Wts));
